### 2025-03-20 forecastflows.tex
Didn't finish transcribing the board to an outline today but at least I will finish the repo!

\documentclass[11pt]{article}
\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{parskip}
\usepackage{todonotes}

\def\labelitemi{--}
\parindent=0pt

\begin{document}
\renewcommand{\refname}{\CHead{}}

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}\\
 \maketitle

{\bf Deadline:} 1 May 2025 (was 1 April 2025)

\emph{For:} Scientific and Statistical Workflow theme issue for \emph{Phil Trans A} as an \emph{Opinion}
% They mention a tex template here: https://royalsocietypublishing.org/rsta/for-authors#question4
% Word length: I should check what they emailed but they say no more than 13 printed pages with 650 per page (whoa, we should be well below that! I am thinking 3-5K would be good)

\begin{abstract}
Increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Relevant datasets to meet these demands have also increased in size and complexity, including in their geographical, temporal and phylogenetic scales. While new research often suggests that accounting for these complexities variously increases, removes or otherwise alters major trends, I argue that the fundamental approach to model fitting in ecology makes it impossible to evaluate and compare models. These problems stem in part from continuing gaps between statistical workflows -- where the data processing and model development are often addressed separately from the ecological question and aim -- and scientific workflows, where all steps are integrated. Yet, as ecologists become increasingly computational, and new tools make it easier to share data, the opportunity to close this gap has never been greater. I outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights. Combining these changes with more open model and data sharing -- and developing new efforts to race the same data -- could be transformative for ecological forecasting. 
\end{abstract}

\begin{enumerate}
\item Problem
\begin{enumerate}
\item Climate change, biodiversity crisis etc. has made it critical to understand trends to date and be able to forecast future trends (for policy)
\item Current workflows in ecology are not up to the task
\begin{enumerate}
\item For trends: lots of different stuff reported for seemingly same data/question, makes people outside ecology wonder if we can even well document them, let alone understand them enough to suggest policy
\item For forecasting
\end{enumerate}
\item Here we outline the problem and provide a solution!
\end{enumerate}
\item What are current workflows and where are they limiting us?
\begin{enumerate}
\item For trends ...
\begin{enumerate}
\item easy to find different trends through small model tweaks to analyses and/or different data
\item For example, right now many different papers report different biodiversity trends (LPI example?)
\item New workflow would make ecologists understand uncertainty in their model data/combo (and perhaps not see/publish results as so divergent?)
\end{enumerate}
\item For forecasting ... (somehow jump to our focus on process-based models PBMs here? Something like, forecasting is big and there are diverse methods! Near-term iterative, correlative niche models, but PBMs are often considered the gold standard ... mention machine learning?)
\end{enumerate}
\item Better workflows to the rescue! 
\begin{enumerate}
\item General overview of new workflows
\item New vision of each workflow
\end{enumerate}
\item Conclusion: world is better
\end{enumerate}

\citep{ospreebbms}

\bibliographystyle{/Users/Lizzie/Documents/EndnoteRelated/Bibtex/styles/besjournals}
\bibliography{/Users/Lizzie/Documents/git/bibtex/LizzieMainMinimal}

\clearpage
\section{Figures}

\begin{figure}[h!]
\includegraphics[width=0.4\textwidth]{..//..//..//..//Professional/images/ColylusVineFruit2.png}
\caption{I am a caption.} 
\label{fig:figname}
\end{figure}

\end{document}

\begin{enumerate}
\item
\end{enumerate}

### 2025-03-23 forecastflows.tex
Small notes from Friday's meeting

@@
 
+{\bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
+
 \begin{enumerate}
@@
 \item General overview of new workflows
+\begin{enumerate}
+\item Step 0: Research Qs and hypotheses (with a mechanism) lets you ...
+\item Step 1: Build a model!
+\item Step 2: Simulate data (and priors or something like priors that forces you to put numbers on stuff)
+\item Step 3: Design experiments/data collection (maybe you go back to Step 1 here)
+\item Step 4: Simulate data from actual design/collection
+\item Step 5: Fit the model to empirical data
+\item Step 6: Retrodictive checks (feed back to 0 and 1)
+\end{enumerate}
 \item New vision of each workflow
@@
 \item Conclusion: world is better
+\end{enumerate}
+
+Current workflows
+\begin{enumerate}
+\item 
+\end{enumerate}
+
+How much of forecasting do we cover?
+\begin{enumerate}
+\item PBMs
+\item Near-term iterative
+\item SDMs
+\item Machine learning
+\end{enumerate}
+
+Miscellaneous notes/points without a home
+\begin{enumerate}
+\item Ecologists need to race the same data to make progress for trends and for forecasting (point to make at end of paper maybe?
+And what is the workflow for this?) ...
+though LPI is used a lot, perhaps it is  sign that ecology is ready to start racing the same data, but then we need `analysis-ready' data so we're not all slightly differently cleaning the data etc..
+\item Machine learning threatens utility of PBMs
+\item We need more uncertainty propagation for trends and forecasting (uncertainty is esp. ignored in PBMs)
+\item This workflow should lead to less model comparison (AIC, stepwise)
+\item This workflow works for machine learning!
+\item PBM: workflow should require estimating all parameters together; data simulation should reduce parameter number and highlight non-biol results
+\item CUrrent trend workflow: SHould be research questi
+\end{enumerate}
+
+\end{document}
+
+\begin{enumerate}
+\item
 \end{enumerate}
@@
 \end{figure}
-
-\end{document}
-
-\begin{enumerate}
-\item
-\end{enumerate}

### 2025-03-24 forecastflows.tex
add some ideas for current limitations in PBM workflow

@@
 \item For trends: lots of different stuff reported for seemingly same data/question, makes people outside ecology wonder if we can even well document them, let alone understand them enough to suggest policy
-\item For forecasting
+\item For forecasting: also high divergence between models, criticism against process-based approaches (supposedly the most robust?) for lack of transparency (in the model building and calibration) and increasing complexity, with many parameters not supported by data
 \end{enumerate}
@@
 mention machine learning?)
+\begin{enumerate}
+\item as many models as researchers working on process-based models 
+\item + accumulation of successive layers in the development of models = significant challenge to scientific transparency, reproducibility, interpretability\\
+models often draw inspiration from each other, which is good (way to do science), but not always explicit...
+(some issues: arbitrarily established parameter value in one model then transmitted to multiple models)
+\item focus of researchers: always integrate new mechanisms, new parameters, to increase "realism"...
+they intuitively "feel" what kinds of adjustments is needed...
+but opaque from an external perspective ("black box" of model building and calibration)
+\item models rarely fitted as a whole, dozens of parameters without explicitly quantifying parameter uncertainty, and often neglect to propagate this uncertainty
+\item simulations of models themselves became a subject of study to disentangle all the processes modelled and understand model sensitivity 
+\end{enumerate}
 \end{enumerate}

### 2025-03-24 forecastflows.tex
Just some random notes from thinking this weekend.

@@
 \date{\today}
-\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}\\
+\author{Victor Van der Meersch, J. Regetz, T. J. Davies$^*$ \& EM Wolkovich}\\
  \maketitle
+$^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author.
 
@@
 \item PBM: workflow should require estimating all parameters together; data simulation should reduce parameter number and highlight non-biol results
-\item CUrrent trend workflow: SHould be research questi
+\item Current trend workflow: Should be research question focused?
+\end{enumerate}
+
+Miscellaneous notes/points from thinking over 22-23 March weekend ...
+\begin{enumerate}
+\item Workflows emphasize there is no easy fix to better science or better stats
+\item Maybe do a retrodictive check example with LPI? Including showing how even the Freckleton work could do more? 
+\item What's the pathway from model comparison of many covariates to something else?
+Sometimes it seems like it's just prediction shoved into a mechanistic study, but if the goal is mechanism, there needs to be more work that either models these covariates together in a useful way (sort of approaching process-models!) or gets down to the fewer extremely relevant ones.
+\item When do we need to open up the black box? Something about we need better training for what science is and our goals; machine learning is not often helping with advancing \emph{science}
+\item Scott Collins point that forecasting is not science (he claims a bunch of economists, weather folks etc.
+came to an ecological forecasting meeting and tried to explain that forecasting is an outcome of science, but it's not something you do science ON)
 \end{enumerate}

### 2025-03-25 forecastflows_outline.tex
outline notes from whiteboard, 24 March

\documentclass[11pt]{article}
\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{hyperref}

\def\labelitemi{--}
\parindent=0pt

\begin{document}
\renewcommand{\refname}{\CHead{}}

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch, J. Regetz, T. J. Davies$^*$ \& EM Wolkovich}
\maketitle
$^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author.

\begin{enumerate}
\item Intro
\begin{itemize}
\item Ecology super challenged to predict stuff for decision making (kind of a whole new world of relevance?)
$leftarrow$ Example of stuff to predict: populations, policy-relevant questions...
\item General way to do this so far... (bifurcated?)
\begin{itemize}
\item long term data (GBIF, Biotime...) for estimation of trends
\item PBM, SDMs for forecast
\end{itemize}
\item Gap, problem: None of this is going well
\begin{itemize}
\item debates over trends, not only on the significance but also the direction!
\item predictive modeling trapped in overly complex models, hard to get scientific insights? (same as GCMs?)
\end{itemize}
\item Here we introduce a universal workflow to adress this (say more)
\end{itemize}
\item Overview
\begin{itemize}
\item General scientific method we all learn stresses: RQ $\rightarrow$ study design $\rightarrow$ collect data $\rightarrow$ build model $\rightarrow$ answer
\item Divergences from this are common: 
\begin{itemize}
\item bad science (data lead to question)
\item important question for which we cannot get data quickly, we have to use existing
\item complexity makes this simple workflow hard/impossible
\end{itemize}
\item this workflow works across these realities by
\begin{itemize}
\item stressing the need to think about model before study design
\item advancing data simulation
% \item (outcome: spending more time in current rare quadrats will improve forecasts!)
\end{itemize}
\item More details on ideal workflow:  walk through the different steps
\begin{itemize}
\item spend more time in critical quadrat, post-model pre-data
\item feedbacks
\item uncertainty
\end{itemize}
\end{itemize}
\item How to address current issues (two case studies)
\begin{enumerate}
\item Trends!
\begin{itemize}
\item Outline current problem: different answers from different analysis (and slightly different datasets?)\\
$\rightarrow$ It is a problem because we can't make decisions on +/- debate... (and it degrades trust in science?)
\item two big missing parts are data simulation steps:
\begin{itemize}
\item retrodictive check: would highlight missing pieces (speed up current process, because without this step each slightly different model is a paper... whereas each big iteration of the workflow---including multiple feedbacks---should be a paper)
\item simulated data: you would know that you have low data sooner => the debate is actually not jus models but really limited data
\end{itemize}
\end{itemize}
\item Forecasts!
\begin{itemize}
\item Outline current problem:
\begin{itemize}
\item black box: intrication between model build and data fitting (calibration), everything is mixed 
\item complexity trap [mention uncertainty]
\item developping a model has become the goal, whereas it should be a way to answer a research question!
\end{itemize}
\item We need the workflow to open the black box! Simulating data would allow to add a necessary step between model building and data fitting, which would highlight strong degeneracies
\item Workflow would also force you to clearly express a research question, define a limited context in which the model should apply
\end{itemize}
\item Step back
\begin{itemize}
\item we need more data, and better question (relate this to both previous study cases)
\item where can we best reduce uncertainties through new scientific insights?
\item machine learning! If we do nothing, ML $>$ process-based without question, and ML $>$ trends without mechanisms
\end{itemize}
\end{enumerate}
\item Wrap up: how to make it happen?
\begin{itemize}
\item usual issues: publication pressure, low standards (especially for models)\\
"\emph{we allow far more hand-waving in the presentation of modeling results than we do for experimental data}" (\href{https://harvardforest1.fas.harvard.edu/publications/pdfs/Aber_BulletinEcoSocAmerica_1997.pdf}{J. Aber, 1997})
\item growing concerns, leading to increase in reproducibility and data sharing practices [mention uncertainty]
\item need a little more here...
\item better training! on BOTH estimation AND prediction
\begin{itemize}
\item prediction: being aware of what is a parameter, and [mention uncertainty propagation]
\item prediction: should be a natural outcome, not a finality
\end{itemize}
\item ML, benchmarking models are probably useful but should not be the core of our scientific practice, not the spirit!
\end{itemize}
\end{enumerate}

\end{document}

### 2025-03-25 forecastflows_outline.tex
small tweaks on outline

@@
 \usepackage{hyperref}
-
-\def\labelitemi{--}
-\parindent=0pt
 
@@
 \begin{itemize}
-\item Ecology super challenged to predict stuff for decision making (kind of a whole new world of relevance?)
-$leftarrow$ Example of stuff to predict: populations, policy-relevant questions...
+\item Ecology super challenged to predict stuff for decision making (kind of a whole new world of relevance?)\\
+$\rightarrow$ Example of stuff to predict: populations, policy-relevant questions...
 \item General way to do this so far... (bifurcated?)
@@
 \item where can we best reduce uncertainties through new scientific insights?
-\item machine learning! If we do nothing, ML $>$ process-based without question, and ML $>$ trends without mechanisms
+\item machine learning! If we do nothing, what's the point of not doing ML? ML $>$ process-based without question, and ML $>$ trends without mechanisms
 \end{itemize}
@@
 \begin{itemize}
-\item prediction: being aware of what is a parameter, and [mention uncertainty propagation]
+\item estimation: being aware of what is a parameter, and mention uncertainty propagation
 \item prediction: should be a natural outcome, not a finality

### 2025-03-25 forecastflows.tex
working on intro

--- 2025-03-23 forecastflows.tex
@@
 \usepackage{amsmath}
-\usepackage{parskip}
-\usepackage{todonotes}
-
-\def\labelitemi{--}
-\parindent=0pt
 
@@
 \date{\today}
-\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}\\
+\author{Victor Van der Meersch, J. Regetz, T. J. Davies$^*$ \& EM Wolkovich}
 \maketitle
+$^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author.
 
@@
 {\bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
+\vspace*{0.5cm}
 
-\begin{enumerate}
-\item Problem
-\begin{enumerate}
-\item Climate change, biodiversity crisis etc. has made it critical to understand trends to date and be able to forecast future trends (for policy)
-\item Current workflows in ecology are not up to the task
-\begin{enumerate}
-\item For trends: lots of different stuff reported for seemingly same data/question, makes people outside ecology wonder if we can even well document them, let alone understand them enough to suggest policy
-\item For forecasting
-\end{enumerate}
-\item Here we outline the problem and provide a solution!
-\end{enumerate}
-\item What are current workflows and where are they limiting us?
-\begin{enumerate}
-\item For trends ...
-\begin{enumerate}
-\item easy to find different trends through small model tweaks to analyses and/or different data
-\item For example, right now many different papers report different biodiversity trends (LPI example?)
-\item New workflow would make ecologists understand uncertainty in their model data/combo (and perhaps not see/publish results as so divergent?)
-\end{enumerate}
-\item For forecasting ...
+\textbf{Introduction}
+
+%\begin{itemize}
+%\item 
+%Increasing impact upon nature, facing multiple drivers of change, with a dominant influence of human activities (land-use, pollution, invasive species, climate change)\\
+%Biodiversity crisis etc. has made it critical to understand trends to date and be able to forecast future trends\\
+%\item Ecology thus has become super challenged to predict stuff for decision making, and meet multiple goals among the socioeconomic and environmental dimensions --- global sustainability more broadly (kind of a whole new world of relevance?)\\
+Nature is increasingly threatened by multiple drivers of change, with a largely dominant influence of human activities.
+This ongoing biodiversity crisis is expected to increase in the next decades, and will continue to alter ecosystem services and human well-being.
+To support implementation of sustainable policies among the socioeconomic and environmental dimensions, it is critical to understand trends to date and be able to forecast future dynamics.
+% $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...
+
+%\item General way to do this so far... (bifurcated?)
+%\begin{itemize}
+%\item long term data (GBIF, Biotime...) for estimation of trends
+%\item PBMs, SDMs for plausible forecast
+%\end{itemize}
+Estimation of global biodiversity indicators and current trends depends on large-scale and long-term datasets, across terrestrial, freshwater, and marine ecosystems.
+These data, gathered opportunistically and from multiple sources, are often unbalanced and have geographic, temporal and taxonomic biases.
+Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models.
+The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach.
+
+%$\rightarrow$ to meet these needs: proliferation of methods from all sides, developed by different researchers, without any coherence
+%\item Gap, problem: None of this is going well. Though there is no doubt nature is declining globally (IPBES), great deal of uncertainty 
+%\begin{itemize}
+%\item Debates over current trends, lots of different stuff reported (differ widely), including different trend directions!
+(makes people outside ecology wonder if we can even well document them, let alone understand them enough to suggest policy)
+%\item For forecasting: also high divergence between projections, predictive modeling is slowly becoming trapped in overly complex models (too many parameters not supported by data), hard to get new scientific insights ($\parallel$ GCMs?)
+%\end{itemize}
+The urgent need to answer policy-relevant questions has favored the proliferation of diverse methods developed by different researchers, and lacking an overall coherence.
+Though there is no doubt nature is declining globally, significant uncertainty thus remains.
+There is no consensus on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions.
+Future projections also diverge considerably.
+Predictive modeling is increasingly relying on overly complex models (with a huge number of parameters), making it less adequate to generate new scientific insights.
+
+%\item Current workflows in ecology are not up to the task\\
+%Here we introduce a universal workflow to adress this (say more)
+%\end{itemize}
+
+%\item What are current workflows and where are they limiting us?
+%\begin{enumerate}
+%\item For trends ...
+%\begin{enumerate}
+%\item easy to find different trends through small model tweaks to analyses and/or different data
+%\item For example, right now many different papers report different biodiversity trends (LPI example?)
+%\item New workflow would make ecologists understand uncertainty in their model data/combo (and perhaps not see/publish results as so divergent?)
+%\end{enumerate}
+%\item For forecasting ...
 (somehow jump to our focus on process-based models PBMs here?
@@
 mention machine learning?)
-\end{enumerate}
-\item Better workflows to the rescue! 
-\begin{enumerate}
-\item General overview of new workflows
-\begin{enumerate}
-\item Step 0: Research Qs and hypotheses (with a mechanism) lets you ...
-\item Step 1: Build a model!
-\item Step 2: Simulate data (and priors or something like priors that forces you to put numbers on stuff)
-\item Step 3: Design experiments/data collection (maybe you go back to Step 1 here)
-\item Step 4: Simulate data from actual design/collection
-\item Step 5: Fit the model to empirical data
-\item Step 6: Retrodictive checks (feed back to 0 and 1)
-\end{enumerate}
-\item New vision of each workflow
-\end{enumerate}
-\item Conclusion: world is better
-\end{enumerate}
-
-Current workflows
-\begin{enumerate}
-\item 
-\end{enumerate}
-
-How much of forecasting do we cover?
-\begin{enumerate}
-\item PBMs
-\item Near-term iterative
-\item SDMs
-\item Machine learning
-\end{enumerate}
-
-Miscellaneous notes/points without a home
-\begin{enumerate}
-\item Ecologists need to race the same data to make progress for trends and for forecasting (point to make at end of paper maybe?
+%\begin{enumerate}
+%\item as many models as researchers working on process-based models 
+%\item + accumulation of successive layers in the development of models = significant challenge to scientific transparency, reproducibility, interpretability\\
+%models often draw inspiration from each other, which is good (way to do science), but not always explicit...
+(some issues: arbitrarily established parameter value in one model then transmitted to multiple models)
+%\item focus of researchers: always integrate new mechanisms, new parameters, to increase "realism"...
+they intuitively "feel" what kinds of adjustments is needed...
+but opaque from an external perspective ("black box" of model building and calibration)
+%\item models rarely fitted as a whole, dozens of parameters without explicitly quantifying parameter uncertainty, and often neglect to propagate this uncertainty
+%\item simulations of models themselves became a subject of study to disentangle all the processes modelled and understand model sensitivity 
+%\end{enumerate}
+%\end{enumerate}
+%\item Better workflows to the rescue! 
+%\begin{enumerate}
+%\item General overview of new workflows
+%\begin{enumerate}
+%\item Step 0: Research Qs and hypotheses (with a mechanism) lets you ...
+%\item Step 1: Build a model!
+%\item Step 2: Simulate data (and priors or something like priors that forces you to put numbers on stuff)
+%\item Step 3: Design experiments/data collection (maybe you go back to Step 1 here)
+%\item Step 4: Simulate data from actual design/collection
+%\item Step 5: Fit the model to empirical data
+%\item Step 6: Retrodictive checks (feed back to 0 and 1)
+%\end{enumerate}
+%\item New vision of each workflow
+%\end{enumerate}
+%\item Conclusion: world is better
+%\end{enumerate}
+%
+%Current workflows
+%\begin{enumerate}
+%\item 
+%\end{enumerate}
+%
+%How much of forecasting do we cover?
+%\begin{enumerate}
+%\item PBMs
+%\item Near-term iterative
+%\item SDMs
+%\item Machine learning
+%\end{enumerate}
+%
+%
+%Miscellaneous notes/points without a home
+%\begin{enumerate}
+%\item Ecologists need to race the same data to make progress for trends and for forecasting (point to make at end of paper maybe?
 And what is the workflow for this?) ...
 though LPI is used a lot, perhaps it is  sign that ecology is ready to start racing the same data, but then we need `analysis-ready' data so we're not all slightly differently cleaning the data etc..
-\item Machine learning threatens utility of PBMs
-\item We need more uncertainty propagation for trends and forecasting (uncertainty is esp. ignored in PBMs)
-\item This workflow should lead to less model comparison (AIC, stepwise)
-\item This workflow works for machine learning!
-\item PBM: workflow should require estimating all parameters together; data simulation should reduce parameter number and highlight non-biol results
-\item CUrrent trend workflow: SHould be research questi
-\end{enumerate}
+%\item Machine learning threatens utility of PBMs
+%\item We need more uncertainty propagation for trends and forecasting (uncertainty is esp. ignored in PBMs)
+%\item This workflow should lead to less model comparison (AIC, stepwise)
+%\item This workflow works for machine learning!
+%\item PBM: workflow should require estimating all parameters together; data simulation should reduce parameter number and highlight non-biol results
+%\item Current trend workflow: Should be research question focused?
+%\end{enumerate}
+%
+%Miscellaneous notes/points from thinking over 22-23 March weekend ...
+%\begin{enumerate}
+%\item Workflows emphasize there is no easy fix to better science or better stats
+%\item Maybe do a retrodictive check example with LPI? Including showing how even the Freckleton work could do more? 
+%\item What's the pathway from model comparison of many covariates to something else?
+Sometimes it seems like it's just prediction shoved into a mechanistic study, but if the goal is mechanism, there needs to be more work that either models these covariates together in a useful way (sort of approaching process-models!) or gets down to the fewer extremely relevant ones.
+%\item When do we need to open up the black box? Something about we need better training for what science is and our goals; machine learning is not often helping with advancing \emph{science}
+%\item Scott Collins point that forecasting is not science (he claims a bunch of economists, weather folks etc.
+came to an ecological forecasting meeting and tried to explain that forecasting is an outcome of science, but it's not something you do science ON)
+%\end{enumerate}
 

### 2025-03-26 forecastflows.tex
slooowly making progress

@@
 
-\textbf{Introduction}
+\noindent \textbf{Introduction}
+% Still need some work...
 
-%\begin{itemize}
-%\item 
-%Increasing impact upon nature, facing multiple drivers of change, with a dominant influence of human activities (land-use, pollution, invasive species, climate change)\\
-%Biodiversity crisis etc. has made it critical to understand trends to date and be able to forecast future trends\\
-%\item Ecology thus has become super challenged to predict stuff for decision making, and meet multiple goals among the socioeconomic and environmental dimensions --- global sustainability more broadly (kind of a whole new world of relevance?)\\
+% Ecology has become super challenged to predict stuff for decision making
 Nature is increasingly threatened by multiple drivers of change, with a largely dominant influence of human activities.
@@
 
-%\item General way to do this so far... (bifurcated?)
-%\begin{itemize}
-%\item long term data (GBIF, Biotime...) for estimation of trends
-%\item PBMs, SDMs for plausible forecast
-%\end{itemize}
-Estimation of global biodiversity indicators and current trends depends on large-scale and long-term datasets, across terrestrial, freshwater, and marine ecosystems.
+% General way to do this so far... (bifurcated?)
+Estimation of global biodiversity indicators and current trends depends on large-scale and long-term datasets---across terrestrial, freshwater, and marine ecosystems.
 These data, gathered opportunistically and from multiple sources, are often unbalanced and have geographic, temporal and taxonomic biases.
+Addressing these biases requires the use of appropriate statistical inference.
 Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models.
@@
 
-%$\rightarrow$ to meet these needs: proliferation of methods from all sides, developed by different researchers, without any coherence
-%\item Gap, problem: None of this is going well. Though there is no doubt nature is declining globally (IPBES), great deal of uncertainty 
-%\begin{itemize}
-%\item Debates over current trends, lots of different stuff reported (differ widely), including different trend directions!
-(makes people outside ecology wonder if we can even well document them, let alone understand them enough to suggest policy)
-%\item For forecasting: also high divergence between projections, predictive modeling is slowly becoming trapped in overly complex models (too many parameters not supported by data), hard to get new scientific insights ($\parallel$ GCMs?)
-%\end{itemize}
-The urgent need to answer policy-relevant questions has favored the proliferation of diverse methods developed by different researchers, and lacking an overall coherence.
-Though there is no doubt nature is declining globally, significant uncertainty thus remains.
+% Gap, problem: None of this is going well
+The urgent need to answer policy-relevant questions has favored the proliferation of diverse methods developed by different researchers, lacking an overall coherence.
+Though there is no doubt nature is declining globally, significant uncertainty remains.
 There is no consensus on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions.
-Future projections also diverge considerably.
+Future projections also diverge considerably, due to a high model uncertainty at the ecological level.
 Predictive modeling is increasingly relying on overly complex models (with a huge number of parameters), making it less adequate to generate new scientific insights.
 
-%\item Current workflows in ecology are not up to the task\\
-%Here we introduce a universal workflow to adress this (say more)
-%\end{itemize}
+% Current workflows in ecology are not up to the task
+% Here we introduce a universal workflow to adress this...
+Ongoing controversies and focus on methodological aspects stem from the lack of coherence between current workflows in ecology.
+Each new model development is added as a separate layer, disconnected from the original research aim, the data stream and the previous scientific insights.
+Workflows should fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections.
+Here, we introduce an universal workflow that proposes to iteratively build upon all these steps, harmonizing both trend estimation and forecasting, with the aim of refocusing the debate on ecological questions.
 

### 2025-03-26 forecastflows.tex
adding some refs

@@
 \usepackage{amsmath}
+\bibliographystyle{besjournals}
 
 \begin{document}
-\renewcommand{\refname}{\CHead{}}
 
@@
 % Ecology has become super challenged to predict stuff for decision making
-Nature is increasingly threatened by multiple drivers of change, with a largely dominant influence of human activities.
-This ongoing biodiversity crisis is expected to increase in the next decades, and will continue to alter ecosystem services and human well-being.
+Nature is increasingly threatened by multiple drivers of change, with a largely dominant influence of human activities \citep{Diaz2019}.
+This ongoing biodiversity crisis is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}.
 To support implementation of sustainable policies among the socioeconomic and environmental dimensions, it is critical to understand trends to date and be able to forecast future dynamics.
@@
 % General way to do this so far... (bifurcated?)
-Estimation of global biodiversity indicators and current trends depends on large-scale and long-term datasets---across terrestrial, freshwater, and marine ecosystems.
+Estimation of global biodiversity indicators and current trends depends on large-scale and long-term datasets---across terrestrial, freshwater, and marine ecosystems \citep[e.g.][]{Dornelas2018}.
 These data, gathered opportunistically and from multiple sources, are often unbalanced and have geographic, temporal and taxonomic biases.
 Addressing these biases requires the use of appropriate statistical inference.
-Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models.
-The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach.
+Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models \citep{IPBES2019}.
+The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022}.
+% ...
+lacks something
 
@@
 Though there is no doubt nature is declining globally, significant uncertainty remains.
-There is no consensus on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions.
+There is no consensus on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}.
 Future projections also diverge considerably, due to a high model uncertainty at the ecological level.
-Predictive modeling is increasingly relying on overly complex models (with a huge number of parameters), making it less adequate to generate new scientific insights.
+Predictive modeling is increasingly relying on overly complex models (with a huge number of parameters), making it less adequate to generate new scientific insights \citep{Franklin2020}.
 
@@
 % Here we introduce a universal workflow to adress this...
-Ongoing controversies and focus on methodological aspects stem from the lack of coherence between current workflows in ecology.
+Current controversies and focus on methodological aspects stem from the lack of coherence between current workflows in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
 Each new model development is added as a separate layer, disconnected from the original research aim, the data stream and the previous scientific insights.
 Workflows should fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections.
-Here, we introduce an universal workflow that proposes to iteratively build upon all these steps, harmonizing both trend estimation and forecasting, with the aim of refocusing the debate on ecological questions.
+Here, we introduce an universal workflow that proposes to iteratively build upon all these steps, harmonizing both trend estimation and forecasting, with the aim of refocusing the debate on ecological questions and increasing the speed of scientific progress.
 
@@
 
+\clearpage
+\bibliography{forecastflows}
+
 \end{document}
-
-\begin{enumerate}
-\item
-\end{enumerate}
-
-\citep{ospreebbms}
-
-\bibliographystyle{/Users/Lizzie/Documents/EndnoteRelated/Bibtex/styles/besjournals}
-\bibliography{/Users/Lizzie/Documents/git/bibtex/LizzieMainMinimal}
-
-\clearpage
-\section{Figures}
-
-\begin{figure}[h!]
-\includegraphics[width=0.4\textwidth]{..//..//..//..//Professional/images/ColylusVineFruit2.png}
-\caption{I am a caption.} 
-\label{fig:figname}
-\end{figure}

### 2025-03-27 forecastflows_outline.tex
add figures in the outline

@@
 \item General scientific method we all learn stresses: RQ $\rightarrow$ study design $\rightarrow$ collect data $\rightarrow$ build model $\rightarrow$ answer
-\item Divergences from this are common: 
+\item Divergences from this are common \textbf{(Figure 1)}
 \begin{itemize}
 \item bad science (data lead to question)
-\item important question for which we cannot get data quickly, we have to use existing
-\item complexity makes this simple workflow hard/impossible
+\item important question for which we cannot get data quickly, we have to use existing data: e.g. global datasets for trends, many various small datasets for process-based (experimental...)
+\item complexity makes the ideal workflow hard/impossible
 \end{itemize}
-\item this workflow works across these realities by
+\item the workflow we present here works across these realities by
 \begin{itemize}
@@
 \end{itemize}
-\item More details on ideal workflow:  walk through the different steps
+\item More details on ideal workflow:  walk through the different steps \textbf{(Figure 2)}
 \begin{itemize}
@@
 \end{itemize}
-\item How to address current issues (two case studies)
+\item How to address current issues (two case studies, \textbf{Figure 3})
 \begin{enumerate}
@@
 \item Outline current problem: different answers from different analysis (and slightly different datasets?)\\
-$\rightarrow$ It is a problem because we can't make decisions on +/- debate... (and it degrades trust in science?)
+$\rightarrow$ It is a problem because we can't make decisions on +/- trends debate... (and it degrades trust in science?)
 \item two big missing parts are data simulation steps:

### 2025-03-27 forecastflows.tex
some writing

@@
 Here, we introduce an universal workflow that proposes to iteratively build upon all these steps, harmonizing both trend estimation and forecasting, with the aim of refocusing the debate on ecological questions and increasing the speed of scientific progress.
+
+\vspace{0.2cm}
+\noindent \textbf{Scientific practice and current workflows}
+
+% General scientific method we all learn
+Quantitative science rely on a model-based framework to confront hypothesis and data, making some approximations \citep{}.
+The general scientific method stresses out that the research question should guide both the design of the experiment and the corresponding model building.
+The experiment should then be conducted---according to this specific design.
+Finally, the resulting experimental data should be used to inform our model and differentiate between hypotheses---hopefully answering our initial question.
+However, this is often an idealized view and does not reflect the complexity of ecology \citep{}.
+
+% Divergences from this are common 
+Divergences from this scientific method are common.
+One explanation is the lack of rigor, leading to questionable practices, such as retrospectively crafting hypothesis to explain the results of a model rather than testing a clear pre-defined question (data-driven analysis, Fig.
+1).
+But the reality of ecological research is also a major driver of the divergences from the ideal scientific method.
+Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective.
+However, this does not explain the persistent flaws and lack of overall coherence in ecological modeling.
+Trend estimation still often relies on a one-way model fitting, and most of the time is spent fitting the model to empirical data (Fig.
+1).
+For forecasting, researchers focus on making predictions with complex models, but the steps of model building and parameterization are not very transparent and not clearly delineated (Fig.
+1).
+The different parts of the model are often calibrated separately rather than as a whole, and some parameter values are just fixed based on experiments and expert knowledge.
+
+% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
 

### 2025-03-27 forecastflows_outline.tex
adding some stuff not to forget

@@
 \end{itemize}
-\item More details on ideal workflow:  walk through the different steps \textbf{(Figure 2)}
+\item More details on ideal workflow:  walk through the different steps \textbf{(Figure 2)}, and emphasizes: not rigid structure, integration of everything,  iterate the process of science, explicit effort to recognize the uncertainties,
 \begin{itemize}
@@
 \item black box: intrication between model build and data fitting (calibration), everything is mixed 
-\item complexity trap [mention uncertainty]
+\item complexity trap [mention uncertainty], counterproductive, not always better performance
 \item developping a model has become the goal, whereas it should be a way to answer a research question!
@@
 \item We need the workflow to open the black box! Simulating data would allow to add a necessary step between model building and data fitting, which would highlight strong degeneracies
+\item a clear framework to support (or not) additional complexity and new parameters
 \item Workflow would also force you to clearly express a research question, define a limited context in which the model should apply
@@
 \item where can we best reduce uncertainties through new scientific insights?
-\item machine learning! If we do nothing, what's the point of not doing ML? ML $>$ process-based without question, and ML $>$ trends without mechanisms
+\item machine learning!
+If we change nothing, what's the point of not doing ML?
+ML $>$ process-based without question, and ML $>$ trends without mechanisms!
+Where theory is lacking, or where we are far from mechanistic understanding, you might as well do ML!
+\item (ML has a way to collect and interpret large datasets...)
 \end{itemize}
@@
 \end{itemize}
-\item ML, benchmarking models are probably useful but should not be the core of our scientific practice, not the spirit!
+\item ML (short-time forecast), benchmarking models are probably useful but should not be the core of our scientific practice, not the spirit!\\
+$\rightarrow$ moving ecology in the right direction!
 \end{itemize}

### 2025-03-28 forecastflows.tex
add some comments
continue writing...

@@
 
-{\bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
+{\noindent \bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
 \vspace*{0.5cm}
@@
 
-\vspace{0.2cm}
-\noindent \textbf{Scientific practice and current workflows}
+\vspace{0.5cm}
+\noindent \textbf{Scientific method and workflows}
+%  too much blah blah blah?
 
@@
 Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective.
+% I feel something is missing
 However, this does not explain the persistent flaws and lack of overall coherence in ecological modeling.
-Trend estimation still often relies on a one-way model fitting, and most of the time is spent fitting the model to empirical data (Fig.
+Trend estimation still often relies on a one-way 'inference',  % need a better word?
+to stress out that there is no feedback, eah paper applied a slightly different model
+and most of the time is spent fitting the model to empirical data (Fig.
 1).
@@
 % There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
+% Moving along the data-model space
+% \item stressing the need to think about model before study design
+% \item advancing data simulation
+[...]
+
+% More details on the workflow:  walk through the different steps
+[...]
+
+\vspace{0.5cm}
+\noindent \textbf{The workflow in pratice}
+% How to address current issues (two case studies)
+
+% Trends - outline current problem
+[...]
+
+% Trends - improvements!
+[...]
+
+% Forecasting - outline current problem
+Forecasting is a broad field with a diverse range of methods.
+Here, we focus on process-based modeling, which is often considered the gold standard in ecology and beyond.
+Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses.
+They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted.
+Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions.
+Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective.
+The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration, where parameter values are inferred.
+Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes.
+Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e.
+fix to a value found in the literature).
+
+% Forecasting - potential improvments (do I say too much ideas)
+Applying the workflow to process-based models is a key for opening the black box.
+It would serve as a guide through the successive steps of model development.
+In particular, incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing strong degeneracies in the model design.
+This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model.
+Data fitting would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights.
+The workflow could also refocus attention on the research question, defining a clear and limited context in which the model should apply.
+Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
+Finally, an universal workflow provides an opportunity to merge statistical and process-based approaches, integrating mechanistic knowledge and leveraging robust statistical approaches.
+
+% Step back
+%\item we need more data, and better question (relate this to both previous study cases)
+%\item where can we best reduce uncertainties through new scientific insights?
+%\item machine learning!
+If we change nothing, what's the point of not doing ML?
+ML $>$ process-based without question, and ML $>$ trends without mechanisms!
+Where theory is lacking, or where we are far from mechanistic understanding, you might as well do ML!
+%\item (ML has a way to collect and interpret large datasets...)
+
+\vspace{0.5cm}
+\noindent \textbf{Wrap up: how to make it happen?}
 

### 2025-03-28 forecastflows_outline.tex
continue writing...

@@
 \end{itemize}
-\item More details on ideal workflow:  walk through the different steps \textbf{(Figure 2)}, and emphasizes: not rigid structure, integration of everything,  iterate the process of science, explicit effort to recognize the uncertainties,
+\item More details on the workflow:  walk through the different steps \textbf{(Figure 2)}, and emphasizes: not rigid structure, integration of everything,  iterate the process of science, explicit effort to recognize the uncertainties, \\
+forecast is just a step:  jointly model the new circumstance along with the original data
 \begin{itemize}

### 2025-03-28 forecastflows.tex
step back, mentioning machine learning

@@
 %\item (ML has a way to collect and interpret large datasets...)
+Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow holds the promise to highlight the opportunities to best reduce uncertainties through new scientific insights, toward the most critical steps.
+This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.
+In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made.
+Machine learning may surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
+Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
+% If we do not take care to apply this workflow, machine learning might be the more pragmatic approach...
 

### 2025-03-29 forecastflows.tex
starting wrap-up

@@
 
+The misuse of models and misinterpretation of their outcomes (including by the authors) have multiple origins.
+One cause, almost tautological, is the pressure to publish academic work, which can lower research standards.
+This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
+This a growing concern, leading to increase in reproducibility and data sharing practices.
+However, model development in ecology is still rarely transparent, which limits model understandability and prevents peers from properly identifying potential issues.
+A significant portion of scientific debate thus becomes lost in methodological considerations rather than advancing our ecological understanding.
+
 %\item What are current workflows and where are they limiting us?

### 2025-03-30 forecastflows.tex
Started work on this! Must focus more on writing than sifting through references.

@@
 \vspace{0.5cm}
-\noindent \textbf{The workflow in pratice}
+\noindent \textbf{The workflow in practice}
 % How to address current issues (two case studies)
@@
 % Trends - outline current problem
-[...]
+%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology.
+Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing concerns of how pollution affects ecosystems and their services gave rise to a growing public concern about protecting and maintaining the environment, and made ecology as a field more relevant \cite{soule1999conserving,soule1991conservation}.
+The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}.
+Such trends drove number of new subfields within ecology---some of which are now complete disciplines within themselves, such as conservation biology \citep{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES.
+Yet as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overhavesting, pollution only increase, and anthropogenic climate change now clearly driving species loss (CITE bramble cay meomys), so has the data and its complexity.
+
+% \cite{loh2005living} is first LPI paper as best I know
+% \cite{collen2009monitoring,ledger2023past,leung2022reply,puurtinen2022living,toszogyova2024mathematical} to review!
+As ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction (CITES).
+The Living Planet Index (LPI), which aims to include long-term data on populations of species across the globe is emblematic of these debates (which also include whether insects are declining and ...
+CITES).
+With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline (CITES), with each paper taking a slightly different analytical approach.
+For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by \citet{Buschke2021} who focused on adding random population variation.
+Recently, \citep{Johnson2024} suggested a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results.
+These apparently conflicting results have made it hard for policy-makers to do advance initiatives aimed at slowing declines, and has led to a debates within ecology about whether such analyses undermine public confidence in science  \citep{gonzalez2016estimating}.
+While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.
+In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their model with simulated data, or their model performance fit to empirical data.
+% While this debate captures part of the process of science---refining our approaches and thus estimates over time---it also makes it hard for policy-makers to do (??
+something) and has led to a debates within ecology about whether such analyses undermine public confidence in science  (CITES Biodiversity debate).
+
+We argue than an improved workflow that required retrodictive checks and data simulation would lead to larger model advances and highlight consistency in estimates across models, which could better aid policy.
+This is because the workflow would make uncertainty in each trend estimate from each unique model more obvious.
+This would make what now appear as major discrepancies  more obviously shifts in point estimates that are generally all in the same uncertainty space---and because it would challenge modelers to show major predictive advances, which is not currently part of the process.
+Explanatory power in most models of observational data is usually very low (CITES) and thus tests of models' predictions rarely expected.
+But the workflow highlights that predictions from the model---what we call retrodictive checks (or whatever we call them)--are part of the process of science, and critical to testing for what may be missing in a model.
+We expect retrodictive checks on most published trend analyses would highlight major missing components in these models (expand here??
+ADD example?), and drive changes both in the models themselves and in the simulated data to check the models.
+% Mention how low R2 are in ecology?
+
+% Note to self for Lizzie: START HERE! And no reading/reviewing papers until more text is written...
+Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models 
+% Mention elephants rebounding...
+Indeed, entire papers are devoted to what we consider this one step (data simulation) ....
+% I need to review these papers! \cite{hourdin2023toward,mcrae2025utility} ... also \cite{Buschke2021} uses some simulations
 
@@
 % If we do not take care to apply this workflow, machine learning might be the more pragmatic approach...
+
+% Misc possibly useful citations -- the GCM people seem to talk a lot about their issues lately on tuning etc. Maybe worth citing?
+% \cite{balaji2022general}mAre general circulation models obsolete? -- covers a lot and good citations within I think
+% \cite{simpson2025confronting} this is the 'Confronting Earth System Model trends with observations' which I have not fully read but Table 1 is interesting to me and something maybe the PBM and NTIF (near-term iterative) people should be doing if they are not doing?
+% \cite{rounce2020quantifying} 'Quantifying parameter uncertainty in a large-scale glacier evolution model using Bayesian inference: application to High Mountain Asia' talks about identifablity problems and uses Bayesian, again I have not fully read, but may be interesting example?
 

### 2025-03-31 forecastflows.tex
Finished first draft of the trends section

--- 2025-03-29 forecastflows.tex
@@
 % Trends - outline current problem
-[...]
+%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology.
+Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \cite{soule1999conserving,soule1991conservation}.
+The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}.
+Such trends drove a number of new subfields within ecology---some of which are now complete disciplines within themselves \citep[such as conservation biology,][]{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES.
+Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity.
+
+% \cite{loh2005living} is first LPI paper as best I know
+% \cite{collen2009monitoring,ledger2023past,leung2022reply,puurtinen2022living,toszogyova2024mathematical} to review!
+Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
+The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these debates.
+With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
+For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}.
+
+Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines, and has led to a debates within ecology about whether such analyses undermine public confidence in science  \citep{gonzalez2016estimating}.
+While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.
+In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data.
+Functionally, research on the LPI has somewhat reverse-engineered the recommended workflow: after a series of papers debating different estimates from different models, more recent papers have focused on simulated data to highlight uncertainty given the model and data togethers \citep[though I don't think they link their simulations to the model they use that well,][]{dove2023quantifying,toszogyova2024mathematical}, but this should have been part of the process for the very first papers.
+% While this debate captures part of the process of science---refining our approaches and thus estimates over time---it also makes it hard for policy-makers to do (??
+something) and has led to a debates within ecology about whether such analyses undermine public confidence in science  (CITES Biodiversity debate).
+% \citet{Buschke2021} who focused on adding random population variation
+
+We argue than an improved workflow that required retrodictive checks and data simulation would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
+Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process.
+Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected.
+But the workflow highlights that predictions from the model---what we call retrodictive checks (or whatever we call them)---are part of the process of science, and critical to testing for what may be missing in a model.
+We expect retrodictive checks on most published trend analyses would highlight major missing components in these models (expand here??
+ADD example?), and drive changes both in the models themselves and in the simulated data to check the models.
+While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility
+% Mention how low R2 are in ecology?
+% Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models and in turn debate them.. 
+% Mention elephants rebounding...
 
 % Trends - improvements!
-[...]
+Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates.
+Returning to the example of a global estimate of trends in vertebrate populations of species over time and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
+Or a need to also find which species are declining most, including those that may have poor or no data?
+From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data.
+Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model.
+For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled.
+From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematicall}.
 

### 2025-04-01 forecastflows.tex
new additions from today huuuuge effort

@@
 Finally, the resulting experimental data should be used to inform our model and differentiate between hypotheses---hopefully answering our initial question.
-However, this is often an idealized view and does not reflect the complexity of ecology \citep{}.
+However, this is often an idealized view.
+% and does not reflect the complexity of ecology \citep{}. I already say it below
 
@@
 But the reality of ecological research is also a major driver of the divergences from the ideal scientific method.
-Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective.
+Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective \citep{Hilborn1997}.
 % I feel something is missing
@@
 % There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
-% Moving along the data-model space
+% Moving along the data-model space, step-by-step
 % \item stressing the need to think about model before study design
-% \item advancing data simulation
-[...]
+% \item 
+To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed to move along the data-model space, following a coherent sequence of steps.
+In particular, more efforts should be placed on model evaluation before incorporating any real data.
+This would force the modeler to acknowledge that some parameters might be non-identifiable and to reconsider the model structure.
+Similarly, it is essential to assess whether model predictions---once parameters are informed by data---are consistent with observations.
+The strength of such workflow lies in its flexibility, making it applicable to a wide range of modeling approaches, from simple trend analyses to more complex process-based models.
+At each step, the modeler need to critically examine its understanding of ecological processes, questioning previous assumptions, and explicitly acknowledge sources of uncertainty.
+This approach has the potential to enhance model interpretability and allow for a more transparent evaluation of model strengths and limitations.
+It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
 
 % More details on the workflow:  walk through the different steps
-[...]
+% the paragraph is a bit loooong
+The first step of the workflow is to define an explicit research question and formulate some hypotheses.
+This involves making assumptions about the most influential drivers, within the specific context of our study.
+This step help us to think about the mechanisms that could generate the data we observe, including the observational error.
+Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question.
+The general idea is to start with a relatively simple model, that we could refine later.
+At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
+The next step is to simulate data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable).
+We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
+Some issues might arise at this step.
+Structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable.
+This is the first feedback loop: if the model is not able to recover known parameter values, we must reconsider its structure.
+This way, we ensure the model works well on simulated data before incorporating real observations.
+We can then fit it to real data, to obtain parameter estimates constrained by observations.
+Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question.
+This could lead us to either simplify our research question or---ideally---launch new data collection efforts.
+The next step is a new data simulation step, this time using our fitted model to generate predictions.
+This retrodictive check allows model output to be compared to observations.
+Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic.
+This is a second feedback loop.
+We can refine the model to integrate the missing process and restart the workflow.
+Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake input data, such as phylogenetic structure or observational biases (e.g.
+unbalanced data).
+This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics.
+At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.
+Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
 
@@
 %  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology.
-Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \cite{soule1999conserving,soule1991conservation}.
+Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \citep{soule1999conserving,soule1991conservation}.
 The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}.
@@
 For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled.
-From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematicall}.
+From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.
 
 % Forecasting - outline current problem
-Forecasting is a broad field with a diverse range of methods.
-Here, we focus on process-based modeling, which is often considered the gold standard in ecology and beyond.
+Ecological forecasting is a broad field with a diverse range of methods.
+Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond.
 Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses.
@@
 Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective.
-The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration, where parameter values are inferred.
+The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred.
 Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes.
 Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e.
-fix to a value found in the literature).
+fix to a value found in the literature) or tuned to reproduce some observations or theory.
+The way models are currently calibrated is not accidental, but rather a constraint imposed by their structure, where many parameters compensate for one another.
+
+These limitations are central to current debates.
+In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty.
+For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{}.
+This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}.
+Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{}.
+A promising avenue is the simplification of models, avoiding over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
+If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
+Each additional process and parameter can increase overall uncertainty to the point where predictions lose their usefulness for decision makers \citep{Saltelli2020}.
 
@@
 This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model.
-Data fitting would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights.
+% => quantifying parameter uncertainty
+% incorporate additional observations and their respective uncertainties
+Model calibration would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights. 
+% It would help properly take into account for any issues regarding non-identifiability.
 The workflow could also refocus attention on the research question, defining a clear and limited context in which the model should apply.
 Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
-Finally, an universal workflow provides an opportunity to merge statistical and process-based approaches, integrating mechanistic knowledge and leveraging robust statistical approaches.
+An universal workflow provides an opportunity to merge statistical and process-based approaches, integrating mechanistic knowledge and leveraging robust statistical approaches.
 
@@
 The misuse of models and misinterpretation of their outcomes (including by the authors) have multiple origins.
-One cause, almost tautological, is the pressure to publish academic work, which can lower research standards.
+One cause, well known, is the pressure to publish academic work, which can lower research standards.
 This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
@@
 A significant portion of scientific debate thus becomes lost in methodological considerations rather than advancing our ecological understanding.
+
+% better training! on BOTH estimation AND prediction
+% learn how to simulate data will allow students to understad what is the link between the maths (eg a distribution) and the ecological processes that could be a l'oeuvre to generate the observations, make better links bewteen disciplines
+% \item estimation: being aware of what is a parameter, and mention uncertainty propagation
+% \item forecasting: should be a natural outcome, not a finality
+
+% \item ML (short-time forecast), benchmarking models are probably useful but should not be the core of our scientific practice, not the spirit!\\
+% $\rightarrow$ moving ecology in the right direction!
 

### 2025-04-01 forecastflows.tex
add some refs

@@
 % \item 
-To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed to move along the data-model space, following a coherent sequence of steps.
+To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed to move along the data-model space, following a coherent sequence of steps (Fig 1).
 In particular, more efforts should be placed on model evaluation before incorporating any real data.
@@
 It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
+% I think we would need to explain more this idea about parameters...
+of keep it for later
 
@@
 % the paragraph is a bit loooong
-The first step of the workflow is to define an explicit research question and formulate some hypotheses.
+The first step of the workflow is to define an explicit research question and formulate some hypotheses (Fig 2).
 This involves making assumptions about the most influential drivers, within the specific context of our study.
@@
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
-Some issues might arise at this step.
+% Some issues might arise at this step
 Structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable.
@@
 This way, we ensure the model works well on simulated data before incorporating real observations.
-We can then fit it to real data, to obtain parameter estimates constrained by observations.
+We can then fit the model to real data, to obtain parameter estimates constrained by observations.
 Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question.
@@
 We can refine the model to integrate the missing process and restart the workflow.
-Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake input data, such as phylogenetic structure or observational biases (e.g.
+Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g.
 unbalanced data).
@@
 fix to a value found in the literature) or tuned to reproduce some observations or theory.
-The way models are currently calibrated is not accidental, but rather a constraint imposed by their structure, where many parameters compensate for one another.
+The way models are currently calibrated is not a coincidence, % is coincidence the right word?
+but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
 
@@
 In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty.
-For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{}.
+For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
 This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}.
-Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{}.
-A promising avenue is the simplification of models, avoiding over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
+Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{Cheaib2012}.
+Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
 If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.

### 2025-04-02 forecastflows.tex
forgot to push last change yesterday

@@
 If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
-Each additional process and parameter can increase overall uncertainty to the point where predictions lose their usefulness for decision makers \citep{Saltelli2020}.
+Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
 
@@
 Model calibration would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights. 
-% It would help properly take into account for any issues regarding non-identifiability.
+% It would help properly take into account for any issues regarding non-identifiability, and the model would be tractable! I think it's a word
 The workflow could also refocus attention on the research question, defining a clear and limited context in which the model should apply.
@@
 % better training! on BOTH estimation AND prediction
-% learn how to simulate data will allow students to understad what is the link between the maths (eg a distribution) and the ecological processes that could be a l'oeuvre to generate the observations, make better links bewteen disciplines
+% learn how to simulate data will allow students to understand what is the link between the maths (e.g. a distribution) and the ecological processes generating the observations 
+% mobilizing their ecological knowledge while thinking of a model
+% => foster stronger links between disciplines
 % \item estimation: being aware of what is a parameter, and mention uncertainty propagation

### 2025-04-02 forecastflows.tex
Some text for the end but not complete yet.

@@
 
-The misuse of models and misinterpretation of their outcomes (including by the authors) have multiple origins.
-One cause, well known, is the pressure to publish academic work, which can lower research standards.
-This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
-This a growing concern, leading to increase in reproducibility and data sharing practices.
-However, model development in ecology is still rarely transparent, which limits model understandability and prevents peers from properly identifying potential issues.
-A significant portion of scientific debate thus becomes lost in methodological considerations rather than advancing our ecological understanding.
+We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science. 
+One well-known hurdle is that the pressure to publish academic work, which can lower research standards and make the added effort of this workflow seem ill-placed.
+This may be especially true for those who see science rewarded mostly through the shear number of publications.
+But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help.
+Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).
+
+% This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
+Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues.
+Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations, but we believe our workflow provides an tractable step to fixing this.
+By focusing on model development more tightly tied to ecological expertise, we ague this workflow should broader the community that contributes to model development.
+As ecologists are increasingly expanding their computational toolkits, many field, and lab and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological, and---most importantly---to interrogate them.
+
+Advancing ecology to where most researchers see models built more flexibly from their understanding of their ecological systems will not happen rapidly without a major shift in training, however.
+Much of ecology still divides the world into training for those focused on several groups.
+First, those who will conduct field and lab studies often learn a suite of particular tests matched to particular experiment designs or to simply match their predictor and response data to a flow diagram of variable types (add EXAMPLE and reference, something like: categorical $y$ and binary $y$ means ...
+I dunno, chi square?) and are left adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is.
+Instead, they are expected to collaborate with others when they need more complex models, which is a second group of ecological researchers---those who focus on complex statistical models and were thus trained more in model development, but often for highly specific applications (e.g., wildlife population estimates) where they may rarely build totally new generative models.
+Separate from these groups, process-based modelers do their own thing.
+Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.
+While specialization is valuable, we argue the fundamental training in ecology has overly-siloed these groups and prevented more rapid progress.
+
+Training all ecologists in our proposed workflow would break down barriers between different groups of ecologists today with likely major benefits for science.
+Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in our workflow would focus on learning to generate questions and then models, and then to simulate data from them.
+This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them.
+Through this and retrodictive checks most ecologists would more easily think through what parameters are most critical to their question and or aim (e.g., management) and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
+Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models.
+Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting.
+% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
+Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches to better forecasts or something.
+
+This type of workflow-focused training would make forecasting from models far more tractable ...
 

### 2025-04-03 forecastflows.tex
writing a new paragraph for process-based forecasting

@@
 
-% Forecasting - potential improvments (do I say too much ideas)
+% Forecasting - potential improvments 
+% I was saying too much in one paragraph, so I cut in two
 Applying the workflow to process-based models is a key for opening the black box.
@@
 This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model.
-% => quantifying parameter uncertainty
-% incorporate additional observations and their respective uncertainties
 Model calibration would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights.
-% It would help properly take into account for any issues regarding non-identifiability, and the model would be tractable! I think it's a word
-The workflow could also refocus attention on the research question, defining a clear and limited context in which the model should apply.
+It would help properly take into account any issues regarding non-identifiability.
+This could involve reformulating the mathematical structure of some processes, making new hypotheses to target the right level of complexity, or incorporating more expert knowledge to better constrain calibration.
+Making the model more tractable would also naturally facilitate the recognition and quantification of parameter uncertainty, as well as its propagation into model projections
+
+Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
+The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model.
+It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.
+% or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza!
+disease!
+pathogens!)...
+or "rather than constantly expanding models by adding ever more processes"?
 Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
-An universal workflow provides an opportunity to merge statistical and process-based approaches, integrating mechanistic knowledge and leveraging robust statistical approaches.
+An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches. 
+Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty.
+Instead, it would guide a step-by-step model fitting, parameter identification, and uncertainty quantification.
+Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
+This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference.
+However, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.
+And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
 
@@
 Instead, they are expected to collaborate with others when they need more complex models, which is a second group of ecological researchers---those who focus on complex statistical models and were thus trained more in model development, but often for highly specific applications (e.g., wildlife population estimates) where they may rarely build totally new generative models.
-Separate from these groups, process-based modelers do their own thing.
+Separate from these groups, process-based modelers
+% ---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
+% hmm some people talk about simulation models (for process-based models), I agree but I'm afraid in would not be clear with the data simulation we try to highlight
+%do their own thing. 
 Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.

### 2025-04-03 forecastflows.tex
adding some refs

@@
 This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}.
-Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{Cheaib2012}.
+Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
+These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
 Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
@@
 Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
-An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches. 
+An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
 Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty.

### 2025-04-04 forecastflows.tex
start incorporating Lizie feedback

@@
 % Ecology has become super challenged to predict stuff for decision making
-Nature is increasingly threatened by multiple drivers of change, with a largely dominant influence of human activities \citep{Diaz2019}.
-This ongoing biodiversity crisis is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}.
-To support implementation of sustainable policies among the socioeconomic and environmental dimensions, it is critical to understand trends to date and be able to forecast future dynamics.
+Nature is increasingly threatened by multiple, mainly human-driven, drivers of change \citep{Diaz2019}.
+The ongoing biodiversity loss is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}.
+To support implementation of sustainable policies, it is critical to understand trends to date and be able to forecast future dynamics.
 % $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...
@@
 % General way to do this so far... (bifurcated?)
-Estimation of global biodiversity indicators and current trends depends on large-scale and long-term datasets---across terrestrial, freshwater, and marine ecosystems \citep[e.g.][]{Dornelas2018}.
-These data, gathered opportunistically and from multiple sources, are often unbalanced and have geographic, temporal and taxonomic biases.
+In an era of rapidly expanding big data, robust models are essential both to estimate biodiversity trends and to forecast future changes.
+Estimating global biodiversity trends depends on large-scale and long-term datasets \citep[e.g.][]{Dornelas2018}.
+These data, gathered opportunistically from multiple sources, are often unbalanced with massive geographic, temporal and taxonomic biases.
 Addressing these biases requires the use of appropriate statistical inference.
@@
 The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022}.
-% ...
-lacks something
 
 % Gap, problem: None of this is going well
-The urgent need to answer policy-relevant questions has favored the proliferation of diverse methods developed by different researchers, lacking an overall coherence.
-Though there is no doubt nature is declining globally, significant uncertainty remains.
-There is no consensus on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}.
+The current workflow has not led to consensus, and significant uncertainty remains---undermining the policy relevance of current results.
+% need a better topic sentence, as it is still a bit similar to topic sentence of the next paragraph?
+There is no clear agreement on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}.
 Future projections also diverge considerably, due to a high model uncertainty at the ecological level.
@@
 Workflows should fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections.
-Here, we introduce an universal workflow that proposes to iteratively build upon all these steps, harmonizing both trend estimation and forecasting, with the aim of refocusing the debate on ecological questions and increasing the speed of scientific progress.
+In particular, we should incorporate data simulation early in the workflow to identify flaws in model structure and constraints in data.
+Here, we introduce a universal workflow that iteratively builds upon all these steps.
+We argue it would harmonize both trend estimation and forecasting, with the aim of refocusing the debate on ecological % and conservation?
+questions and increasing the speed of scientific progress.
 
@@
 \noindent \textbf{Scientific method and workflows}
-%  too much blah blah blah?
 
 % General scientific method we all learn
-Quantitative science rely on a model-based framework to confront hypothesis and data, making some approximations \citep{}.
-The general scientific method stresses out that the research question should guide both the design of the experiment and the corresponding model building.
-The experiment should then be conducted---according to this specific design.
-Finally, the resulting experimental data should be used to inform our model and differentiate between hypotheses---hopefully answering our initial question.
-However, this is often an idealized view.
-% and does not reflect the complexity of ecology \citep{}. I already say it below
+Quantitative science relies on a model-based framework, to confront hypothesis and data, making some approximations \citep{}.
+The general scientific method stresses we should formulate a research question, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses.
+However, this idealized scientific method often does not apply to the reality of ecological research.
+Many important questions cannot be addressed through controlled experiments and replications.
+Instead, we must often rely on existing, heterogeneous datasets to have a large-scale and long-term perspective \citep{Hilborn1997}.
 
-% Divergences from this are common 
-Divergences from this scientific method are common.
-One explanation is the lack of rigor, leading to questionable practices, such as retrospectively crafting hypothesis to explain the results of a model rather than testing a clear pre-defined question (data-driven analysis, Fig.
-1).
-But the reality of ecological research is also a major driver of the divergences from the ideal scientific method.
-Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective \citep{Hilborn1997}.
-% I feel something is missing
-However, this does not explain the persistent flaws and lack of overall coherence in ecological modeling.
-Trend estimation still often relies on a one-way 'inference',  % need a better word?
-to stress out that there is no feedback, eah paper applied a slightly different model
+% I can't remember precisely what we say about this paragraph
+% Should it be the first paragraph of section 'The workflow in practice'? But I think it's nice to mention each part of the Fig. 1 here
+This reality should drive researchers to use more robust and coherent methods, but this is rather leading to persistent flaws. 
+Trend estimation still often relies on a one-way 'inference',  % need a better word? to stress out that there is no feedback, eah paper applied a slightly different model 
 and most of the time is spent fitting the model to empirical data (Fig.
@@
 The different parts of the model are often calibrated separately rather than as a whole, and some parameter values are just fixed based on experiments and expert knowledge.
+% should we keep the sentence below
+The lack of rigor can also lead to questionable practices---such as retrospectively crafting hypotheses to explain the results of a model rather than testing a clear predefined question (data-driven analysis, Fig.
+1).
+% mention the disconnection between trends and forecasts?
 
 % There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
-% Moving along the data-model space, step-by-step
-% \item stressing the need to think about model before study design
-% \item 
-To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed to move along the data-model space, following a coherent sequence of steps (Fig 1).
+% foreshadow stuf about simulation and feedback loops
+To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed that moves along the data-model space, in a coherent sequence of steps (Fig 1).
 In particular, more efforts should be placed on model evaluation before incorporating any real data.
@@
 It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
-% I think we would need to explain more this idea about parameters...
-of keep it for later
 
 % More details on the workflow:  walk through the different steps
-% the paragraph is a bit loooong
 The first step of the workflow is to define an explicit research question and formulate some hypotheses (Fig 2).
@@
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
-The next step is to simulate data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable).
+The next step is the first data simulation step.
+The idea is to craft some fake data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable).
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
-% Some issues might arise at this step
-Structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable.
-This is the first feedback loop: if the model is not able to recover known parameter values, we must reconsider its structure.
-This way, we ensure the model works well on simulated data before incorporating real observations.
-We can then fit the model to real data, to obtain parameter estimates constrained by observations.
+Once we are confident about our model structure, we can incorporate real data.
+This way, we obtain parameter estimates constrained by observations.
 Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question.
 This could lead us to either simplify our research question or---ideally---launch new data collection efforts.
-The next step is a new data simulation step, this time using our fitted model to generate predictions.
+This lead to the second data simulation step, this time using our fitted model to generate predictions.
 This retrodictive check allows model output to be compared to observations.
+% adding something about parameters
+Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
+
+% More details, what's special! Feedbacks!
+A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
+The first feedback arises when we evaluate the model on simulated data: structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable.
+If the model fails to recover known parameter values, this flags % not sure about this word
+the need to reconsider its structure---before incorporating real observations. 
+The second feedback loop comes from the retrodictive check. 
 Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic.
-This is a second feedback loop.
 We can refine the model to integrate the missing process and restart the workflow.
@@
 At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.
-Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
 
@@
 \noindent \textbf{The workflow in practice}
+
 % How to address current issues (two case studies)
+% Need an introduction here
+Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow holds the promise to highlight the opportunities to best reduce uncertainties through new scientific insights, toward the most critical steps.
+This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.
 
@@
 %\item (ML has a way to collect and interpret large datasets...)
-Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow holds the promise to highlight the opportunities to best reduce uncertainties through new scientific insights, toward the most critical steps.
-This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.
+
 In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made.

### 2025-04-05 forecastflows.tex
continue, small modifications

@@
 
-\noindent \textbf{Introduction}
-% Still need some work...
+\section{Introduction}
 
 % Ecology has become super challenged to predict stuff for decision making
-Nature is increasingly threatened by multiple, mainly human-driven, drivers of change \citep{Diaz2019}.
+Nature is increasingly threatened by multiple, mainly anthropogenic, drivers of change \citep{Diaz2019}.
 The ongoing biodiversity loss is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}.
@@
 % General way to do this so far... (bifurcated?)
-In an era of rapidly expanding big data, robust models are essential both to estimate biodiversity trends and to forecast future changes.
+% the topic sentence is really a repetition of previous sentence, that's baaad
+In an era of rapidly expanding big data, robust models are essential for trend estimation and forecasting.
 Estimating global biodiversity trends depends on large-scale and long-term datasets \citep[e.g.][]{Dornelas2018}.
@@
 
-\vspace{0.5cm}
-\noindent \textbf{Scientific method and workflows}
+\section{Scientific method and workflows}
 
@@
 The next step is the first data simulation step.
-The idea is to craft some fake data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable).
+The idea is to craft some fake data that reflect our assumptions about the data structure and biases, by fixing parameters to some values (which is straightforward if the parameters are interpretable).
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
@@
 
-\vspace{0.5cm}
-\noindent \textbf{The workflow in practice}
+\section{The workflow in practice}
 
@@
 This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.
+Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.
+
+\subsection{Trends}
+% emw03april: could shorten the first 3 paragraphs, move some text to a box?
 
@@
 
+\subsection{Forecasting}
+
 % Forecasting - outline current problem
+% emw03april: move also this to a box?
 Ecological forecasting is a broad field with a diverse range of methods.
@@
 Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
-An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
 Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty.
-Instead, it would guide a step-by-step model fitting, parameter identification, and uncertainty quantification.
-Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
+Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.
 This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference.
-However, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.
-And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
-
-% Step back
-%\item we need more data, and better question (relate this to both previous study cases)
-%\item where can we best reduce uncertainties through new scientific insights?
-%\item machine learning!
-If we change nothing, what's the point of not doing ML?
-ML $>$ process-based without question, and ML $>$ trends without mechanisms!
-Where theory is lacking, or where we are far from mechanistic understanding, you might as well do ML!
-%\item (ML has a way to collect and interpret large datasets...)
-
-In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made.
-Machine learning may surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
-Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
-% If we do not take care to apply this workflow, machine learning might be the more pragmatic approach...
+But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.
 
@@
 
-\vspace{0.5cm}
-\noindent \textbf{Wrap up: how to make it happen?}
+\section{Bareers and opportunities}
 
@@
 % This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
+% emw03april: move this somewhere else?
 Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues.
@@
 
+An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
+Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
+And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
+In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made.
+Machine learning may surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
+Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
+
+% emw03april: massively shorten?
 Advancing ecology to where most researchers see models built more flexibly from their understanding of their ecological systems will not happen rapidly without a major shift in training, however.
@@
 Instead, they are expected to collaborate with others when they need more complex models, which is a second group of ecological researchers---those who focus on complex statistical models and were thus trained more in model development, but often for highly specific applications (e.g., wildlife population estimates) where they may rarely build totally new generative models.
-Separate from these groups, process-based modelers
-% ---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
+Separate from these groups, process-based modelers do their own thing.
+% process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
 % hmm some people talk about simulation models (for process-based models), I agree but I'm afraid in would not be clear with the data simulation we try to highlight
-%do their own thing. 
+%
 Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.

### 2025-04-06 forecastflows.tex
Some edits but stopping at the case studies for now

@@
 % Word length: I should check what they emailed but they say no more than 13 printed pages with 650 per page (whoa, we should be well below that! I am thinking 3-5K would be good)
+% Already 5K!
+
+% To do:
+% Fix the abstract so it matches the paper better (I think mainly the last few sentences need work, or we just need to say more about data sharing in the text)
+% Add in Figures
+% Delete out unnecessary comments so easier for co-authors to read?
 
@@
 Relevant datasets to meet these demands have also increased in size and complexity, including in their geographical, temporal and phylogenetic scales.
-While new research often suggests that accounting for these complexities variously increases, removes or otherwise alters major trends, I argue that the fundamental approach to model fitting in ecology makes it impossible to evaluate and compare models.
-These problems stem in part from continuing gaps between statistical workflows -- where the data processing and model development are often addressed separately from the ecological question and aim -- and scientific workflows, where all steps are integrated.
-Yet, as ecologists become increasingly computational, and new tools make it easier to share data, the opportunity to close this gap has never been greater.
-I outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights.
-Combining these changes with more open model and data sharing -- and developing new efforts to race the same data -- could be transformative for ecological forecasting.
+While new research often suggests that accounting for these complexities variously increases, removes or otherwise alters major trends, we argue that the fundamental approach to model fitting in ecology makes it impossible to evaluate and compare models.
+These problems stem in part from continuing gaps between statistical workflows---where the data processing and model development are often addressed separately from the ecological question and aim---and scientific workflows, where all steps are integrated.
+Yet, as ecologists become increasingly computational the opportunity to close this gap has never been greater.
+We outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights.
+Combining these changes with more open model and data sharing---and developing new efforts to race the same data---could be transformative for ecological forecasting.
 \end{abstract}
@@
 % Ecology has become super challenged to predict stuff for decision making
-Nature is increasingly threatened by multiple, mainly anthropogenic, drivers of change \citep{Diaz2019}.
-The ongoing biodiversity loss is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}.
-To support implementation of sustainable policies, it is critical to understand trends to date and be able to forecast future dynamics.
+Anthropogenic drivers are reshaping natural systems, with impacts expected to increase in coming decades \citep{Diaz2019}.
+Climate change will likely accelerate biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}.
+Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics.
+% These requirements in turn have created a growing need for more global ecological data, and the types of models that can robustly handle such data.
 % $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...
@@
 % General way to do this so far... (bifurcated?)
-% the topic sentence is really a repetition of previous sentence, that's baaad
-In an era of rapidly expanding big data, robust models are essential for trend estimation and forecasting.
-Estimating global biodiversity trends depends on large-scale and long-term datasets \citep[e.g.][]{Dornelas2018}.
-These data, gathered opportunistically from multiple sources, are often unbalanced with massive geographic, temporal and taxonomic biases.
-Addressing these biases requires the use of appropriate statistical inference.
-Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models \citep{IPBES2019}.
-The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022}.
+Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets and another focused on forecasting from generally distinct datasets or mechanistic models based on less data.
+Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced with massive geographic, temporal and taxonomic biases.
+Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
+Instead forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative  or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning.
+The current outcome from these approaches is no clear agreement on current species trends, with ongoing debates on the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and forecasts that diverge due to high model uncertainty at the ecological level.
 
-% Gap, problem: None of this is going well
-The current workflow has not led to consensus, and significant uncertainty remains---undermining the policy relevance of current results.
-% need a better topic sentence, as it is still a bit similar to topic sentence of the next paragraph?
-There is no clear agreement on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}.
-Future projections also diverge considerably, due to a high model uncertainty at the ecological level.
-Predictive modeling is increasingly relying on overly complex models (with a huge number of parameters), making it less adequate to generate new scientific insights \citep{Franklin2020}.
+% This appears to be only increasing in newer models that generally have more complexity (and an ever-increasing number of parameters), making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.
 
@@
 % Here we introduce a universal workflow to adress this...
-Current controversies and focus on methodological aspects stem from the lack of coherence between current workflows in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
-Each new model development is added as a separate layer, disconnected from the original research aim, the data stream and the previous scientific insights.
-Workflows should fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections.
-In particular, we should incorporate data simulation early in the workflow to identify flaws in model structure and constraints in data.
-Here, we introduce a universal workflow that iteratively builds upon all these steps.
-We argue it would harmonize both trend estimation and forecasting, with the aim of refocusing the debate on ecological % and conservation?
-questions and increasing the speed of scientific progress.
+We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
+Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting.
+At the same time, forecasting-focused process-based models often develop by adding a new separate layer, disconnected from the original research aim, the data stream and the previous scientific insights, because they fail to examine the model as a functioning whole and thus ignore major degeneracies.
+
+Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections, could reduce many of these problems.
+In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data.
+Towards this aim, here we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
+% with the aim of refocusing the debate on ecological questions and increasing the speed of scientific progress.
 
@@
 % General scientific method we all learn
-Quantitative science relies on a model-based framework, to confront hypothesis and data, making some approximations \citep{}.
-The general scientific method stresses we should formulate a research question, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses.
-However, this idealized scientific method often does not apply to the reality of ecological research.
+Quantitative science relies on a model-based framework, to confront hypothesis and data \citep{}.
+In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses.
+But this idealized method often does not apply to the reality of ecological research.
 Many important questions cannot be addressed through controlled experiments and replications.
-Instead, we must often rely on existing, heterogeneous datasets to have a large-scale and long-term perspective \citep{Hilborn1997}.
+Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.
 
@@
 % Should it be the first paragraph of section 'The workflow in practice'? But I think it's nice to mention each part of the Fig. 1 here
-This reality should drive researchers to use more robust and coherent methods, but this is rather leading to persistent flaws. 
-Trend estimation still often relies on a one-way 'inference',  % need a better word? to stress out that there is no feedback, eah paper applied a slightly different model 
-and most of the time is spent fitting the model to empirical data (Fig.
+This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead appear to be leading to persistent problems.
+% Trend estimation still relies on a one-way `inference',  and most of the time is spent fitting the model to empirical data (Fig. 1).
+Trend estimation has focused mostly on fitting the model to empirical data in a uni-directional way that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig.
 1).
-For forecasting, researchers focus on making predictions with complex models, but the steps of model building and parameterization are not very transparent and not clearly delineated (Fig.
+For forecasting, researchers focus on making predictions with increasingly complex models, making the steps for model building and parameterization increasingly obscured (Fig.
 1).
-The different parts of the model are often calibrated separately rather than as a whole, and some parameter values are just fixed based on experiments and expert knowledge.
-% should we keep the sentence below
-The lack of rigor can also lead to questionable practices---such as retrospectively crafting hypotheses to explain the results of a model rather than testing a clear predefined question (data-driven analysis, Fig.
+Today, researchers often calibrate different parts of the model separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
+Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.
+% should we keep the sentence below % %emw6Apr: I suggest we delete. 
+% The lack of rigor can also lead to questionable practices---such as retrospectively crafting hypotheses to explain the results of a model rather than testing a clear predefined question (data-driven analysis, Fig.
 1).
@@
 
-% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
+% There is thus room for improvment, with a workflow that addresses these flaws while taking account the reality of working with ecological data.
 % foreshadow stuf about simulation and feedback loops
-To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed that moves along the data-model space, in a coherent sequence of steps (Fig 1).
-In particular, more efforts should be placed on model evaluation before incorporating any real data.
+We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig 1) could reduce many of these problems and thus improve ecological science.
+%emw6Apr: I wonder if we can see how it goes to just cut this?
+We can see what co-authors say and add it back in later ...
+I sometimes start a file to toss this type of text (e.g., forecastflows_extras.tex) in case I need it but keep this file cleaner for co-authors.
+% In particular, more efforts should be placed on model evaluation before incorporating any real data.
 This would force the modeler to acknowledge that some parameters might be non-identifiable and to reconsider the model structure.
@@
 It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
-
-% More details on the workflow:  walk through the different steps
-The first step of the workflow is to define an explicit research question and formulate some hypotheses (Fig 2).
-This involves making assumptions about the most influential drivers, within the specific context of our study.
+The first step of the workflow is to define an explicit research question and formulate hypotheses (Fig 2).
+This involves making clear assumptions about the most influential drivers, within the specific context of our study.
 This step help us to think about the mechanisms that could generate the data we observe, including the observational error.
@@
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
-The next step is the first data simulation step.
-The idea is to craft some fake data that reflect our assumptions about the data structure and biases, by fixing parameters to some values (which is straightforward if the parameters are interpretable).
+
+With a model in place the next step focuses on testing and understanding it via data simulation.
+`Fake' or `test' data are generated directly from the model by fixing parameters to some values (which is straightforward if the parameters are interpretable) and from fake predictor data.
+This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model.
+For example, if a researcher realizes their empirical is geographically biased, that should be built into the model and thus then into this data simulation step.
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. 
@@
 This could lead us to either simplify our research question or---ideally---launch new data collection efforts.
-This lead to the second data simulation step, this time using our fitted model to generate predictions.
+This leads to the second data simulation step, this time using our fitted model parameters to generate predictions.
 This retrodictive check allows model output to be compared to observations.
@@
 % More details, what's special! Feedbacks!
+%emw6Apr: I think there's more than just degeneracy here, often people build in more complexity to their simulated data than their model allows, but this means they should change the model (or the question).
+This needs to be clearer above and below.
+I think this is the first point (since it's a little more obvious to readers and it relates to trends, which we always handle first), and the next one could be about structural degeneracies  but I think we need to be careful here in what we mean and explain it more to readers who will not get it.
 A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
@@
 % Need an introduction here
-Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow holds the promise to highlight the opportunities to best reduce uncertainties through new scientific insights, toward the most critical steps.
+%emw6Apr: I like the shift to 'parameter estimation' instead of trend estimation, but I wonder if we could make that shift clearer above.
+Maybe somewhere we say that the workflow would  encourage a focus on the full model and its parameters: For trend estimates, the trend becomes just one parameter that must be carefully interpreted alongside others such as the  global mean or intercept and the estimate variance.
+Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties through new scientific insights, toward the most critical steps.
 This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.
@@
 
-\section{Bareers and opportunities}
+\section{Barriers and opportunities}
 

### 2025-04-07 forecastflows.tex
Start moving text to a box

@@
 \usepackage{amsmath}
+\usepackage{multicol}
+\usepackage{tcolorbox}
+\usepackage{capt-of}
+
 \bibliographystyle{besjournals}
@@
 We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig 1) could reduce many of these problems and thus improve ecological science.
-%emw6Apr: I wonder if we can see how it goes to just cut this?
-We can see what co-authors say and add it back in later ...
-I sometimes start a file to toss this type of text (e.g., forecastflows_extras.tex) in case I need it but keep this file cleaner for co-authors.
-% In particular, more efforts should be placed on model evaluation before incorporating any real data.
-This would force the modeler to acknowledge that some parameters might be non-identifiable and to reconsider the model structure.
-Similarly, it is essential to assess whether model predictions---once parameters are informed by data---are consistent with observations.
-The strength of such workflow lies in its flexibility, making it applicable to a wide range of modeling approaches, from simple trend analyses to more complex process-based models.
-At each step, the modeler need to critically examine its understanding of ecological processes, questioning previous assumptions, and explicitly acknowledge sources of uncertainty.
-This approach has the potential to enhance model interpretability and allow for a more transparent evaluation of model strengths and limitations.
-It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
 The first step of the workflow is to define an explicit research question and formulate hypotheses (Fig 2).
@@
 This retrodictive check allows model output to be compared to observations.
-% adding something about parameters
+The workflow replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
+Any parameter (such as a trend estimate) must be carefully interpreted alongside others.
+%such as the  global mean or intercept and the estimate variance
 Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
@@
 I think this is the first point (since it's a little more obvious to readers and it relates to trends, which we always handle first), and the next one could be about structural degeneracies  but I think we need to be careful here in what we mean and explain it more to readers who will not get it.
-A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
-The first feedback arises when we evaluate the model on simulated data: structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable.
-If the model fails to recover known parameter values, this flags % not sure about this word
-the need to reconsider its structure---before incorporating real observations. 
+A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
+The failure of the model to handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
+Further, % structural degeneracies in the model might become apparent, 
+this step might reveal that some parameters are highly non-identifiable, % If the model fails to recover known parameter values, this flags % not sure about this word
+flagging the need to change the model structure---before incorporating real observations. 
 The second feedback loop comes from the retrodictive check. 
@@
 %emw6Apr: I like the shift to 'parameter estimation' instead of trend estimation, but I wonder if we could make that shift clearer above.
-Maybe somewhere we say that the workflow would  encourage a focus on the full model and its parameters: For trend estimates, the trend becomes just one parameter that must be carefully interpreted alongside others such as the  global mean or intercept and the estimate variance.
+Maybe somewhere we say that the workflow would  encourage a focus on the full model and its parameters: For trend estimates, the trend becomes just one parameter that must be carefully interpreted alongside others .
 Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties through new scientific insights, toward the most critical steps.
@@
 % emw03april: move also this to a box?
-Ecological forecasting is a broad field with a diverse range of methods.
-Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond.
-Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses.
-They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted.
-Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions.
-Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective.
-The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred.
-Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes.
-Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e.
-fix to a value found in the literature) or tuned to reproduce some observations or theory.
-The way models are currently calibrated is not a coincidence, % is coincidence the right word?
-but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
+Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. [...]
 
@@
 % \cite{rounce2020quantifying} 'Quantifying parameter uncertainty in a large-scale glacier evolution model using Bayesian inference: application to High Mountain Asia' talks about identifablity problems and uses Bayesian, again I have not fully read, but may be interesting example?
+
+% \setlength{\columnseprule}{0.1pt}
+
+\begin{tcolorbox}[
+sharp corners=all,
+colback=white,
+colframe=black,
+size=tight,
+boxrule=0.1mm,
+left=3mm,right=3mm, top = 2mm, bottom = 2mm,
+fontupper=\small
+]
+{\begin{multicols}{2}
+
+\centerline{\bf Trend estimation}
+
+\begin{minipage}[t]{\linewidth}
+    \includegraphics[width=\linewidth]{figures/trendestimation_details}
+
+    \vspace{-0.6cm}\captionof{figure}{Caption}\label{fig:trends}
+\end{minipage}
+\vspace{0.1cm}
+
+\noindent Blabla \newline
+Blabla \newline
+ \ \newline
+Blabla \newline
+\vfill
+
+\columnbreak
+
+\centerline{\bf Mechanistic forecasting}
+\bigskip
+
+\noindent Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses.
+They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted.
+Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions.
+Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective.
+The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred.
+Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes.
+Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e.
+fix to a value found in the literature) or tuned to reproduce some observations or theory.
+The way models are currently calibrated is not a coincidence, % is coincidence the right word?
+but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
+
+\vfill
+
+\end{multicols}}
+
+\centerline{\bf Common solution!}
+\vspace{0.2cm}
+\begin{minipage}[t]{\linewidth}
+    \includegraphics[width=\linewidth]{figures/iterativeworkflow_details}
+
+    \vspace{-0.6cm}\captionof{figure}{Caption}\label{fig:trends}
+\end{minipage}
+
+\end{tcolorbox}
 

### 2025-04-07 forecastflows.tex
start shrotening study cases, will need more reformulation

@@
 \usepackage{tcolorbox}
-\usepackage{capt-of}
+\usepackage{caption}
 
@@
 
-% Ecology has become super challenged to predict stuff for decision making
 Anthropogenic drivers are reshaping natural systems, with impacts expected to increase in coming decades \citep{Diaz2019}.
@@
 % These requirements in turn have created a growing need for more global ecological data, and the types of models that can robustly handle such data.
-% $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...
 
-% General way to do this so far... (bifurcated?)
 Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets and another focused on forecasting from generally distinct datasets or mechanistic models based on less data.
@@
 
-% This appears to be only increasing in newer models that generally have more complexity (and an ever-increasing number of parameters), making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.
-
-% Current workflows in ecology are not up to the task
-% Here we introduce a universal workflow to adress this...
 We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
@@
 Towards this aim, here we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
-% with the aim of refocusing the debate on ecological questions and increasing the speed of scientific progress.
 
@@
 
-% General scientific method we all learn
 Quantitative science relies on a model-based framework, to confront hypothesis and data \citep{}.
@@
 
-% I can't remember precisely what we say about this paragraph
-% Should it be the first paragraph of section 'The workflow in practice'? But I think it's nice to mention each part of the Fig. 1 here
 This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead appear to be leading to persistent problems.
-% Trend estimation still relies on a one-way `inference',  and most of the time is spent fitting the model to empirical data (Fig. 1).
 Trend estimation has focused mostly on fitting the model to empirical data in a uni-directional way that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig.
@@
 Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.
-% should we keep the sentence below % %emw6Apr: I suggest we delete. 
-% The lack of rigor can also lead to questionable practices---such as retrospectively crafting hypotheses to explain the results of a model rather than testing a clear predefined question (data-driven analysis, Fig.
-1).
 % mention the disconnection between trends and forecasts?
 
-% There is thus room for improvment, with a workflow that addresses these flaws while taking account the reality of working with ecological data.
-% foreshadow stuf about simulation and feedback loops
 We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig 1) could reduce many of these problems and thus improve ecological science.
@@
 A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
-The failure of the model to handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
+The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
 Further, % structural degeneracies in the model might become apparent, 
-this step might reveal that some parameters are highly non-identifiable, % If the model fails to recover known parameter values, this flags % not sure about this word
-flagging the need to change the model structure---before incorporating real observations. 
+this step might reveal that some parameters are highly non-identifiable,
+flagging % not sure about this word
+the need to change the model structure---before incorporating real observations. 
 The second feedback loop comes from the retrodictive check. 
@@
 
-% How to address current issues (two case studies)
-% Need an introduction here
 %emw6Apr: I like the shift to 'parameter estimation' instead of trend estimation, but I wonder if we could make that shift clearer above.
@@
 % emw03april: could shorten the first 3 paragraphs, move some text to a box?
-
-% Trends - outline current problem
-%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology.
-Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \citep{soule1999conserving,soule1991conservation}.
-The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}.
-Such trends drove a number of new subfields within ecology---some of which are now complete disciplines within themselves \citep[such as conservation biology,][]{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES.
-Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity.
-
-% \cite{loh2005living} is first LPI paper as best I know
-% \cite{collen2009monitoring,ledger2023past,leung2022reply,puurtinen2022living,toszogyova2024mathematical} to review!
-Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
-The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these debates.
-With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
-For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}.
-
-Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines, and has led to a debates within ecology about whether such analyses undermine public confidence in science  \citep{gonzalez2016estimating}.
-While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.
-In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data.
-Functionally, research on the LPI has somewhat reverse-engineered the recommended workflow: after a series of papers debating different estimates from different models, more recent papers have focused on simulated data to highlight uncertainty given the model and data togethers \citep[though I don't think they link their simulations to the model they use that well,][]{dove2023quantifying,toszogyova2024mathematical}, but this should have been part of the process for the very first papers.
-% While this debate captures part of the process of science---refining our approaches and thus estimates over time---it also makes it hard for policy-makers to do (??
-something) and has led to a debates within ecology about whether such analyses undermine public confidence in science  (CITES Biodiversity debate).
-% \citet{Buschke2021} who focused on adding random population variation
 
@@
 % emw03april: move also this to a box?
-Ecological forecasting is a broad field with a diverse range of methods. Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond. [...]
+Ecological forecasting is a broad field with a diverse range of methods.
+Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond.
+Newer models generally incorporate greater complexity and an ever-growing number of parameters, making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.
 
-These limitations are central to current debates.
+% These limitations are central to current debates. 
 In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty.
@@
 % Forecasting - potential improvments 
-% I was saying too much in one paragraph, so I cut in two
 Applying the workflow to process-based models is a key for opening the black box.
@@
 
-% Misc possibly useful citations -- the GCM people seem to talk a lot about their issues lately on tuning etc. Maybe worth citing?
-% \cite{balaji2022general}mAre general circulation models obsolete? -- covers a lot and good citations within I think
-% \cite{simpson2025confronting} this is the 'Confronting Earth System Model trends with observations' which I have not fully read but Table 1 is interesting to me and something maybe the PBM and NTIF (near-term iterative) people should be doing if they are not doing?
-% \cite{rounce2020quantifying} 'Quantifying parameter uncertainty in a large-scale glacier evolution model using Bayesian inference: application to High Mountain Asia' talks about identifablity problems and uses Bayesian, again I have not fully read, but may be interesting example?
-
 % \setlength{\columnseprule}{0.1pt}
@@
 boxrule=0.1mm,
+grow to left by=+1cm, grow to right by=+1cm,
+enlarge top by=-1.2cm,
 left=3mm,right=3mm, top = 2mm, bottom = 2mm,
-fontupper=\small
+fontupper=\footnotesize
 ]
@@
 
-    \vspace{-0.6cm}\captionof{figure}{Caption}\label{fig:trends}
+    \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
+    \vspace*{1mm}
 \end{minipage}
-\vspace{0.1cm}
 
-\noindent Blabla \newline
-Blabla \newline
- \ \newline
-Blabla \newline
+% Trends - outline current problem
+%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology.
+% Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \citep{soule1999conserving,soule1991conservation}.
+The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}.
+% Such trends drove a number of new subfields within ecology---some of which are now complete disciplines within themselves \citep[such as conservation biology,][]{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES.
+Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity.
+Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
+The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these debates.
+With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
+For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}.
+
+% Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines, and has led to a debates within ecology about whether such analyses undermine public confidence in science  \citep{gonzalez2016estimating}.
+While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.
+In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data.
+Functionally, research on the LPI has somewhat reverse-engineered the recommended workflow: after a series of papers debating different estimates from different models, more recent papers have focused on simulated data to highlight uncertainty given the model and data togethers \citep[though I don't think they link their simulations to the model they use that well,][]{dove2023quantifying,toszogyova2024mathematical}, but this should have been part of the process for the very first papers.
+% While this debate captures part of the process of science---refining our approaches and thus estimates over time---it also makes it hard for policy-makers to do (??
+something) and has led to a debates within ecology about whether such analyses undermine public confidence in science  (CITES Biodiversity debate).
+% \citet{Buschke2021} who focused on adding random population variation
 \vfill
@@
 \centerline{\bf Mechanistic forecasting}
-\bigskip
 
-\noindent Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses.
-They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted.
+\begin{minipage}[t]{\linewidth}
+    \includegraphics[width=\linewidth]{figures/trendestimation_details}
+
+    \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
+    \vspace*{1mm}
+\end{minipage}
+
+\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
+% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted.
 Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions.
@@
 fix to a value found in the literature) or tuned to reproduce some observations or theory.
-The way models are currently calibrated is not a coincidence, % is coincidence the right word?
+The way models are currently calibrated is not a coincidence,
 but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
@@
 \vspace{0.2cm}
-\begin{minipage}[t]{\linewidth}
+\noindent\makebox[\textwidth][c]{
+\begin{minipage}{0.8\linewidth}
     \includegraphics[width=\linewidth]{figures/iterativeworkflow_details}
 
-    \vspace{-0.6cm}\captionof{figure}{Caption}\label{fig:trends}
-\end{minipage}
+    \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
+    \vspace*{1mm}
+\end{minipage}}
 
@@
 
-% This challenge extends beyond the scope of this paper, yet we believe that applying an intelligible workflow can promote better research ethics.
 % emw03april: move this somewhere else?
@@
 % hmm some people talk about simulation models (for process-based models), I agree but I'm afraid in would not be clear with the data simulation we try to highlight
-%
 Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.
@@
 
-This type of workflow-focused training would make forecasting from models far more tractable ...
-
-% better training! on BOTH estimation AND prediction
-% learn how to simulate data will allow students to understand what is the link between the maths (e.g. a distribution) and the ecological processes generating the observations 
-% mobilizing their ecological knowledge while thinking of a model
-% => foster stronger links between disciplines
-% \item estimation: being aware of what is a parameter, and mention uncertainty propagation
-% \item forecasting: should be a natural outcome, not a finality
-
-% \item ML (short-time forecast), benchmarking models are probably useful but should not be the core of our scientific practice, not the spirit!\\
-% $\rightarrow$ moving ecology in the right direction!
+% This type of workflow-focused training would make forecasting from models far more tractable ...
 
@@
 %\end{enumerate}
-%\item Better workflows to the rescue! 
-%\begin{enumerate}
-%\item General overview of new workflows
-%\begin{enumerate}
-%\item Step 0: Research Qs and hypotheses (with a mechanism) lets you ...
-%\item Step 1: Build a model!
-%\item Step 2: Simulate data (and priors or something like priors that forces you to put numbers on stuff)
-%\item Step 3: Design experiments/data collection (maybe you go back to Step 1 here)
-%\item Step 4: Simulate data from actual design/collection
-%\item Step 5: Fit the model to empirical data
-%\item Step 6: Retrodictive checks (feed back to 0 and 1)
-%\end{enumerate}
-%\item New vision of each workflow
-%\end{enumerate}
-%\item Conclusion: world is better
-%\end{enumerate}
-%
-%Current workflows
-%\begin{enumerate}
-%\item 
-%\end{enumerate}
-%
-%How much of forecasting do we cover?
-%\begin{enumerate}
-%\item PBMs
-%\item Near-term iterative
-%\item SDMs
-%\item Machine learning
-%\end{enumerate}
-%
-%
+
 %Miscellaneous notes/points without a home

### 2025-04-08 forecastflows.tex
working on figures and box

@@
 Yet, as ecologists become increasingly computational the opportunity to close this gap has never been greater.
-We outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights.
-Combining these changes with more open model and data sharing---and developing new efforts to race the same data---could be transformative for ecological forecasting.
+We outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights for both trend estimation and forecasting.
+%Combining these changes with more open model and data sharing---and developing new efforts to race the same data---could be transformative for ecological forecasting. 
+A shift toward universal training in a more robust model building could bridge the gap between process-based and statistical approaches and be transformative for ecological modeling.
 \end{abstract}
@@
 Instead forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning.
-The current outcome from these approaches is no clear agreement on current species trends, with ongoing debates on the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and forecasts that diverge due to high model uncertainty at the ecological level.
+The current outcome from these approaches is no clear agreement on current species trends, with ongoing debates on the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.
 
@@
 
+\begin{figure}[h]
+	\centering
+	\includegraphics{figures/modeldataspaces}
+	\caption{} 
+	\label{fig:modeldata}
+\end{figure}
+
 We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig 1) could reduce many of these problems and thus improve ecological science.
@@
 Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
+
+\begin{figure}[h]
+	\centering
+    \hspace*{-1.5cm}
+	\includegraphics{figures/figure_worflow}
+	\caption{} 
+	\label{fig:modeldata}
+\end{figure}
 
@@
 \subsection{Trends}
-% emw03april: could shorten the first 3 paragraphs, move some text to a box?
 
-We argue than an improved workflow that required retrodictive checks and data simulation would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
+Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
+Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines, and has led to a debates within ecology about whether such analyses undermine public confidence in science  \citep{gonzalez2016estimating}.
+While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.
+
+An improved workflow that required retrodictive checks and data simulation would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
 Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process.
@@
 
-% Forecasting - outline current problem
-% emw03april: move also this to a box?
+% vvdm07april: need to work on a better intro paragraph
 Ecological forecasting is a broad field with a diverse range of methods.
@@
 
-% These limitations are central to current debates. 
-In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty.
-For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
+% vvdm07april: need a topic sentence
+Increasing model complexity has not necessarily led to reduced uncertainty.
+In climate modeling, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
 This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}.
@@
 
-% Forecasting - potential improvments 
 Applying the workflow to process-based models is a key for opening the black box.
@@
 
+% Box
 % \setlength{\columnseprule}{0.1pt}
-
 \begin{tcolorbox}[
@@
 \centerline{\bf Trend estimation}
-
+\vspace*{2mm}
 \begin{minipage}[t]{\linewidth}
@@
 
-    \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
+    % \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
     \vspace*{1mm}
@@
 Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity.
-Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
-The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these debates.
+% Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
+In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data.
+The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results.
 With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
@@
 While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.
-In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data.
-Functionally, research on the LPI has somewhat reverse-engineered the recommended workflow: after a series of papers debating different estimates from different models, more recent papers have focused on simulated data to highlight uncertainty given the model and data togethers \citep[though I don't think they link their simulations to the model they use that well,][]{dove2023quantifying,toszogyova2024mathematical}, but this should have been part of the process for the very first papers.
+
+% Functionally, research on the LPI has somewhat reverse-engineered the recommended workflow: after a series of papers debating different estimates from different models, more recent papers have focused on simulated data to highlight uncertainty given the model and data togethers \citep[though I don't think they link their simulations to the model they use that well,][]{dove2023quantifying,toszogyova2024mathematical}, but this should have been part of the process for the very first papers.
 % While this debate captures part of the process of science---refining our approaches and thus estimates over time---it also makes it hard for policy-makers to do (??
@@
 \centerline{\bf Mechanistic forecasting}
+\vspace*{2mm}
+\begin{minipage}[t]{\linewidth}
+    \includegraphics[width=\linewidth]{figures/forecasting_details}
 
-\begin{minipage}[t]{\linewidth}
-    \includegraphics[width=\linewidth]{figures/trendestimation_details}
-
-    \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
+    % \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
     \vspace*{1mm}
@@
 Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions.
-Model development is the central step, typically requiring several years, yet it often remains opaque from an external perspective.
+Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque from an external perspective.
 The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred.
@@
 
-% emw03april: move this somewhere else?
-Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues.
-Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations, but we believe our workflow provides an tractable step to fixing this.
-By focusing on model development more tightly tied to ecological expertise, we ague this workflow should broader the community that contributes to model development.
-As ecologists are increasingly expanding their computational toolkits, many field, and lab and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological, and---most importantly---to interrogate them.
-
-An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
-Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
-And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
-In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made.
-Machine learning may surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
-Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
-
+\subsection{Comprehensive training}
 % emw03april: massively shorten?
@@
 Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches to better forecasts or something.
+
+% \subsection{Tractable models with mechanistic understanding as an alternative to machine learning} % not a good subheader title
+\subsection{A tractable alternative to machine learning}
+
+An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
+Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
+And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
+In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made.
+Machine learning will likely surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
+Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
+
+% emw03april: move this somewhere else?
+Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues.
+Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations, but we believe our workflow provides an tractable step to fixing this.
+By focusing on model development more tightly tied to ecological expertise, we ague this workflow should broader the community that contributes to model development.
+As ecologists are increasingly expanding their computational toolkits, many field, and lab and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological, and---most importantly---to interrogate them.
 

### 2025-04-08 forecastflows.tex
modify box, reduce common workflow fig size
Some edits but seems good to get feedback on!

@@
 \maketitle
-$^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author.
+$^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author (though Lizzie used his LPI example).
 
@@
 
-Anthropogenic drivers are reshaping natural systems, with impacts expected to increase in coming decades \citep{Diaz2019}.
-Climate change will likely accelerate biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}.
+Anthropogenic drivers are reshaping natural systems \citep{Diaz2019}.
+Impacts are projected to increase in coming decades, as climate change accelerates biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}.
 Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics.
@@
 Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets and another focused on forecasting from generally distinct datasets or mechanistic models based on less data.
-Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced with massive geographic, temporal and taxonomic biases.
+Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced with massive geographic, temporal and taxonomic biases.
 Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
@@
 Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting.
-At the same time, forecasting-focused process-based models often develop by adding a new separate layer, disconnected from the original research aim, the data stream and the previous scientific insights, because they fail to examine the model as a functioning whole and thus ignore major degeneracies.
+At the same time, forecasting-focused process-based models often develop by adding new separate layers or components.
+These new parts are often disconnected from the original research aim, its data stream and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major degeneracies.
 
@@
 In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data.
-Towards this aim, here we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
+Towards this aim,  we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
 
@@
 
-Quantitative science relies on a model-based framework, to confront hypothesis and data \citep{}.
+Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}.
 In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses.
@@
 
-This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead appear to be leading to persistent problems.
-Trend estimation has focused mostly on fitting the model to empirical data in a uni-directional way that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig.
-1).
-For forecasting, researchers focus on making predictions with increasingly complex models, making the steps for model building and parameterization increasingly obscured (Fig.
-1).
+This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead may lead to persistent problems. 
+Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig.
+\ref{fig:modeldata}).
+For forecasting, researchers have focused on making predictions with increasingly complex models, increasingly obscuring the steps underlying model building and parameterization (Fig.
+\ref{fig:modeldata}).
 Today, researchers often calibrate different parts of the model separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
@@
 
-We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig 1) could reduce many of these problems and thus improve ecological science.
-The first step of the workflow is to define an explicit research question and formulate hypotheses (Fig 2).
+We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig.
+\ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
+The first step of this workflow is to define an explicit research question and formulate hypotheses (Fig.
+\ref{fig:workflowy}).
 This involves making clear assumptions about the most influential drivers, within the specific context of our study.
-This step help us to think about the mechanisms that could generate the data we observe, including the observational error.
-Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question.
-The general idea is to start with a relatively simple model, that we could refine later.
+This step should help focus on the mechanisms that could generate the data we observe, including the observational error.
+Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and is designed to answer our research question.
+The general idea is to start with a relatively simple model that we could refine later.
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
@@
 With a model in place the next step focuses on testing and understanding it via data simulation.
-`Fake' or `test' data are generated directly from the model by fixing parameters to some values (which is straightforward if the parameters are interpretable) and from fake predictor data.
+`Fake' or `test' data are generated directly from the model by fixing parameters to some values (which is straightforward if the parameters are interpretable) and from fake predictor data (Fig.
+\ref{fig:workflowy}).
 This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model.
-For example, if a researcher realizes their empirical is geographically biased, that should be built into the model and thus then into this data simulation step.
+For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step.
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
+
 Once we are confident about our model structure, we can incorporate real data.
@@
 	\includegraphics{figures/figure_worflow}
-	\caption{} 
-	\label{fig:modeldata}
+	\caption{Here we could have a caption like ...
+Our workflow builds on others (with citations) to highlight the integrated nature of stuff in ecology.
+We can work on this once we get feedback from mighty co-authors.
+Also, should we list the steps on the figure?
+What do people think?}
+	\label{fig:workflowy}
 \end{figure}
@@
 
-Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
-Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines, and has led to a debates within ecology about whether such analyses undermine public confidence in science  \citep{gonzalez2016estimating}.
+Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
+While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced through use of an improved workflow.
+% Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.
+% Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines \citep{gonzalez2016estimating}.
 While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.
 
-An improved workflow that required retrodictive checks and data simulation would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
+An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
 Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process.
 Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected.
-But the workflow highlights that predictions from the model---what we call retrodictive checks (or whatever we call them)---are part of the process of science, and critical to testing for what may be missing in a model.
-We expect retrodictive checks on most published trend analyses would highlight major missing components in these models (expand here??
-ADD example?), and drive changes both in the models themselves and in the simulated data to check the models.
-While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility
-% Mention how low R2 are in ecology?
+But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model.
+We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models.
+While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility.
+% expand on 'highlight major missing components in these models'? ADD example?
 % Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models and in turn debate them.. 
@@
 % vvdm07april: need a topic sentence
-Increasing model complexity has not necessarily led to reduced uncertainty.
-In climate modeling, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
+Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome.
+In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
 This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}.
@@
 
-Applying the workflow to process-based models is a key for opening the black box.
-It would serve as a guide through the successive steps of model development.
-In particular, incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing strong degeneracies in the model design.
-This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model.
-Model calibration would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights.
-It would help properly take into account any issues regarding non-identifiability.
-This could involve reformulating the mathematical structure of some processes, making new hypotheses to target the right level of complexity, or incorporating more expert knowledge to better constrain calibration.
-Making the model more tractable would also naturally facilitate the recognition and quantification of parameter uncertainty, as well as its propagation into model projections
+Applying the workflow to process-based models would help open black box.
+Each successive step of model development in the workflow may highlight current problems and a path to solutions.
+Incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing potentially large degeneracies in the model design.
+Uncovering these degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
+The workflow may also highlight degeneracies by requiring more explicit model calibration, which is currently hidden within opaque `model building,' making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated).
+Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
+The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.
 
@@
 Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.
+
 This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference.
@@
 
-\centerline{\bf Common solution!}
-\vspace{0.2cm}
-\noindent\makebox[\textwidth][c]{
-\begin{minipage}{0.8\linewidth}
+\centerline{\bf A common workflow to bridge trend estimation and forecasting}
+\vspace*{-3mm}
+{\begin{multicols}{2}
+\begin{minipage}[t]{\linewidth}
 \includegraphics[width=\linewidth]{figures/iterativeworkflow_details}
@@
 \vspace*{1mm}
-\end{minipage}}
+\end{minipage}
+
+\columnbreak
+\vspace*{1mm}
+Blabla \\
+We say here briefly how great the new approach is or something.
+
+\end{multicols}}
 
@@
 We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
-One well-known hurdle is that the pressure to publish academic work, which can lower research standards and make the added effort of this workflow seem ill-placed.
+One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed.
 This may be especially true for those who see science rewarded mostly through the shear number of publications.
@@
 \subsection{Comprehensive training}
-% emw03april: massively shorten?
-Advancing ecology to where most researchers see models built more flexibly from their understanding of their ecological systems will not happen rapidly without a major shift in training, however.
-Much of ecology still divides the world into training for those focused on several groups.
-First, those who will conduct field and lab studies often learn a suite of particular tests matched to particular experiment designs or to simply match their predictor and response data to a flow diagram of variable types (add EXAMPLE and reference, something like: categorical $y$ and binary $y$ means ...
-I dunno, chi square?) and are left adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is.
-Instead, they are expected to collaborate with others when they need more complex models, which is a second group of ecological researchers---those who focus on complex statistical models and were thus trained more in model development, but often for highly specific applications (e.g., wildlife population estimates) where they may rarely build totally new generative models.
-Separate from these groups, process-based modelers do their own thing.
+Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training.
+Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models.
+In ecological training today may researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
+Such training leaves a large number of ecologists adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is.
+
+Instead, ecologists trained in a limited set of tests are expected to collaborate with others when they need more complex models, who were trained more in model development (though often for highly specific applications, such as wildlife population estimates, where they may rarely develop entirely new generative models).
+These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models.
 % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms
-% hmm some people talk about simulation models (for process-based models), I agree but I'm afraid in would not be clear with the data simulation we try to highlight
 Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.
@@
 
-Training all ecologists in our proposed workflow would break down barriers between different groups of ecologists today with likely major benefits for science.
+Training all ecologists in our proposed workflow would break down barriers between these different groups with likely major benefits for science.
 Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in our workflow would focus on learning to generate questions and then models, and then to simulate data from them.

### 2025-04-08 forecastflows.tex
move some text to the box (maybe we will need more ideas in the last section at some point?)

@@
 This retrodictive check allows model output to be compared to observations.
-The workflow replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
+The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
 Any parameter (such as a trend estimate) must be carefully interpreted alongside others.
@@
 
-% vvdm07april: need a topic sentence
 Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome.
@@
 \vspace*{1mm}
-Blabla \\
-We say here briefly how great the new approach is or something.
+Blabla... \\
+% We say here briefly how great the new approach is or something.
+An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
+Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
+And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
 
@@
 
-An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
-Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
-And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
 In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made.
@@
 
-% This type of workflow-focused training would make forecasting from models far more tractable ...
-
-%\item What are current workflows and where are they limiting us?
-%\begin{enumerate}
-%\item For trends ...
-%\begin{enumerate}
-%\item easy to find different trends through small model tweaks to analyses and/or different data
-%\item For example, right now many different papers report different biodiversity trends (LPI example?)
-%\item New workflow would make ecologists understand uncertainty in their model data/combo (and perhaps not see/publish results as so divergent?)
-%\end{enumerate}
-%\item For forecasting ...
-(somehow jump to our focus on process-based models PBMs here?
-Something like, forecasting is big and there are diverse methods!
-Near-term iterative, correlative niche models, but PBMs are often considered the gold standard ...
-mention machine learning?)
-%\begin{enumerate}
-%\item as many models as researchers working on process-based models 
-%\item + accumulation of successive layers in the development of models = significant challenge to scientific transparency, reproducibility, interpretability\\
-%models often draw inspiration from each other, which is good (way to do science), but not always explicit...
-(some issues: arbitrarily established parameter value in one model then transmitted to multiple models)
-%\item focus of researchers: always integrate new mechanisms, new parameters, to increase "realism"...
-they intuitively "feel" what kinds of adjustments is needed...
-but opaque from an external perspective ("black box" of model building and calibration)
-%\item models rarely fitted as a whole, dozens of parameters without explicitly quantifying parameter uncertainty, and often neglect to propagate this uncertainty
-%\item simulations of models themselves became a subject of study to disentangle all the processes modelled and understand model sensitivity 
-%\end{enumerate}
-%\end{enumerate}
-
-%Miscellaneous notes/points without a home
-%\begin{enumerate}
-%\item Ecologists need to race the same data to make progress for trends and for forecasting (point to make at end of paper maybe?
-And what is the workflow for this?) ...
-though LPI is used a lot, perhaps it is  sign that ecology is ready to start racing the same data, but then we need `analysis-ready' data so we're not all slightly differently cleaning the data etc..
-%\item Machine learning threatens utility of PBMs
-%\item We need more uncertainty propagation for trends and forecasting (uncertainty is esp. ignored in PBMs)
-%\item This workflow should lead to less model comparison (AIC, stepwise)
-%\item This workflow works for machine learning!
-%\item PBM: workflow should require estimating all parameters together; data simulation should reduce parameter number and highlight non-biol results
-%\item Current trend workflow: Should be research question focused?
-%\end{enumerate}
-%
-%Miscellaneous notes/points from thinking over 22-23 March weekend ...
-%\begin{enumerate}
-%\item Workflows emphasize there is no easy fix to better science or better stats
-%\item Maybe do a retrodictive check example with LPI? Including showing how even the Freckleton work could do more? 
-%\item What's the pathway from model comparison of many covariates to something else?
-Sometimes it seems like it's just prediction shoved into a mechanistic study, but if the goal is mechanism, there needs to be more work that either models these covariates together in a useful way (sort of approaching process-models!) or gets down to the fewer extremely relevant ones.
-%\item When do we need to open up the black box? Something about we need better training for what science is and our goals; machine learning is not often helping with advancing \emph{science}
-%\item Scott Collins point that forecasting is not science (he claims a bunch of economists, weather folks etc.
-came to an ecological forecasting meeting and tried to explain that forecasting is an outcome of science, but it's not something you do science ON)
-%\end{enumerate}
-
 \clearpage

### 2025-04-16 forecastflows.tex
removing some comments before sending
start to incorporate J&J feedback

@@
 \date{\today}
-\author{Victor Van der Meersch, J. Regetz, T. J. Davies$^*$ \& EM Wolkovich}
+\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}
 \maketitle
-$^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author (though Lizzie used his LPI example).
+% $^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author (though Lizzie used his LPI example).
 
@@
 
-% To do:
-% Fix the abstract so it matches the paper better (I think mainly the last few sentences need work, or we just need to say more about data sharing in the text)
-% Add in Figures
-% Delete out unnecessary comments so easier for co-authors to read?
-
 \begin{abstract}
@@
 Relevant datasets to meet these demands have also increased in size and complexity, including in their geographical, temporal and phylogenetic scales.
-While new research often suggests that accounting for these complexities variously increases, removes or otherwise alters major trends, we argue that the fundamental approach to model fitting in ecology makes it impossible to evaluate and compare models.
+While new research often suggests that accounting for these complexities has a major impact on projected trends and forecasts, we argue that the typical approach to model fitting in ecology makes it impossible to evaluate and compare the models used to generate these insights and predictions.
+% jdavies15april: I feel this could be a critical step somehow - sharing simulations could allow cross-validation between models or perhaps cross-calibration ...
 These problems stem in part from continuing gaps between statistical workflows---where the data processing and model development are often addressed separately from the ecological question and aim---and scientific workflows, where all steps are integrated.
@@
 Impacts are projected to increase in coming decades, as climate change accelerates biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}.
+% jdavies15april: not sure what models you are referring to here, but perhaps look up models of 'mean species abundance'.
 Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics.
@@
 Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections, could reduce many of these problems.
-In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data.
+In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to undertand when, where and why different models diverge.
 Towards this aim,  we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
@@
 In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses.
-But this idealized method often does not apply to the reality of ecological research.
+% jdavies15april: could refer to the 'pre-registration' movement here ...
+But this idealized method often does not apply to the reality of ecological research, and most of macroecological insights emerge from exploring patterns in data.
 Many important questions cannot be addressed through controlled experiments and replications.
@@
 This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead may lead to persistent problems.
-Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig.
+% jdavies15april: what challenges?
+forecasting from imperfect data or the urgent need to generate forecasts?
+Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way % jdavies15april: not obvious what this means
+that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig.
 \ref{fig:modeldata}).
-For forecasting, researchers have focused on making predictions with increasingly complex models, increasingly obscuring the steps underlying model building and parameterization (Fig.
+For forecasting, researchers have focused on making predictions with increasingly complex models, frequently obscuring the steps underlying model building and parameterization (Fig.
 \ref{fig:modeldata}).
 Today, researchers often calibrate different parts of the model separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
+% jdavies15april: is this particularly specific to process based models?
 Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow. 
@@
 	\includegraphics{figures/modeldataspaces}
-	\caption{} 
+	\caption{Here the caption could explain a bit more the pre-model pre-data to the post-model post-data...} 
 	\label{fig:modeldata}
 \end{figure}
+% jdavies15april: Could the bottom right be where exploratory data analysis fits in, generating hypotheses that inform the research question?
 
-We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig.
-\ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
+We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure?
+Or wait to mention simulation until referencing Fig 2?
+(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
 The first step of this workflow is to define an explicit research question and formulate hypotheses (Fig.
@@
 This involves making clear assumptions about the most influential drivers, within the specific context of our study.
-This step should help focus on the mechanisms that could generate the data we observe, including the observational error.
-Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and is designed to answer our research question.
+This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error.
+% vvdm16april: to answer Jonathan comment about verbal model
+From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question.
 The general idea is to start with a relatively simple model that we could refine later.
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
+This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires to think about alternative parameterizations).
 
 With a model in place the next step focuses on testing and understanding it via data simulation.
-`Fake' or `test' data are generated directly from the model by fixing parameters to some values (which is straightforward if the parameters are interpretable) and from fake predictor data (Fig.
+`Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data (Fig.
 \ref{fig:workflowy}).
 This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model.
+% jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e.
+simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model.
 For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step.
@@
 
-Once we are confident about our model structure, we can incorporate real data.
+Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step.
 This way, we obtain parameter estimates constrained by observations.
-Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question.
+Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works.
 This could lead us to either simplify our research question or---ideally---launch new data collection efforts.
@@
 This retrodictive check allows model output to be compared to observations.
+% jregetz16april: do we need to define “retrodictive”?
 The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
-Any parameter (such as a trend estimate) must be carefully interpreted alongside others.
-%such as the  global mean or intercept and the estimate variance
+% jregetz16april: Not sure what this means?
+jdavies did not like it either!
+(he thought it was a typo)
+Any parameter (such as a trend estimate) must be carefully interpreted alongside others. %such as the  global mean or intercept and the estimate variance
 Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
@@
 % More details, what's special! Feedbacks!
-%emw6Apr: I think there's more than just degeneracy here, often people build in more complexity to their simulated data than their model allows, but this means they should change the model (or the question).
-This needs to be clearer above and below.
-I think this is the first point (since it's a little more obvious to readers and it relates to trends, which we always handle first), and the next one could be about structural degeneracies  but I think we need to be careful here in what we mean and explain it more to readers who will not get it.
 A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
@@
 
+%jregetz16april: One thing I’m stuck on (and could’ve written this note earlier too).
+Is “trends” really essential?
+Or is this just shorthand for “simple, one/few parameter, non-mechanistic forecasting models”?
+If yes, can we generalize that in the MS overall, or at least say so up front and then say e.g.
+“we’ll refer to these as trend models for brevity”?
+If no, and something about trend estimates is specifically key to the paper, can we say more about *that*?
+
 Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
-While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced through use of an improved workflow.
+While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced and more rapidly resolved through use of an improved workflow.
 % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.
-% Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines \citep{gonzalez2016estimating}.
-While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.
 
@@
 While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility.
+% jdavies15april: but there is a history of using null models (e.g.
+Gotelli &co) that is perhaps not so different - here the 'simulated' data is some sampled process of the observed data, but we can estimate standard effect sizes that are part of our statistical model.
+I know this is different to the simulations you propose based on some generative process, but ecologists should be able to able to make this additional step.
+
 % expand on 'highlight major missing components in these models'? ADD example?
@@
 Returning to the example of a global estimate of trends in vertebrate populations of species over time and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
+% jdavies15april: This is what the LPI wants to be (like a index fund in the stock market tracking trends over time) , but this is often overlooked.
 Or a need to also find which species are declining most, including those that may have poor or no data?
 From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data.
-Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model.
+Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends.
 For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled.
@@
 Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
+% jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
 If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
 Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
+% jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty ....
 
-Applying the workflow to process-based models would help open black box.
+Applying the workflow to process-based models would help open the black box. % jdavies15april: what is the black box?
 Each successive step of model development in the workflow may highlight current problems and a path to solutions.
-Incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing potentially large degeneracies in the model design.
-Uncovering these degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
-The workflow may also highlight degeneracies by requiring more explicit model calibration, which is currently hidden within opaque `model building,' making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated).
+Incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and help expose potential degeneracies in the model design.
+%jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below)
+Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
+The workflow may also highlight degeneracies by requiring more explicit model calibration, % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
+which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated).
 Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
 The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.
+% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties?
+This could help illustrate potential degeneracies etc.
 
+% jdavies15april: might be better grouped together in concluding section
 Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
@@
 The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results.
+% jdavies16april: could highlight how influential the LPI has become (recent reports suggest a 73% ave.
+pop decline since 1970.
+This would be massive!
 With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
 For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}.
+% jdavies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations.
+We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends.
+For this we need to predict outside of the set ...
 
@@
 fix to a value found in the literature) or tuned to reproduce some observations or theory.
-The way models are currently calibrated is not a coincidence,
-but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
+The way models are currently calibrated is not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
+% jdavies15april: is this more specific to process based  models?
 
@@
 
-\centerline{\bf A common workflow to bridge trend estimation and forecasting}
+\centerline{\bf A common workflow to bridge trend estimation and forecasting} % jdavies15april: would it be possible to label this panel with components of the LPI to provide a clearer illustration of how the framework is implemented?
+% vvdm16april: not sure I understood
 \vspace*{-3mm}
@@
 But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help.
+% jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
 Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).
 
-\subsection{Comprehensive training}
+\subsection{Adopting the workflow}
 Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training.
 Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models.
-In ecological training today may researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
+In ecological training today researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
 Such training leaves a large number of ecologists adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is.
+% jdavies15april: this may be true, but I am not sure we want to broadly disparage ecologists
 
@@
 
-Training all ecologists in our proposed workflow would break down barriers between these different groups with likely major benefits for science.
+Training all ecologists in our proposed workflow would break down barriers between these different groups with likely major benefits for science. % jdavies15april: this feels a little OTT
 Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in our workflow would focus on learning to generate questions and then models, and then to simulate data from them.
@@
 Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches to better forecasts or something.
+% jdavies15april: and parameter estimation to improve forecasts.
 
-% \subsection{Tractable models with mechanistic understanding as an alternative to machine learning} % not a good subheader title
 \subsection{A tractable alternative to machine learning}
@@
 Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
+% jdavies15april: nothing intrinsically bad about ML, but sometimes the goal is just to understand the system, not make predictions or forecasts.
 
-% emw03april: move this somewhere else?
+% jdavies15april: This could be the Concluding section highlighting the general advances provided by adopting your framework (and perhaps a better place for text currently at end of section 3.2).
 Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues.

### 2025-04-17 forecastflows.tex
making some progress

@@
 Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}.
-In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses.
+In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses---following the principles of pre-registration, where hypotheses and methodology are defined prior to data collection.
 % jdavies15april: could refer to the 'pre-registration' movement here ...
-But this idealized method often does not apply to the reality of ecological research, and most of macroecological insights emerge from exploring patterns in data.
+But this idealized method often does not apply to the reality of ecological research.
 Many important questions cannot be addressed through controlled experiments and replications.
-Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.
+Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}, and most of macroecological insights emerge from exploring patterns in these datasets (exploratory data analysis).
 
-This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead may lead to persistent problems.
+This reality should drive researchers to use more robust and coherent methods.
+But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems.
 % jdavies15april: what challenges?
 forecasting from imperfect data or the urgent need to generate forecasts?
-Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way % jdavies15april: not obvious what this means
+Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way % jdavies15april: not obvious what this means... [FLAG]
 that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig.
 \ref{fig:modeldata}).
-For forecasting, researchers have focused on making predictions with increasingly complex models, frequently obscuring the steps underlying model building and parameterization (Fig.
+For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (Fig.
 \ref{fig:modeldata}).
-Today, researchers often calibrate different parts of the model separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
+Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
 % jdavies15april: is this particularly specific to process based models?
@@
 Or wait to mention simulation until referencing Fig 2?
+[FLAG: what do you think?
+I agree you should choose on of the two options]
 (Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
@@
 \ref{fig:workflowy}).
-This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model.
+% This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model.
 % jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e.
 simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model.
-For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step.
+[FLAG: yes I can understand this point...]
+% For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step.
+[FLAG: so I moved this sentence below, when we say "Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data"
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
@@
 % jregetz16april: do we need to define “retrodictive”?
-The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
+% The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
 % jregetz16april: Not sure what this means?
 jdavies did not like it either!
-(he thought it was a typo)
-Any parameter (such as a trend estimate) must be carefully interpreted alongside others. %such as the  global mean or intercept and the estimate variance
+(he thought it was a typo), vvdm16april: reformulate:
+The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.
 Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
+The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to come out with a research question.
 
@@
 The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
-Further, % structural degeneracies in the model might become apparent, 
-this step might reveal that some parameters are highly non-identifiable,
-flagging % not sure about this word
-the need to change the model structure---before incorporating real observations. 
+Further, this step might reveal that some parameters are highly non-identifiable,
+flagging the need to change the model structure---before incorporating real observations. 
 The second feedback loop comes from the retrodictive check. 
@@
 unbalanced data).
+For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step.
 This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics.
@@
 If no, and something about trend estimates is specifically key to the paper, can we say more about *that*?
+[FLAG]
 
@@
 
-% vvdm07april: need to work on a better intro paragraph
 Ecological forecasting is a broad field with a diverse range of methods.
@@
 % jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
-If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
+If a model becomes too complex, it may become a black box, and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
 Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
 % jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty ....
+[FLAG: agree but not really relevant to PBM?
+Maybe it's not clear enough we are focusing on PBM in this section...]
 

### 2025-04-17 forecastflows.tex
that's all for today

@@
 % jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
-If a model becomes too complex, it may become a black box, and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
+If a model becomes too complex, it may become a black box, % mention black box here to line up with the beginning of next paragraph
+and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
 Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
@@
 Each successive step of model development in the workflow may highlight current problems and a path to solutions.
-Incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and help expose potential degeneracies in the model design.
-%jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below)
+Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential degeneracies in the model design.
+%jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below) [FLAG: agree, and made me realize it was not 100% clear in my mind.
+As I understand it, non-identifiability concerns one parameter, and degeneracy concerns the all model?...
+but not sure]
 Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
@@
 % Trends - outline current problem
-%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology.
-% Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \citep{soule1999conserving,soule1991conservation}.
-The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}.
-% Such trends drove a number of new subfields within ecology---some of which are now complete disciplines within themselves \citep[such as conservation biology,][]{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES.
-Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity.
-% Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
+
 In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data.
@@
 For this we need to predict outside of the set ...
+[FLAG]
 
-% Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines, and has led to a debates within ecology about whether such analyses undermine public confidence in science  \citep{gonzalez2016estimating}.
-While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.
-
-% Functionally, research on the LPI has somewhat reverse-engineered the recommended workflow: after a series of papers debating different estimates from different models, more recent papers have focused on simulated data to highlight uncertainty given the model and data togethers \citep[though I don't think they link their simulations to the model they use that well,][]{dove2023quantifying,toszogyova2024mathematical}, but this should have been part of the process for the very first papers.
-% While this debate captures part of the process of science---refining our approaches and thus estimates over time---it also makes it hard for policy-makers to do (??
-something) and has led to a debates within ecology about whether such analyses undermine public confidence in science  (CITES Biodiversity debate).
-% \citet{Buschke2021} who focused on adding random population variation
 \vfill
@@
 % jdavies15april: is this more specific to process based  models?
+[FLAG: hmm yes again?
+Same point as above, maybe it's not clear we're talking mostly about PBM?]
 
@@
 \centerline{\bf A common workflow to bridge trend estimation and forecasting} % jdavies15april: would it be possible to label this panel with components of the LPI to provide a clearer illustration of how the framework is implemented?
-% vvdm16april: not sure I understood
+% [FLAG: not sure I understood]
 \vspace*{-3mm}
@@
 We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
-One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed.
+% One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed.
 This may be especially true for those who see science rewarded mostly through the shear number of publications.
@@
 % jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
+Good model fitting is inherently iterative and takes time, at odds with the pressure to publish quickly.
 Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).

### 2025-04-17 forecastflows.tex
continue making very small progress

@@
 (Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
-The first step of this workflow is to define an explicit research question and formulate hypotheses (Fig.
-\ref{fig:workflowy}).
+The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, fig.
+\ref{fig:workflow}).
 This involves making clear assumptions about the most influential drivers, within the specific context of our study.
@@
 % vvdm16april: to answer Jonathan comment about verbal model
-From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question.
+From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, fig.
+\ref{fig:workflow}).
 The general idea is to start with a relatively simple model that we could refine later.
@@
 
-With a model in place the next step focuses on testing and understanding it via data simulation.
-`Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data (Fig.
-\ref{fig:workflowy}).
+With a model in place the next step focuses on testing and understanding it via data simulation (step 3, fig.
+\ref{fig:workflow}).
+`Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data.
 % This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model.
@@
 
-Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step.
+Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, fig.
+\ref{fig:workflow}).
 This way, we obtain parameter estimates constrained by observations.
@@
 This could lead us to either simplify our research question or---ideally---launch new data collection efforts.
-This leads to the second data simulation step, this time using our fitted model parameters to generate predictions.
+This leads to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, fig.
+\ref{fig:workflow}).
 This retrodictive check allows model output to be compared to observations.
@@
     \hspace*{-1.5cm}
-	\includegraphics{figures/figure_worflow}
+	\includegraphics{figures/figure_worflow_wsteps}
 	\caption{Here we could have a caption like ...
@@
 What do people think?}
-	\label{fig:workflowy}
+	\label{fig:workflow}
 \end{figure}
@@
 % jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
+[FLAG: agree but not really relevant to PBM?
+Maybe it's not clear enough we are focusing on PBM in this section...]
 If a model becomes too complex, it may become a black box, % mention black box here to line up with the beginning of next paragraph
@@
 
-Applying the workflow to process-based models would help open the black box. % jdavies15april: what is the black box?
+Applying the workflow to process-based models would help open the black box.
 Each successive step of model development in the workflow may highlight current problems and a path to solutions.
@@
 Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
-The workflow may also highlight degeneracies by requiring more explicit model calibration, % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
+The workflow may also highlight degeneracies by requiring more explicit model calibration (i.e.
+data fitting), % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
 which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated).

### 2025-04-17 forecastflows.tex
adding some stuff about pbmodels (but not great)

@@
 	\centering
-	\includegraphics{figures/modeldataspaces}
+	\includegraphics{figures/modeldataspaces_rotation}
 	\caption{Here the caption could explain a bit more the pre-model pre-data to the post-model post-data...} 
@@
 
+% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties?
+This could help illustrate potential degeneracies etc.
+% [FLAG: I tried something, but I'm not really illustrating degeneracies + I don't like citing myself as much + I'm not really satisfied]
+An illustration of the issues with the current forecast workflow is the process-based models used to project species or ecosystem distributions.
+Most of the focus is on the model projections---often represented as maps without uncertainties.
+Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only of the many different output variables of the model.
+The model building and calibration steps are usually limited to the Methods section, and estimated parameter values are rarely discussed \citep[e.g.][]{Conradi2024, VanderMeersch2025a}.
+The complexity of these models often makes it difficult to assess the robustness of their calibration, whether for a reviewer or a reader.
+If ever done, the assesment of the model degeneracies is often done as a a separate work, exposing model degeneracies later on \citep{VanderMeersch2025b}, and not required as a preliminary step.
+
 Applying the workflow to process-based models would help open the black box.
@@
 The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.
-% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties?
-This could help illustrate potential degeneracies etc.
 
 % jdavies15april: might be better grouped together in concluding section
+% [FLAG: I feel if we want to put this in the concluding section, we would need to talk also about the "opportunities to bridge statistical and process-based", rather than keeping it in the box?
+So, what should we put in the box?]
 Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.

### 2025-04-17 forecastflows.tex
Small changes in last section

@@
 \subsection{Adopting the workflow}
+
 Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training.
@@
 In ecological training today researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
-Such training leaves a large number of ecologists adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is.
+% Such training leaves a large number of ecologists adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is.
 % jdavies15april: this may be true, but I am not sure we want to broadly disparage ecologists
@@
 Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.
-While specialization is valuable, we argue the fundamental training in ecology has overly-siloed these groups and prevented more rapid progress.
+While specialization is valuable, we argue the fundamental % training
+separation in ecology has overly-siloed these groups and prevented more rapid progress.
 
-Training all ecologists in our proposed workflow would break down barriers between these different groups with likely major benefits for science. % jdavies15april: this feels a little OTT
-Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in our workflow would focus on learning to generate questions and then models, and then to simulate data from them.
+% Training all ecologists in our proposed workflow would break down barriers between these different groups with likely major benefits for science. % jdavies15april: this feels a little OTT
+Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them.
 This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them.
@@
 % Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
-Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches to better forecasts or something.
-% jdavies15april: and parameter estimation to improve forecasts.
+Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.
 

### 2025-04-18 analyses/sampleworkflow.Rmd
Added in some files to start the workflow example.

# A sample workflow

*Hi Victor, it's Lizzie. I will just put my notes to you in italics. You really should remove them. In general, I suggest you focus on getting some code/plots and basic text done that you then share next week with everyone.*

To show our workflow in action, we examine a subset of the Living Planet Index (LPI, download at: https://www.livingplanetindex.org/data_portal. This download should include a csv file, here: LPD2022_public.csv, that we use below). For more information on the LPI, check out the main text and www.livingplanetindex.org/.

## Step 1: Research question

*I started this, but it would be perfect for JD to fix/add to, so I did not finish it. (As mentioned above, I suggest you focus on getting some code/plots and basic text done that you then share next week with everyone.)*

Here we need to focus on exactly what question we are trying to answer, which will guide us in the next steps. This step is sometimes skipped over or moved through so quickly that the exact question is not clear, but is critical for developing a useful model. We might start with: What is the global trend for biodiversity in vetebrate species? This is rather broad and might mean we should develop a model that will yield one global estimate. In thinking about this, we might realize we don't want just that, we actually want to understand the variation across different species and perhaps populations and also a global estimate. So we might refine the question to:

What is global trend in vertebrate species population numbers, both overall and how does it vary acrosss species and populations?

## Step 2: Model building

Next we need to think about how we would build a model to address this question, with the goal to build the model and evaluate it using simulated data. There is a lot of demographic models in population ecology, so we might revisit those and consider a discrete time model based on population size and stepping through each year of data. We might also start with a simple linear model, where populations can go up, down or do nothing. That gets us started, but now we need to consider how we will model species and populations. Populations usually are nested in species in ecological models, so we might consider that to start. Our resulting model might then be:

y_i ~ normal(mu, sigma)

mu = alpha[sp[pop]] + beta[sp[pop]]*year

*Need to finish the notation as I always have to look up the nesting notation.*

## Step 3: Model evaluation

Now we need to simulate data to test our model. As we start to think about how to do this, we may become overwhelmed with the scale of the question we started with. We likely would not feel immediately sure how to simulate for all different vertebrate species.

### Stepping backward...

Here we would likely feedback one to two steps. We could step back to Step 2 and consider adding a phylogeny or other structure from which we could then simulate data from, or we could step back to Step 1 and narrow the question (before potentially building it back up.) We believe the latter approach is often better as working on a small scale first can point out more fundamental problems versus starting big and then spending much longer to identify fundamental problems.

*Obviously we can charge forward with the full model if you want, I just thought this would be (a) easier for us, (b) way better to show as it's how you should do it and (c) related to a, I suspect JD can be more help, though any country with a good amount of data could work.* 
So here we refine our question to:

What is trend in terrestrial vertebrate species population numbers in South Africa, both overall and how does it vary acrosss species and populations?

Now we start to simulate! 
*Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.* 
```{r}

```
*Add some text here showing the model output returning paramaters.*

## Step 4: Fit model to empirical data

Here we finally get to see how our model performs on the empirical data and are a step closer to answering our question, but we also start to notice some differences in our empirical data from our simulated data. In preparing our data for the model, we realize we have major gaps in sampling years and ... something else?

Here, we might step back and adjust our model (re-doing Step 2-3 again), but in this case we go ahead.

*See starterLPI.R for what I have so far with the data...*

## Step 5: Retrodictive check

Now we check how well our model predicts our data ...

## Feedbacks

At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including questioning our assumption of one sigma for all populations and potentially our linearity assumption. We also may want to add predictors for the slope such as  ... We would do this, returning to Steps 2-5 likely several times until we have a model that is good enough for our purposes here.

And then we would likely add in more countries and species....

And then we would realize we really don't have great data for this question and we would ...

### 2025-04-18 analyses/starterLPI.R

@@ -1,0 +1,50 @@
+## Started 17 April 2025 ##
+## By Lizzie so far ... ## 
+
+# housekeeping
+rm(list=ls()) 
+options(stringsAsFactors=FALSE)
+
+# set your directory if you're me
+setwd("~/Documents/git/projects/misc/miscmisc/workflowsPhilTrans/analyses")
+
+# packages 
+library(data.table)
+# What?! These two are because I was cheap and copied from Johnson et al. 2024
+library(countrycode)
+library(tidyverse)
+
+
+lpi <- fread("input/LPD2022_public.csv", na = c("", "NA", "NULL"))
+
+# Below is copied from :
+# Johnson_etal_2024_code_GitTFJ-correlated_effect_model-21c0d95/code/data_compile.Rmd
+custom_match <- c(`Cura??ao` = "CUW", `?Óland Islands` = "ALA", `International Waters` = "ABNJ")
+
+lpi <- subset(lpi, Replicate == 0)
+lpi <- lpi %>% 
+  mutate(
+    citation_year = str_replace_all(Citation, "\\\"|\\.|-|\\\'", ""),
+    citation_year = parse_number(citation_year),
+    site = paste(word(Citation, 1, sep = " |,"),
+                 citation_year,
+                 Location,
+                 sep = "_")
+  ) %>%
+  select(site, Country, Latitude, Longitude, Binomial, Units, `1950`:`2020`) %>% 
+  pivot_longer(cols = `1950`:`2020`,
+               names_to = "date",
+               values_to = "abundance")
+lpi <- as.data.frame(lpi)
+
+
+## Thinking we need to subset for sanity ... 
+length(unique(lpi$site))
+sort(unique(lpi$Country))
+table(lpi$Country, lpi$site)
+
+sa <- subset(lpi, Country=="South Africa")
+
+ggplot(sa, aes(x=date, y=abundance)) +
+	geom_point() + 
+	facet_wrap(.~site)

### 2025-04-19 forecastflows.tex
Small edits to current draft with some call-out notes

@@
 \usepackage{caption}
+\usepackage{todonotes}
 
@@
 \maketitle
-% $^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author (though Lizzie used his LPI example).
 
-{\bf Deadline:} 1 May 2025 (was 1 April 2025)
+{\bf Deadline:} 1-ish May 2025 (was 1 April 2025, getting in by May 8th would be good!)
 
@@
 % Already 5K!
+
+% Jim, if you want to add any of this into the abstract, please do!
+%There's a fundamental problem in how a lot of empirically but observationally driven (i.e., non-experimental) research is done in ecology and related fields.
+The crux of the issue comes down to a couple of things.
+First, the research community has two very distinct paradigms for modeling the system or phenomenon or interest: either fitting simple trend-like models with only one or a few parameters, or developing forecast models that often include potentially complex, mechanistic, and/or black box submodels largely unconstrained by data.
+Second, regardless of the approach, these models tend to be developed not only in isolation of one another, but also without a coherent strategy for linking modeling/statistical decisions, data, and scientific understanding (both as input and output) throughout the entire modeling process.
+We propose a unified, principled workflow for end-to-end empirical model development, incorporating sound statistical and scientific practices, and especially relying on data simulation to inform decisions at multiple steps in the process.
 
@@
 At the same time, forecasting-focused process-based models often develop by adding new separate layers or components.
-These new parts are often disconnected from the original research aim, its data stream and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major degeneracies.
+These new parts are often disconnected from the original research aim, its data stream and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).
 
-Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections, could reduce many of these problems.
-In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to undertand when, where and why different models diverge.
+Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and potential problems, before estimating its parameters and making projections, could reduce many of these problems.
+In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where and why different models diverge.
 Towards this aim,  we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
@@
 Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}.
-In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses---following the principles of pre-registration, where hypotheses and methodology are defined prior to data collection.
-% jdavies15april: could refer to the 'pre-registration' movement here ...
+In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses.
+This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection.
+%add cites for prereg?
 But this idealized method often does not apply to the reality of ecological research.
@@
 forecasting from imperfect data or the urgent need to generate forecasts?
-Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way % jdavies15april: not obvious what this means... [FLAG]
-that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig.
+Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Fig.
 \ref{fig:modeldata}).
@@
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
-This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires to think about alternative parameterizations).
+This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).
 
@@
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
+At this stage the focus is understanding the model, so researchers may have to spend a while to think through what parameters are reasonable or not.
+This is also a step for making sure the model is working as we expect, so simulated data often would have low error and high sample sizes (to make sure the model returns the parameter values used to generate the simulated data),
 
@@
 This way, we obtain parameter estimates constrained by observations.
-Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works.
+%emwApr19: I commented out the below as I think linking difficulty in fitting the model to problems with the data is pretty tricky (difficulty in fitting the model can point to many things!) and thus too confusing to include here.
+% Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works.
 This could lead us to either simplify our research question or---ideally---launch new data collection efforts.
-This leads to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, fig.
+These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, fig.
 \ref{fig:workflow}).
@@
 Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
-The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to come out with a research question.
+The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to more clearly develop a research question before extensive model fitting.
 
@@
 The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
-Further, this step might reveal that some parameters are highly non-identifiable,
+Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
 flagging the need to change the model structure---before incorporating real observations. 
 The second feedback loop comes from the retrodictive check. 
-Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic.
-We can refine the model to integrate the missing process and restart the workflow.
+Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic.
+We can refine the model to integrate the missing process(es) (if we can identify them) and restart the workflow.
 Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g.
@@
 [FLAG]
+%emwApr19: I think the only key part is that biodiversity trends and trends over time are big in forecasting ...
+as long as we don't lose that I think it would be great to make the suggested change -- Jim, can you see if you can edit it on the next round?
 
-Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
+Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends over time, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
 While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced and more rapidly resolved through use of an improved workflow.
@@
 We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models.
-While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility.
-% jdavies15april: but there is a history of using null models (e.g.
-Gotelli &co) that is perhaps not so different - here the 'simulated' data is some sampled process of the observed data, but we can estimate standard effect sizes that are part of our statistical model.
-I know this is different to the simulations you propose based on some generative process, but ecologists should be able to able to make this additional step.
+This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand.
+Ecologists have started to use simulated data more to understand potential limitations of their models and data combined, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.
 
@@
 
-\subsection{Forecasting}
+\subsection{Forecasting from process-based models}
 
 Ecological forecasting is a broad field with a diverse range of methods.
-Process-based modeling is often considered the gold standard in ecology \citep{Urban2016, Pilowsky2022} and beyond.
+Our second case-study focuses on process-based modeling, which is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond.
 Newer models generally incorporate greater complexity and an ever-growing number of parameters, making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.
@@
 Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome.
-In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
+In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
 This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}.
@@
 Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
-% jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
-[FLAG: agree but not really relevant to PBM?
-Maybe it's not clear enough we are focusing on PBM in this section...]
+\todo[size=\tiny]{jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
+Reply: we agree but not really relevant to PBM?
+Maybe it's not clear enough we are focusing on PBM in this section, which we tried to change above, see if it works}
 If a model becomes too complex, it may become a black box, % mention black box here to line up with the beginning of next paragraph
@@
 Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
-% jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty ....
-[FLAG: agree but not really relevant to PBM?
-Maybe it's not clear enough we are focusing on PBM in this section...]
+\todo[size=\tiny]{jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty ....
+Reply: Same as just above.}
 
-% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties?
+Process-based models used to project species or ecosystem distributions highlight some of these problems.
+Most of the focus is on the model projections---often represented as maps without uncertainties.
+Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only one of the many different output variables of the model.
+%The model building and calibration steps are usually limited to the Methods section, and estimated parameter values are rarely discussed \citep[e.g.][]{Conradi2024, VanderMeersch2025a}.
+The complexity of these models often makes it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability---such work is generally done as a separate effort, exposing potential model problems later on \citep{VanderMeersch2025b}, and not required as a preliminary step.
+\todo[inline, size=\tiny]{For above, addressing: jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties?
 This could help illustrate potential degeneracies etc.
-% [FLAG: I tried something, but I'm not really illustrating degeneracies + I don't like citing myself as much + I'm not really satisfied]
-An illustration of the issues with the current forecast workflow is the process-based models used to project species or ecosystem distributions.
-Most of the focus is on the model projections---often represented as maps without uncertainties.
-Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only of the many different output variables of the model.
-The model building and calibration steps are usually limited to the Methods section, and estimated parameter values are rarely discussed \citep[e.g.][]{Conradi2024, VanderMeersch2025a}.
-The complexity of these models often makes it difficult to assess the robustness of their calibration, whether for a reviewer or a reader.
-If ever done, the assesment of the model degeneracies is often done as a a separate work, exposing model degeneracies later on \citep{VanderMeersch2025b}, and not required as a preliminary step.
+Reply: Tried something, but not sure it works so feedback much appreciated!}
 
+%emw19Apr -- Can we just get away with using non-identifiable and save the reader also needing to digest degenerate?
+(https://betanalpha.github.io/assets/case_studies/identifiability.html)?
+I tried that on this draft, see how it works!
 Applying the workflow to process-based models would help open the black box.
 Each successive step of model development in the workflow may highlight current problems and a path to solutions.
-Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential degeneracies in the model design.
-%jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below) [FLAG: agree, and made me realize it was not 100% clear in my mind.
-As I understand it, non-identifiability concerns one parameter, and degeneracy concerns the all model?...
-but not sure]
-Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
-The workflow may also highlight degeneracies by requiring more explicit model calibration (i.e.
+Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential parameter identifiability issues in the model design.
+%jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below)
+Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
+The workflow may also highlight model problems by requiring more explicit model calibration (i.e.
 data fitting), % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
-which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated).
+which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model.
 Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
@@
 
-% jdavies15april: might be better grouped together in concluding section
-% [FLAG: I feel if we want to put this in the concluding section, we would need to talk also about the "opportunities to bridge statistical and process-based", rather than keeping it in the box?
-So, what should we put in the box?]
+\todo[inline, size=\tiny]{For below: jdavies15april: might be better grouped together in concluding section.
+Reply: if we want to put this in the concluding section, we would need to talk also about the "opportunities to bridge statistical and process-based", rather than keeping it in the box?
+So, what should we put in the box?}
 Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
@@
 
-This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference.
+This shift would present a significant challenge---as it would likely reveal many issues related to non-identifiability in models and data limitations before achieving robust inference.
 But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.
@@
 The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results.
-% jdavies16april: could highlight how influential the LPI has become (recent reports suggest a 73% ave.
+% \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave.
 pop decline since 1970.
-This would be massive!
+This would be massive!}
 With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
 For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}.
-% jdavies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations.
+% \todo[]{davies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations.
 We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends.
-For this we need to predict outside of the set ...
-[FLAG]
+For this we need to predict outside of the set}
 
@@
 In ecological training today researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
-% Such training leaves a large number of ecologists adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is.
-% jdavies15april: this may be true, but I am not sure we want to broadly disparage ecologists
 
@@
 
-% Training all ecologists in our proposed workflow would break down barriers between these different groups with likely major benefits for science. % jdavies15april: this feels a little OTT
 Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them.

### 2025-04-22 analyses/african_mammals_workflow.qmd
start working on workflow with African mammals

--- 2025-04-18 analyses/sampleworkflow.Rmd
@@
-# A sample workflow
+---
+title: "A sample workflow"
+authors: "Lizzie, Victor"
+date: "April 2025"
+date-format: "MMMM YYYY"
+toc: true
+number-sections: true
+highlight: pygments
+execute:
+  enabled: true
+format:
+  html:
+    html-math-method: katex
+    theme:
+      - lux
+    embed-resources: true
+    code-overflow: wrap
+    linkcolor: "#B97C7C"
+    page-layout: full
+    fig-width: 8
+    fig-height: 4
+  pdf:
+    keep-tex: true
+    fig-width: 5.5
+    fig-height: 5.5
+    code-overflow: wrap
+    monofontoptions:
+      - Scale=0.5
+knitr:
+  opts_chunk:
+    comment: ''
+  opts_knit:
+    global.par: TRUE
+format-links: false
+---
 
@@
 For more information on the LPI, check out the main text and www.livingplanetindex.org/.
+
+**Note**: data from Jonathan comes from *LivingPlanetIndexDatabase_2025-02-01_19.19.51*.
 
@@
 What is global trend in vertebrate species population numbers, both overall and how does it vary acrosss species and populations?
+
+```{r}
+set.seed(20012029)
+
+library(ggplot2)
+
+lpi_subset <- read.table("/home/victor/projects/forecastflows/analyses/jdavies/African_mammals.txt")
+
+```
 
@@
 
-What is trend in terrestrial vertebrate species population numbers in South Africa, both overall and how does it vary acrosss species and populations?
+What is trend in mammal species population numbers in Africa, both overall and how does it vary acrosss species and populations? 
+**Note**: we will actually focus on African mammals with 10+ populations with 5+ observations per population.
 
+### Simulating data
 Now we start to simulate!
@@
 
+nspecies <- 9
+npops_persp <- round(runif(nspecies, 1, 3)) # no. of different populations per species
+npops <- sum(npops_persp) # total no. of different populations
+nobs_perpop <- round(runif(npops, 5, 15)) # no. of different observations per population
+
+spid_perpop <- rep(1:nspecies, times = npops_persp)
+spid_perobs <- rep(spid_perpop, times = nobs_perpop)
+popid_perobs <- rep(1:npops, times = nobs_perpop)
+
+years <- c()
+for(np in nobs_perpop){
+  years_p <- sample(1960:2020, size = np, replace = FALSE)
+  years <- c(years, years_p)
+}
+
+# nested intercepts
+mu_species_alpha <- rnorm(n = nspecies, mean = 200, sd = 25)
+sig_species_alpha <- rnorm(n = nspecies, mean = 5, sd = 2)
+mu_pop_alpha <- rnorm(n = npops, mean = mu_species_alpha[spid_perpop])
+sig_pop_alpha <- abs(rnorm(n = npops))
+alpha <- rnorm(n = npops, mean = mu_pop_alpha, sd = sig_pop_alpha)
+
+# nested slopes
+mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.1)
+sig_species_beta <- rnorm(n = nspecies, mean =0, sd = 0.5)
+mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = 0.2)
+sig_pop_beta <- abs(rnorm(n = npops))
+beta <- rnorm(n = npops, mean = mu_pop_beta, sd = sig_pop_beta)
+
+sigma_obs <- abs(rnorm(1, mean = 10, sd = 1))
+
+y <- c()
+for(p in 1:npops){
+  nobs_p <- nobs_perpop[p]
+  alpha_p <- alpha[p]
+  beta_p <- beta[p]
+
+  years_p <- years[popid_perobs == p]
+
+  yhat_p <- alpha_p + beta_p * (years_p - 1980)
+  # yhat_p <- alpha_p
+  y_p <- rnorm(n = nobs_p, mean = yhat_p, sd = sigma_obs)
+  y <- c(y, y_p)
+}
+
+dataplot <- data.frame(
+  y,
+  year = years - 1980,
+  species = as.character(spid_perobs),
+  population = as.character(popid_perobs)
+)
+
+ggplot(data = dataplot) +
+  geom_line(aes(x = year+1980, y = y, 
+                group = paste0(species, population),
+                color = species)) +
+  theme_bw()
+
+nestlm <- lm(y ~ year * species/population, data = dataplot)
+coefs <- summary(nestlm)$coefficients
+
+species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
+species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp), "Std. Error"]}))
+
+species_slope <- c(coefs["year", "Estimate"], sapply(2:nspecies, function(sp){coefs["year", "Estimate"] + coefs[paste0("year:species", sp), "Estimate"]}))
+species_slope_sd <- c(coefs["year", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("year:species", sp), "Std. Error"]}))
+
+intercepts <- data.frame(obs=mu_species_alpha, est=species_intercept, est.sd=species_intercept_sd, species = as.character(1:nspecies))
+ggplot(data = intercepts) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
+  geom_pointrange(aes(ymin = est-est.sd, ymax = est+est.sd, x = obs, y = est, color = species)) +
+  theme_bw()
+
+slopes <- data.frame(obs=mu_species_beta, est=species_slope, est.sd=species_slope_sd, species = as.character(1:nspecies))
+ggplot(data = slopes) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
+  geom_pointrange(aes(ymin = est-est.sd, ymax = est+est.sd, x = obs, y = est, color = species)) +
+  theme_bw()
+
 ```
-*Add some text here showing the model output returning paramaters.*
+*Add some text here showing the model output returning parameters.*
 
@@
 
+I might also add a short paragraph about moving away from linear regression and towards models for species abudance (via mixtures I think?
+I suspect the articles I posted for class have some good ideas) and thinking more about linking to demography models.
+
 And then we would realize we really don't have great data for this question and we would ...

### 2025-04-22 analyses/jdavies/LPI subset African mammals.R

@@ -1,0 +1,33 @@
+################################################################################################
+################################################################################################
+
+setwd("C:/Jonathan/Dropbox/BLUE/Zurich/LPI/LivingPlanetIndexDatabase_2025-02-01_19.19.51")
+data<-read.csv("LPD_2024_public.csv", header=T)
+
+Af1<-subset(data, Region == "Africa")
+Af2<-subset(Af1, Included.in.LPR2024 == 1)
+Af3<-subset(Af2, Class == "Mammalia")
+
+#ignoring units for now
+#unique(Af$Units)
+
+# subset 5+ counts
+obs.cnt<-NULL
+for (n in 1:dim(Af3)[1]){
+obs.cnt[n]<-sum(Af3[n, 31:101]!="NULL")
+}
+
+Af4<-Af3[obs.cnt>4,]
+
+# subset species 10+ populations
+sp<-unique(Af4$Binomial)#
+
+sp.cnt<-NULL
+for(x in 1:length(sp)){
+Af.tmp<-subset(Af4, Binomial == sp[x])
+sp.cnt[x]<-dim(Af.tmp)[1]
+}
+
+keep.sp<-sp[sp.cnt>9]
+Af5<-Af4[Af4$Binomial %in% keep.sp,]
+#14 sp remaining

### 2025-04-23 analyses/african_mammals_workflow.qmd
slowly continuing workflow example, but LPI is a mess

@@
 lpi_subset <- read.table("/home/victor/projects/forecastflows/analyses/jdavies/African_mammals.txt")
+
+lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Units", paste0("X", 1950:2020))]
+lpi_subset$genus_species <- paste(lpi_subset$Genus, lpi_subset$Species)
+
+lpi_subset_rsh <- 
+  reshape(lpi_subset, direction = "long", idvar = "ID",
+          varying =  paste0("X", 1950:2020), v.names = c("value"), timevar = "year", times = 1950:2020)
+lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != "NULL",]
 
@@
 
+```{r}
+
+lpi_subset <- read.table("/home/victor/projects/forecastflows/analyses/jdavies/African_mammals.txt")
+
+lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Units", paste0("X", 1950:2020))]
+lpi_subset$genus_species <- paste(lpi_subset$Genus, lpi_subset$Species)
+
+lpi_subset_rsh <- 
+  reshape(lpi_subset, direction = "long", idvar = "ID",
+          varying =  paste0("X", 1950:2020), v.names = c("value"), timevar = "year", times = 1950:2020)
+lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != "NULL",]
+lpi_subset_rsh$value <- as.numeric(lpi_subset_rsh$value)
+
+units_to_keep <- c("Estimate - entire population", "Estimate - entire population (August)",
+                   "Individual", "individuals", "Individuals", "Number of adults", "number of individuals",
+                   "Number of individuals", "Number of Individuals", "Number of rhinos", "numbers of indviduals",
+                   "population estimate", "Population estimate", "Population estimate (individuals)", 
+                   "Population estimates", "Total population size")
+
+lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$Units %in% units_to_keep,]
+
+ggplot(data = lpi_subset_rsh) +
+  geom_line(aes(x = year, y = value, 
+                group = paste0(genus_species, ID),
+                color = genus_species)) +
+  theme_bw()
+
+```
+
+```{r}
+
+lpi_subset_rsh$year <- lpi_subset_rsh$year-1980
+nestlm <- lm(value ~ year * genus_species/ID, data = lpi_subset_rsh)
+coefs <- summary(nestlm)$coefficients
+
+species <- unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]
+
+species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(species[2:length(species)], function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("genus_species", sp), "Estimate"]}))
+species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(species[2:length(species)], function(sp){coefs[paste0("genus_species", sp), "Std. Error"]}))
+
+intercepts <- data.frame(est=species_intercept, est.sd=species_intercept_sd, species)
+
+species_slope <- c(coefs["year", "Estimate"], sapply(species[2:length(species)], function(sp){coefs["year", "Estimate"] + coefs[paste0("year:genus_species", sp), "Estimate"]}))
+species_slope_sd <- c(coefs["year", "Std. Error"], sapply(species[2:length(species)], function(sp){coefs[paste0("year:genus_species", sp), "Std. Error"]}))
+
+slopes <- data.frame(est=species_slope, est.sd=species_slope_sd, species)
+
+```
+
 Here we finally get to see how our model performs on the empirical data and are a step closer to answering our question, but we also start to notice some differences in our empirical data from our simulated data.
@@
 
-Now we check how well our model predicts our data ...
+```{r}
+#| fig-height: 10
+#| fig-width: 12
+
+retrodata <- data.frame(obs = lpi_subset_rsh$value, pred = predict(nestlm),
+                        species = lpi_subset_rsh$genus_species, id = lpi_subset_rsh$ID)
+
+ggplot(data = retrodata) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
+  facet_wrap(~ species, scales = "free") +
+  geom_point(aes(x = obs, y = pred, group = id, color = as.character(id)),
+             size = 0.5) + 
+  theme_bw() + theme(legend.position = 'none')
+
+```
 

### 2025-04-24 analyses/african_mammals_workflow.qmd
some stuff I worked on, on the plane, still pretty much a draft

@@
     fig-width: 5.5
-    fig-height: 5.5
+    fig-height: 7.5
     code-overflow: wrap
@@
 ```{r}
+wd <- "/home/victor/projects/forecastflows/analyses"
+
 set.seed(20012029)
@@
 
-lpi_subset <- read.table("/home/victor/projects/forecastflows/analyses/jdavies/African_mammals.txt")
+lpi_subset <- read.table(file.path(wd, "jdavies/African_mammals.txt"))
 
@@
 *Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.* 
-```{r}
+```{r sim data}
 
 nspecies <- 9
-npops_persp <- round(runif(nspecies, 1, 3)) # no. of different populations per species
+npops_persp <- round(runif(nspecies, 2, 15)) # no. of different populations per species
 npops <- sum(npops_persp) # total no. of different populations
-nobs_perpop <- round(runif(npops, 5, 15)) # no. of different observations per population
+nobs_perpop <- round(runif(npops, 10, 40)) # no. of different observations per population
 
@@
 for(np in nobs_perpop){
-  years_p <- sample(1960:2020, size = np, replace = FALSE)
+  years_p <- sample(1950:2020, size = np, replace = FALSE)
   years <- c(years, years_p)
@@
 # nested intercepts
-mu_species_alpha <- rnorm(n = nspecies, mean = 200, sd = 25)
-sig_species_alpha <- rnorm(n = nspecies, mean = 5, sd = 2)
+mu_species_alpha <- rnorm(n = nspecies, mean = log(200), sd = log(5))
+sig_species_alpha <- rnorm(n = nspecies, mean = 5, sd = 5)
 mu_pop_alpha <- rnorm(n = npops, mean = mu_species_alpha[spid_perpop])
@@
 # nested slopes
-mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.1)
-sig_species_beta <- rnorm(n = nspecies, mean =0, sd = 0.5)
-mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = 0.2)
-sig_pop_beta <- abs(rnorm(n = npops))
+mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.01)
+sig_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.05)
+mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = 0.05)
+sig_pop_beta <- abs(rnorm(n = npops, mean = 0, sd = 0.01))
 beta <- rnorm(n = npops, mean = mu_pop_beta, sd = sig_pop_beta)
 
-sigma_obs <- abs(rnorm(1, mean = 10, sd = 1))
+sigma_obs <- abs(rnorm(1, mean = 0.25, sd = 0.25))
 
@@
   # yhat_p <- alpha_p
-  y_p <- rnorm(n = nobs_p, mean = yhat_p, sd = sigma_obs)
+  y_p <- exp(rnorm(n = nobs_p, mean = yhat_p, sd = sigma_obs))
   y <- c(y, y_p)
@@
   species = as.character(spid_perobs),
-  population = as.character(popid_perobs)
+  population = paste0(spid_perobs,popid_perobs)
 )
+
+dataplot$logy <- log(dataplot$y)
 
@@
 
-nestlm <- lm(y ~ year * species/population, data = dataplot)
+ggplot(data = dataplot) +
+  geom_line(aes(x = year+1980, y = logy, 
+                group = paste0(species, population),
+                color = species)) +
+  theme_bw()
+
+```
+
+### Fit model on simulated data!
+
+```{r fit sim data}
+
+nestlm <- lm(logy ~ species/population + year * species/population, data = dataplot)
 coefs <- summary(nestlm)$coefficients
 
+```
+
+#### Retrodictive check with model fitted on simulated data
+
+```{r retro sim data}
+#| fig-height: 6
+#| fig-width: 8
+
+retrodata <- data.frame(obs = dataplot$logy, pred = predict(nestlm),
+                        year = dataplot$year,
+                        species = dataplot$species, id = dataplot$population)
+
+ggplot(data = retrodata) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
+  facet_wrap(~ species, scales = "free") +
+  geom_point(aes(x = obs, y = pred, group = id, color = as.character(id)),
+             size = 0.5) + 
+  theme_bw() + theme(legend.position = 'none')
+
+ggplot(data = retrodata) +
+  facet_wrap(~ paste0("Species ", species), scales = "free") +
+  geom_point(aes(x = year, y = obs, group = id, color = as.character(id)),
+             size = 1) + 
+  geom_line(aes(x = year, y = pred, group = id, color = as.character(id))) +
+  theme_bw() + theme(legend.position = 'none')
+
+ggplot(data = retrodata) +
+  facet_wrap(~ paste0("Species ", species), scales = "free") +
+  geom_point(aes(x = year, y = exp(obs), group = id, color = as.character(id)),
+             size = 1) + 
+  geom_line(aes(x = year, y = exp(pred), group = id, color = as.character(id))) +
+  theme_bw() + theme(legend.position = 'none')
+
+```
+
+#### Inference with model fitted on simulated data
+
+The model does a poor job at recovering species-level parameters.
+
+```{r inference sim data}
 species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
@@
 
-species_slope <- c(coefs["year", "Estimate"], sapply(2:nspecies, function(sp){coefs["year", "Estimate"] + coefs[paste0("year:species", sp), "Estimate"]}))
-species_slope_sd <- c(coefs["year", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("year:species", sp), "Std. Error"]}))
+species_slope <- c(coefs["year", "Estimate"], sapply(2:nspecies, function(sp){coefs["year", "Estimate"] + coefs[paste0("species", sp, ":year"), "Estimate"]}))
+species_slope_sd <- c(coefs["year", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp,":year"), "Std. Error"]}))
 
@@
   theme_bw()
-
 ```
-*Add some text here showing the model output returning parameters.*
 
@@
 
-```{r}
+```{r fitdata}
+#| eval: true
 
@@
 
+ggplot(data = lpi_subset_rsh[lpi_subset_rsh$value != 0,]) +
+  geom_line(aes(x = year, y = log(value), 
+                group = paste0(genus_species, ID),
+                color = genus_species)) +
+  theme_bw()
+
 ```
@@
 ```{r}
+#| eval: false
 
+runmodel <- TRUE
+
+lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,]
 lpi_subset_rsh$year <- lpi_subset_rsh$year-1980
-nestlm <- lm(value ~ year * genus_species/ID, data = lpi_subset_rsh)
+lpi_subset_rsh$logvalue <- log(lpi_subset_rsh$value)
+lpi_subset_rsh$ID <- as.character(lpi_subset_rsh$ID)
+
+if(runmodel){
+  nestlm <- lm(logvalue ~ genus_species/ID + year * genus_species/ID, data = lpi_subset_rsh)
+
+  saveRDS(nestlm, file = file.path(wd, "output/fit/nestmodel1.rds"))
+}else{
+  nestlm <- readRDS(file.path(wd, "output/fit/nestmodel1.rds"))
+}
+
 coefs <- summary(nestlm)$coefficients
@@
 #| fig-width: 12
+#| eval: false
 
-retrodata <- data.frame(obs = lpi_subset_rsh$value, pred = predict(nestlm),
+retrodata <- data.frame(obs = lpi_subset_rsh$logvalue, pred = predict(nestlm),
+                        year = lpi_subset_rsh$year, units = lpi_subset_rsh$Units,
                         species = lpi_subset_rsh$genus_species, id = lpi_subset_rsh$ID)
@@
              size = 0.5) + 
+  theme_bw() + theme(legend.position = 'none')
+
+```
+
+```{r}
+#| fig-height: 8
+#| fig-width: 8
+#| eval: false
+
+ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
+  facet_wrap(~ id, scales = "free") +
+  geom_point(aes(x = year, y = obs, group = id, color = id),
+             size = 1) + 
+  geom_line(aes(x = year, y = pred, group = id, color = id)) +
   theme_bw() + theme(legend.position = 'none')

### 2025-04-25 analyses/african_mammals_workflow.qmd
retro and inference on simulated data with marginal means

@@
 *Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.*
+
 ```{r sim data}
@@
 
+# save simulated intercepts and slopes
+pops_fit <- data.frame()
+pid <- 1
+for(sp in 1:nspecies){
+  for(p in 1:npops_persp[sp]){
+    pops_fit <- rbind(pops_fit, data.frame(species = sp, pop = pid))
+    pid <- pid + 1
+  }
+}
+pops_fit$int <- mu_pop_alpha
+pops_fit$slp <- mu_pop_beta
+
 sigma_obs <- abs(rnorm(1, mean = 0.25, sd = 0.25))
@@
 
-dataplot <- data.frame(
+datasim <- data.frame(
   y, 
@@
 
-dataplot$logy <- log(dataplot$y)
+datasim$logy <- log(datasim$y)
 
-ggplot(data = dataplot) +
+ggplot(data = datasim) +
   geom_line(aes(x = year+1980, y = y, 
-                group = paste0(species, population),
+                group = population,
                 color = species)) +
-  theme_bw()
+  theme_bw() +
+  labs(x = 'Year', y = 'Simulated counts', colour = 'Species')
 
-ggplot(data = dataplot) +
+ggplot(data = datasim) +
   geom_line(aes(x = year+1980, y = logy, 
-                group = paste0(species, population),
+                group = population,
                 color = species)) +
-  theme_bw()
+  theme_bw() +
+  labs(x = 'Year', y = 'Simulated counts (log-scale)', colour = 'Species')
 
@@
 
-nestlm <- lm(logy ~ species/population + year * species/population, data = dataplot)
+nestlm <- lm(logy ~ species/population + year * species/population, data = datasim)
 coefs <- summary(nestlm)$coefficients
@@
 
-retrodata <- data.frame(obs = dataplot$logy, pred = predict(nestlm),
-                        year = dataplot$year,
-                        species = dataplot$species, id = dataplot$population)
+retrodata <- data.frame(obs = datasim$logy, pred = predict(nestlm),
+                        year = datasim$year,
+                        species = datasim$species, id = datasim$population)
 
@@
   geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
-  facet_wrap(~ species, scales = "free") +
+  facet_wrap(~ paste0("Species ", species), scales = "free") +
   geom_point(aes(x = obs, y = pred, group = id, color = as.character(id)),
              size = 0.5) + 
-  theme_bw() + theme(legend.position = 'none')
+  theme_bw() + theme(legend.position = 'none') +
+  labs(x = 'Simulated counts (log-scale)', y = 'Estimated counts (log-scale)')
 
@@
   geom_line(aes(x = year, y = pred, group = id, color = as.character(id))) +
-  theme_bw() + theme(legend.position = 'none')
+  theme_bw() + theme(legend.position = 'none') +
+  labs(x = 'Year', y = 'Counts (log-scale)')
 
@@
   geom_line(aes(x = year, y = exp(pred), group = id, color = as.character(id))) +
-  theme_bw() + theme(legend.position = 'none')
+  theme_bw() + theme(legend.position = 'none') +
+   labs(x = 'Year', y = 'Counts')
 
@@
 
-The model does a poor job at recovering species-level parameters.
+The model does a fair job at recovering population-level parameters.
 
 ```{r inference sim data}
+#| eval: false
+#| echo: false
+#| 
 species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
@@
   theme_bw()
+
+```
+
+```{r inference pop sim data}
+
+em <- data.frame(emmeans::emmeans(nestlm, ~ species/population))
+pops_fit$int_fit <- em$emmean
+pops_fit$int_fit.lcl <- em$lower.CL
+pops_fit$int_fit.ucl <- em$upper.CL
+
+em <- data.frame(emmeans::emtrends(nestlm,  ~ species/population, var = 'year'))
+pops_fit$slp_fit <- em$year.trend
+pops_fit$slp_fit.lcl <- em$lower.CL
+pops_fit$slp_fit.ucl <- em$upper.CL
+
+ggplot(data = pops_fit) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
+  geom_pointrange(aes(ymin = int_fit.lcl, ymax = int_fit.ucl, x = int, y = int_fit, color = as.character(species))) +
+  theme_bw() +
+  labs(x = 'Simulated intercepts', y = 'Estimated intercepts', colour = 'Species')
+
+ggplot(data = pops_fit) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
+  geom_pointrange(aes(ymin = slp_fit.lcl, ymax = slp_fit.ucl, x = slp, y = slp_fit, color = as.character(species))) +
+  theme_bw() +
+  labs(x = 'Simulated slopes', y = 'Estimated slopes', colour = 'Species')
+
 ```
@@
 
-```{r fitdata}
+```{r lookdata}
 #| eval: true
@@
 
-```{r}
-#| eval: false
+```{r fitdata}
+#| eval: true
 
@@
 lpi_subset_rsh$logvalue <- log(lpi_subset_rsh$value)
-lpi_subset_rsh$ID <- as.character(lpi_subset_rsh$ID)
+lpi_subset_rsh$ID <- as>character(lpi_subset_rsh$ID)
 

### 2025-04-25 analyses/african_mammals_workflow.qmd
some small progress

@@
 ---
-
-*Hi Victor, it's Lizzie.
-I will just put my notes to you in italics.
-You really should remove them.
-In general, I suggest you focus on getting some code/plots and basic text done that you then share next week with everyone.*
 
@@
 
-y_i ~ normal(mu, sigma)
+$$\log(y_i) \sim \text{normal}(\mu_i, \sigma)$$
 
-mu = alpha[sp[pop]] + beta[sp[pop]]*year
+$$\mu_i = \alpha_{\text{sp}[i],\text{pop}[i]} + \beta_{\text{sp}[i],\text{pop}[i]} * \text{year}_i$$
 
-*Need to finish the notation as I always have to look up the nesting notation.*
+$$\alpha_{\text{sp}[i],\text{pop}[i]} \sim \text{normal}(\mu_{\text{pop}[i]}^\alpha, \sigma_{\text{pop}[i]}^\alpha) \\
+\mu_{\text{pop}[i]}^\alpha \sim \text{normal}(\mu_{\text{sp}[i]}^\alpha, \sigma_{\text{sp}[i]}^\alpha)$$
+
+$$\beta{\text{sp}[i],\text{pop}[i]} \sim \text{normal}(\mu_{\text{pop}[i]}^\beta, \sigma_{\text{pop}[i]}^\beta) \\
+\mu_{\text{pop}[i]}^\beta \sim \text{normal}(\mu_{\text{sp}[i]}^\beta, \sigma_{\text{sp}[i]}^\beta)$$
+
+*I'm not 100% sure with the notation here.*
 
@@
 
-*Obviously we can charge forward with the full model if you want, I just thought this would be (a) easier for us, (b) way better to show as it's how you should do it and (c) related to a, I suspect JD can be more help, though any country with a good amount of data could work.*
 So here we refine our question to:
 
-What is trend in mammal species population numbers in Africa, both overall and how does it vary acrosss species and populations? 
-**Note**: we will actually focus on African mammals with 10+ populations with 5+ observations per population.
+What is trend in mammal species population numbers in Africa, both overall and how does it vary across species and populations?
 
 ### Simulating data
+
 Now we start to simulate!
-*Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.*
 
@@
 
-nspecies <- 9
+nspecies <- 9 # no. of species
 npops_persp <- round(runif(nspecies, 2, 15)) # no. of different populations per species
@@
 # nested intercepts
-mu_species_alpha <- rnorm(n = nspecies, mean = log(200), sd = log(5))
-sig_species_alpha <- rnorm(n = nspecies, mean = 5, sd = 5)
-mu_pop_alpha <- rnorm(n = npops, mean = mu_species_alpha[spid_perpop])
+mu_species_alpha <- rnorm(n = nspecies, mean = log(200), sd = log(5)) # simulate some species intercepts
+sig_species_alpha <- abs(rnorm(n = nspecies, mean = 0, sd = 2))
+mu_pop_alpha <- rnorm(n = npops, mean = mu_species_alpha[spid_perpop], sd = sig_species_alpha[spid_perpop]) # population intercepts, nested in species
 sig_pop_alpha <- abs(rnorm(n = npops))
@@
 # nested slopes
-mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.01)
-sig_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.05)
-mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = 0.05)
+mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.01) # simulate some species slopes
+sig_species_beta <- abs(rnorm(n = nspecies, mean = 0, sd = 0.05))
+mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = sig_species_beta[spid_perpop]) # population slopes, nested in species
 sig_pop_beta <- abs(rnorm(n = npops, mean = 0, sd = 0.01))
@@
 
-sigma_obs <- abs(rnorm(1, mean = 0.25, sd = 0.25))
+sigma_obs <- abs(rnorm(1, mean = 0.25, sd = 0.25)) # common observational error
 
@@
 #| echo: false
-#| 
+
 species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
@@
 
-lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Units", paste0("X", 1950:2020))]
+lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Country", "Units", paste0("X", 1950:2020))]
 lpi_subset$genus_species <- paste(lpi_subset$Genus, lpi_subset$Species)
@@
                 color = genus_species)) +
-  theme_bw()
+  theme_bw() +
+  labs(x = 'Year', y = 'Observed counts')
 
@@
                 color = genus_species)) +
-  theme_bw()
+  theme_bw() +
+  labs(x = 'Year', y = 'Observed counts (log-scale)')
 
@@
 
-runmodel <- TRUE
+runmodel <- FALSE
 
-lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,]
+lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,] # poor way to avoid log(0)
 lpi_subset_rsh$year <- lpi_subset_rsh$year-1980
 lpi_subset_rsh$logvalue <- log(lpi_subset_rsh$value)
-lpi_subset_rsh$ID <- as>character(lpi_subset_rsh$ID)
+lpi_subset_rsh$ID <- as.character(lpi_subset_rsh$ID)
 
@@
   nestlm <- lm(logvalue ~ genus_species/ID + year * genus_species/ID, data = lpi_subset_rsh)
-
   saveRDS(nestlm, file = file.path(wd, "output/fit/nestmodel1.rds"))
@@
 }
-
-coefs <- summary(nestlm)$coefficients
-
-species <- unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]
-
-species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(species[2:length(species)], function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("genus_species", sp), "Estimate"]}))
-species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(species[2:length(species)], function(sp){coefs[paste0("genus_species", sp), "Std. Error"]}))
-
-intercepts <- data.frame(est=species_intercept, est.sd=species_intercept_sd, species)
-
-species_slope <- c(coefs["year", "Estimate"], sapply(species[2:length(species)], function(sp){coefs["year", "Estimate"] + coefs[paste0("year:genus_species", sp), "Estimate"]}))
-species_slope_sd <- c(coefs["year", "Std. Error"], sapply(species[2:length(species)], function(sp){coefs[paste0("year:genus_species", sp), "Std. Error"]}))
-
-slopes <- data.frame(est=species_slope, est.sd=species_slope_sd, species)
 
@@
 #| fig-width: 12
-#| eval: false
+#| eval: true
 
@@
                         year = lpi_subset_rsh$year, units = lpi_subset_rsh$Units,
-                        species = lpi_subset_rsh$genus_species, id = lpi_subset_rsh$ID)
+                        species = lpi_subset_rsh$genus_species, id = lpi_subset_rsh$ID,
+                        country = lpi_subset_rsh$Country)
 
@@
              size = 0.5) + 
-  theme_bw() + theme(legend.position = 'none')
+  theme_bw() + theme(legend.position = 'none') +
+  labs(x = 'Observed counts (log-scale)', y = 'Estimated counts (log-scale)')
 
@@
 #| fig-height: 8
-#| fig-width: 8
-#| eval: false
+#| fig-width: 10
+#| eval: true
 
 ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
-  facet_wrap(~ id, scales = "free") +
-  geom_point(aes(x = year, y = obs, group = id, color = id),
+  facet_wrap(~ country, scales = "free") +
+  geom_point(aes(x = year + 1980, y = obs, group = id, color = id),
              size = 1) + 
-  geom_line(aes(x = year, y = pred, group = id, color = id)) +
-  theme_bw() + theme(legend.position = 'none')
+  geom_line(aes(x = year + 1980, y = pred, group = id, color = id)) +
+  theme_bw() + theme(legend.position = 'none') +
+  labs(y = 'Counts (log-scale)', x = 'Year')
+
+ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
+  facet_wrap(~ country, scales = "free") +
+  geom_point(aes(x = year + 1980, y = exp(obs), group = id, color = id),
+             size = 1) + 
+  geom_line(aes(x = year + 1980, y = exp(pred), group = id, color = id)) +
+  theme_bw() + theme(legend.position = 'none') +
+  labs(y = 'Counts', x = 'Year')
 

### 2025-04-25 analyses/african_mammals_workflow.qmd
additional plots

@@
 lpi_subset_rsh$value <- as.numeric(lpi_subset_rsh$value)
+```
 
+The values in the database are not necessarily in the same unit, so I imagine we need to take a decision here..?
+
+```{r units}
 units_to_keep <- c("Estimate - entire population", "Estimate - entire population (August)",
@@
 Here we finally get to see how our model performs on the empirical data and are a step closer to answering our question, but we also start to notice some differences in our empirical data from our simulated data.
-In preparing our data for the model, we realize we have major gaps in sampling years and ...
-something else?
+In preparing our data for the model, we realize we have quite different amplitudes of variations across populations and species (on a log-scale).
+
+```{r}
+#| fig-height: 4
+#| fig-width: 5
+
+ggplot(data = lpi_subset_rsh[lpi_subset_rsh$value != 0 & lpi_subset_rsh$genus_species == "Connochaetes taurinus",]) +
+  geom_line(aes(x = year, y = log(value), 
+                group = paste0(genus_species, ID),
+                color = genus_species)) +
+  theme_bw() + theme(legend.position = 'none') +
+  labs(x = 'Year', y = 'Observed counts (log-scale)')
+
+```
 
@@
 
-*See starterLPI.R for what I have so far with the data...*
+## Step 5: Retrodictive check
 
-## Step 5: Retrodictive check
+Now we use our fitted model parameters to generate predictions, and compare them to observations.
 
@@
 
+We might examine species separately, e.g. here the African bush elephant.
+
 ```{r}
-#| fig-height: 8
-#| fig-width: 10
+#| fig-height: 7
+#| fig-width: 9
 #| eval: true
@@
 ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
-  facet_wrap(~ country, scales = "free") +
+  facet_wrap(~ country, scales = "free_y") +
   geom_point(aes(x = year + 1980, y = obs, group = id, color = id),
@@
 ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
-  facet_wrap(~ country, scales = "free") +
+  facet_wrap(~ country, scales = "free_y") +
   geom_point(aes(x = year + 1980, y = exp(obs), group = id, color = id),
@@
 
+We can focus on populations in South Africa.
+
+```{r}
+#| fig-height: 3
+#| fig-width: 9
+#| eval: true
+
+ggplot(data = retrodata[retrodata$species == "Loxodonta africana" & retrodata$country == "South Africa", ]) +
+  facet_wrap( ~ id, scales = "free_y") +
+  geom_point(aes(x = year + 1980, y = obs, group = id, color = units),
+             size = 0.7) + 
+  geom_line(aes(x = year + 1980, y = obs, group = id, color = units),
+             linewidth = 0.3) + 
+  geom_line(aes(x = year + 1980, y = pred, group = id),
+            linewidth = 0.5) +
+  theme_bw() + 
+  labs(y = 'Counts (log-scale)', x = 'Year', color = 'Units')
+
+```
+
+```{r}
+#| fig-height: 3
+#| fig-width: 9
+#| eval: true
+
+ggplot(data = retrodata[retrodata$species == "Loxodonta africana" & retrodata$country == "South Africa", ]) +
+  facet_wrap( ~ id, scales = "free_y") +
+  geom_point(aes(x = year + 1980, y = exp(obs), group = id, color = units),
+             size = 0.7) + 
+  geom_line(aes(x = year + 1980, y = exp(obs), group = id, color = units),
+             linewidth = 0.3) + 
+  geom_line(aes(x = year + 1980, y = exp(pred), group = id),
+            linewidth = 0.5) +
+  theme_bw() + 
+  labs(y = 'Counts', x = 'Year', color = 'Units')
+
+```
+
+The three populations are measured in the same units, but likely represent different spatial scales (ranging from a few hundred individuals to more than 10000).
+
+The same holds for other species and other countries, e.g. the blue wildebeest in Tanzania.
+
+```{r}
+#| fig-height: 5
+#| fig-width: 9
+#| eval: true
+
+ggplot(data = retrodata[retrodata$species == "Connochaetes taurinus" & retrodata$country == "Tanzania, United Republic Of", ]) +
+  facet_wrap( ~ id, scales = "free_y") +
+  geom_point(aes(x = year + 1980, y = exp(obs), group = id, color = units),
+             size = 0.7) + 
+  geom_line(aes(x = year + 1980, y = exp(obs), group = id, color = units),
+             linewidth = 0.3) + 
+  geom_line(aes(x = year + 1980, y = exp(pred), group = id),
+            linewidth = 0.5) +
+  theme_bw() + 
+  labs(y = 'Counts', x = 'Year', color = 'Units')
+
+```
+The four populations are also measured in the same units (I guess?!), but represents different scales (so the blue one is likely a population, while the others represent region/country level?).
+
 ## Feedbacks
@@
 At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including questioning our assumption of one sigma for all populations and potentially our linearity assumption.
-We also may want to add predictors for the slope such as  ...
-We would do this, returning to Steps 2-5 likely several times until we have a model that is good enough for our purposes here.
+One possibility would be to move to a non-linear regression, and use a more flexible model
+Quick example for one of the wildbeest dataset in Tanzania:
+
+```{r}
+#| fig-height: 2.5
+#| fig-width: 3.5
+
+gompodel <- nls(logvalue ~ SSgompertz(year, Asym, b2, b3), data=lpi_subset_rsh[lpi_subset_rsh$ID == 5589,])
+fit <-data.frame(pred = predict(gompodel, newdata = data.frame(year = seq(-25,30,1))), year = seq(-25,30,1))
+
+ggplot(data = lpi_subset_rsh[lpi_subset_rsh$ID == 5589,]) +
+  facet_wrap( ~ ID, scales = "free_y") +
+  geom_point(aes(x = year + 1980, y = value),
+             size = 0.7, color = "#F8766D") + 
+  geom_line(aes(x = year + 1980, y = value),
+             linewidth = 0.3, color = "#F8766D") + 
+  geom_line(aes(x = year + 1980, y = exp(pred)),
+            linewidth = 0.5, data = fit) +
+  theme_bw() + 
+  labs(y = 'Counts', x = 'Year', color = 'Units')
+
+```
+
+We also may want to add predictors for the slope such as  ... We would do this, returning to Steps 2-5 likely several times until we have a model that is good enough for our purposes here.
 

### 2025-04-28 forecastflows.tex
Jim modifications

@@
 
-% Jim, if you want to add any of this into the abstract, please do!
-%There's a fundamental problem in how a lot of empirically but observationally driven (i.e., non-experimental) research is done in ecology and related fields.
-The crux of the issue comes down to a couple of things.
-First, the research community has two very distinct paradigms for modeling the system or phenomenon or interest: either fitting simple trend-like models with only one or a few parameters, or developing forecast models that often include potentially complex, mechanistic, and/or black box submodels largely unconstrained by data.
-Second, regardless of the approach, these models tend to be developed not only in isolation of one another, but also without a coherent strategy for linking modeling/statistical decisions, data, and scientific understanding (both as input and output) throughout the entire modeling process.
-We propose a unified, principled workflow for end-to-end empirical model development, incorporating sound statistical and scientific practices, and especially relying on data simulation to inform decisions at multiple steps in the process.
+\begin{abstract}
+Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts.
+Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions.
+While new research often suggests that models accounting for these complexities can yield more accurate trends and predictions, we argue that the usual workflows for model fitting in ecology make it impossible to evaluate and compare the models used to generate insights and predictions.
+First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data.
+Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process.
+We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process.
+We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could be transformative for ecological modeling.
 
-\begin{abstract}
-Increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts.
-Relevant datasets to meet these demands have also increased in size and complexity, including in their geographical, temporal and phylogenetic scales.
-While new research often suggests that accounting for these complexities has a major impact on projected trends and forecasts, we argue that the typical approach to model fitting in ecology makes it impossible to evaluate and compare the models used to generate these insights and predictions.
 % jdavies15april: I feel this could be a critical step somehow - sharing simulations could allow cross-validation between models or perhaps cross-calibration ...
@@
 
-Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets and another focused on forecasting from generally distinct datasets or mechanistic models based on less data.
-Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced with massive geographic, temporal and taxonomic biases.
+Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on less data.
+Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from massive geographic, temporal and taxonomic biases.
 Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
-Instead forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning.
-The current outcome from these approaches is no clear agreement on current species trends, with ongoing debates on the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.
+Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning.
+These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.
 
 We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
-Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting.
-At the same time, forecasting-focused process-based models often develop by adding new separate layers or components.
-These new parts are often disconnected from the original research aim, its data stream and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).
+Research estimating biodiversity trends has become focused on methodological aspects; the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting.
+At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components.
+These new parts are often disconnected from the original research aim, its data stream, and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).
 
-Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and potential problems, before estimating its parameters and making projections, could reduce many of these problems.
-In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where and why different models diverge.
+Workflows that fully integrate all the steps required to build a model from an ecological question, with evaluation of limitations and potential problems before estimating its parameters and making projections, could reduce many of these problems.
+In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge.
 Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
@@
 
-Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}.
-In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses.
+Quantitative science relies on a model-based framework to confront hypothesis with data \citep{Chamberlin:1965cd}.
+In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses.
 This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection.
@@
 Many important questions cannot be addressed through controlled experiments and replications.
-Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}, and most of macroecological insights emerge from exploring patterns in these datasets (exploratory data analysis).
+In such cases, we must rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.
+Indeed, most macroecological insights have emerged from exploring patterns in these datasets (exploratory data analysis).
 
@@
 
-With a model in place the next step focuses on testing and understanding it via data simulation (step 3, fig.
+With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, fig.
 \ref{fig:workflow}).
@@
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
-At this stage the focus is understanding the model, so researchers may have to spend a while to think through what parameters are reasonable or not.
-This is also a step for making sure the model is working as we expect, so simulated data often would have low error and high sample sizes (to make sure the model returns the parameter values used to generate the simulated data),
+At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is reasonable or not.
+This is also a step for making sure the model is working as expected, so simulated data often would have low error and high sample sizes (to make sure the model returns the parameter values used to generate the simulated data).
+%jr24april not sure what this final sentence means?
 
@@
 
-% More details, what's special! Feedbacks!
+% More details, what's special! Feedbacks! %jrApr25 I think this paragraph reads really well!
 A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
@@
 An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
-Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process.
+Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process.
 Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected.
@@
 In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data.
-The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results.
+The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results.
 % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave.
@@
 Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes.
-Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e.
-fix to a value found in the literature) or tuned to reproduce some observations or theory.
+Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce some observations or theory.
 The way models are currently calibrated is not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
@@
 % We say here briefly how great the new approach is or something.
-An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
+A universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
 Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
@@
 % jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
-Good model fitting is inherently iterative and takes time, at odds with the pressure to publish quickly.
-Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).
+The first is pressure to publish quickly, which can be at odds with the reality that good model fitting is inherently iterative and takes time.
+The second is reluctance to embrace adoption of open science practices that ensure modeling efforts are fully transparent and reproducible.
+It is imperative that we continue to place increasing value on research that is carefully developed, openly collaborative (including both data and code), and transparent about areas of uncertainty.
+Especially in ecology where samples sizes are low and effects are likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, both for individual researchers focused on making enduring contributions, and for the community as a whole focused on improving future data collection and model development (see Fig).
 
@@
 Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models.
-In ecological training today researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
+In ecological training today, researchers who conduct field and lab studies often learn a highly limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
 
@@
 % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms
-Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.
+Few of these groups have integrated data simulation into their statistical or scientific workflows.
+Simulation is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.
 While specialization is valuable, we argue the fundamental % training
@@
 This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them.
-Through this and retrodictive checks most ecologists would more easily think through what parameters are most critical to their question and or aim (e.g., management) and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
+Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
 Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models.
@@
 % Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
-Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.
+Those focused on learning complex statistical models may find many others in the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.
 
@@
 Machine learning will likely surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
-Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
+Similarly, engaging in statistical trend analysis, especially when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
 % jdavies15april: nothing intrinsically bad about ML, but sometimes the goal is just to understand the system, not make predictions or forecasts.
@@
 Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues.
-Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations, but we believe our workflow provides an tractable step to fixing this.
-By focusing on model development more tightly tied to ecological expertise, we ague this workflow should broader the community that contributes to model development.
-As ecologists are increasingly expanding their computational toolkits, many field, and lab and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological, and---most importantly---to interrogate them.
+Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations.
+However, we believe our workflow provides an tractable step to fixing this.
+By focusing on model development more tightly tied to ecological expertise, we argue this workflow should broaden the community that contributes to model development.
+As ecologists are increasingly expanding their computational toolkits, many field, and lab, and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological domains of interest, and---most importantly---to interrogate those models.
 

### 2025-04-28 forecastflows.tex
integrate JD comments

@@
 
-We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure?
+We argue a workflow that moves along the data-model space in a coherent sequence of steps % with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure?
 Or wait to mention simulation until referencing Fig 2?
@@
 Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
-And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
+And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimations.
 
@@
 It is imperative that we continue to place increasing value on research that is carefully developed, openly collaborative (including both data and code), and transparent about areas of uncertainty.
-Especially in ecology where samples sizes are low and effects are likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, both for individual researchers focused on making enduring contributions, and for the community as a whole focused on improving future data collection and model development (see Fig).
+Especially in ecology where samples sizes are low and effects are likely non-linear and complex, we believe adopting a workflow such as the one we propose here will pay off in the longer term, both for individual researchers focused on making enduring contributions, and for the community as a whole focused on improving future data collection and model development (see Fig).
 
@@
 Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training.
-Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models.
+Much of ecology still divides the world into training for those who gather data and learn a limited set of pre-built models versus those who develop more complex models.
 In ecological training today, researchers who conduct field and lab studies often learn a highly limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
 
-Instead, ecologists trained in a limited set of tests are expected to collaborate with others when they need more complex models, who were trained more in model development (though often for highly specific applications, such as wildlife population estimates, where they may rarely develop entirely new generative models).
+When ecologists trained in a limited set of tests require more complex models, they are expected to collaborate with others trained more in model development (though often for highly specific applications, such as wildlife population estimates, where generative models are often predefined).
 These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models.
 % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms
-Few of these groups have integrated data simulation into their statistical or scientific workflows.
+Neither group has fully integrated data simulation into their statistical or scientific workflows.
 Simulation is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.
@@
 
-Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them.
-This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them.
+Instead of siloing ecological training into experimental design and tests that may match certain designs (though rarely do for ecological data, CITES) versus model design and development, we suggest unified training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them.
+This would mean training to link the ecological processes they study with the mathematical models that may describe them.
 Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
-Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models.
+Empiricists % change suggested by JD, was empirical ecologists before...
+ would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models.
 Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting.
@@
 % jdavies15april: This could be the Concluding section highlighting the general advances provided by adopting your framework (and perhaps a better place for text currently at end of section 3.2).
-Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues.
+Today model development in ecology is rarely transparent, which limits how easily the research community can understand models, and thus identify potential issues.
 Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations.
-However, we believe our workflow provides an tractable step to fixing this.
+However, we believe our workflow provides a tractable step to fixing this.
 By focusing on model development more tightly tied to ecological expertise, we argue this workflow should broaden the community that contributes to model development.
+It may also help resolve apparent conflicts by identifying where divergences in model predictions emerge, whether from differences in assumptions, model structures, or other aspects of the modeling process.
 As ecologists are increasingly expanding their computational toolkits, many field, and lab, and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological domains of interest, and---most importantly---to interrogate those models.

### 2025-04-28 forecastflows.tex
move text

@@
 So, what should we put in the box?}
-Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
-The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model.
-It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.
-% or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza!
-disease!
-pathogens!)...
-or "rather than constantly expanding models by adding ever more processes"?
-Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
-Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty.
-Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.
-
-This shift would present a significant challenge---as it would likely reveal many issues related to non-identifiability in models and data limitations before achieving robust inference.
-But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.
 
@@
 \begin{minipage}[t]{\linewidth}
+	\vfill
 \includegraphics[width=\linewidth]{figures/iterativeworkflow_details}
-
+\vfill
 % \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
-\vspace*{1mm}
+\vspace*{3mm}
 \end{minipage}
@@
 \vspace*{1mm}
-Blabla... \\
-% We say here briefly how great the new approach is or something.
 A universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
 Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
-And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimations.
+It would be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimations.
+For both trend estimation and forecasting, the workflow would refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model.
 
@@
 
+Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
+Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
+Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty.
+Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.
+It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.
+
 % jdavies15april: This could be the Concluding section highlighting the general advances provided by adopting your framework (and perhaps a better place for text currently at end of section 3.2).
@@
 
+This shift would present a significant challenge---as it would likely reveal many issues related to non-identifiability in models and data limitations before achieving robust inference.
+But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.
+
 \clearpage

### 2025-04-29 analyses/african_mammals_workflow.qmd
Added some text to workflow, in progress

@@
 This download should include a csv file, here: LPD2022_public.csv, that we use below).
-For more information on the LPI, check out the main text and www.livingplanetindex.org/.
+We discuss the LPI in the main text; for more information, check out: www.livingplanetindex.org/.
 
@@
 
-*I started this, but it would be perfect for JD to fix/add to, so I did not finish it.
-(As mentioned above, I suggest you focus on getting some code/plots and basic text done that you then share next week with everyone.)*
-
 Here we need to focus on exactly what question we are trying to answer, which will guide us in the next steps.
 This step is sometimes skipped over or moved through so quickly that the exact question is not clear, but is critical for developing a useful model.
-We might start with: What is the global trend for biodiversity in vetebrate species?
+Many early papers on the LPI seem focused on something similar to: Are vertebrate species overall declining on average?
+And, if so, by how much?
+We, however might want to have a question that more clearly allows any directionality, so we could start with: What is the global trend for biodiversity in vetebrate species?
 This is rather broad and might mean we should develop a model that will yield one global estimate.
-In thinking about this, we might realize we don't want just that, we actually want to understand the variation across different species and perhaps populations and also a global estimate.
+In thinking about this, we would likely realize we don't want just that, we actually want to understand the variation across different species and perhaps populations and also a global estimate.
 So we might refine the question to:
 
-What is global trend in vertebrate species population numbers, both overall and how does it vary acrosss species and populations?
+What is global trend over time in vertebrate species population numbers, both overall and how does it vary acrosss species and populations?
 
@@
 We might also start with a simple linear model, where populations can go up, down or do nothing.
-That gets us started, but now we need to consider how we will model species and populations.
+
+Using a simple linear gets us started, but now we need to consider how we will model species and populations.
 Populations usually are nested in species in ecological models, so we might consider that to start.
-Our resulting model might then be:
+We also need to make sure our model yields one global estimate, so for this we decide to model species trends over time as part of larger distribution.
+Our resulting model would then be:
 
@@
 We likely would not feel immediately sure how to simulate for all different vertebrate species.
+For example, what are reasonable starting population sizes (intercepts) for different groups?
+As we start to answer this we likely would realize different clades should have different starting population sizes.
+We could thus try to build in increasing complexity immediately, but then we will likely struggle to see how our model is performing at the scales we care about -- for each population in each species.
+It is almost always better to scale back the question (and potentially dataset) when confronted with these sorts of problems.
 
@@
 
-Here we would likely feedback one to two steps.
-We could step back to Step 2 and consider adding a phylogeny or other structure from which we could then simulate data from, or we could step back to Step 1 and narrow the question (before potentially building it back up.) We believe the latter approach is often better as working on a small scale first can point out more fundamental problems versus starting big and then spending much longer to identify fundamental problems.
+Here we would likely feedback (step back) one to two steps.
+We could step back to Step 2 and consider adding a phylogeny or other structure from which we could then simulate data from, or we could step back to Step 1 and narrow the question (before potentially building it back up.) As mentioned above, we believe the latter approach is usually better as working on a small scale first can point out more fundamental problems versus starting big and then spending much longer to identify fundamental problems (or potentially missing them).
 
@@
 
-What is trend in mammal species population numbers in Africa, both overall and how does it vary across species and populations?
+What is trend in mammal species population numbers in southern Africa, both overall and how does it vary across species and populations?
+
+We chose this because mammals are a well studied group compared to other clades/groups in the dataset, so we will have a better idea of how to simulate data, and whether our model makes sense.
 
@@
 
-Now we start to simulate!
+Now we start to simulate! We follow the mathematical model we built above and simulate data from it.
 
@@
 
-### Fit model on simulated data!
+### Fit and check model on simulated data!
+
+Now we code the model from our mathematical notation (above) and fit it to our simulated data (also built from our mathematical model).
 
@@
 
-#### Retrodictive check with model fitted on simulated data
+#### Checking model fitted on simulated data
 
-```{r retro sim data}
-#| fig-height: 6
-#| fig-width: 8
+Our first goal with simulated data is a simple check of our code (both the model and simulated data code), given a high enough sample size relative to our effect sizes and error, the model should return our paramaters.
+So our first step is to check this.
 
-retrodata <- data.frame(obs = datasim$logy, pred = predict(nestlm),
-                        year = datasim$year,
-                        species = datasim$species, id = datasim$population)
+First, we start with the parameters, which include an overall estimate of the trend across species (and thus would address our question):
 
-ggplot(data = retrodata) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
-  facet_wrap(~ paste0("Species ", species), scales = "free") +
-  geom_point(aes(x = obs, y = pred, group = id, color = as.character(id)),
-             size = 0.5) + 
-  theme_bw() + theme(legend.position = 'none') +
-  labs(x = 'Simulated counts (log-scale)', y = 'Estimated counts (log-scale)')
-
-ggplot(data = retrodata) +
-  facet_wrap(~ paste0("Species ", species), scales = "free") +
-  geom_point(aes(x = year, y = obs, group = id, color = as.character(id)),
-             size = 1) + 
-  geom_line(aes(x = year, y = pred, group = id, color = as.character(id))) +
-  theme_bw() + theme(legend.position = 'none') +
-  labs(x = 'Year', y = 'Counts (log-scale)')
-
-ggplot(data = retrodata) +
-  facet_wrap(~ paste0("Species ", species), scales = "free") +
-  geom_point(aes(x = year, y = exp(obs), group = id, color = as.character(id)),
-             size = 1) + 
-  geom_line(aes(x = year, y = exp(pred), group = id, color = as.character(id))) +
-  theme_bw() + theme(legend.position = 'none') +
-   labs(x = 'Year', y = 'Counts')
-
+```{r inference sim data params}
 ```
 
-#### Inference with model fitted on simulated data
+We find the model does a fair job at recovering population-level parameters (hyperparameters).
 
-The model does a fair job at recovering population-level parameters.
-
-```{r inference sim data}
+```{r inference sim data hyperparams}
 #| eval: false
@@
 ```
+
+What if the model did not do a good job at recovering our parameters? Then we should likely adjust our sample size up and/or our error lower to make sure it is can (as a basic code test).
+
+But what if the model can recover parameters only at sample sizes much higher than our data (or at error sizes much lower than our data)?
+Then we would need to take this important information forward in our analysis and be careful in how we intepret estimates from our model fit to empirical data (as we have a hint at this stage that getting robust estimates from our data and model together may be difficult).
+
+For now, we also take a look at how model fits look against the simulated data:
+
+```{r retro sim data}
+#| fig-height: 6
+#| fig-width: 8
+
+retrodata <- data.frame(obs = datasim$logy, pred = predict(nestlm),
+                        year = datasim$year,
+                        species = datasim$species, id = datasim$population)
+
+ggplot(data = retrodata) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
+  facet_wrap(~ paste0("Species ", species), scales = "free") +
+  geom_point(aes(x = obs, y = pred, group = id, color = as.character(id)),
+             size = 0.5) + 
+  theme_bw() + theme(legend.position = 'none') +
+  labs(x = 'Simulated counts (log-scale)', y = 'Estimated counts (log-scale)')
+
+ggplot(data = retrodata) +
+  facet_wrap(~ paste0("Species ", species), scales = "free") +
+  geom_point(aes(x = year, y = obs, group = id, color = as.character(id)),
+             size = 1) + 
+  geom_line(aes(x = year, y = pred, group = id, color = as.character(id))) +
+  theme_bw() + theme(legend.position = 'none') +
+  labs(x = 'Year', y = 'Counts (log-scale)')
+
+ggplot(data = retrodata) +
+  facet_wrap(~ paste0("Species ", species), scales = "free") +
+  geom_point(aes(x = year, y = exp(obs), group = id, color = as.character(id)),
+             size = 1) + 
+  geom_line(aes(x = year, y = exp(pred), group = id, color = as.character(id))) +
+  theme_bw() + theme(legend.position = 'none') +
+   labs(x = 'Year', y = 'Counts')
+
+```
+
+These look pretty good, so we move onto the next step.
 

### 2025-04-29 analyses/african_mammals_workflow.qmd
Adding in more text (some of which I am about to cut)

@@
 
-Our first goal with simulated data is a simple check of our code (both the model and simulated data code), given a high enough sample size relative to our effect sizes and error, the model should return our paramaters.
+Our first goal with simulated data is a simple check of our code (both the model code and simulated data code), given a high enough sample size relative to our effect sizes and error, the model should return our paramaters.
 So our first step is to check this.
 
-First, we start with the parameters, which include an overall estimate of the trend across species (and thus would address our question):
+First, we start with the parameters, which include an overall estimate of the trend across species (this parameter would address our question that we decided on above):
 
@@
 
-We find the model does a fair job at recovering population-level parameters (hyperparameters).
+Now we check additional parameters. We find the model does a fair job at recovering population-level parameters (aka hyperparameters).
 
@@
 But what if the model can recover parameters only at sample sizes much higher than our data (or at error sizes much lower than our data)?
-Then we would need to take this important information forward in our analysis and be careful in how we intepret estimates from our model fit to empirical data (as we have a hint at this stage that getting robust estimates from our data and model together may be difficult).
+Then we would need to take this important information forward in our analysis and be careful in how we intepret estimates from our model fit to empirical data (as we have a hint at this stage that getting robust estimates from our data and model together may be difficult, but we do not know the actual values of our parameters -- including error and variance parameters -- from our empirical data so we cannot be sure of whether we may have these issues yet).
 
@@
 
+Now that we have designed our question, built our model for it, and tested our model on simulated data, we can fit the model to our empirical data.
+We start by some basic clean-up of the dataframe so we have years in one column.
+
 ```{r lookdata}
@@
 
-The values in the database are not necessarily in the same unit, so I imagine we need to take a decision here..?
+The values in the database are not necessarily in the same units, so we subset to units related to absolute population sizes, which is what we have mostly been thinking of (some units are in density and we would need to build our model to incorporate those, but here we start first with a simpler model and can build up later).
+We take a quick look at the data to make sure it looks reasonable based on our domaine knowledge.
 
@@
 ```
+
+Now, we make a few final adjustments and fit the model:
 
@@
 
-## Step 5: Retrodictive check
+## Step 5: Retrodictive and related checks
 
 Now we use our fitted model parameters to generate predictions, and compare them to observations.
+This is similar to the simulating data above, but differs in two major ways: (1) we simulate from the paramaters estimated by the model fit to the empirical data (versus coming up with them based on our expert knowledge) and (2) we need to check across many predictions from the model to understand the full uncertainty of the model (and not just one draw from the model).
+
+Here we start by examining how well predictions from the mean estimated values from the model compare to the observed data; this is similar to many plots where we simply want to see how well the model seems to predict the data.
+
+To do this, we start by extracting the model parameter estimates:
 
@@
 ```
-
-We might examine species separately, e.g. here the African bush elephant.
+ADD HERE (and adjust code above): Something about what the overall trends appears to be, but that to see if that might be a good estimate we need to check how the model predicts the data on finer scales.
+So we start by examining some species separately, e.g.
+here the African bush elephant:
 
@@
 
-We can focus on populations in South Africa.
+Focusing on populations in South Africa, we see some places where the linear model apears well (ID 7673), but also two other populations where it does not appear as a very good fit.
 
@@
 
-The three populations are measured in the same units, but likely represent different spatial scales (ranging from a few hundred individuals to more than 10000).
+The three populations are measured in the same units, but are very different in their sizses (ranging from a few hundred individuals to more than 10000), thus they likely represent populations measured across different spatial scales.
 
-The same holds for other species and other countries, e.g. the blue wildebeest in Tanzania.
+The same holds for other species and other countries, e.g. the blue wildebeest in Tanzania:
 
@@
 ```
-The four populations are also measured in the same units (I guess?!), but represents different scales (so the blue one is likely a population, while the others represent region/country level?).
+The four populations are also measured in the same basic units, but represent likely represent very different populations, spanning ones in small parks that cannot migrate to the large migrating herds (e.g., Serengeti Masai-Mara), though gathering additional metadata to fully check this would be another important component to add to our workflow for these data.
 
@@
 
-At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including questioning our assumption of one sigma for all populations and potentially our linearity assumption.
-One possibility would be to move to a non-linear regression, and use a more flexible model
-Quick example for one of the wildbeest dataset in Tanzania:
+At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one sigma for all populations (CHECK HERE: do we really use one sigma for all populations?) and potentially our linearity assumption.
+One possibility would be to move to a non-linear regression, and use a more flexible model that would allow population numbers to stabilize, such as shown using a Gompertz model for the wildbeest dataset in Tanzania:
 
@@
 
-We also may want to add predictors for the slope such as  ... We would do this, returning to Steps 2-5 likely several times until we have a model that is good enough for our purposes here.
+We also have seen variability in population trends within species (e.g., two apparently increasing populations and two apparently decreasing populations in Tanzania), so we would want to think a lot more about how we scale up from population to species to capture this variation in a way that is most useful to our question.
+Here we might realize we want to try to understand the drivers of this variability -- for this we would need to return to Step 1 and reformulate our question, then go through the workflow again, likely adding in potential predictors such as some metric of conservation efforts for that species and or population (e.g., stopping the Ivory trade for elephants, metrics of poaching in different parks) and possibly also country, which may capture differences in efforts related to conservation that are not yet measured.
+We would do this, returning to Steps 2-5 likely several times until we have a model that is good enough for our purposes here.
+And then we would likely add in more countries and species.
 
-And then we would likely add in more countries and species....
+Alternatively, we may feel this model is good enough for our question and think about how to robustly add in the species and geography we originally excluded.
 

### 2025-04-29 analyses/african_mammals_workflow.qmd
Last edits for the day, I am done for now with workflow.

@@
 
-At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one sigma for all populations (CHECK HERE: do we really use one sigma for all populations?) and potentially our linearity assumption.
+At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one sigma for all populations (CHECK HERE: do we really use one sigma for all populations?), how we model population trends (e.g., there are many ways to model abundance) and (relatedly) our linearity assumption.
 One possibility would be to move to a non-linear regression, and use a more flexible model that would allow population numbers to stabilize, such as shown using a Gompertz model for the wildbeest dataset in Tanzania:
@@
 We also have seen variability in population trends within species (e.g., two apparently increasing populations and two apparently decreasing populations in Tanzania), so we would want to think a lot more about how we scale up from population to species to capture this variation in a way that is most useful to our question.
-Here we might realize we want to try to understand the drivers of this variability -- for this we would need to return to Step 1 and reformulate our question, then go through the workflow again, likely adding in potential predictors such as some metric of conservation efforts for that species and or population (e.g., stopping the Ivory trade for elephants, metrics of poaching in different parks) and possibly also country, which may capture differences in efforts related to conservation that are not yet measured.
-We would do this, returning to Steps 2-5 likely several times until we have a model that is good enough for our purposes here.
-And then we would likely add in more countries and species.
+For example, may want to think about adding country, which may capture differences in efforts related to conservation and thus predict trends (or we may have better metrics of what drives these trends, but given our global aim, we would need it for most/all populations) and we may want to think about how or if we partially pool species and populations within species (potentially using models other than a normal distribution for pooling).
+As we address these potential problems, we would likely begin to add in more countries, species, and populations.
 
-Alternatively, we may feel this model is good enough for our question and think about how to robustly add in the species and geography we originally excluded.
+We expect scaling up to more countries, species and populations may cause us to think carefully about the data we have in hand versus the question we aimed to answer.
+While many recent papers treat the LPI data as effectively a well stratified and randomized sample or even effectively a good cencus, it is clearly not.
+Thus, models that try to deal with such uneven sampling and build in the more of the structure of data, would be far more appropriate.
+For example, most models for election forecasts now include regularization via partial pooling, post-stratificatiob and other methods designed to adjust for the fact that many polls are 'non-representative.' Such polls may be more similar to LPI data.
+Our model includes some regularization via partial pooling, but we may need far more of an observational model added to our population demography model to accommodate the uneven sampling.
+Estimates from such a model may be far more uncertain, but likely also more correct given our combination of question and data.
 
-I might also add a short paragraph about moving away from linear regression and towards models for species abudance (via mixtures I think?
-I suspect the articles I posted for class have some good ideas) and thinking more about linking to demography models.
+This may lead us to realize we really don't have great data for this question and we would then need to prioritize better data collection efforts if we truly want good estimates to answer this question.
 
-And then we would realize we really don't have great data for this question and we would ...
+## The full process
+
+The small workflow we should here may seem like extra work to those who have used it, and it is compared to many ecological workflows in practice.
+We argue that the extra effort is worth it given the critical questions we are addressing, and also that the process becomes much quicker with practice.
+The first time may be slow, but the resulting model will be better for it (even if it is still arguably simple, as the one we show here is), but the following times will go faster with likely improved models.

### 2025-04-29 analyses/african_mammals_workflow.qmd
Fixing a lot of typos and removing superlatives (and -- eek! -- this was in two paragraphs, so many more exist)

@@
 We expect scaling up to more countries, species and populations may cause us to think carefully about the data we have in hand versus the question we aimed to answer.
-While many recent papers treat the LPI data as effectively a well stratified and randomized sample or even effectively a good cencus, it is clearly not.
-Thus, models that try to deal with such uneven sampling and build in the more of the structure of data, would be far more appropriate.
-For example, most models for election forecasts now include regularization via partial pooling, post-stratificatiob and other methods designed to adjust for the fact that many polls are 'non-representative.' Such polls may be more similar to LPI data.
-Our model includes some regularization via partial pooling, but we may need far more of an observational model added to our population demography model to accommodate the uneven sampling.
-Estimates from such a model may be far more uncertain, but likely also more correct given our combination of question and data.
+While many recent papers treat the LPI data as (effectively) a well stratified and randomized sample or even a good census, the LPI data is clearly neither.
+Thus, models that try to deal with such uneven sampling and build in the more of the structure of data, would be more appropriate.
+For example, most models for election forecasts now include regularization via partial pooling, post-stratification and other methods designed to adjust for the fact that many polls are 'non-representative.' Such non-representative polls may be very similar to LPI data.
+Our model includes some regularization via partial pooling, but we may need much more of an observational model added to our population demography model to accommodate the uneven sampling.
+Estimates from such a model may be more uncertain, but likely also more correct given our combination of question and data.
 
-This may lead us to realize we really don't have great data for this question and we would then need to prioritize better data collection efforts if we truly want good estimates to answer this question.
+This may lead us to realize we really don't have great data for this question and we would then need to prioritize better data collection efforts if we want good estimates to answer this question.
 
@@
 
-The small workflow we should here may seem like extra work to those who have used it, and it is compared to many ecological workflows in practice.
+The small workflow we show here may seem like extra work to those who have used it, and it is compared to many ecological workflows in practice.
 We argue that the extra effort is worth it given the critical questions we are addressing, and also that the process becomes much quicker with practice.
-The first time may be slow, but the resulting model will be better for it (even if it is still arguably simple, as the one we show here is), but the following times will go faster with likely improved models.
+The first time may be slow, but the resulting model will be better for it (even if it is still arguably simple, as the one we show here is), but the following uses of this iterative workflow will go faster with likely improved models.

### 2025-04-30 analyses/lpiworkflow.qmd
progress on workflow

---
title: "A sample workflow"
authors: "Lizzie, Victor"
date: "April 2025"
date-format: "MMMM YYYY"
toc: true
number-sections: true
highlight: pygments
execute:
  enabled: true
format:
  html:
    html-math-method: katex
    theme:
      - lux
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
    page-layout: full
    fig-width: 8
    fig-height: 4
  pdf:
    keep-tex: true
    fig-width: 5.5
    fig-height: 7.5
    code-overflow: wrap
    monofontoptions:
      - Scale=0.5
knitr:
  opts_chunk:
    comment: ''
  opts_knit:
    global.par: TRUE
format-links: false
---

To show our workflow in action, we examine a subset of the Living Planet Index (LPI, download at: https://www.livingplanetindex.org/data_portal. This download should include a csv file, here: LPD2022_public.csv, that we use below). We discuss the LPI in the main text; for more information, check out: www.livingplanetindex.org/.

**Note**: data from Jonathan comes from *LivingPlanetIndexDatabase_2025-02-01_19.19.51*.

## Step 1: Research question

Here we need to focus on exactly what question we are trying to answer, which will guide us in the next steps. This step is sometimes skipped over or moved through so quickly that the exact question is not clear, but is critical for developing a useful model. Many early papers on the LPI seem focused on something similar to: Are vertebrate species overall declining on average? And, if so, by how much? We, however might want to have a question that more clearly allows any directionality, so we could start with: What is the global trend for biodiversity in vetebrate species? This is rather broad and might mean we should develop a model that will yield one global estimate. In thinking about this, we would likely realize we don't want just that, we actually want to understand the variation across different species and perhaps populations and also a global estimate. So we might refine the question to:

What is global trend over time in vertebrate species population numbers, both overall and how does it vary acrosss species and populations?

```{r}
wd <- "/home/victor/projects/forecastflows/analyses"

set.seed(20012029)

library(rstan)
library(ggplot2)

lpi_subset <- read.table(file.path(wd, "jdavies/African_mammals.txt"))

lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Units", paste0("X", 1950:2020))]
lpi_subset$genus_species <- paste(lpi_subset$Genus, lpi_subset$Species)

lpi_subset_rsh <- 
  reshape(lpi_subset, direction = "long", idvar = "ID",
          varying =  paste0("X", 1950:2020), v.names = c("value"), timevar = "year", times = 1950:2020)
lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != "NULL",]

```

## Step 2: Model building

Next we need to think about how we would build a model to address this question, with the goal to build the model and evaluate it using simulated data. There is a lot of demographic models in population ecology, so we might revisit those and consider a discrete time model based on population size and stepping through each year of data. We might also start with a simple linear model, where populations can go up, down or do nothing.

Using a simple linear gets us started, but now we need to consider how we will model species and populations. Populations usually are nested in species in ecological models, so we might consider that to start. We also need to make sure our model yields one global estimate, so for this we decide to model species trends over time as part of larger distribution. Our resulting model would then be:

$$\log(y_i) \sim \text{normal}(\mu_i, \sigma)$$

$$\mu_i = \alpha_{\text{pop}[\text{sp}[i]]} + \beta_{\text{pop}[\text{sp}[i]]} * \text{year}_i$$

$$\alpha_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\alpha, \text{sp}} \ , \ \sigma_{\alpha, \text{sp}}) \\
\mu_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_1} \ , \ \sigma_{\alpha_1}) \\
\sigma_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_2} \ , \ \sigma_{\alpha_2}) $$

$$\beta_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\beta, \text{sp}} \ , \ \sigma_{\beta, \text{sp}}) \\
\mu_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_1} \ , \ \sigma_{\beta_1}) \\
\sigma_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_2} \ , \ \sigma_{\beta_2}) $$

## Step 3: Model evaluation

Now we need to simulate data to test our model. As we start to think about how to do this, we may become overwhelmed with the scale of the question we started with. We likely would not feel immediately sure how to simulate for all different vertebrate species. For example, what are reasonable starting population sizes (intercepts) for different groups? As we start to answer this we likely would realize different clades should have different starting population sizes. We could thus try to build in increasing complexity immediately, but then we will likely struggle to see how our model is performing at the scales we care about -- for each population in each species. It is almost always better to scale back the question (and potentially dataset) when confronted with these sorts of problems.

### Stepping backward...

Here we would likely feedback (step back) one to two steps. We could step back to Step 2 and consider adding a phylogeny or other structure from which we could then simulate data from, or we could step back to Step 1 and narrow the question (before potentially building it back up.) As mentioned above, we believe the latter approach is usually better as working on a small scale first can point out more fundamental problems versus starting big and then spending much longer to identify fundamental problems (or potentially missing them).

So here we refine our question to:

What is trend in mammal species population numbers in southern Africa, both overall and how does it vary across species and populations?

We chose this because mammals are a well studied group compared to other clades/groups in the dataset, so we will have a better idea of how to simulate data, and whether our model makes sense.

### Simulating data

Now we start to simulate! We follow the mathematical model we built above and simulate data from it.

```{r sim data}

nspecies <- 9 # no. of species
npops_persp <- round(runif(nspecies, 2, 15)) # no. of different populations per species
npops <- sum(npops_persp) # total no. of different populations
nobs_perpop <- round(runif(npops, 20, 60)) # no. of different observations per population

spid_perpop <- rep(1:nspecies, times = npops_persp)
spid_perobs <- rep(spid_perpop, times = nobs_perpop)
popid_perobs <- rep(1:npops, times = nobs_perpop)

years <- c()
for(np in nobs_perpop){
  years_p <- sample(1950:2020, size = np, replace = FALSE)
  years <- c(years, years_p)
}

# nested intercepts
mu_alpha1 <- log(200)
sigma_alpha1 <-  log(50)/2.57
mu_alpha_sp <- rnorm(n = nspecies, mean = mu_alpha1, sd = sigma_alpha1)
mu_alpha2 <- log(20)
sigma_alpha2 <- log(50)/2.57
sig_species_alpha <- abs(rnorm(n = nspecies, mean = mu_alpha2, sd = sigma_alpha2))
alpha_pop_sp <- rnorm(n = npops, mean = mu_alpha_sp[spid_perpop], sd = sig_species_alpha[spid_perpop])

# nested slopes
mu_beta1 <- 0.05
sigma_beta1 <- 0.25/2.57
mu_beta_sp <- rnorm(n = nspecies, mean = mu_beta1, sd = sigma_beta1)
mu_beta2 <- 0
sigma_beta2 <- 0.05/2.57
sig_species_beta <- abs(rnorm(n = nspecies, mean = mu_beta2, sd = sigma_beta2))
beta_pop_sp <- rnorm(n = npops, mean = mu_beta_sp[spid_perpop], sd = sig_species_beta[spid_perpop])

# save simulated intercepts and slopes
pops_fit <- data.frame()
pid <- 1
for(sp in 1:nspecies){
  for(p in 1:npops_persp[sp]){
    pops_fit <- rbind(pops_fit, data.frame(species = sp, pop = pid))
    pid <- pid + 1
  }
}
pops_fit$int <- alpha_pop_sp
pops_fit$slp <- beta_pop_sp

sigma_obs <- 0.75 # common observational error

y <- c()
for(p in 1:npops){
  nobs_p <- nobs_perpop[p]
  alpha_p <- alpha_pop_sp[p]
  beta_p <- beta_pop_sp[p]

  years_p <- years[popid_perobs == p]

  yhat_p <- alpha_p + beta_p * (years_p - 1980)
  # yhat_p <- alpha_p
  y_p <- exp(rnorm(n = nobs_p, mean = yhat_p, sd = sigma_obs))
  y <- c(y, y_p)
}

datasim <- data.frame(
  y, 
  year = years - 1980,
  species = as.character(spid_perobs),
  population = paste0(spid_perobs,popid_perobs)
)

datasim$logy <- log(datasim$y)

ggplot(data = datasim) +
  geom_line(aes(x = year+1980, y = y, 
                group = population,
                color = species)) +
  theme_bw() +
  labs(x = 'Year', y = 'Simulated counts', colour = 'Species')

ggplot(data = datasim) +
  geom_line(aes(x = year+1980, y = logy, 
                group = population,
                color = species)) +
  theme_bw() +
  labs(x = 'Year', y = 'Simulated counts (log-scale)', colour = 'Species')

```

### Fit and check model on simulated data!

Now we code the model from our mathematical notation (above) and fit it to our simulated data (also built from our mathematical model).

```{r fit sim data}

runmodel <- FALSE

data <- list(
  N = nrow(datasim),
  Nsp = nspecies,
  Npop = npops,
  y = y,
  year = years,
  spid_perpop = spid_perpop,
  popid = popid_perobs
)

if(runmodel){
  fit <- stan(file = "~/projects/forecastflows/analyses/stan/model1_nc2.stan",
            data = data,  iter=4000, warmup=3000, cores = 4, seed = 20012029)
  saveRDS(fit, file = file.path(wd, "output/fit/simulateddatafit.rds"))
}else{
  fit <- readRDS(file = file.path(wd, "output/fit/simulateddatafit.rds"))
}

```

#### Checking model fitted on simulated data

Our first goal with simulated data is a simple check of our code (both the model code and simulated data code), given a high enough sample size relative to our effect sizes and error, the model should return our parameters. So our first step is to check this.

First, we start with the parameters, which include an overall estimate of the trend across species (this parameter would address our question that we decided on above):

```{r inference sim data params}
#| echo: false
#| fig-height: 1
#| fig-width: 4

summ_fit <- data.frame(summary(fit)$summary)

ggplot() +
  geom_pointrange(data = summ_fit["mu_beta1", c('X50.', 'X2.5.', 'X97.5.')],
                  aes(xmin = X2.5., xmax = X97.5., x = X50., y = 0), size = 0.25) +
  geom_vline(aes(xintercept = mu_beta1), color = "darkred", linetype = "dashed") +
  lims(y = c(-0.02,0.02),  x = c(-0.5,0.5)) +
  labs(x = "Overall trend across species") +
  theme_bw() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.minor.y = element_blank(), panel.grid.major.y = element_blank())

```

Now we check additional parameters. We find the model does a fair job at recovering population-level parameters (aka hyperparameters).

```{r inference pop sim data}
#| echo: false
#| fig-height: 4
#| fig-width: 5
#| layout-ncol: 2

pops_fit$int_fit <- summ_fit[paste0("alpha_pop_sp[",1:npops,"]"), c('X50.')]
pops_fit$int_fit.lcl <- summ_fit[paste0("alpha_pop_sp[",1:npops,"]"), c('X2.5.')]
pops_fit$int_fit.ucl <- summ_fit[paste0("alpha_pop_sp[",1:npops,"]"), c('X97.5.')]

pops_fit$slp_fit <- summ_fit[paste0("beta_pop_sp[",1:npops,"]"), c('X50.')]
pops_fit$slp_fit.lcl <- summ_fit[paste0("beta_pop_sp[",1:npops,"]"), c('X2.5.')]
pops_fit$slp_fit.ucl <- summ_fit[paste0("beta_pop_sp[",1:npops,"]"), c('X97.5.')]

ggplot(data = pops_fit) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = int_fit.lcl, ymax = int_fit.ucl, x = int, y = int_fit, color = as.character(species)),
                  size = 0.2) +
  theme_bw() +
  labs(x = 'Simulated population intercepts', y = 'Estimated population intercepts', colour = 'Species')

ggplot(data = pops_fit) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = slp_fit.lcl, ymax = slp_fit.ucl, x = slp, y = slp_fit, color = as.character(species)),
                  size = 0.2) +
  theme_bw() +
  labs(x = 'Simulated population trends', y = 'Estimated population trends', colour = 'Species')

```

What if the model did not do a good job at recovering our parameters? Then we should likely adjust our sample size up and/or our error lower to make sure it is can (as a basic code test).

But what if the model can recover parameters only at sample sizes much higher than our data (or at error sizes much lower than our data)? Then we would need to take this important information forward in our analysis and be careful in how we interpret estimates from our model fit to empirical data (as we have a hint at this stage that getting robust estimates from our data and model together may be difficult, but we do not know the actual values of our parameters -- including error and variance parameters -- from our empirical data so we cannot be sure of whether we may have these issues yet).

For now, we also take a look at how model fits look against the simulated data:

```{r retro sim data}
#| fig-height: 6
#| fig-width: 7
#| echo: false

summ_fit <- data.frame(summary(fit)$summary)
retrodata <- data.frame(obs = datasim$logy, 
                        pred = summ_fit[paste0("logy_pred[",1:data$N,"]"), c('X50.')],
                        predlc = summ_fit[paste0("logy_pred[",1:data$N,"]"), c('X2.5.')],
                        preduc = summ_fit[paste0("logy_pred[",1:data$N,"]"), c('X97.5.')],
                        year = datasim$year,
                        species = datasim$species, id = datasim$population)

ggplot(data = retrodata) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  facet_wrap(~ paste0("Species ", species), scales = "free") +
  geom_pointrange(aes(x = obs, y = pred, ymin = predlc, ymax = preduc, group = id, color = as.character(species)),
             size = 0.01, linewidth = 0.2) + 
  theme_bw() + theme(legend.position = 'none') +
  labs(x = 'Simulated counts (log-scale)', y = 'Estimated counts (log-scale)')

```

```{r retro sim data time}
#| fig-height: 6
#| fig-width: 7
#| echo: false

ggplot(data = retrodata) +
  facet_wrap(~ paste0("Species ", species), scales = "free") +
  geom_ribbon(aes(x = year, ymin = predlc, ymax = preduc, group = id, fill = as.character(species)),alpha = 0.1) +
  geom_line(aes(x = year, y = pred, group = id, color = as.character(species))) +
  geom_point(aes(x = year, y = obs, group = id, ),
             size = 1, color = 'white') + 
  geom_point(aes(x = year, y = obs, group = id, color = as.character(species)),
             size = 0.2) + 
  theme_bw() + theme(legend.position = 'none') +
  labs(x = 'Year', y = 'Counts (log-scale)')

```

These look pretty good, so we move onto the next step.

## Step 4: Fit model to empirical data

Now that we have designed our question, built our model for it, and tested our model on simulated data, we can fit the model to our empirical data. We start by some basic clean-up of the dataframe so we have years in one column.

```{r lookdata}
#| eval: true

lpi_subset <- read.table("/home/victor/projects/forecastflows/analyses/jdavies/African_mammals.txt")

lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Country", "Units", paste0("X", 1950:2020))]
lpi_subset$genus_species <- paste(lpi_subset$Genus, lpi_subset$Species)

lpi_subset_rsh <- 
  reshape(lpi_subset, direction = "long", idvar = "ID",
          varying =  paste0("X", 1950:2020), v.names = c("value"), timevar = "year", times = 1950:2020)
lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != "NULL",]
lpi_subset_rsh$value <- as.numeric(lpi_subset_rsh$value)

```

The values in the database are not necessarily in the same units, so we subset to units related to absolute population sizes, which is what we have mostly been thinking of (some units are in density and we would need to build our model to incorporate those, but here we start first with a simpler model and can build up later). We take a quick look at the data to make sure it looks reasonable based on our domaine knowledge.

```{r units}
#| eval: true
#| layout-nrow: 2

units_to_keep <- c("Estimate - entire population", "Estimate - entire population (August)",
                   "Individual", "individuals", "Individuals", "Number of adults", "number of individuals",
                   "Number of individuals", "Number of Individuals", "Number of rhinos", "numbers of indviduals",
                   "population estimate", "Population estimate", "Population estimate (individuals)", 
                   "Population estimates", "Total population size")

lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$Units %in% units_to_keep,]

ggplot(data = lpi_subset_rsh) +
  geom_line(aes(x = year, y = value, 
                group = paste0(genus_species, ID),
                color = genus_species)) +
  scale_color_manual(values = hcl.colors(14, palette = "Dark2"),
                     breaks = unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]) +
  theme_bw() +
  labs(x = 'Year', y = 'Observed counts')

ggplot(data = lpi_subset_rsh[lpi_subset_rsh$value != 0,]) +
  geom_line(aes(x = year, y = log(value), 
                group = paste0(genus_species, ID),
                color = genus_species)) +
  scale_color_manual(values = hcl.colors(14, palette = "Dark2"),
                     breaks = unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]) +
  theme_bw() +
  labs(x = 'Year', y = 'Observed counts (log-scale)', color = "Species")

```

Now, we make a few final adjustments and fit the model:

```{r fitdata}
#| eval: true

runmodel <- FALSE

lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,] # poor way to avoid log(0)

popids <- data.frame(ID = unique(lpi_subset_rsh$ID), popid_num = 1:length(unique(lpi_subset_rsh$ID)))
spids <- data.frame(genus_species = unique(lpi_subset_rsh$genus_species), spid_num = 1:length(unique(lpi_subset_rsh$genus_species)))

lpi_subset_rsh <- merge(lpi_subset_rsh, spids)
lpi_subset_rsh <- merge(lpi_subset_rsh, popids)

spid_perpop <- unique(lpi_subset_rsh[c("popid_num", "spid_num")])
spid_perpop <- spid_perpop[order(spid_perpop$popid_num),]

data <- list(
  N = nrow(lpi_subset_rsh),
  Nsp = max(spids$spid_num),
  Npop = max(popids$popid_num),
  y = lpi_subset_rsh$value,
  year = lpi_subset_rsh$year,
  spid_perpop = spid_perpop$spid_num,
  popid = lpi_subset_rsh$popid_num
)

if(runmodel){
  fit <- stan(file = "~/projects/forecastflows/analyses/stan/model1_nc2.stan",
            data = data,  iter=4000, warmup=3000, cores = 4, seed = 20012029)
  saveRDS(fit, file = file.path(wd, "output/fit/truedatafit.rds"))
}else{
  fit <- readRDS(file = file.path(wd, "output/fit/truedatafit.rds"))
}

```

Here we finally get to see how our model performs on the empirical data and are a step closer to answering our question, but we also start to notice some differences in our empirical data from our simulated data. In preparing our data for the model, we realize we have quite different amplitudes of variations across populations and species (on a log-scale), e.g. for blue wildebeest (each color represents a different population):

```{r}
#| fig-height: 3.5
#| fig-width: 5
#| echo: false

ggplot(data = lpi_subset_rsh[lpi_subset_rsh$value != 0 & lpi_subset_rsh$genus_species == "Connochaetes taurinus",]) +
  geom_line(aes(x = year, y = log(value), 
                group = paste0(genus_species, ID),
                color = as.character(ID))) +
  theme_bw() + theme(legend.position = 'none') +
  labs(x = 'Year', y = 'Observed counts (log-scale)')

```

Here, we might step back and adjust our model (re-doing Step 2-3 again), but in this case we go ahead.

## Step 5: Retrodictive and related checks

Now we use our fitted model parameters to generate predictions, and compare them to observations. This is similar to the simulating data above, but differs in two major ways: (1) we simulate from the parameters estimated by the model fit to the empirical data (versus coming up with them based on our expert knowledge) and (2) we need to check across many predictions from the model to understand the full uncertainty of the model (and not just one draw from the model).

Here we start by visually comparing the posterior predictive distribution (red line) to observed data (black dashed line):
```{r}
#| fig-height: 5
#| fig-width: 5
#| echo: false

names <- paste0('logy_pred[', 1:data$N, ']')
retrodictions <- unlist(extract(fit, pars = names))

ggplot() +
  geom_density(data = data.frame(obs = log(data$y)), aes(x = obs), alpha = 0.5, linetype = 'dashed') +
  geom_density(data = data.frame(retrodictions), aes(x = retrodictions), alpha = 0.5, col = 'darkred') +
  theme_bw() +
  labs(y = 'Density', x = 'Counts (log-scale)')

```

We then examine, by species, how well predictions from the model compare to the observed data; this is similar to many plots where we simply want to see how well the model seems to predict the data.

```{r}
#| fig-height: 10
#| fig-width: 12
#| echo: false

summ_fit <- data.frame(summary(fit)$summary)
retrodata <- data.frame(obs = data$y, 
                        pred = summ_fit[paste0("logy_pred[",1:data$N,"]"), c('X50.')],
                        predlc = summ_fit[paste0("logy_pred[",1:data$N,"]"), c('X2.5.')],
                        preduc = summ_fit[paste0("logy_pred[",1:data$N,"]"), c('X97.5.')],
                        year = lpi_subset_rsh$year, units = lpi_subset_rsh$Units,
                        species = lpi_subset_rsh$genus_species, 
                        id = lpi_subset_rsh$ID,
                        country = lpi_subset_rsh$Country)

ggplot(data = retrodata) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  facet_wrap(~ species, scales = "free") +
  geom_pointrange(aes(x = log(obs), y = pred, ymin = predlc, ymax = preduc, group = id, color = as.character(species)),
             size = 0.1, linewidth = 0.2) + 
  scale_color_manual(values = hcl.colors(14, palette = "Dark2"),
                     breaks = unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]) +
  theme_bw() + theme(legend.position = 'none') +
  labs(x = 'Observed counts (log-scale)', y = 'Estimated counts (log-scale)')

```

We may want to check what the overall and species trends appear to be:

```{r}
#| echo: false
#| fig-height: 4
#| fig-width: 5

species_trends <- summ_fit[paste0("mu_beta_sp[",1:data$Nsp,"]"), c('X50.', 'X2.5.', 'X97.5.')]
species_trends$spid_num <- 1:data$Nsp
species_trends <- merge(species_trends, spids)

ggplot() +
  geom_pointrange(data = summ_fit["mu_beta1", c('X50.', 'X2.5.', 'X97.5.')],
                  aes(xmin = X2.5., xmax = X97.5., x = X50., y = "Z"), size = 0.25) +
  geom_text(data = summ_fit["mu_beta1", c('X50.', 'X2.5.', 'X97.5.')],
                  aes(x = -0.045, y = "Z", label = "Overall"), size = 3) +
  geom_pointrange(data = species_trends,
                  aes(xmin = X2.5., xmax = X97.5., x = X50., y = genus_species, color = genus_species), 
                  size = 0.25) +
  scale_color_manual(values = hcl.colors(14, palette = "Dark2"),
                     breaks = unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]) +
  lims(x = c(-0.1,0.1)) +
  labs(x = "Trends", color = "Species") +
  theme_bw() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.minor.y = element_blank(), panel.grid.major.y = element_blank())

```

To see if that might be a good estimate, we need to check how the model predicts the data on finer scales. So we start by examining some species separately, e.g. here the African bush elephant (each color within a country represents a different population):

```{r}
#| fig-height: 7
#| fig-width: 9
#| echo: false

# ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
#   facet_wrap(~ country, scales = "free_y") +
#   geom_point(aes(x = year, y = log(obs), group = id,  color = as.character(species)),
#              size = 1) + 
#   geom_line(aes(x = year, y = pred, group = id)) +
#   scale_color_manual(values = hcl.colors(14, palette = "Dark2"),
#                      breaks = unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]) +
#   theme_bw() + theme(legend.position = 'none') +
#   labs(y = 'Counts (log-scale)', x = 'Year')

retrodata_elephant <- retrodata[retrodata$species == "Loxodonta africana", ]
retrodata_elephant$idpop_percountry <- as.character(ave(retrodata_elephant$id, retrodata_elephant$country, FUN = function(x) match(x, unique(x))))

ggplot(data = retrodata_elephant) +
  facet_wrap(~ country, scales = "free") +
  geom_ribbon(aes(x = year, ymin = predlc, ymax = preduc, group = id, fill = idpop_percountry),alpha = 0.1) +
  geom_line(aes(x = year, y = pred, group = id, color = idpop_percountry)) +
  geom_point(aes(x = year, y = log(obs), group = id, ),
             size = 1, color = 'white') + 
  geom_point(aes(x = year, y = log(obs), group = id, color = idpop_percountry),
             size = 0.2) + 
  # scale_color_manual(values = hcl.colors(14, palette = "Dark2"),
  #                    breaks = unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]) +
  theme_bw() + theme(legend.position = 'none') +
  labs(y = 'Counts (log-scale)', x = 'Year')

```

Focusing on populations in South Africa, we see some places where the linear model appears well (ID 7673), but also two other populations where it does not appear as a very good fit.

```{r}
#| fig-height: 3
#| fig-width: 9
#| echo: false

retrodata_elephant <- retrodata[retrodata$species == "Loxodonta africana" & retrodata$country == "South Africa", ]
retrodata_elephant$idpop_percountry <- as.character(ave(retrodata_elephant$id, retrodata_elephant$country, FUN = function(x) match(x, unique(x))))

ggplot(data = retrodata_elephant) +
  geom_ribbon(aes(x = year, ymin = predlc, ymax = preduc, group = id),alpha = 0.1) +
  geom_line(aes(x = year, y = pred, group = id),
            linewidth = 0.5) +
  facet_wrap( ~ id, scales = "free") +
  geom_point(aes(x = year, y = log(obs), group = id, color = units),
             size = 0.7) + 
  geom_line(aes(x = year, y = log(obs), group = id, color = units),
             linewidth = 0.3) +

  theme_bw() + 
  labs(y = 'Counts (log-scale)', x = 'Year', color = 'Units')

# ggplot(data = retrodata[retrodata$species == "Loxodonta africana" & retrodata$country == "South Africa", ]) +
#   facet_wrap( ~ id, scales = "free_y") +
#   geom_point(aes(x = year + 1980, y = obs, group = id, color = units),
#              size = 0.7) + 
#   geom_line(aes(x = year + 1980, y = obs, group = id, color = units),
#              linewidth = 0.3) + 
#   geom_line(aes(x = year + 1980, y = exp(pred), group = id),
#             linewidth = 0.5) +
#   theme_bw() + 
#   labs(y = 'Counts', x = 'Year', color = 'Units')

```

The three populations are measured in the same units, but are very different in their sizes (ranging from a few hundred individuals to more than 10000), thus they likely represent populations measured across different spatial scales.

The same holds for other species and other countries, e.g. the blue wildebeest in Tanzania:

```{r}
#| fig-height: 5
#| fig-width: 9
#| echo: false

retrodata_wildebeest <- retrodata[retrodata$species == "Connochaetes taurinus" & retrodata$country == "Tanzania, United Republic Of", ]
retrodata_wildebeest$idpop_percountry <- as.character(ave(retrodata_wildebeest$id, retrodata_wildebeest$country, FUN = function(x) match(x, unique(x))))

ggplot(data = retrodata_wildebeest) +
  geom_ribbon(aes(x = year, ymin = predlc, ymax = preduc, group = id),alpha = 0.1) +
  geom_line(aes(x = year, y = pred, group = id),
            linewidth = 0.5) +
  facet_wrap( ~ id, scales = "free") +
  geom_point(aes(x = year, y = log(obs), group = id, color = units),
             size = 0.7) + 
  geom_line(aes(x = year, y = log(obs), group = id, color = units),
             linewidth = 0.3) +

  theme_bw() + 
  labs(y = 'Counts (log-scale)', x = 'Year', color = 'Units')

```
The four populations are also measured in the same basic units, but represent likely represent very different populations, spanning ones in small parks that cannot migrate to the large migrating herds (e.g., Serengeti Masai-Mara), though gathering additional metadata to fully check this would be another important component to add to our workflow for these data.

## Feedbacks

At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one sigma for all populations (CHECK HERE: do we really use one sigma for all populations?), how we model population trends (e.g., there are many ways to model abundance) and (relatedly) our linearity assumption. One possibility would be to move to a non-linear regression, and use a more flexible model that would allow population numbers to stabilize, such as shown using a logistic model for one wildbeest population in Tanzania:

```{r}
#| fig-height: 2.5
#| fig-width: 3.5

subset_wildebeest <- lpi_subset_rsh[lpi_subset_rsh$ID == 5589,]
subset_wildebeest$year <- subset_wildebeest$year-1980
gompodel <- nls(log(value) ~ SSlogis(year, Asym, b2, b3), data=subset_wildebeest)
logisfit <-data.frame(pred = predict(gompodel, newdata = data.frame(year = seq(-25,30,1))), year = seq(-25,30,1))

ggplot(data = lpi_subset_rsh[lpi_subset_rsh$ID == 5589,]) +
  facet_wrap( ~ ID, scales = "free_y") +
  geom_point(aes(x = year, y = value),
             size = 0.7, color = "grey") + 
  geom_line(aes(x = year, y = value),
             linewidth = 0.3, color = "grey") + 
  geom_line(aes(x = year + 1980, y = exp(pred)),
            linewidth = 0.5, data = logisfit, linetype = 'dashed') +
  theme_bw() + 
  labs(y = 'Counts', x = 'Year')

```

We also have seen variability in population trends within species (e.g., two apparently increasing populations and two apparently decreasing populations in Tanzania), so we would want to think a lot more about how we scale up from population to species to capture this variation in a way that is most useful to our question. For example, may want to think about adding country, which may capture differences in efforts related to conservation and thus predict trends (or we may have better metrics of what drives these trends, but given our global aim, we would need it for most/all populations) and we may want to think about how or if we partially pool species and populations within species (potentially using models other than a normal distribution for pooling). As we address these potential problems, we would likely begin to add in more countries, species, and populations.

We expect scaling up to more countries, species and populations may cause us to think carefully about the data we have in hand versus the question we aimed to answer. While many recent papers treat the LPI data as effectively a well stratified and randomized sample or even effectively a good cencus, it is clearly not. Thus, models that try to deal with such uneven sampling and build in the more of the structure of data, would be far more appropriate. For example, most models for election forecasts now include regularization via partial pooling, post-stratificatiob and other methods designed to adjust for the fact that many polls are 'non-representative.' Such polls may be more similar to LPI data. Our model includes some regularization via partial pooling, but we may need far more of an observational model added to our population demography model to accommodate the uneven sampling. Estimates from such a model may be far more uncertain, but likely also more correct given our combination of question and data.

This may lead us to realize we really don't have great data for this question and we would then need to prioritize better data collection efforts if we truly want good estimates to answer this question.

## The full process

The small workflow we should here may seem like extra work to those who have used it, and it is compared to many ecological workflows in practice. We argue that the extra effort is worth it given the critical questions we are addressing, and also that the process becomes much quicker with practice. The first time may be slow, but the resulting model will be better for it (even if it is still arguably simple, as the one we show here is), but the following times will go faster with likely improved models.

### 2025-04-30 forecastflows.tex
cleaning

@@
 
-\todo[inline, size=\tiny]{For below: jdavies15april: might be better grouped together in concluding section.
-Reply: if we want to put this in the concluding section, we would need to talk also about the "opportunities to bridge statistical and process-based", rather than keeping it in the box?
-So, what should we put in the box?}
-
 % Box

### 2025-04-30 analyses/Johnsonetal2024/code/modelling_abundances.R

@@ -1,116 +1,0 @@
-library(INLA)
-library(brinla)
-library(data.table)
-analysis_list = readRDS("data/derived_data/analysis_list.rds")
-
-
-
-prior_prec = "expression:
-  log_dens = 0 - log(2) - theta / 2;
-  return(log_dens);
-"
-
-# Run suite of models
-
-
-#Random intercept (different intercept at multiple levels) - a model which assumes trends are similar across space/species
-
-for(a in names(analysis_list)[1:10]){ 
-  
-  tmp_df = analysis_list[[a]][[1]]
-  tmp_df$log_abundance_trim = tmp_df$log_abundance
-  tmp_df$cent_abundance_trim = tmp_df$cent_abundance
-  
-  uni_sites = unique(tmp_df$site_spec_code)
-  uni_sites_sample = sample(uni_sites,(length(uni_sites)/2))
-  remove_abundances = list()
-  for(c in c(1:(length(uni_sites)))){
-    trim_tmp_df = subset(tmp_df, site_spec_code == uni_sites[c])
-    if(uni_sites[c] %in% uni_sites_sample){
-      trim_tmp_df$log_abundance_trim[c(nrow(trim_tmp_df))]  = NA
-      trim_tmp_df$cent_abundance_trim[c(nrow(trim_tmp_df))]  = NA
-    } else {
-      
-    }
-    remove_abundances[[c]] = trim_tmp_df
-  }
-  remove_abundances = rbindlist(remove_abundances)
-  analysis_list[[a]][[1]] = remove_abundances
-  
-  print(a)
-  message("   Model 1 - Intercept model")
-  m1 = inla(log_abundance_trim ~
-              year_centre +
-              f(site_spec_code,  model = "iid", constr = F, hyper = prior_prec) +
-              f(tips_code, model = "iid", constr = F, hyper = prior_prec) +
-              f(genus_code, model = "iid", constr = F, hyper = prior_prec) +
-              f(site_code, model = "iid", constr = F, hyper = prior_prec) +
-              f(region_code, model = "iid", constr = F, hyper = prior_prec),
-            data = analysis_list[[a]][[1]],  family = "gaussian", 
-            control.predictor=list(compute=TRUE),
-            control.fixed = list(mean.intercept = 0, prec.intercept = 0.001,
-                                 mean = 0, prec = 1),
-            num.threads = "1:1")
-    m1 = inla.rerun(m1)
-    m1 = inla.rerun(m1)
-  
[…]

### 2025-04-30 analyses/Johnsonetal2024/code/modelling_core.R

@@ -1,219 +1,0 @@
-library(INLA)
-library(brinla)
-analysis_list = readRDS("data/derived_data/analysis_list.rds")
-
-
-
-prior_prec = "expression:
-  log_dens = 0 - log(2) - theta / 2;
-  return(log_dens);
-"
-
-
-
-# Run suite of models
-
-
-#Random intercept (different intercept at multiple levels) - a model which assumes trends are similar across space/species
-
-cmb_df = NULL
-cmb_df2 = NULL
-for(a in names(analysis_list)[c(1:10)]){ 
-  for(b in c(1:10)){
-  print(a)
-  print(b)
-  message("   Model 1 - Intercept model")
-  if(b == 1){
-    m1 = inla(log_abundance ~
-                year_centre +
-                f(site_spec_code,  model = "iid", constr = F, hyper = prior_prec) +
-                f(tips_code, model = "iid", constr = F, hyper = prior_prec) +
-                f(genus_code, model = "iid", constr = F, hyper = prior_prec) +
-                f(site_code, model = "iid", constr = F, hyper = prior_prec) +
-                f(region_code, model = "iid", constr = F, hyper = prior_prec),
-              data = analysis_list[[a]][[1]],  family = "gaussian", 
-              control.predictor=list(compute=TRUE),
-              control.fixed = list(mean.intercept = 0, prec.intercept = 0.001,
-                                   mean = 0, prec = 1),
-              num.threads = "1:1")
-    m1_coef_mn = m1$summary.fixed[2,4]
-    m1_coef_lc5 = inla.hpdmarginal(0.5, m1$marginals.fixed[[2]])[1]
-    m1_coef_uc5 = inla.hpdmarginal(0.5, m1$marginals.fixed[[2]])[2]
-    
-    
-    
-    #Random slope (different slopes at multiple levels) - a model which captures dissilmar trends across space/species, but assumes independence instead of covariance
-    
-    
-    message("   Model 2 - Slope model")
-    m2 = inla(cent_abundance ~ 
-                year_centre +
[…]

### 2025-04-30 analyses/Johnsonetal2024/code/modelling_lpi_assumption.R

@@ -1,74 +1,0 @@
-library(INLA)
-library(brinla)
-
-
-
-prior_prec = "expression:
-  log_dens = 0 - log(2) - theta / 2;
-  return(log_dens);
-"
-
-  m1 = inla(log_abundance ~
-              year_centre +
-              f(site_spec_code,  model = "iid", constr = F, hyper = prior_prec) +
-              f(tips_code, model = "iid", constr = F, hyper = prior_prec) +
-              f(genus_code, model = "iid", constr = F, hyper = prior_prec) +
-              f(site_code, model = "iid", constr = F, hyper = prior_prec) +
-              f(region_code, model = "iid", constr = F, hyper = prior_prec),
-            data = analysis_list_lpi[[1]],  family = "gaussian", 
-            control.predictor=list(compute=TRUE),
-            control.fixed = list(mean.intercept = 0, prec.intercept = 0.001,
-                                 mean = 0, prec = 1),
-            num.threads = 4)
-
-  
-  
-  #Random slope (different slopes at multiple levels) - a model which captures dissilmar trends across space/species, but assumes independence instead of covariance
-  
-  
-  message("   Model 2 - Slope model")
-  m2 = inla(cent_abundance ~ 
-              year_centre +
-              f(site_spec_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(tips_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(genus_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(site_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(region_code, year_centre, model = "iid", constr = F, hyper = prior_prec),
-            data = analysis_list_lpi[[1]],  family = "gaussian", 
-            control.predictor=list(compute=TRUE),
-            control.fixed = list(mean.intercept = 0, prec.intercept = 0.001,
-                                 mean = 0, prec = 1),
-            num.threads = 4)
-  m2_sig = bri.hyperpar.summary(m2)[1,4]
-
-  
-  ar_prior = list(theta1 = list(prior="pc.prec", param=c(m2_sig*3, 0.01)),
-                  theta2 = list(prior="pc.cor1", param=c(0, 0.9), initial = 0))
-  
-  #Correlated effect - a model which captures covariances and similarties in trends between neighbouring sites and species
-  message("   Model 3 - Correlation model")
-  m3 = inla(cent_abundance ~ 
[…]

### 2025-04-30 analyses/Johnsonetal2024/code/modelling_phylo.R

@@ -1,163 +1,0 @@
-library(INLA)
-library(brinla)
-analysis_list = readRDS("data/derived_data/analysis_list_phylo.rds")
-
-
-
-
-prior_prec = "expression:
-  log_dens = 0 - log(2) - theta / 2;
-  return(log_dens);
-"
-
-
-# Run suite of models
-
-
-#Random intercept (different intercept at multiple levels) - a model which assumes trends are similar across space/species
-
-cmb_df = NULL
-for(a in names(analysis_list)[c(1:10)]){
-  print(a)
-  
-  
-  m2 = inla(cent_abundance ~ 
-              year_centre +
-              f(site_spec_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(tips_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(genus_code1, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(site_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(region_code, year_centre, model = "iid", constr = F, hyper = prior_prec),
-            data = analysis_list[[a]][[1]],  family = "gaussian", 
-            control.predictor=list(compute=TRUE),
-            control.fixed = list(mean.intercept = 0, prec.intercept = 0.001,
-                                 mean = 0, prec = 1),
-            num.threads = "1:1")
-  m2_sig = bri.hyperpar.summary(m2)[1,4]
-  ar_prior = list(theta1 = list(prior="pc.prec", param=c(m2_sig*3, 0.01)),
-                  theta2 = list(prior="pc.cor1", param=c(0, 0.9), initial = 0))
-  
-  message("   Model 3 - Correlation model with the trimmed rotl data")
-  
-  m3a = inla(cent_abundance ~ 
-               year_centre +
-               f(year3, model = "ar1", replicate = site_spec_code2, hyper = ar_prior) +
-               f(site_spec_code, year_centre, model = "iid", 
-                 constr = F, hyper = prior_prec) +
[…]

### 2025-04-30 analyses/Johnsonetal2024/code/modelling_predict2.R

@@ -1,86 +1,0 @@
-library(INLA)
-library(brinla)
-analysis_list = readRDS("data/derived_data/analysis_list_predict2.rds")
-
-
-
-prior_prec = "expression:
-  log_dens = 0 - log(2) - theta / 2;
-  return(log_dens);
-"
-
-
-# Run suite of models
-
-
-#Random intercept (different intercept at multiple levels) - a model which assumes trends are similar across space/species
-
-cmb_df = NULL
-for(a in names(analysis_list)){ 
-  print(a)
-  
-  m1 = inla(log_abundance ~
-              year_centre +
-              f(site_spec_code,  model = "iid", constr = F, hyper = prior_prec) +
-              f(tips_code, model = "iid", constr = F, hyper = prior_prec) +
-              f(genus_code, model = "iid", constr = F, hyper = prior_prec) +
-              f(site_code, model = "iid", constr = F, hyper = prior_prec) +
-              f(region_code, model = "iid", constr = F, hyper = prior_prec),
-            data = analysis_list[[a]][[1]],  family = "gaussian", 
-            control.predictor=list(compute=TRUE),
-            control.fixed = list(mean.intercept = 0, prec.intercept = 0.001,
-                                 mean = 0, prec = 1),
-            num.threads = "1:1")
-  
-  #Random slope (different slopes at multiple levels) - a model which captures dissilmar trends across space/species, but assumes independence instead of covariance
-  
-  
-  message("   Model 2 - Slope model")
-  m2 = inla(cent_abundance ~ 
-              year_centre +
-              f(site_spec_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(tips_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(genus_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(site_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(region_code, year_centre, model = "iid", constr = F, hyper = prior_prec),
-            data = analysis_list[[a]][[1]],  family = "gaussian", 
-            control.predictor=list(compute=TRUE),
-            control.fixed = list(mean.intercept = 0, prec.intercept = 0.001,
-                                 mean = 0, prec = 1),
-            num.threads = "1:1")
-  
-  m2_sig = bri.hyperpar.summary(m2)[1,4]
-  ar_prior = list(theta1 = list(prior="pc.prec", param=c(m2_sig*3, 0.01)),
[…]

### 2025-04-30 analyses/Johnsonetal2024/code/modelling_structure.R

@@ -1,189 +1,0 @@
-library(INLA)
-library(brinla)
-analysis_list = readRDS("data/derived_data/analysis_list.rds")
-
-
-
-prior_prec = "expression:
-  log_dens = 0 - log(2) - theta / 2;
-  return(log_dens);
-"
-
-
-# Run suite of models
-
-
-#Random intercept (different intercept at multiple levels) - a model which assumes trends are similar across space/species
-
-cmb_df = NULL
-for(a in names(analysis_list)[1:10]){
-  print(a)
-  
-  m2 = inla(cent_abundance ~ 
-              year_centre +
-              f(site_spec_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(tips_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(genus_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(site_code, year_centre, model = "iid", constr = F, hyper = prior_prec) +
-              f(region_code, year_centre, model = "iid", constr = F, hyper = prior_prec),
-            data = analysis_list[[a]][[1]],  family = "gaussian", 
-            control.predictor=list(compute=TRUE),
-            control.fixed = list(mean.intercept = 0, prec.intercept = 0.001,
-                                 mean = 0, prec = 1),
-            num.threads = 4)
-  m2_sig = bri.hyperpar.summary(m2)[1,4]
-  ar_prior = list(theta1 = list(prior="pc.prec", param=c(m2_sig*3, 0.01)),
-                  theta2 = list(prior="pc.cor1", param=c(0, 0.9), initial = 0))
-  
-  print("   Model 3 - Correlation model without time")
-  m3_a = inla(cent_abundance ~ 
-                year_centre +
-                f(site_spec_code, year_centre, model = "iid", 
-                  constr = F, hyper = prior_prec) +
-                f(tips_code2, year_centre, model = "generic0",
-                  constr = F, Cmatrix = analysis_list[[a]][[2]], hyper = prior_prec) +
-                f(tips_code, year_centre, model = "iid", 
-                  constr = F, hyper = prior_prec) +
-                f(genus_code, year_centre, model = "iid", 
-                  constr = F, hyper = prior_prec) +
-                f(site_code2, year_centre, model = "generic0", 
[…]

### 2025-04-30 analyses/Johnsonetal2024/code/modelling_trends.R

@@ -1,251 +1,0 @@
-library(INLA)
-library(brinla)
-library(data.table)
-library(dplyr)
-analysis_list = readRDS("data/derived_data/analysis_list.rds")
-
-
-
-prior_prec = "expression:
-  log_dens = 0 - log(2) - theta / 2;
-  return(log_dens);
-"
-
-
-# Run suite of models
-cmb_df = NULL
-for(a in names(analysis_list)[c(1:10)]){ 
-  
-  m_tmp1 = inla(cent_abundance ~
-                  f(site_spec_code, year_centre, model = "iid", constr = F, hyper = prior_prec), data = analysis_list[[a]][[1]])
-  sig = bri.hyperpar.summary(m_tmp1)[1,4]
-  ar_prior = list(theta1 = list(prior="pc.prec", param=c(sig*3, 0.01)),
-                  theta2 = list(prior="pc.cor1", param=c(0, 0.9), initial = 0))
-  m_tmp2 = inla(cent_abundance ~
-                  f(year3, model = "ar1", replicate = site_spec_code2, hyper = ar_prior) +
-                  f(site_spec_code, year_centre, model = "iid", 
-                    constr = F, hyper = prior_prec),
-                data = analysis_list[[a]][[1]])
-  sites = m_tmp2$summary.random$site_spec_code
-  sites$sig = ifelse(sites$`0.025quant` > 0 | sites$`0.975quant` < 0, "sig", "ns")
-  
-  cmb_trends = NULL
-  for(b in c(1:50)){
-    print(b)
-    tmp_df = analysis_list[[a]][[1]]
-    tmp_df$log_abundance_trim = tmp_df$log_abundance
-    tmp_df$cent_abundance_trim = tmp_df$cent_abundance
-    
-    uni_sites = sites[which(sites$sig == "sig"),]$ID
-    uni_sites_sample = sample(uni_sites,1)
-    remove_trends = list()
-    for(c in uni_sites){
-      trim_tmp_df = subset(tmp_df, site_spec_code == c)
-      if(uni_sites[c] %in% uni_sites_sample){
-        trim_tmp_df$log_abundance_trim = NA
-        trim_tmp_df$cent_abundance_trim = NA
-      } else {
-        
-      }
-      remove_trends[[c]] = trim_tmp_df
-    }
-    remove_trends = rbindlist(remove_trends)
-    
-    
[…]

### 2025-04-30 analyses/jdavies/LPI subset African mammals.R

@@ -1,33 +1,0 @@
-################################################################################################
-################################################################################################
-
-setwd("C:/Jonathan/Dropbox/BLUE/Zurich/LPI/LivingPlanetIndexDatabase_2025-02-01_19.19.51")
-data<-read.csv("LPD_2024_public.csv", header=T)
-
-Af1<-subset(data, Region == "Africa")
-Af2<-subset(Af1, Included.in.LPR2024 == 1)
-Af3<-subset(Af2, Class == "Mammalia")
-
-#ignoring units for now
-#unique(Af$Units)
-
-# subset 5+ counts
-obs.cnt<-NULL
-for (n in 1:dim(Af3)[1]){
-obs.cnt[n]<-sum(Af3[n, 31:101]!="NULL")
-}
-
-Af4<-Af3[obs.cnt>4,]
-
-# subset species 10+ populations
-sp<-unique(Af4$Binomial)#
-
-sp.cnt<-NULL
-for(x in 1:length(sp)){
-Af.tmp<-subset(Af4, Binomial == sp[x])
-sp.cnt[x]<-dim(Af.tmp)[1]
-}
-
-keep.sp<-sp[sp.cnt>9]
-Af5<-Af4[Af4$Binomial %in% keep.sp,]
-#14 sp remaining

### 2025-04-30 analyses/starterLPI.R

@@ -1,50 +1,0 @@
-## Started 17 April 2025 ##
-## By Lizzie so far ... ## 
-
-# housekeeping
-rm(list=ls()) 
-options(stringsAsFactors=FALSE)
-
-# set your directory if you're me
-setwd("~/Documents/git/projects/misc/miscmisc/workflowsPhilTrans/analyses")
-
-# packages 
-library(data.table)
-# What?! These two are because I was cheap and copied from Johnson et al. 2024
-library(countrycode)
-library(tidyverse)
-
-
-lpi <- fread("input/LPD2022_public.csv", na = c("", "NA", "NULL"))
-
-# Below is copied from :
-# Johnson_etal_2024_code_GitTFJ-correlated_effect_model-21c0d95/code/data_compile.Rmd
-custom_match <- c(`Cura??ao` = "CUW", `?Óland Islands` = "ALA", `International Waters` = "ABNJ")
-
-lpi <- subset(lpi, Replicate == 0)
-lpi <- lpi %>% 
-  mutate(
-    citation_year = str_replace_all(Citation, "\\\"|\\.|-|\\\'", ""),
-    citation_year = parse_number(citation_year),
-    site = paste(word(Citation, 1, sep = " |,"),
-                 citation_year,
-                 Location,
-                 sep = "_")
-  ) %>%
-  select(site, Country, Latitude, Longitude, Binomial, Units, `1950`:`2020`) %>% 
-  pivot_longer(cols = `1950`:`2020`,
-               names_to = "date",
-               values_to = "abundance")
-lpi <- as.data.frame(lpi)
-
-
-## Thinking we need to subset for sanity ... 
-length(unique(lpi$site))
-sort(unique(lpi$Country))
-table(lpi$Country, lpi$site)
-
-sa <- subset(lpi, Country=="South Africa")
-
-ggplot(sa, aes(x=date, y=abundance)) +
-	geom_point() + 
-	facet_wrap(.~site)

### 2025-04-30 analyses/lpiworkflow.qmd
small change

@@
 
-What is global trend over time in vertebrate species population numbers, both overall and how does it vary acrosss species and populations?
+What is global trend over time in vertebrate species population numbers, both overall and how does it vary across species and populations?
 
@@
 
-lpi_subset <- read.table(file.path(wd, "jdavies/African_mammals.txt"))
-
-lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Units", paste0("X", 1950:2020))]
-lpi_subset$genus_species <- paste(lpi_subset$Genus, lpi_subset$Species)
-
-lpi_subset_rsh <- 
-  reshape(lpi_subset, direction = "long", idvar = "ID",
-          varying =  paste0("X", 1950:2020), v.names = c("value"), timevar = "year", times = 1950:2020)
-lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != "NULL",]
+lpi_data <- read.csv(file.path(wd, "input", "LPD_2024_public.csv"), header=T)
 
@@
 We chose this because mammals are a well studied group compared to other clades/groups in the dataset, so we will have a better idea of how to simulate data, and whether our model makes sense.
+
+```{r}
+
+lpi_subset <-subset(lpi_data, Region == "Africa" & Class == "Mammalia" & Included.in.LPR2024 == 1 )
+
+# subset 5+ counts
+obs.cnt<-NULL
+for (n in 1:dim(lpi_subset)[1]){
+  obs.cnt[n]<-sum(lpi_subset[n, 31:101]!="NULL")
+}
+lpi_subset<-lpi_subset[obs.cnt>4,]
+
+# subset species 10+ populations
+sp <- unique(lpi_subset$Binomial)
+sp.cnt<-NULL
+for(x in 1:length(sp)){
+  Af.tmp<-subset(lpi_subset, Binomial == sp[x])
+  sp.cnt[x]<-dim(Af.tmp)[1]
+}
+keep.sp<-sp[sp.cnt>9]
+lpi_subset<-lpi_subset[lpi_subset$Binomial %in% keep.sp,]
+
+```
 
@@
 
-First, we start with the parameters, which include an overall estimate of the trend across species (this parameter would address our question that we decided on above):
+First, we start with the parameters, which include an overall estimate of the trend across species (this parameter would address our question that we decided on above).
+The red dashed line represents the true simulated value:
 
@@
 #| eval: true
-
-lpi_subset <- read.table("/home/victor/projects/forecastflows/analyses/jdavies/African_mammals.txt")
 
@@
 
-At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one sigma for all populations (CHECK HERE: do we really use one sigma for all populations?), how we model population trends (e.g., there are many ways to model abundance) and (relatedly) our linearity assumption.
+At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one observational error $\sigma$ for all populations, how we model population trends (e.g., there are many ways to model abundance) and (relatedly) our linearity assumption.
 One possibility would be to move to a non-linear regression, and use a more flexible model that would allow population numbers to stabilize, such as shown using a logistic model for one wildbeest population in Tanzania:

### 2025-05-01 analyses/lpiworkflow.qmd
small additions

@@
 To show our workflow in action, we examine a subset of the Living Planet Index (LPI, download at: https://www.livingplanetindex.org/data_portal.
-This download should include a csv file, here: LPD2022_public.csv, that we use below).
+This download should include a csv file, here: LPD_2024_public.csv, that we use below).
 We discuss the LPI in the main text; for more information, check out: www.livingplanetindex.org/.
-
-**Note**: data from Jonathan comes from *LivingPlanetIndexDatabase_2025-02-01_19.19.51*.
 
@@
 
+::: {.callout-tip appearance="simple" icon=false}
+
 What is global trend over time in vertebrate species population numbers, both overall and how does it vary across species and populations?
 
+:::
+
 ```{r}
+
 wd <- "/home/victor/projects/forecastflows/analyses"
@@
 
+::: {.callout-tip appearance="simple" icon=false}
+
 What is trend in mammal species population numbers in southern Africa, both overall and how does it vary across species and populations?
+
+:::
 
@@
 
-For now, we also take a look at how model fits look against the simulated data:
+For now, we also take a look at how model fits look against the simulated data (error bars and ribbon represent the 95% credible intervals):
 
@@
 
-units_to_keep <- c("Estimate - entire population", "Estimate - entire population (August)",
-                   "Individual", "individuals", "Individuals", "Number of adults", "number of individuals",
-                   "Number of individuals", "Number of Individuals", "Number of rhinos", "numbers of indviduals",
-                   "population estimate", "Population estimate", "Population estimate (individuals)", 
-                   "Population estimates", "Total population size")
+units_to_keep <- c("Estimate - entire population", "Estimate - entire population (August)", "Individual", "individuals", "Individuals", "Number of adults", "number of individuals", "Number of individuals", "Number of Individuals", "Number of rhinos", "numbers of indviduals", "population estimate", "Population estimate", "Population estimate (individuals)", "Population estimates", "Total population size")
 

### 2025-05-01 forecastflows.tex
caption fig 1

@@
 \usepackage{todonotes}
+\usepackage[font=sf]{caption}
 
@@
 % jdavies15april: I feel this could be a critical step somehow - sharing simulations could allow cross-validation between models or perhaps cross-calibration ...
-These problems stem in part from continuing gaps between statistical workflows---where the data processing and model development are often addressed separately from the ecological question and aim---and scientific workflows, where all steps are integrated.
+% These problems stem in part from continuing gaps between statistical workflows---where the data processing and model development are often addressed separately from the ecological question and aim---and scientific workflows, where all steps are integrated.
 Yet, as ecologists become increasingly computational the opportunity to close this gap has never been greater.
@@
 %Combining these changes with more open model and data sharing---and developing new efforts to race the same data---could be transformative for ecological forecasting. 
-A shift toward universal training in a more robust model building could bridge the gap between process-based and statistical approaches and be transformative for ecological modeling.
+%A shift toward universal training in a more robust model building could bridge the gap between process-based and statistical approaches and be transformative for ecological modeling.
 \end{abstract}
@@
 Impacts are projected to increase in coming decades, as climate change accelerates biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}.
-% jdavies15april: not sure what models you are referring to here, but perhaps look up models of 'mean species abundance'.
 Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics.
@@
 
+\begin{figure}[h]
+	\centering
+	\includegraphics{figures/modeldataspaces_rotation}
+	\caption{Different trajectories through model-data space during model development.
+Ideally, this begins in the pre-model/pre-data quadrant, where the research question is defined.
+The next step moves to the post-model/pre-data quadrant, where a model is build---independently of the data.
+This is where we should spent most of the time, evaluating model behavior using simulated data.
+Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data.
+Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedbacks to refine our model.}
+	\label{fig:modeldata}
+\end{figure}
+
 This reality should drive researchers to use more robust and coherent methods.
@@
 Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Fig.
-\ref{fig:modeldata}).
+\ref{fig:modeldata}a).
 For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (Fig.
-\ref{fig:modeldata}).
+\ref{fig:modeldata}b).
 Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
@@
 
-\begin{figure}[h]
-	\centering
-	\includegraphics{figures/modeldataspaces_rotation}
-	\caption{Here the caption could explain a bit more the pre-model pre-data to the post-model post-data...} 
-	\label{fig:modeldata}
-\end{figure}
 % jdavies15april: Could the bottom right be where exploratory data analysis fits in, generating hypotheses that inform the research question?

### 2025-05-01 analyses/lpiworkflow.qmd
Small edits to workflow text (and 0 issue)

@@
 title: "A sample workflow"
-authors: "Lizzie, Victor"
+authors: "Victor, Lizzie"
 date: "April 2025"
@@
 This download should include a csv file, here: LPD_2024_public.csv, that we use below).
-We discuss the LPI in the main text; for more information, check out: www.livingplanetindex.org/.
+We discuss the LPI in the main text; for more information, check out [their website](www.livingplanetindex.org/).
 
@@
 And, if so, by how much?
-We, however might want to have a question that more clearly allows any directionality, so we could start with: What is the global trend for biodiversity in vetebrate species?
+We, however, might want to have a question that more clearly allows any directionality, so we could start with: What is the global trend for biodiversity in vetebrate species?
 This is rather broad and might mean we should develop a model that will yield one global estimate.
-In thinking about this, we would likely realize we don't want just that, we actually want to understand the variation across different species and perhaps populations and also a global estimate.
+In thinking about this, we would likely realize we don't want just that, we actually want to understand the variation across different species and perhaps populations and also have a global estimate.
 So we might refine the question to:
@@
 Next we need to think about how we would build a model to address this question, with the goal to build the model and evaluate it using simulated data.
-There is a lot of demographic models in population ecology, so we might revisit those and consider a discrete time model based on population size and stepping through each year of data.
+There are a lot of demographic models in population ecology, so we might revisit those and consider a discrete time model based on population size and stepping through each year of data.
 We might also start with a simple linear model, where populations can go up, down or do nothing.
@@
 Populations usually are nested in species in ecological models, so we might consider that to start.
-We also need to make sure our model yields one global estimate, so for this we decide to model species trends over time as part of larger distribution.
+We also need to make sure our model yields one global estimate; for this we decide to model species trends over time as part of larger distribution.
 Our resulting model would then be:
@@
 
-For now, we also take a look at how model fits look against the simulated data (error bars and ribbon represent the 95% credible intervals):
+For now, we also take a look at how model fits look against the simulated data (error bars and ribbon represent the 95% uncertainty intervals):
 
@@
 The values in the database are not necessarily in the same units, so we subset to units related to absolute population sizes, which is what we have mostly been thinking of (some units are in density and we would need to build our model to incorporate those, but here we start first with a simpler model and can build up later).
-We take a quick look at the data to make sure it looks reasonable based on our domaine knowledge.
+We take a quick look at the data to make sure it looks reasonable based on our domain knowledge.
 
@@
 
-lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,] # poor way to avoid log(0)
+lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,] # We now also find we have counts of 0, which is tricky for our approach (as we cannot evaluate log(0)), we would also flag this to return.
+As we iteratively improve the model, we would need to consider what processes generate these zeroes, and include them in our model.
 
@@
 
-Here, we might step back and adjust our model (re-doing Step 2-3 again), but in this case we go ahead.
+Our model may accommodate this, as we have allowed different populations to have different intercepts (starting population sizes), but the scale of this variation is much larger than we simulated so we would again want to flag this and check how our model performs with these data.
 
@@
 Now we use our fitted model parameters to generate predictions, and compare them to observations.
-This is similar to the simulating data above, but differs in two major ways: (1) we simulate from the parameters estimated by the model fit to the empirical data (versus coming up with them based on our expert knowledge) and (2) we need to check across many predictions from the model to understand the full uncertainty of the model (and not just one draw from the model).
+This is similar to the simulating data step above, but differs in two major ways: (1) we simulate from the parameters estimated by the model fit to the empirical data (versus coming up with them based on our expert knowledge) and (2) we need to check across many predictions from the model to understand the full uncertainty of the model (and not just one draw from the model).
 
@@
 
-Focusing on populations in South Africa, we see some places where the linear model appears well (ID 7673), but also two other populations where it does not appear as a very good fit.
+Focusing on populations in South Africa, we see some places where the linear model appears to fit reasonably well (ID 7673), but also one population where it does not appear as a very good fit (ID 10010).
 
@@
 
-At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one observational error $\sigma$ for all populations, how we model population trends (e.g., there are many ways to model abundance) and (relatedly) our linearity assumption.
+At this point, we would have a lot of ideas of ways to improve this model that we would want to follow up on, including (but not limited to) questioning our assumption of one observational error $\sigma$ for all populations, how we model population trends (e.g., there are many ways to model abundance), and (relatedly) our linearity assumption.
 One possibility would be to move to a non-linear regression, and use a more flexible model that would allow population numbers to stabilize, such as shown using a logistic model for one wildbeest population in Tanzania:
@@
 We also have seen variability in population trends within species (e.g., two apparently increasing populations and two apparently decreasing populations in Tanzania), so we would want to think a lot more about how we scale up from population to species to capture this variation in a way that is most useful to our question.
-For example, may want to think about adding country, which may capture differences in efforts related to conservation and thus predict trends (or we may have better metrics of what drives these trends, but given our global aim, we would need it for most/all populations) and we may want to think about how or if we partially pool species and populations within species (potentially using models other than a normal distribution for pooling).
+For example, may want to think about adding country, which may capture differences in efforts related to conservation and thus predict trends (or we may have better metrics of what drives these trends, but given our global aim, we would ideally need the predictor for most/all populations) and we may want to think about how or if we partially pool species and populations within species (potentially using models other than a normal distribution for pooling).
 As we address these potential problems, we would likely begin to add in more countries, species, and populations.
 
-We expect scaling up to more countries, species and populations may cause us to think carefully about the data we have in hand versus the question we aimed to answer.
+We expect scaling up to more countries, species and populations may cause us to think carefully about the data we have in hand and our approach versus the question we aimed to answer.
 While many recent papers treat the LPI data as (effectively) a well stratified and randomized sample or even a good census, the LPI data is clearly neither.
-Thus, models that try to deal with such uneven sampling and build in the more of the structure of data, would be more appropriate.
+Thus, models that try to deal with such uneven sampling and build in more of the structure of data would be more appropriate.
 For example, most models for election forecasts now include regularization via partial pooling, post-stratification and other methods designed to adjust for the fact that many polls are 'non-representative.' Such non-representative polls may be very similar to LPI data.
@@
 
-The small workflow we show here may seem like extra work to those who have used it, and it is compared to many ecological workflows in practice.
+The small workflow we show here may seem like extra work to those who have not used it, and it is compared to many ecological workflows in practice.
 We argue that the extra effort is worth it given the critical questions we are addressing, and also that the process becomes much quicker with practice.
-The first time may be slow, but the resulting model will be better for it (even if it is still arguably simple, as the one we show here is), but the following uses of this iterative workflow will go faster with likely improved models.
+The first time may be slow, but the resulting model will be better for it (even if it is still arguably simple, as the one we show here is), but the following uses of this iterative workflow will go faster and very likely yield improved models.

### 2025-05-01 forecastflows.tex
integrate Lizzie edits and other stuff

@@
 \documentclass[11pt]{article}
-\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
+\usepackage[top=0.70in, bottom=.80in, left=1in, right=1in]{geometry}
 \renewcommand{\baselinestretch}{1.1}
@@
 \usepackage[font=sf]{caption}
+\usepackage{hyperref}
 
@@
 \date{\today}
-\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}
+\author{Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
 \maketitle
+\noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
+$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
+$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
+$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}\\
 
-{\bf Deadline:} 1-ish May 2025 (was 1 April 2025, getting in by May 8th would be good!)
-
-\emph{For:} Scientific and Statistical Workflow theme issue for \emph{Phil Trans A} as an \emph{Opinion}
+\noindent \emph{For:} Scientific and Statistical Workflow theme issue for \emph{Phil Trans A} as an \emph{Opinion}
 % They mention a tex template here: https://royalsocietypublishing.org/rsta/for-authors#question4
@@
 Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions.
-While new research often suggests that models accounting for these complexities can yield more accurate trends and predictions, we argue that the usual workflows for model fitting in ecology make it impossible to evaluate and compare the models used to generate insights and predictions.
+New research often suggests that models accounting for these complexities can yield more accurate trends and predictions.
+We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons.
 First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data.
@@
 We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process.
-We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could be transformative for ecological modeling.
+We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
 
@@
 
-{\noindent \bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
-\vspace*{0.5cm}
+% {\noindent \bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
 
@@
 Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on less data.
-Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from massive geographic, temporal and taxonomic biases.
+Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic biases.
 Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
 Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning.
-These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.
+These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012, Thuiller2019}.
 
@@
 Workflows that fully integrate all the steps required to build a model from an ecological question, with evaluation of limitations and potential problems before estimating its parameters and making projections, could reduce many of these problems.
-In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge.
+In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge \citep{McElreath2018, betanworkflow, Gelman2020, Wolkovich2024}.
 Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
@@
 In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses.
-This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection.
-%add cites for prereg?
+This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection \citep{Nosek2018}.
 But this idealized method often does not apply to the reality of ecological research.
@@
 The next step moves to the post-model/pre-data quadrant, where a model is build---independently of the data.
-This is where we should spent most of the time, evaluating model behavior using simulated data.
+This is where we should spend a good deal of time evaluating model behavior using simulated data.
 Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data.
@@
 But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems.
-% jdavies15april: what challenges?
-forecasting from imperfect data or the urgent need to generate forecasts?
-Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Fig.
-\ref{fig:modeldata}a).
-For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (Fig.
-\ref{fig:modeldata}b).
+Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}a).
+For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (figure \ref{fig:modeldata}b).
 Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
-% jdavies15april: is this particularly specific to process based models?
 Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.
-% mention the disconnection between trends and forecasts?
 
-% jdavies15april: Could the bottom right be where exploratory data analysis fits in, generating hypotheses that inform the research question?
-
-We argue a workflow that moves along the data-model space in a coherent sequence of steps % with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure?
-Or wait to mention simulation until referencing Fig 2?
-[FLAG: what do you think?
-I agree you should choose on of the two options]
-(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
-The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, fig.
-\ref{fig:workflow}).
+We argue a workflow that moves along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
+The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}).
 This involves making clear assumptions about the most influential drivers, within the specific context of our study.
 This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error.
-% vvdm16april: to answer Jonathan comment about verbal model
-From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, fig.
-\ref{fig:workflow}).
-The general idea is to start with a relatively simple model that we could refine later.
+From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}).
+The general idea is to start with a relatively simple model that we could refine later (see example worklow we provide).
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
@@
 
-With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, fig.
-\ref{fig:workflow}).
+With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}).
 `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data.
-% This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model.
-% jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e.
-simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model.
-[FLAG: yes I can understand this point...]
-% For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step.
-[FLAG: so I moved this sentence below, when we say "Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data"
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
 At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is reasonable or not.
-This is also a step for making sure the model is working as expected, so simulated data often would have low error and high sample sizes (to make sure the model returns the parameter values used to generate the simulated data).
-%jr24april not sure what this final sentence means?
+This is also a step for making sure the model is working as expected (see example worklow).
 
-Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, fig.
-\ref{fig:workflow}).
+Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}).
 This way, we obtain parameter estimates constrained by observations.
-%emwApr19: I commented out the below as I think linking difficulty in fitting the model to problems with the data is pretty tricky (difficulty in fitting the model can point to many things!) and thus too confusing to include here.
-% Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works.
-This could lead us to either simplify our research question or---ideally---launch new data collection efforts.
-These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, fig.
-\ref{fig:workflow}).
-This retrodictive check allows model output to be compared to observations.
-% jregetz16april: do we need to define “retrodictive”?
-% The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
-% jregetz16april: Not sure what this means?
-jdavies did not like it either!
-(he thought it was a typo), vvdm16april: reformulate:
+These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}).
+This---which we call a retrodictive check (\citealp{betanworkflow}, also sometimes called a posterior predictive check, \citealp{Gelman2020})---allows model output to be compared to observations.
 The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.
+
 Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
@@
 
-\begin{figure}[h]
-	\centering
-    \hspace*{-1.5cm}
-	\includegraphics{figures/figure_worflow_wsteps}
-	\caption{Here we could have a caption like ...
-Our workflow builds on others (with citations) to highlight the integrated nature of stuff in ecology.
-We can work on this once we get feedback from mighty co-authors.
-Also, should we list the steps on the figure?
-What do people think?}
-	\label{fig:workflow}
-\end{figure}
-
-% More details, what's special! Feedbacks! %jrApr25 I think this paragraph reads really well!
 A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
-The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
+The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
 Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
-flagging the need to change the model structure---before incorporating real observations. 
+flagging the need to change the model structure---before incorporating empirical observations. 
 The second feedback loop comes from the retrodictive check. 
 Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic.
-We can refine the model to integrate the missing process(es) (if we can identify them) and restart the workflow.
+We can refine the model to integrate the missing process(es) (if we can identify them) and return to the start of the workflow.
 Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g.
@@
 
+\begin{figure}[h]
+	\centering
+	\vspace*{-0.1cm}
+    \hspace*{-1.5cm}
+	\includegraphics{figures/figure_worflow_wsteps}
+	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data.
+Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5).
+Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
+The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.}
+	\vspace*{-0.5cm}
+	\label{fig:workflow}
+\end{figure}
+
 \section{The workflow in practice}
 
-%emw6Apr: I like the shift to 'parameter estimation' instead of trend estimation, but I wonder if we could make that shift clearer above.
-Maybe somewhere we say that the workflow would  encourage a focus on the full model and its parameters: For trend estimates, the trend becomes just one parameter that must be carefully interpreted alongside others .
-Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties through new scientific insights, toward the most critical steps.
+Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties and yield new scientific insights.
 This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.
@@
 
-%jregetz16april: One thing I’m stuck on (and could’ve written this note earlier too).
-Is “trends” really essential?
-Or is this just shorthand for “simple, one/few parameter, non-mechanistic forecasting models”?
-If yes, can we generalize that in the MS overall, or at least say so up front and then say e.g.
-“we’ll refer to these as trend models for brevity”?
-If no, and something about trend estimates is specifically key to the paper, can we say more about *that*?
-[FLAG]
-%emwApr19: I think the only key part is that biodiversity trends and trends over time are big in forecasting ...
-as long as we don't lose that I think it would be great to make the suggested change -- Jim, can you see if you can edit it on the next round?
-
 Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends over time, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
-While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced and more rapidly resolved through use of an improved workflow.
+While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be fewer and they would be more rapidly resolved through use of an improved workflow.
 % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.
@@
 
-% expand on 'highlight major missing components in these models'? ADD example?
-% Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models and in turn debate them.. 
-% Mention elephants rebounding...
-
-% Trends - improvements!
 Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates.
-Returning to the example of a global estimate of trends in vertebrate populations of species over time and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
-% jdavies15april: This is what the LPI wants to be (like a index fund in the stock market tracking trends over time) , but this is often overlooked.
+Returning to the example of a global estimate of trends in vertebrate populations of species over time (see Box) and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
 Or a need to also find which species are declining most, including those that may have poor or no data?
 From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data.
-Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends.
+Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends (see workflow example).
 For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled.
@@
 Our second case-study focuses on process-based modeling, which is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond.
-Newer models generally incorporate greater complexity and an ever-growing number of parameters, making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.
+Newer models, however, generally incorporate greater complexity (and an ever-growing number of parameters), which can make it difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.
 
@@
 These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
-Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
-\todo[size=\tiny]{jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
-Reply: we agree but not really relevant to PBM?
-Maybe it's not clear enough we are focusing on PBM in this section, which we tried to change above, see if it works}
-If a model becomes too complex, it may become a black box, % mention black box here to line up with the beginning of next paragraph
-and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
+Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. 
+If a model becomes too complex, it may become a black box,
+and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
 Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
-\todo[size=\tiny]{jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty ....
-Reply: Same as just above.}
 
@@
 Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only one of the many different output variables of the model.
-%The model building and calibration steps are usually limited to the Methods section, and estimated parameter values are rarely discussed \citep[e.g.][]{Conradi2024, VanderMeersch2025a}.
-The complexity of these models often makes it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability---such work is generally done as a separate effort, exposing potential model problems later on \citep{VanderMeersch2025b}, and not required as a preliminary step.
-\todo[inline, size=\tiny]{For above, addressing: jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties?
-This could help illustrate potential degeneracies etc.
-Reply: Tried something, but not sure it works so feedback much appreciated!}
+The complexity of these models can make it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability---such work is generally done as a separate effort, exposing potential model problems later on \citep{VanderMeersch2025b}, and not required as a preliminary step.
 
-%emw19Apr -- Can we just get away with using non-identifiable and save the reader also needing to digest degenerate?
-(https://betanalpha.github.io/assets/case_studies/identifiability.html)?
-I tried that on this draft, see how it works!
 Applying the workflow to process-based models would help open the black box.
 Each successive step of model development in the workflow may highlight current problems and a path to solutions.
-Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential parameter identifiability issues in the model design.
-%jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below)
+Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two, and help expose potential parameter identifiability issues in the model design.
 Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
 The workflow may also highlight model problems by requiring more explicit model calibration (i.e.
-data fitting), % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
-which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model.
+data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model.
 Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
@@
 % Trends - outline current problem
-
-In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data.
-The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results.
+In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data.
+The Living Planet Index (LPI, \url{www.livingplanetindex.org}), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results.
 % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave.
@@
 This would be massive!}
-With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
-For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}.
-% \todo[]{davies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations.
-We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends.
-For this we need to predict outside of the set}
+With updated data released semi-annually alongside new datasets and new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
+For examples using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.
 
@@
 Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions.
-Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque from an external perspective.
+Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque for anyone who has not worked the model itself.
 The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred.
-Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes.
-Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce some observations or theory.
-The way models are currently calibrated is not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
-% jdavies15april: is this more specific to process based  models?
-[FLAG: hmm yes again?
-Same point as above, maybe it's not clear we're talking mostly about PBM?]
+Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see figure B above).
+Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory.
+The way models are currently calibrated is likely not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
 
@@
 Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
-It would be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimations.
+It would be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimates.
 For both trend estimation and forecasting, the workflow would refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model.
@@
 
+\clearpage
 \section{Barriers and opportunities}
@@
 We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
-% One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed.
-This may be especially true for those who see science rewarded mostly through the shear number of publications.
-But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help.
-% jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
 The first is pressure to publish quickly, which can be at odds with the reality that good model fitting is inherently iterative and takes time.
 The second is reluctance to embrace adoption of open science practices that ensure modeling efforts are fully transparent and reproducible.
-It is imperative that we continue to place increasing value on research that is carefully developed, openly collaborative (including both data and code), and transparent about areas of uncertainty.
-Especially in ecology where samples sizes are low and effects are likely non-linear and complex, we believe adopting a workflow such as the one we propose here will pay off in the longer term, both for individual researchers focused on making enduring contributions, and for the community as a whole focused on improving future data collection and model development (see Fig).
+We believe adopting a workflow such as the one we propose may change this reluctance, and aligns with ongoing efforts to place increasing value on research that is carefully developed, openly collaborative (including both data and code), and transparent about areas of uncertainty.
 
@@
 Much of ecology still divides the world into training for those who gather data and learn a limited set of pre-built models versus those who develop more complex models.
-In ecological training today, researchers who conduct field and lab studies often learn a highly limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
+In ecological training today, researchers who conduct field and lab studies often learn a limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi-squared test).
 
@@
 These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models.
-% process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms
 Neither group has fully integrated data simulation into their statistical or scientific workflows.
@@
 
-Instead of siloing ecological training into experimental design and tests that may match certain designs (though rarely do for ecological data, CITES) versus model design and development, we suggest unified training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them.
-This would mean training to link the ecological processes they study with the mathematical models that may describe them.
+We suggest unified training in our proposed workflow, or a similar one \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, would focus on learning to generate questions and then models, and then how to simulate data from them.
 Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
-Empiricists % change suggested by JD, was empirical ecologists before...
- would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models.
-Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting.
-% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
-Those focused on learning complex statistical models may find many others in the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.
+Empiricists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models.
+Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may see new bridges from their work to empirical data and forecasting.
 
@@
 
-In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made.
-Machine learning will likely surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
-Similarly, engaging in statistical trend analysis, especially when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
-% jdavies15april: nothing intrinsically bad about ML, but sometimes the goal is just to understand the system, not make predictions or forecasts.
+With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important.
+Machine learning will likely surpass process-based models for forecasting accuracy if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
+Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.
 
-Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
-Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
+Beyond improving model building and evaluation, our proposed workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
+Process-based models could once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
 Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty.
-Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.
-It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.
+Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification---preventing modelers from making biased inferences and unfounded assumptions beyond what the data can support.
+It would thus define a clear and limited context in which the model should apply, and limit discussion of adding increasing complexity.
 
-% jdavies15april: This could be the Concluding section highlighting the general advances provided by adopting your framework (and perhaps a better place for text currently at end of section 3.2).
+\emph{Conclusions:} 
 Today model development in ecology is rarely transparent, which limits how easily the research community can understand models, and thus identify potential issues.
-Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations.
+Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate has become mired in methodological considerations.
 However, we believe our workflow provides a tractable step to fixing this.
@@
 
-This shift would present a significant challenge---as it would likely reveal many issues related to non-identifiability in models and data limitations before achieving robust inference.
+% This shift would present a significant challenge---as it would likely reveal many issues related to non-identifiability in models and data limitations before achieving robust inference.
 But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.

### 2025-05-01 analyses/lpiworkflow.qmd
small edits to example workflow + add one sentence to manuscript

@@
 title: "A sample workflow"
-authors: "Victor, Lizzie"
+author:   
+  - Victor
+  - Jonathan
+  - Lizzie
 date: "April 2025"
@@
 
-library(rstan)
+suppressPackageStartupMessages(library(rstan))
 library(ggplot2)
@@
 
-Now we check additional parameters. We find the model does a fair job at recovering population-level parameters (aka hyperparameters).
+Now we check additional parameters. We find the model does a fair job at recovering population-level parameters.
 
@@
 
-We may want to check what the overall and species trends appear to be:
-
-```{r}
-#| echo: false
-#| fig-height: 4
-#| fig-width: 5
-
-species_trends <- summ_fit[paste0("mu_beta_sp[",1:data$Nsp,"]"), c('X50.', 'X2.5.', 'X97.5.')]
-species_trends$spid_num <- 1:data$Nsp
-species_trends <- merge(species_trends, spids)
-
-ggplot() +
-  geom_pointrange(data = summ_fit["mu_beta1", c('X50.', 'X2.5.', 'X97.5.')],
-                  aes(xmin = X2.5., xmax = X97.5., x = X50., y = "Z"), size = 0.25) +
-  geom_text(data = summ_fit["mu_beta1", c('X50.', 'X2.5.', 'X97.5.')],
-                  aes(x = -0.045, y = "Z", label = "Overall"), size = 3) +
-  geom_pointrange(data = species_trends,
-                  aes(xmin = X2.5., xmax = X97.5., x = X50., y = genus_species, color = genus_species), 
-                  size = 0.25) +
-  scale_color_manual(values = hcl.colors(14, palette = "Dark2"),
-                     breaks = unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]) +
-  lims(x = c(-0.1,0.1)) +
-  labs(x = "Trends", color = "Species") +
-  theme_bw() +
-  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
-        panel.grid.minor.y = element_blank(), panel.grid.major.y = element_blank())
-
-```
-
-To see if that might be a good estimate, we need to check how the model predicts the data on finer scales.
+To evaluate our model more precisely, we need to check how the model predicts the data on finer scales.
 So we start by examining some species separately, e.g.
@@
 
+## Inference
+
+Once we are confident with our model (which is not truly the case here), we can look at the parameter values and check what the overall and species trends appear to be:
+
+```{r}
+#| echo: false
+#| fig-height: 4
+#| fig-width: 5
+
+species_trends <- summ_fit[paste0("mu_beta_sp[",1:data$Nsp,"]"), c('X50.', 'X2.5.', 'X97.5.')]
+species_trends$spid_num <- 1:data$Nsp
+species_trends <- merge(species_trends, spids)
+
+ggplot() +
+  geom_pointrange(data = summ_fit["mu_beta1", c('X50.', 'X2.5.', 'X97.5.')],
+                  aes(xmin = X2.5., xmax = X97.5., x = X50., y = "Z"), size = 0.25) +
+  geom_text(data = summ_fit["mu_beta1", c('X50.', 'X2.5.', 'X97.5.')],
+                  aes(x = -0.045, y = "Z", label = "Overall"), size = 3) +
+  geom_pointrange(data = species_trends,
+                  aes(xmin = X2.5., xmax = X97.5., x = X50., y = genus_species, color = genus_species), 
+                  size = 0.25) +
+  scale_color_manual(values = hcl.colors(14, palette = "Dark2"),
+                     breaks = unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]) +
+  scale_y_discrete(limits=rev) +
+  lims(x = c(-0.1,0.1)) +
+  labs(x = "Trends", color = "Species") +
+  theme_bw() +
+  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        panel.grid.minor.y = element_blank(), panel.grid.major.y = element_blank())
+
+```
+
 ## Feedbacks

### 2025-05-01 forecastflows.tex
small edits to example workflow + add one sentence to manuscript
tweak one sentence
small change to fig2

@@
 
-\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
+\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
 \date{\today}
@@
 First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data.
+Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process.
+We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process.
+We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
+
+Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models.
+Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions.
+New research often suggests that models accounting for these complexities can yield more accurate trends and predictions.
+We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons.
+First, the research community is split between two disconnected paradigms: using data to fit simple, trend-like models with few parameters, or developing forecasting models that include complex, mechanistic submodels only indirectly informed by data.
 Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process.
@@
 This---which we call a retrodictive check (\citealp{betanworkflow}, also sometimes called a posterior predictive check, \citealp{Gelman2020})---allows model output to be compared to observations.
+It's only once all steps have been completed that we can interpret parameter values with respect to our research question.
 The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.
@@
 The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to more clearly develop a research question before extensive model fitting.
+
+\begin{figure}[h]
+	\centering
+	\vspace*{-0.1cm}
+    \hspace*{-1.5cm}
+	\includegraphics{figures/figure_worflow_wsteps}
+	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data.
+Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5).
+Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
+The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.}
+	\vspace*{-0.5cm}
+	\label{fig:workflow}
+\end{figure}
 
@@
 At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.
-
-\begin{figure}[h]
-	\centering
-	\vspace*{-0.1cm}
-    \hspace*{-1.5cm}
-	\includegraphics{figures/figure_worflow_wsteps}
-	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data.
-Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5).
-Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
-The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.}
-	\vspace*{-0.5cm}
-	\label{fig:workflow}
-\end{figure}
 

### 2025-05-01 forecastflows.tex
add line numbers

@@
 \usepackage{hyperref}
+
+\usepackage{lineno}
+\renewcommand\linenumberfont{\normalfont\tiny\color{gray}}
 
@@
 \begin{abstract}
-Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts.
+% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts.
 Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions.
@@
 We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
-
 Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models.
@@
 % {\noindent \bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
+\linenumbers
 

### 2025-05-02 analyses/lpiworkflow.qmd
add a plot with some replications

@@
 
-## Step 1: Research question
+# Step 1: Research question
 
@@
 
-## Step 2: Model building
+# Step 2: Model building
 
@@
 
-## Step 3: Model evaluation
+# Step 3: Model evaluation
 
@@
 
-### Stepping backward...
+## Stepping backward...
 
@@
 
-lpi_subset <-subset(lpi_data, Region == "Africa" & Class == "Mammalia" & Included.in.LPR2024 == 1 )
+lpi_subset <-subset(lpi_data, Region == "Africa" & Class == "Mammalia" & 
+                      Included.in.LPR2024 == 1 )
 
@@
 
-### Simulating data
+## Simulating data
 
@@
 ```{r sim data}
+#| fig-height: 4
+#| fig-width: 7
 
@@
 sigma_alpha2 <- log(50)/2.57
-sig_species_alpha <- abs(rnorm(n = nspecies, mean = mu_alpha2, sd = sigma_alpha2))
-alpha_pop_sp <- rnorm(n = npops, mean = mu_alpha_sp[spid_perpop], sd = sig_species_alpha[spid_perpop])
+sig_species_alpha <- abs(rnorm(n = nspecies, mean = mu_alpha2, 
+                               sd = sigma_alpha2))
+alpha_pop_sp <- rnorm(n = npops, mean = mu_alpha_sp[spid_perpop], 
+                      sd = sig_species_alpha[spid_perpop])
 
@@
 sigma_beta2 <- 0.05/2.57
-sig_species_beta <- abs(rnorm(n = nspecies, mean = mu_beta2, sd = sigma_beta2))
-beta_pop_sp <- rnorm(n = npops, mean = mu_beta_sp[spid_perpop], sd = sig_species_beta[spid_perpop])
+sig_species_beta <- abs(rnorm(n = nspecies, mean = mu_beta2, 
+                              sd = sigma_beta2))
+beta_pop_sp <- rnorm(n = npops, mean = mu_beta_sp[spid_perpop], 
+                     sd = sig_species_beta[spid_perpop])
 
@@
 
-### Fit and check model on simulated data!
+## Fit and check model on simulated data!
 
@@
 ```
-
-#### Checking model fitted on simulated data
 
@@
 
-## Step 4: Fit model to empirical data
+# Step 4: Fit model to empirical data
 
@@
 
-lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Country", "Units", paste0("X", 1950:2020))]
+lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Country", "Units", 
+                             paste0("X", 1950:2020))]
 lpi_subset$genus_species <- paste(lpi_subset$Genus, lpi_subset$Species)
@@
   reshape(lpi_subset, direction = "long", idvar = "ID",
-          varying =  paste0("X", 1950:2020), v.names = c("value"), timevar = "year", times = 1950:2020)
+          varying =  paste0("X", 1950:2020), v.names = c("value"), 
+          timevar = "year", times = 1950:2020)
 lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != "NULL",]
@@
 #| layout-nrow: 2
+#| fig-height: 4
+#| fig-width: 7
 
-units_to_keep <- c("Estimate - entire population", "Estimate - entire population (August)", "Individual", "individuals", "Individuals", "Number of adults", "number of individuals", "Number of individuals", "Number of Individuals", "Number of rhinos", "numbers of indviduals", "population estimate", "Population estimate", "Population estimate (individuals)", "Population estimates", "Total population size")
+units_to_keep <- c("Estimate - entire population", "Estimate - entire population (August)", 
+                   "Individual", "individuals", "Individuals", "Number of adults", 
+                   "number of individuals", "Number of individuals", "Number of Individuals", 
+                   "Number of rhinos", "numbers of indviduals", "population estimate", 
+                   "Population estimate", "Population estimate (individuals)", 
+                   "Population estimates", "Total population size")
 
@@
 #| eval: true
+#| fig-height: 4
+#| fig-width: 7
 
@@
 
-lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,] # We now also find we have counts of 0, which is tricky for our approach (as we cannot evaluate log(0)), we would also flag this to return.
-As we iteratively improve the model, we would need to consider what processes generate these zeroes, and include them in our model.
+# We now also find we have counts of 0, which is tricky for our approach 
+#(as we cannot evaluate log(0)), we would also flag this to return. 
+# As we iteratively improve the model, we would need to consider what 
+# processes generate these zeroes, and include them in our model.
+lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,]
 
-popids <- data.frame(ID = unique(lpi_subset_rsh$ID), popid_num = 1:length(unique(lpi_subset_rsh$ID)))
-spids <- data.frame(genus_species = unique(lpi_subset_rsh$genus_species), spid_num = 1:length(unique(lpi_subset_rsh$genus_species)))
+popids <- data.frame(ID = unique(lpi_subset_rsh$ID), 
+                     popid_num = 1:length(unique(lpi_subset_rsh$ID)))
+spids <- data.frame(genus_species = unique(lpi_subset_rsh$genus_species), 
+                    spid_num = 1:length(unique(lpi_subset_rsh$genus_species)))
 
@@
 
-## Step 5: Retrodictive and related checks
+# Step 5: Retrodictive and related checks
 
@@
 ```{r}
-#| fig-height: 5
-#| fig-width: 5
+#| fig-height: 3
+#| fig-width: 3
 #| echo: false
@@
   labs(y = 'Density', x = 'Counts (log-scale)')
+
+```
+
+We can also look at some posterior replications separately:
+
+```{r}
+#| fig-height: 5
+#| fig-width: 5
+#| echo: false
+
+n <- sample(1:4000, 9)
+retrodictions <- (extract(fit, pars = names))
+retrodictions_sub <- sapply(names, function(x){retrodictions[[x]][n]})
+
+retrodictions_sub_df <- data.frame(ret = as.numeric(t(retrodictions_sub)), chain = rep(n, each = data$N))
+
+ggplot() +
+  facet_wrap(~ chain) +
+  geom_density(data = data.frame(obs = log(data$y)), aes(x = obs), alpha = 0.5, linetype = 'dashed') +
+  geom_density(data = retrodictions_sub_df, aes(x = ret), alpha = 0.5, col = 'darkred') +
+  theme_bw() +
+  labs(y = 'Density', x = 'Counts (log-scale)') +
+  theme(strip.text.x.top = element_blank())
 
@@
 
-## Inference
+# Inference
 
@@
 
-## Feedbacks
+# Feedbacks
 
@@
 gompodel <- nls(log(value) ~ SSlogis(year, Asym, b2, b3), data=subset_wildebeest)
-logisfit <-data.frame(pred = predict(gompodel, newdata = data.frame(year = seq(-25,30,1))), year = seq(-25,30,1))
+logisfit <-data.frame(pred = predict(gompodel, 
+                                     newdata = data.frame(year = seq(-25,30,1))), 
+                      year = seq(-25,30,1))
 
@@
 
-## The full process
+# The full process
 

### 2025-07-08 figures/retrodic/phylo_nophylo_whocares.R
brute-force approach, 100 sim

@@ -1,0 +1,157 @@
+rm(list = ls());gc()
+if(length(grep("victor", getwd()) > 0)) {
+  wd <- "/home/victor/projects/forecastflows"
+} else if(length(grep("lizzie", getwd()) > 0)) {
+  wd <- "/Users/lizzie/Documents/git/projects/misc/miscmisc/workflowsPhilTrans"
+}  
+library(ggplot2)
+library(rstan)
+library(phytools)
+library(geiger)
+rstan_options(auto_write = TRUE)
+
+phytree <- readRDS(file.path(wd, 'figures/input', 'phytree.rds')) # Lizzie does not have, right?
+# phytree[["edge.length"]][39:40] <- phytree[["edge.length"]][39:40]*100
+# phytree <- force.ultrametric(phytree)
+plot(phytree)
+
+if(length(grep("lizzie", getwd()) > 0)){
+  require(ape)
+  nspecies = 40
+  phytree <- pbtree(n=nspecies, nsim=1, b=1, complete=FALSE,scale=1)
+}
+
+# ------------- #
+# Simulate data #
+# ------------- #
+
+set.seed(1234567)
+seeds <- round(runif(100, 1, 9999999),0)
+
+for(s in seeds){
+  set.seed(s)
+  
+  nspecies <- phytree[["Nnode"]]+1 # no. of species
+  nobs_perspecies <- round(runif(nspecies, 5,10)) # no. of different observations per population
+  nobs_perspecies[c(14,20)] <- round(runif(2, 180,221)) # some bias
+  # nobs_perspecies[c(1)] <- round(runif(1, 90,120)) # some bias
+  spid_perobs <- rep(1:nspecies, times = nobs_perspecies)
+  
+  years <- c()
+  for(np in nobs_perspecies){
+    years_p <- sample(1800:2020, size = np, replace = FALSE)
+    years <- c(years, years_p)
+  }
+  
+  ## species intercepts, with phylogenetic structure
+  lambda <- 0.7
+  aroot <- 5
+  sigma <- 1e-1
+  scaledtree_int <- rescale(phytree, model = "lambda", lambda)
+  speciesnum <- as.numeric(gsub("sp", "", phytree[["tip.label"]]))
+  plot.phylo(scaledtree_int,
+             cex = 1, 
+             tip.color = c("#FF7601", "#00809D")[as.numeric(as.numeric(gsub("sp", "", phytree[["tip.label"]])) %in% 1:6)+1])
[…]

### 2025-07-08 figures/retrodic/plotme.R

@@ -1,0 +1,31 @@
+
+
+slopes <- data.frame(
+  true_slopes = mu_beta_sp_wphylogeny,
+  slp_fit = summ_fit[paste0("mu_beta_sp[",1:nspecies,"]"), c('X50.')],
+  slp_fit.lcl = summ_fit[paste0("mu_beta_sp[",1:nspecies,"]"), c('X2.5.')],
+  slp_fit.ucl = summ_fit[paste0("mu_beta_sp[",1:nspecies,"]"), c('X97.5.')],
+  species = 1:nspecies,
+  mod = 'No phylo.'
+)
+
+slopesphylo <- data.frame(
+  true_slopes = mu_beta_sp_wphylogeny,
+  slp_fit = summ_fitphylo[paste0("mu_beta_sp[",1:nspecies,"]"), c('X50.')],
+  slp_fit.lcl = summ_fitphylo[paste0("mu_beta_sp[",1:nspecies,"]"), c('X2.5.')],
+  slp_fit.ucl = summ_fitphylo[paste0("mu_beta_sp[",1:nspecies,"]"), c('X97.5.')],
+  species = 1:nspecies,
+  mod = 'Phylo'
+)
+
+
+plotslopes <- ggplot(data = rbind(slopes, slopesphylo)) +
+  facet_wrap(~mod) +
+  geom_pointrange(aes(ymin = slp_fit.lcl, ymax = slp_fit.ucl, x = species, 
+                      y = slp_fit, color = species %in% 1:6),
+                  size = 0.2) +
+  geom_point(aes(x = species, y = true_slopes), shape = 4) +
+  theme_classic() +
+  theme(legend.position = 'none')
+
+ggsave(plotslopes, file = file.path(wd, 'figures','retrodic', 'multisim', paste0(s, '.pdf')))

### 2025-07-09 revision/response_r1.tex
start working on reviewer response

\documentclass[11pt,letter]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{graphicx} % Required for inserting images
\usepackage{xcolor} 
\definecolor{darkgrey}{HTML}{4d4d4d}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{gensymb}
\usepackage{xr-hyper}
\makeatletter
  \long\def\myempty{}
  \def\XR@addURL#1{\XR@@dURL#1\myempty{}{}{}{}{}\\}
  \def\XR@@dURL#1#2#3#4#5#6#7\\{%
    {#1}{#2}%
    \ifx\myempty#6\@empty
      {#3}{#4}{\XR@URL}%
    \else
    \fi
}
\makeatother
\usepackage{hyperref}
\hypersetup{hidelinks,breaklinks=true}
\usepackage{cleveref}
\usepackage[labelfont=bf]{caption}
\bibliographystyle{../amnat}

\RequirePackage[labelfont={bf,sf},%
                font={small, sf}]{caption}

\RequirePackage[labelfont={bf,sf},%
                font={small, sf}]{caption}

\externaldocument{forecastflows_r1}
\newcommand{\lref}[1]{}
% (\href{file:forecastflows_r1\#lintarget:SOMELABEL}{line \ref*{lin:SOMELABEL}})

\DeclareEmphSequence{\itshape\color{darkgrey}}
\usepackage[most]{tcolorbox}
% \newtcolorbox{mybox}[1][]{%
%     blanker, 
%     left skip=0em,
%     left=1em,
%     borderline west={1pt}{2pt}{darkgrey},
%     breakable,
%     #1}
\newtcolorbox{mybox}[1][]{%
    blanker, 
    left skip=0em,
    left=0em,
    borderline west={0pt}{0pt}{white},
    breakable,
    #1}

\begin{document}
\setlength{\parindent}{0cm}
\setlength{\parskip}{7pt}

Editor and reviewer comments (we provide below the full context of the two reviewers' comments) are in \emph{italics}, while our responses are in regular text. \\

{\bf Response to editor's comments:}

\begin{mybox}
\emph{...}
\end{mybox}

We are happy to hear that both reviewers recognized the value of our work and we appreciate the opportunity to revise our manuscript. We have worked to clarify the points raised by the editor and the two reviewers in this new version.

{\bf Response to reviewers' comments:}

{\bf Reviewer 1}

\begin{mybox}
\emph{This opinion manuscript (MS) argues that adapting a modern statistical workflow will help resolve some key issues in forecasting models used in ecology to inform policy about biodiversity loss and climatechanges impacts. Specifically, the authors identity two specific “disconnected paradigms” for making such forecasts and argue that a workflow adopted from the Bayesian literature, and that emphasizes data simulation, would help unify these paradigms and make for a better scientific understanding of these important issues. I found the MS well-written, concise and easy to read, and well-organized with clearly laid out challenges and understandable solutions. I agree with the central proposal that the adoption of a “principled” workflow would be beneficial in many areas of ecology, and these types of studies in particular. I have no experience with these specific types of ecological models, but I do have experience with the Bayesian workflow underlying their key recommendation. So, I will focus my review on the recommendation and not the challenge. Below I outline several major concerns/questions, and below that minor comments and corrections.}
\end{mybox}

Response.

\begin{mybox}
\emph{\textbf{Inadequate justification of simulation.}\\
Simulating data and fitting models is key to the step-by-step workflow but I think the MS lacks sufficient
detail and motivation. The methods described come from the Bayesian literature, where simulating
parameters from priors (prior predictive/pushforward) or posterior (posterior predictive or retrodictive)
and then data, are natural. But a quick peek at the cited literature suggests that Bayesian models are not
used. So how would one adapt them? On L76 they say “... by fixing parameters to some reasonable
range of values (which is straightforward if the parameters are interpretable)” but I disagree with this.
Do you suggest analysts develop priors (using, e.g., Betancourt’s cited workflow case study), and then
simulate parameters from those, but ignore them in the subsequent fitting to empirical data? Or take a
range for each parameter and simulate parameters across N equal steps for each one? Posterior
predictive checks (retrodiction) are known to be conservative due to the double-use of the data (see
Conn et al. 2018 [https://doi.org/10.1002/ecm.1314]). How does that affect your recommendations and
how the analyst should use this tool?}
\end{mybox}

Response.

\begin{mybox}
\emph{Furthermore, I think the authors should include a simple illustration of how these retrodiction checks can flag the need for model changes. For instance, I imagine they could take a fitted model from one of the key cited papers like the LPI trend, and simulate data given spatial/phylogenetic structure, but then refit a model without that and show how the model fails to fit the data and how the analysts would
identify (visually, with a new figure) and resolve that by adding model structure/complexity. I am a big proponent of prior/posterior predictive checking and do not disagree with their recommended usage, rather I think more detail is needed for the MS. I think a judicious, simple example here could really help the reader interpret the recommendations and start thinking about how to apply it in their studies. Given the importance of this step in the overall thesis of the MS I believe more detail on how an analyst would do this would be worthwhile.}
\end{mybox}

Response.

\begin{mybox}
\emph{\textbf{Resolving process-based model issues.}\\
I do not see how the proposed workflow would resolve what appears to a key issue in the process-based
research, namely that individual model components are done separately. Box (B) is supposed to address
this but I am left unconvinced how this would resolve that. As I understand it, their workflow would lead
to improved submodels, but would not change how those were linked together. Perhaps I missed a
point here (admittedly I am unfamiliar with these models), but if not, I encourage the authors to try to
make this point clearer. Otherwise will the workflow “bridge the gap” between the paradigms?}
\end{mybox}

Response.

\begin{mybox}
\emph{\textbf{Causal analysis as a framework to develop research questions and elicit expertise.}\\
The authors repeatedly highlight (rightfully so) that more time needs to be spent developing the research question. E.g., L266 “By focusing on model development more tightly tied to ecological expertise…”. However, they give no advice on how to do that. I pose that the causal analysis literature can and should be considered as a good source to bolster this argument and guide readers, as it provides a robust scientific framework for this step of the workflow. For instance, Grace and Irvine (2020) [https://doi.org/10.1002/ecy.2962] is an accessible introduction to this topic with ecological examples. I suggest the authors either incorporate this literature into their recommendation, or justify why not. As Grace and Irvine argue, it is not really fair to ask readers to develop these hypotheses without giving advice on how to do just that.}
\end{mybox}

Response.

\begin{mybox}
\emph{\textbf{Improvements to Figure 1:}\\
I like this figure but improving the caption would really help the reader take in the material more efficiently. I suggest moving the upper right panel to the upper left, as it serves as a sort of legend of the three other panels and should be read first. I would label this one A as well. What is the significance of the figure-8 in B? How is that different than a circle? This caption does not define or discuss the panels A,B,C, and I think that it should. Why are the terms “pre-data” and “pre-model” only used in the capture and nowhere else in the MS? “Retrodictive checks” may need to be further clarified in the caption.}
\end{mybox}

Response.

\begin{mybox}
\emph{“..where a model is built..”\\
“..some feedback to..”}
\end{mybox}

\begin{mybox}
\emph{\textbf{Minor}\\
Page 1, Line 40 in abstract: what is “unified” about the workflow?}
\end{mybox}

\begin{mybox}
\emph{Page 1, Line 45 in abstract: drop “for”}
\end{mybox}

\begin{mybox}
\emph{L6: “date, alongside”}
\end{mybox}

\begin{mybox}
\emph{L36: These workflow citations are all unpublished (arXiv). I’m fine with that, but consider citing some from other fields that have gone through peer review. The caption of Figure 2 includes more, suggest putting them here too.}
\end{mybox}

\begin{mybox}
\emph{L103: What does “change the model structure” mean here? Adding constraints and priors? Removing or adding complexity?}
\end{mybox}

\begin{mybox}
\emph{L110: delete “then”}
\end{mybox}

\begin{mybox}
\emph{Fig 2: Caption should mention this is expanded from Fig. 1C.}
\end{mybox}

\begin{mybox}
\emph{L116, L168: I disagree that “reduce uncertainties” is the goal – instead it is to more accurately quantify it. See L 257 “properly accounting for their uncertainty” for instance.}
\end{mybox}

\begin{mybox}
\emph{L143: FYI the fisheries literature has used the terms pre-data post-model etc. for decades now and have implemented this simulation approach. E.g., https://journal.iwc.int/index.php/jcrm/article/view/239/17. I don’t think it’s necessary to cite this but worth being aware of. You might want to re-cite Hilborn and Mangel (1997) here as they explicitly call for more simulation testing in ecology.}
\end{mybox}

\begin{mybox}
\emph{L201: “..workflow, non…” This sentence partially answers my question about L103 above}
\end{mybox}

\begin{mybox}
\emph{Box A: Please check the logic of the first sentence. More papers because of less time confronting their models? Is that what you meant to say? “For examples” should be “For example”?}
\end{mybox}

\begin{mybox}
\emph{Box B: Please clarify the last sentence if possible. I’m not sure what “inappropriate way to accommodate their complexity” means here.}
\end{mybox}

\begin{mybox}
\emph{L225: I like this paragraph and it makes a point that I hadn’t seen written out so clearly.}
\end{mybox}

\begin{mybox}
\emph{L245: I feel like section 4.2 is slightly out of place. I’d rather see more text devoted to retrodictive checks than this. Consider dropping it and focusing on just a conclusion here}
\end{mybox}

{\bf Reviewer 2}

\begin{mybox}
\emph{The paper is well written and in a style that is accessible to a wide audience. The topic fits
under the theme of “workflow for applied data analysis” by proposing a workflow that if used
thoughtfully could lead to less divergence in forecasts and therefore more trust in the scientific
process. The proposed workflow is similar to the Bayesian workflow outlined in Gelman et
al. (2020) and is translated to two different approaches to ecological forecasting. I think
the proposed workflow could be applied or modified to be used with a wide range of research
questions within ecological forecasting if careful attention is paid to nuances and idiosyncrasies
associated with each unique research question, and I think this necessary attention to question-
specific nuances could be highlighted more in the paper. There are places where I feel more
specificity (or clarity) with respect to these nuances could improve the manuscript and its
potential for impact with respect to building trust in the scientific process. There is also a
lack of specificity in how, exactly, this type of workflow could be implemented because there
is no example application to data. I have outlined these areas under “Major comments” and
provided minor editorial comments in “Minor Comments.”}
\end{mybox}

% Note: I am an applied statistician with experience collaborating with ecologists and developing
% statistical methodology for large-scale long-term monitoring programs. The bulk of my work
% is more concerned with explanatory than predictive inference and my knowledge of building
% ecological forecasting models for inference on a global scale is limited. I reviewed the paper
% through this lens.

\begin{mybox}
\emph{\textbf{Major comments}\\
1. Clarity on inferential goals of “disconnected paradigms”\\
 When the paper discusses “harmonizing both trend estimation and forecasting” what is meant by this? The
title of the paper suggests the inferential goals from both approaches to be predictive
in nature (i.e., “forecasting trends”), but it seems plausible that the simpler models are
developed for more of an explanatory purpose (e.g., investigating relationships between
covariates and mean response) rather than predicting or forecasting future trends, which
would make the questions being asked by each modeling approach fundamentally dif-
ferent (see, Shmueli 2010). The paper could be improved if more clarity was provided
on the inferential goals of both types of modeling paradigms it discusses, and perhaps
provide some discussion why they need to be (or if they should be) harmonized if they
are in pursuit of different types of inference.}
\end{mybox}

\begin{mybox}
\emph{2. Can this workflow be applied to non-Bayesian models\\
The emphasis on simulation-based investigation of model components is highlighted as a key feature of
the workflow, so it seems a generative (or at least partially generative) model is assumed. Does this model need to be Bayesian? Explicit connections to posterior predictive checks and more discussion for how both prior- and posterior-predictive assessment could be used in a non-Bayesian context would strengthen this work. (Also note that prior predictive checks should be made explicit too – see minor comments).}
\end{mybox}

\begin{mybox}
\emph{3. Discussion of Calibration:\\
Lines 98-104 seem to be referring to a check on estimating
“the model” using synthetic data that are generated from the model (either through
optimization or McMC), but there is no mention of various computational approaches
that may be taken to fit “the model.” How does the chosen computational strategy fit
into the workflow if at all? Further, is the proposed workflow to conduct a full-blown
simulation calibration study (i.e., simulate-fit-check for capture) many times and across
all possible parameter values? Assuming yes, I’d be interested in how something like
that would scale for the types of models that are the focus of the paper? Could the
authors provide more guidance on how to conduct such a study in this context?}
\end{mybox}

\begin{mybox}
\emph{4. Gaps in the workflow\\
A big argument in the paper is that the proposed workflow
could lessen discrepancies in results between studies of trends (lines 123-128). However,
in the workflow there is no mention of transparent communication of study design, which
if ignored in analysis, could result in discrepancies in results for studies with similar
research questions. Differences in sampling designs must also be considered prior to
integrating different data sources (if integration is even possible). The importance of
study design/sampling design and its ties to appropriate modeling is not discussed. Is it
implicitly assumed that there is alignment between design, question, and model in this
workflow? Some mention of the importance of study design would strengthen the work.}
\end{mybox}

\begin{mybox}
\emph{Additionally, the choice of prior and posterior predictive checks (PPC) is non-trivial and
depends on the context of the study. There is discussion of these types of simulation-
based investigations, but no mention of the importance of developing checks that assess
key features of the observed data that are tied to the research question. Acknowledging
the complexity of this step, and including some ideas for how to do this in the context
of the types of problems discussed in the paper could strengthen the paper.}
\end{mybox}

\begin{mybox}
\emph{5. Theoretical examples provided, but no data application of the proposed workflow\\
Sections 3.1 and 3.2 and the Box outline theoretical examples of how the workflow
could be applied for trend estimation and forecasting in the context of types of data, but
there are no concrete data examples. A comment on how the proposed workflow could
be applied to real data and how it would scale with complexity of data and research
questions would be a good addition to the paper if it is not feasible to include real-data
examples with the general discussion.}
\end{mybox}

\begin{mybox}
\emph{6. Adopting the workflow\\
ecommendations are made to train ecologists more quantitatively, but there is no suggestion or encouragement of collaboration with statisticians. These types of workflows are generally part of statistical practice, so could an alternative to more quantitative training (which would have to come at the cost of some other aspect of training) be more cross-disciplinary collaboration with statisticians?}
\end{mybox}

\begin{mybox}
\emph{\textbf{Minor comments}\\
Abstract, line 45: remove “for” in, “could transform for ecological modeling.”}
\end{mybox}

\begin{mybox}
\emph{Line 52: change “in term” to “in terms”}
\end{mybox}

\begin{mybox}
\emph{Figure 1 caption, line 34: change “where a model is build” to “where a model is built”}
\end{mybox}

\begin{mybox}
\emph{Lines 75-81, the paper discusses a form of prior predictive checking without explicitly mentioning the connection. The connection should be mentioned and perhaps}
\end{mybox}

\begin{mybox}
\emph{Line 80: typo in “workflow” in “(see example worklow)”}
\end{mybox}

\begin{mybox}
\emph{Line 86: change “also sometimes called a posterior predictive check” to “in a Bayesian workflow this is called a posterior predictive check.”}
\end{mybox}

\begin{mybox}
\emph{Line 87: could be more specific about “model output” as typically a PPC investigates how the entire posterior predictive distribution for the response or some summary statistic that is a function of the response and parameters compares to the observed data (or observed summary stat computed from observed data).}
\end{mybox}

\begin{mybox}
\emph{Line 117: should “hypothesis” be “hypotheses”?}
\end{mybox}

\begin{mybox}
\emph{Lines 198-201: if the goal is accurate and relatively-precise forecasts, do all of the nuisance
parameters need to be identifiable for a model to be useful? That is, does identifiability
of all parameters matter for process-based models with a goal of forecasting?}
\end{mybox}

\bibliography{forecastflows}

\end{document}

### 2025-07-10 revision/response_r1.tex
I was hoping to have more time for this today

@@
 \begin{mybox}
-\emph{This opinion manuscript (MS) argues that adapting a modern statistical workflow will help resolve some key issues in forecasting models used in ecology to inform policy about biodiversity loss and climatechanges impacts.
+\emph{This opinion manuscript (MS) argues that adapting a modern statistical workflow will help resolve some key issues in forecasting models used in ecology to inform policy about biodiversity loss and climate changes impacts.
 Specifically, the authors identity two specific “disconnected paradigms” for making such forecasts and argue that a workflow adopted from the Bayesian literature, and that emphasizes data simulation, would help unify these paradigms and make for a better scientific understanding of these important issues.
@@
 \emph{\textbf{Inadequate justification of simulation.}\\
-Simulating data and fitting models is key to the step-by-step workflow but I think the MS lacks sufficient
-detail and motivation. The methods described come from the Bayesian literature, where simulating
-parameters from priors (prior predictive/pushforward) or posterior (posterior predictive or retrodictive)
-and then data, are natural. But a quick peek at the cited literature suggests that Bayesian models are not
-used. So how would one adapt them? On L76 they say “... by fixing parameters to some reasonable
-range of values (which is straightforward if the parameters are interpretable)” but I disagree with this.
-Do you suggest analysts develop priors (using, e.g., Betancourt’s cited workflow case study), and then
-simulate parameters from those, but ignore them in the subsequent fitting to empirical data? Or take a
-range for each parameter and simulate parameters across N equal steps for each one? Posterior
-predictive checks (retrodiction) are known to be conservative due to the double-use of the data (see
-Conn et al. 2018 [https://doi.org/10.1002/ecm.1314]). How does that affect your recommendations and
-how the analyst should use this tool?}
+Simulating data and fitting models is key to the step-by-step workflow but I think the MS lacks sufficient detail and motivation.
+The methods described come from the Bayesian literature, where simulating parameters from priors (prior predictive/pushforward) or posterior (posterior predictive or retrodictive) and then data, are natural.
+But a quick peek at the cited literature suggests that Bayesian models are not used.
+So how would one adapt them?}
+\end{mybox}
+
+While Bayesian models are more and more frequently (we cite for example ), we clarified that---despite some Bayesian-specific terms---data simulation apply for all (\href{file:forecastflows_r1\#lintarget:datasimforall}{line \ref*{lin:datasimforall}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}).
+
+% Below, AG says not relevant
+\begin{mybox}
+\emph{On L76 they say “...
+by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable)” but I disagree with this.
+Do you suggest analysts develop priors (using, e.g., Betancourt’s cited workflow case study), and then simulate parameters from those, but ignore them in the subsequent fitting to empirical data?
+Or take a range for each parameter and simulate parameters across N equal steps for each one?
+Posterior predictive checks (retrodiction) are known to be conservative due to the double-use of the data (see Conn et al.
+2018 [https://doi.org/10.1002/ecm.1314]).
+How does that affect your recommendations and how the analyst should use this tool?}
 \end{mybox}
@@
 \emph{Furthermore, I think the authors should include a simple illustration of how these retrodiction checks can flag the need for model changes.
-For instance, I imagine they could take a fitted model from one of the key cited papers like the LPI trend, and simulate data given spatial/phylogenetic structure, but then refit a model without that and show how the model fails to fit the data and how the analysts would
-identify (visually, with a new figure) and resolve that by adding model structure/complexity.
+For instance, I imagine they could take a fitted model from one of the key cited papers like the LPI trend, and simulate data given spatial/phylogenetic structure, but then refit a model without that and show how the model fails to fit the data and how the analysts would identify (visually, with a new figure) and resolve that by adding model structure/complexity.
 I am a big proponent of prior/posterior predictive checking and do not disagree with their recommended usage, rather I think more detail is needed for the MS.
@@
 \emph{\textbf{Resolving process-based model issues.}\\
-I do not see how the proposed workflow would resolve what appears to a key issue in the process-based
-research, namely that individual model components are done separately. Box (B) is supposed to address
-this but I am left unconvinced how this would resolve that. As I understand it, their workflow would lead
-to improved submodels, but would not change how those were linked together. Perhaps I missed a
-point here (admittedly I am unfamiliar with these models), but if not, I encourage the authors to try to
-make this point clearer. Otherwise will the workflow “bridge the gap” between the paradigms?}
+I do not see how the proposed workflow would resolve what appears to a key issue in the process-based research, namely that individual model components are done separately.
+Box (B) is supposed to address this but I am left unconvinced how this would resolve that.
+As I understand it, their workflow would lead to improved submodels, but would not change how those were linked together.
+Perhaps I missed a point here (admittedly I am unfamiliar with these models), but if not, I encourage the authors to try to make this point clearer.
+Otherwise will the workflow “bridge the gap” between the paradigms?}
 \end{mybox}
 
-Response.
+Sorry it was not clear.
+Applying the workflow to a process-based model would imply applying it to the model as a whole.
+This would make researchers calibrate the full model jointly, rather than treating submodels separately.
+We clarify this in Box (C).
 
@@
 
-Response.
+Thank you for this reference.
+We added it and now mention that the conceptualization of explanatory hypothesis has gaining attention in ecology (\href{file:forecastflows_r1\#lintarget:causalinf}{line \ref*{lin:causalinf}}).
 
@@
 \begin{mybox}
-\emph{The paper is well written and in a style that is accessible to a wide audience. The topic fits
-under the theme of “workflow for applied data analysis” by proposing a workflow that if used
-thoughtfully could lead to less divergence in forecasts and therefore more trust in the scientific
-process. The proposed workflow is similar to the Bayesian workflow outlined in Gelman et
-al. (2020) and is translated to two different approaches to ecological forecasting. I think
-the proposed workflow could be applied or modified to be used with a wide range of research
-questions within ecological forecasting if careful attention is paid to nuances and idiosyncrasies
-associated with each unique research question, and I think this necessary attention to question-
-specific nuances could be highlighted more in the paper. There are places where I feel more
-specificity (or clarity) with respect to these nuances could improve the manuscript and its
-potential for impact with respect to building trust in the scientific process. There is also a
-lack of specificity in how, exactly, this type of workflow could be implemented because there
-is no example application to data. I have outlined these areas under “Major comments” and
-provided minor editorial comments in “Minor Comments.”}
+\emph{The paper is well written and in a style that is accessible to a wide audience.
+The topic fits under the theme of “workflow for applied data analysis” by proposing a workflow that if used thoughtfully could lead to less divergence in forecasts and therefore more trust in the scientific process.
+The proposed workflow is similar to the Bayesian workflow outlined in Gelman et al.
+(2020) and is translated to two different approaches to ecological forecasting.
+I think the proposed workflow could be applied or modified to be used with a wide range of research questions within ecological forecasting if careful attention is paid to nuances and idiosyncrasies associated with each unique research question, and I think this necessary attention to question-specific nuances could be highlighted more in the paper.
+There are places where I feel more specificity (or clarity) with respect to these nuances could improve the manuscript and its potential for impact with respect to building trust in the scientific process.
+There is also a lack of specificity in how, exactly, this type of workflow could be implemented because there
+is no example application to data. I have outlined these areas under “Major comments” and provided minor editorial comments in “Minor Comments.”}
 \end{mybox}
@@
 \emph{3. Discussion of Calibration:\\
-Lines 98-104 seem to be referring to a check on estimating
-“the model” using synthetic data that are generated from the model (either through
-optimization or McMC), but there is no mention of various computational approaches
-that may be taken to fit “the model.” How does the chosen computational strategy fit
-into the workflow if at all? Further, is the proposed workflow to conduct a full-blown
-simulation calibration study (i.e., simulate-fit-check for capture) many times and across
-all possible parameter values? Assuming yes, I’d be interested in how something like
-that would scale for the types of models that are the focus of the paper? Could the
-authors provide more guidance on how to conduct such a study in this context?}
+Lines 98-104 seem to be referring to a check on estimating “the model” using synthetic data that are generated from the model (either through optimization or McMC), but there is no mention of various computational approaches that may be taken to fit “the model.” How does the chosen computational strategy fit into the workflow if at all?
+Further, is the proposed workflow to conduct a full-blown simulation calibration study (i.e., simulate-fit-check for capture) many times and across all possible parameter values?
+Assuming yes, I’d be interested in how something like that would scale for the types of models that are the focus of the paper?
+Could the authors provide more guidance on how to conduct such a study in this context?}
 \end{mybox}
@@
 A big argument in the paper is that the proposed workflow
-could lessen discrepancies in results between studies of trends (lines 123-128). However,
-in the workflow there is no mention of transparent communication of study design, which
-if ignored in analysis, could result in discrepancies in results for studies with similar
-research questions. Differences in sampling designs must also be considered prior to
-integrating different data sources (if integration is even possible). The importance of
-study design/sampling design and its ties to appropriate modeling is not discussed. Is it
-implicitly assumed that there is alignment between design, question, and model in this
-workflow? Some mention of the importance of study design would strengthen the work.}
+could lessen discrepancies in results between studies of trends (lines 123-128).
+However, in the workflow there is no mention of transparent communication of study design, which if ignored in analysis, could result in discrepancies in results for studies with similar research questions.
+Differences in sampling designs must also be considered prior tointegrating different data sources (if integration is even possible).
+The importance of study design/sampling design and its ties to appropriate modeling is not discussed.
+Is itimplicitly assumed that there is alignment between design, question, and model in this workflow?
+Some mention of the importance of study design would strengthen the work.}
 \end{mybox}
@@
 \begin{mybox}
-\emph{Additionally, the choice of prior and posterior predictive checks (PPC) is non-trivial and
-depends on the context of the study. There is discussion of these types of simulation-
-based investigations, but no mention of the importance of developing checks that assess
-key features of the observed data that are tied to the research question. Acknowledging
-the complexity of this step, and including some ideas for how to do this in the context
-of the types of problems discussed in the paper could strengthen the paper.}
+\emph{Additionally, the choice of prior and posterior predictive checks (PPC) is non-trivial and depends on the context of the study.
+There is discussion of these types of simulation-based investigations, but no mention of the importance of developing checks that assess key features of the observed data that are tied to the research question.
+Acknowledging the complexity of this step, and including some ideas for how to do this in the context of the types of problems discussed in the paper could strengthen the paper.}
 \end{mybox}
@@
 \emph{5. Theoretical examples provided, but no data application of the proposed workflow\\
-Sections 3.1 and 3.2 and the Box outline theoretical examples of how the workflow
-could be applied for trend estimation and forecasting in the context of types of data, but
-there are no concrete data examples. A comment on how the proposed workflow could
-be applied to real data and how it would scale with complexity of data and research
-questions would be a good addition to the paper if it is not feasible to include real-data
-examples with the general discussion.}
+Sections 3.1 and 3.2 and the Box outline theoretical examples of how the workflow could be applied for trend estimation and forecasting in the context of types of data, but there are no concrete data examples.
+A comment on how the proposed workflow could be applied to real data and how it would scale with complexity of data and research questions would be a good addition to the paper if it is not feasible to include real-data examples with the general discussion.}
 \end{mybox}
@@
 \emph{6. Adopting the workflow\\
-ecommendations are made to train ecologists more quantitatively, but there is no suggestion or encouragement of collaboration with statisticians.
+Recommendations are made to train ecologists more quantitatively, but there is no suggestion or encouragement of collaboration with statisticians.
 These types of workflows are generally part of statistical practice, so could an alternative to more quantitative training (which would have to come at the cost of some other aspect of training) be more cross-disciplinary collaboration with statisticians?}

### 2025-07-10 revision/forecastflows_r1.tex
start working on reviewer response
I was hoping to have more time for this today

--- 2025-05-01 forecastflows.tex
@@
 \usepackage{hyperref}
+\newcommand{\llabel}[1]{\hypertarget{lintarget:#1}{}\linelabel{lin:#1}}
 
@@
 
-\noindent \emph{For:} Scientific and Statistical Workflow theme issue for \emph{Phil Trans A} as an \emph{Opinion}
+\noindent \emph{For:} Workflow for Applied Data Analysis theme issue for \emph{Phil Trans A} as an \emph{Opinion}
 % They mention a tex template here: https://royalsocietypublishing.org/rsta/for-authors#question4
@@
 	\centering
-	\includegraphics{figures/modeldataspaces_rotation}
+	\includegraphics{../figures/modeldataspaces_rotation}
 	\caption{Different trajectories through model-data space during model development.
@@
 This involves making clear assumptions about the most influential drivers, within the specific context of our study.
-This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error.
+This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error.
+\llabel{causalinf}This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}.
 From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}).
@@
 With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}).
-`Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data.
+`Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable values (which is straightforward if the parameters are interpretable) and from fake predictor data.
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
@@
 This is also a step for making sure the model is working as expected (see example worklow).
+\llabel{datasimforall}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range (or, in a Bayesian framework, from the prior distribution).
 
@@
 These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}).
-This---which we call a retrodictive check (\citealp{betanworkflow}, also sometimes called a posterior predictive check, \citealp{Gelman2020})---allows model output to be compared to observations.
+This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations.
 It's only once all steps have been completed that we can interpret parameter values with respect to our research question.
@@
 The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to more clearly develop a research question before extensive model fitting.
-
-\begin{figure}[h]
-	\centering
-	\vspace*{-0.1cm}
-    \hspace*{-1.5cm}
-	\includegraphics{figures/figure_worflow_wsteps}
-	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data.
-Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5).
-Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
-The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.}
-	\vspace*{-0.5cm}
-	\label{fig:workflow}
-\end{figure}
 
@@
 At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.
+
+\begin{figure}[h]
+	\centering
+	\vspace*{-0.1cm}
+    \hspace*{-1.5cm}
+	\includegraphics{../figures/figure_worflow_wsteps}
+	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data.
+Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5).
+Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
+The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.}
+	\vspace*{-0.5cm}
+	\label{fig:workflow}
+\end{figure}
 
@@
 \begin{minipage}[t]{\linewidth}
-    \includegraphics[width=\linewidth]{figures/trendestimation_details}
+    \includegraphics[width=\linewidth]{../figures/trendestimation_details}
 
@@
 \begin{minipage}[t]{\linewidth}
-    \includegraphics[width=\linewidth]{figures/forecasting_details}
+    \includegraphics[width=\linewidth]{../figures/forecasting_details}
 
@@
 	\vfill
-\includegraphics[width=\linewidth]{figures/iterativeworkflow_details}
+\includegraphics[width=\linewidth]{../figures/iterativeworkflow_details}
 \vfill
@@
 A universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
-Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
+Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge---and where the full model is fitted jointly.
 It would be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimates.
@@
 \clearpage
-\bibliography{forecastflows}
+\bibliography{../forecastflows.bib}
 

### 2025-07-10 revision/response_r1.tex
continue working on reviewer response

@@
 
-While Bayesian models are more and more frequently (we cite for example ), we clarified that---despite some Bayesian-specific terms---data simulation apply for all (\href{file:forecastflows_r1\#lintarget:datasimforall}{line \ref*{lin:datasimforall}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}).
+While Bayesian models are more and more frequently (we cite for example ), we clarified that some terms are specific to Bayesian inference but that data simulation still apply for all (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).
 
@@
 
+We removed this word.
+
 \begin{mybox}
@@
 
+Done, thank you.
+
 \begin{mybox}
@@
 \end{mybox}
+
+Done.
 
@@
 
+Done. % there were two 'then", I removed both...
+
 \begin{mybox}
@@
 \end{mybox}
+
+We have modified ``reduce" to ``quantify" (\href{file:forecastflows_r1\#lintarget:quant1}{line \ref*{lin:quant1}}, \href{file:forecastflows_r1\#lintarget:quant2}{line \ref*{lin:quant2}}).
 
@@
 \begin{mybox}
-\emph{Box A: Please check the logic of the first sentence. More papers because of less time confronting their models? Is that what you meant to say? “For examples” should be “For example”?}
+\emph{Box A: Please check the logic of the first sentence. More papers because of less time confronting their models? Is that what you meant to say?}
 \end{mybox}
+
+\begin{mybox}
+\emph{“For examples” should be “For example”?}
+\end{mybox}
+
+% Hmm, not sure? "For examples using LPI data" and then we talk about several examples right?
 
@@
 
+We have claried this sentence:
+\begin{quote}
+``The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model.
+Calibrating submodels separately allows to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use."
+\end{quote}
+
 \begin{mybox}
@@
 
+Thank you!
+
 \begin{mybox}
@@
 \end{mybox}
+
+Response. %hmm, he's making a good point, maybe we should at least change the section title?
 
@@
 1. Clarity on inferential goals of “disconnected paradigms”\\
- When the paper discusses “harmonizing both trend estimation and forecasting” what is meant by this? The
-title of the paper suggests the inferential goals from both approaches to be predictive
-in nature (i.e., “forecasting trends”), but it seems plausible that the simpler models are
-developed for more of an explanatory purpose (e.g., investigating relationships between
-covariates and mean response) rather than predicting or forecasting future trends, which
-would make the questions being asked by each modeling approach fundamentally dif-
-ferent (see, Shmueli 2010). The paper could be improved if more clarity was provided
-on the inferential goals of both types of modeling paradigms it discusses, and perhaps
-provide some discussion why they need to be (or if they should be) harmonized if they
-are in pursuit of different types of inference.}
+ When the paper discusses “harmonizing both trend estimation and forecasting” what is meant by this?
+The title of the paper suggests the inferential goals from both approaches to be predictive in nature (i.e., “forecasting trends”), but it seems plausible that the simpler models are developed for more of an explanatory purpose (e.g., investigating relationships between covariates and mean response) rather than predicting or forecasting future trends, which would make the questions being asked by each modeling approach fundamentally different (see, Shmueli 2010).
+The paper could be improved if more clarity was provided on the inferential goals of both types of modeling paradigms it discusses, and perhaps provide some discussion why they need to be (or if they should be) harmonized if they are in pursuit of different types of inference.}
 \end{mybox}
+
+Thank you for this suggestion.
+We added a paragraph in the introduction to discuss the inferential goals of these two approachs (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}).
+We also changed the title to reflect the broader inferential goals.
 
@@
 
+We clarified that some terms are specific to Bayesian inference but that this workflow can be applied more broadly (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).
+
 \begin{mybox}
 \emph{3. Discussion of Calibration:\\
-Lines 98-104 seem to be referring to a check on estimating “the model” using synthetic data that are generated from the model (either through optimization or McMC), but there is no mention of various computational approaches that may be taken to fit “the model.” How does the chosen computational strategy fit into the workflow if at all?
-Further, is the proposed workflow to conduct a full-blown simulation calibration study (i.e., simulate-fit-check for capture) many times and across all possible parameter values?
+Lines 98-104 seem to be referring to a check on estimating “the model” using synthetic data that are generated from the model (either through optimization or McMC), but there is no mention of various computational approaches that may be taken to fit “the model.” How does the chosen computational strategy fit into the workflow if at all?}
+\end{mybox}
+
+While we agree that this is an interesting question, we believe it falls outside the scope of this manuscript.
+
+\begin{mybox}
+\emph{Further, is the proposed workflow to conduct a full-blown simulation calibration study (i.e., simulate-fit-check for capture) many times and across all possible parameter values?
 Assuming yes, I’d be interested in how something like that would scale for the types of models that are the focus of the paper?
@@
 
+We have clarified the data simulaton step.
+We believe a first step would be to simulate from one parameter set (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}), and then to repeat the simulations with several parameter sets (\href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}).
+
 \begin{mybox}
 \emph{4. Gaps in the workflow\\
-A big argument in the paper is that the proposed workflow
-could lessen discrepancies in results between studies of trends (lines 123-128).
+A big argument in the paper is that the proposed workflow could lessen discrepancies in results between studies of trends (lines 123-128).
 However, in the workflow there is no mention of transparent communication of study design, which if ignored in analysis, could result in discrepancies in results for studies with similar research questions.
@@
 
+Response.
+
 \begin{mybox}
@@
 \end{mybox}
+
+Thank you for this suggestion. We now say that the checks should be developped specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).
 
@@
 
+Sorry for this typo.
+
 \begin{mybox}
@@
 
+Done.
+
 \begin{mybox}
@@
 \end{mybox}
+
+We now mention that this step can allow to check our prior assumptions (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}).
 
@@
 \end{mybox}
+
+...
 
@@
 
+We now mention that the summary statistics should be developped specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).
+% I already said this above...
+
 \begin{mybox}
@@
 
+Corrected.
+
 \begin{mybox}
-\emph{Lines 198-201: if the goal is accurate and relatively-precise forecasts, do all of the nuisance
-parameters need to be identifiable for a model to be useful? That is, does identifiability
-of all parameters matter for process-based models with a goal of forecasting?}
+\emph{Lines 198-201: if the goal is accurate and relatively-precise forecasts, do all of the nuisance parameters need to be identifiable for a model to be useful?
+That is, does identifiability of all parameters matter for process-based models with a goal of forecasting?}
 \end{mybox}
+
+We believe that the strength of process-based models (as compared to machine learning for example, \href{file:forecastflows_r1\#lintarget:MLvsPBM}{line \ref*{lin:MLvsPBM}}) is their interpretability.
+Their parameters should represent biological processes, and their values are of interest.
+The inferential goal of process-based models is both predictive and explanatory (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}), and poor identifiability could reduce this.
 

### 2025-07-10 revision/forecastflows_r1.tex
continue working on reviewer response

@@
 
-\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
+% \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
+\title{Closing the gap between statistical and scientific workflows for improved inference in ecology} 
 \date{\today}
@@
 \noindent \emph{For:} Workflow for Applied Data Analysis theme issue for \emph{Phil Trans A} as an \emph{Opinion}
-% They mention a tex template here: https://royalsocietypublishing.org/rsta/for-authors#question4
-% Word length: I should check what they emailed but they say no more than 13 printed pages with 650 per page (whoa, we should be well below that! I am thinking 3-5K would be good)
-% Already 5K!
 
@@
 Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process.
-We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process.
-We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
+We propose a principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process.
+We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform ecological modeling.
 
@@
 Impacts are projected to increase in coming decades, as climate change accelerates biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}.
-Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics.
+Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date, alongside the skill to forecast future dynamics.
 % These requirements in turn have created a growing need for more global ecological data, and the types of models that can robustly handle such data.
@@
 
-We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
+\llabel{inferentialgoal}The inferential goals of these two modeling paradigms---explanation vs.\ prediction---seem disconnected at first glance, but remains in reality much more blurred.
+Trend estimation, often with simpler models, tend to focus on an explanatory purpose, but often lacks a clear link with theories that could provide causal interpretation.
+For forecasting, the use of more complex process-based models for a predictive purpose is often justified by the assumption that ``we must understand to predict", contributing to the confusion between explanation and prediction \citep{Shmueli2010}.
+
+We argue that current debates, confusions and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
 Research estimating biodiversity trends has become focused on methodological aspects; the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting.
@@
 Workflows that fully integrate all the steps required to build a model from an ecological question, with evaluation of limitations and potential problems before estimating its parameters and making projections, could reduce many of these problems.
-In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge \citep{McElreath2018, betanworkflow, Gelman2020, Wolkovich2024}.
+In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge \citep{McElreath2018, betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024}.
 Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
@@
 Ideally, this begins in the pre-model/pre-data quadrant, where the research question is defined.
-The next step moves to the post-model/pre-data quadrant, where a model is build---independently of the data.
+The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data.
 This is where we should spend a good deal of time evaluating model behavior using simulated data.
@@
 This reality should drive researchers to use more robust and coherent methods.
-But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems.
+But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
 Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}a).
@@
 From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}).
-The general idea is to start with a relatively simple model that we could refine later (see example worklow we provide).
+The general idea is to start with a relatively simple model that we could refine later (see example workflow we provide).
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
@@
 With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}).
-`Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable values (which is straightforward if the parameters are interpretable) and from fake predictor data.
+`Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values (which is straightforward if the parameters are interpretable) and from fake predictor data.
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
-At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is reasonable or not.
+At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not.
 This is also a step for making sure the model is working as expected (see example worklow).
-\llabel{datasimforall}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range (or, in a Bayesian framework, from the prior distribution).
+\llabel{repeatsim}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range.
+\llabel{priorPC}In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors, \citealp{Gelman2020}).
 
@@
 This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations.
+\llabel{howtoPPC}There is no general way to do this: the goal is to tailor checks that assess key features of the data in regards to our research question \citep{Gelman2020}.
 It's only once all steps have been completed that we can interpret parameter values with respect to our research question.
@@
 
-A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
+A key feature of this workflow is the central role of data simulation, which introduces two feedback loops---\llabel{broadcontext}and should be applied not only in a Bayesian framework but across broader modeling approaches.
+The first feedback arises when we evaluate the model on simulated data.
 The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
@@
 unbalanced data).
-For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step.
+For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step.
 This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics.
@@
 
-Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties and yield new scientific insights.
-This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.
+Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}quantify uncertainties and yield new scientific insights.
+This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data.
 Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.
@@
 
-Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome.
+Increasing model complexity can be beneficial, especially when it helps better \llabel{quant2}quantify the uncertainty; however, this is not always the outcome.
 In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
@@
 Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory.
-The way models are currently calibrated is likely not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
+The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model.
+Calibrating submodels separately allows to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.
 
@@
 With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important.
-Machine learning will likely surpass process-based models for forecasting accuracy if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
+\llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
 Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.

### 2025-07-11 revision/response_r1.tex
continue reviewer reponses

@@
 \usepackage[labelfont=bf]{caption}
-\bibliographystyle{../amnat}
+\bibliographystyle{besjournals}
 
@@
 
-While Bayesian models are more and more frequently (we cite for example ), we clarified that some terms are specific to Bayesian inference but that data simulation still apply for all (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).
+Bayesian models are more and more frequently used in ecology (we cite for example \citealp{Johnson2024}). % I think this one is Bayesian right?
+We clarified that some terms are specific to Bayesian inference but that data simulation still apply for all (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).
 
@@
 
+% here I hope we will have a PPC figure
 Response.
@@
 
-Response.
+We have reorganized the figure based on your suggestions, and we now explain more clearly the different panels in the caption.
+We also now used the pre- and post-data/model in the main text (\href{file:forecastflows_r1\#lintarget:quad1}{lines \ref*{lin:quad1}}, \href{file:forecastflows_r1\#lintarget:quad2}{\ref*{lin:quad2}}, \href{file:forecastflows_r1\#lintarget:quad3}{\ref*{lin:quad3}}, \href{file:forecastflows_r1\#lintarget:quad4}{\ref*{lin:quad4}}, \href{file:forecastflows_r1\#lintarget:quad5}{\ref*{lin:quad5}}, \href{file:forecastflows_r1\#lintarget:quad6}{line \ref*{lin:quad6}}, and \href{file:forecastflows_r1\#lintarget:quad7}{\ref*{lin:quad7}}).
 
@@
 \end{mybox}
+
+Thank you for spotting these mistakes.
 
@@
 \end{mybox}
+
+Response.
 
@@
 
-\bibliography{forecastflows}
+\clearpage
+\bibliography{../forecastflows.bib}
 

### 2025-07-11 revision/forecastflows_r1.tex
continue reviewer reponses

@@
 	\centering
-	\includegraphics{../figures/modeldataspaces_rotation}
-	\caption{Different trajectories through model-data space during model development.
-Ideally, this begins in the pre-model/pre-data quadrant, where the research question is defined.
+	\includegraphics{../figures/figure_modeldataspaces_revised}
+	\caption{Different trajectories through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
+	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses we want to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
+	Panel B illustrates that in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
+	Panel C illustrates the typical process-based model development.
+A lot of time is spent building and calibrating (fitting) separate submodels, without clear distinction between model building and model calibration and without evaluating the full model structure a priori.
+Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
+	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details).
+Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined.
 The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data.
@@
 Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data.
-Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedbacks to refine our model.}
+Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
 	\label{fig:modeldata}
@@
 But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
-Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}a).
-For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (figure \ref{fig:modeldata}b).
+Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e.
+post-model/post-data, figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}b).
+For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data).
 Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
@@
 
-We argue a workflow that moves along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
+We argue a workflow that moves \llabel{quad4}along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}d) could reduce many of these problems and thus improve ecological science.
 The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}).
 This involves making clear assumptions about the most influential drivers, within the specific context of our study.
-This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error.
+This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, figure \ref{fig:modeldata}d).
 \llabel{causalinf}This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}.
@@
 `Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values (which is straightforward if the parameters are interpretable) and from fake predictor data.
-We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
+We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, figure \ref{fig:modeldata}d).
 At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not.
@@
 Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}).
-This way, we obtain parameter estimates constrained by observations.
+This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, figure \ref{fig:modeldata}d).
 These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}).

### 2025-07-16 revision/forecastflows_r1.tex
small modification

@@
 This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand.
-Ecologists have started to use simulated data more to understand potential limitations of their models and data combined, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.
+Ecologists have started to use simulated data more to understand potential limitations of their models and data combined \citep{Hilborn1997}, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.
 

### 2025-07-17 revision/response_r1.tex
small modification
continue working on revision

@@
 
-Response.
+We thank the reviewer for their positive comments regarding our manuscript. 
+We have worked to clarify the points raised by the reviewer in this new version.
 
@@
 
-Response.
+Thank you for this interesting point.
+We have work to clarify the data simulaton step (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}, \href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}), and now highlight one should develop specific posterior predictive checks for the question of interest (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).
+However, we believe an extended discussion of the PPC advantages and inconvenients is outside the word limit we have.
 
@@
 % here I hope we will have a PPC figure
-Response.
+[Waiting for a nice PPC figure] Response.
 
@@
 
+...
+
 \begin{mybox}
@@
 % Hmm, not sure? "For examples using LPI data" and then we talk about several examples right?
+...
 
@@
 There are places where I feel more specificity (or clarity) with respect to these nuances could improve the manuscript and its potential for impact with respect to building trust in the scientific process.
-There is also a lack of specificity in how, exactly, this type of workflow could be implemented because there
-is no example application to data. I have outlined these areas under “Major comments” and provided minor editorial comments in “Minor Comments.”}
+There is also a lack of specificity in how, exactly, this type of workflow could be implemented because there is no example application to data.
+I have outlined these areas under “Major comments” and provided minor editorial comments in “Minor Comments.”}
 \end{mybox}
@@
 % through this lens.
+
+We are glad that the reviewer found our manuscript well-written. 
+We thank them for the feedback, which has improved the manuscript.
 
@@
 
-Response.
+% Real-data examples... not do-able!
+...
 

### 2025-07-17 revision/forecastflows_r1.tex
small modif. to figure C in box
continue working on revision

@@
 	Panel C illustrates the typical process-based model development.
-A lot of time is spent building and calibrating (fitting) separate submodels, without clear distinction between model building and model calibration and without evaluating the full model structure a priori.
+A lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals), without clear distinction between model building and model calibration and without evaluating the full model structure a priori.
 Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
@@
 Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
-The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.}
+The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.
+This figure is expanded from Figure \ref{fig:modeldata}d.}
 	\vspace*{-0.5cm}
@@
 	\vfill
-\includegraphics[width=\linewidth]{../figures/iterativeworkflow_details}
+\includegraphics[width=\linewidth]{../figures/iterativeworkflow_details_revised}
 \vfill

### 2025-07-22 revision/response_r1.tex
incorporate Jim feedback

@@
 ...
+% Jim: I think the reviewer makes a fair point that this is vague.
+Can we add something here, or even just refer to section 3.2 where (I think) relevant examples are already given in the final paragraph?
 
@@
 
-We have modified ``reduce" to ``quantify" (\href{file:forecastflows_r1\#lintarget:quant1}{line \ref*{lin:quant1}}, \href{file:forecastflows_r1\#lintarget:quant2}{line \ref*{lin:quant2}}).
+We have clarified this (\href{file:forecastflows_r1\#lintarget:quant1}{line \ref*{lin:quant1}}, \href{file:forecastflows_r1\#lintarget:quant2}{line \ref*{lin:quant2}}).
 
@@
 
-Response.
+Sorry it was not clear. We have clarified this first paragraph.
 
@@
 
-Response.
+We have added an entire paragraph to discuss the methodological aspects, including study/sampling design, that should be more transparently shared along with the use of this workflow (\href{file:forecastflows_r1\#lintarget:studydesignandmore}{line \ref*{lin:studydesignandmore}}).
 
@@
 
-Response.
+We have added thus suggestion (\href{file:forecastflows_r1\#lintarget:collab}{line \ref*{lin:collab}}).
 

### 2025-07-22 revision/forecastflows_r1.tex
incorporate Jim feedback

@@
 \documentclass[11pt]{article}
-\usepackage[top=0.70in, bottom=.80in, left=1in, right=1in]{geometry}
+\usepackage[top=0.70in, bottom=.70in, left=1in, right=1in]{geometry}
 \renewcommand{\baselinestretch}{1.1}
@@
 \usepackage{todonotes}
-\usepackage[font=sf]{caption}
+\usepackage[font={sf,footnotesize}]{caption}
 \usepackage{hyperref}
@@
 
-\llabel{inferentialgoal}The inferential goals of these two modeling paradigms---explanation vs.\ prediction---seem disconnected at first glance, but remains in reality much more blurred.
-Trend estimation, often with simpler models, tend to focus on an explanatory purpose, but often lacks a clear link with theories that could provide causal interpretation.
-For forecasting, the use of more complex process-based models for a predictive purpose is often justified by the assumption that ``we must understand to predict", contributing to the confusion between explanation and prediction \citep{Shmueli2010}.
+\llabel{inferentialgoal}At first glance, these two modeling paradigms may appear to pursue independent inferential goals--- explanation vs.\ prediction---but in reality this distinction is blurred.
+Trend estimation may be nominally focused on explanatory outcomes, but especially because it tends to be carried out using simpler models, it often lacks a clear link with theories that facilitate meaningful causal interpretation.
+Moreover, whether implicitly or explicitly, estimated trends are frequently treated as quantities that project forward into the future, taking on a forecasting role.
+Meanwhile, although more complex forecasting models are certainly built for predictive purposes, the typical use of mechanistic, process-based model components is often justified by the assumption that ``we must understand to predict”---and indeed, such models are often judged by how faithfully they mimic the real-world processes they represent---which contributes to the conflation of explanation with prediction \citep{Shmueli2010}.
 
@@
 
-\section{Scientific method and workflows}
-
-Quantitative science relies on a model-based framework to confront hypothesis with data \citep{Chamberlin:1965cd}.
-In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses.
-This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection \citep{Nosek2018}.
-But this idealized method often does not apply to the reality of ecological research.
-Many important questions cannot be addressed through controlled experiments and replications.
-In such cases, we must rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.
-Indeed, most macroecological insights have emerged from exploring patterns in these datasets (exploratory data analysis).
-
-\begin{figure}[h]
+\begin{figure}
 	\centering
@@
 \end{figure}
+
+\section{Scientific method and workflows}
+
+Quantitative science relies on a model-based framework to confront hypothesis with data \citep{Chamberlin:1965cd}.
+In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses.
+This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection \citep{Nosek2018}.
+But this idealized method often does not apply to the reality of ecological research.
+Many important questions cannot be addressed through controlled experiments and replications.
+In such cases, we must rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.
+Indeed, most macroecological insights have emerged from exploring patterns in these datasets (exploratory data analysis).
 
@@
 
+\begin{figure}[h]
+	\centering
+	\vspace*{-0.5cm}
+    \hspace*{-1.5cm}
+	\includegraphics{../figures/figure_worflow_wsteps}
+	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data.
+Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5).
+Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
+The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.
+This figure is expanded from Figure \ref{fig:modeldata}d.}
+	\vspace*{-0cm}
+	\label{fig:workflow}
+\end{figure}
+
 Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}).
@@
 
-\begin{figure}[h]
-	\centering
-	\vspace*{-0.1cm}
-    \hspace*{-1.5cm}
-	\includegraphics{../figures/figure_worflow_wsteps}
-	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data.
-Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5).
-Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
-The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.
-This figure is expanded from Figure \ref{fig:modeldata}d.}
-	\vspace*{-0.5cm}
-	\label{fig:workflow}
-\end{figure}
-
 \section{The workflow in practice}
 
-Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}quantify uncertainties and yield new scientific insights.
+Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}understand and manage uncertainties and yield new scientific insights.
 This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data.
@@
 
-Increasing model complexity can be beneficial, especially when it helps better \llabel{quant2}quantify the uncertainty; however, this is not always the outcome.
+In principle, increasing model complexity can be beneficial by enhancing mechanistic realism, better representing more complex dynamics, and \llabel{quant2}reducing uncertainty by restricting the effective parameter space.
+In practice, however, this is not always the outcome.
 In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
@@
 data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model.
-Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
+Non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
 The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.
@@
 % Trends - outline current problem
-In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data.
+% In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data.
+In the current workflow for estimating trends over time, a model is typically chosen a priori to fit to data, yielding parameter estimates that are treated as true and often translated directly into published findings and conclusions, without interrogating the model using simulated data and carefully evaluating its performance.
+Any such analysis must be treated with caution, because different model variants can yield very different estimates (hence different conclusions), and published results may not prove robust when confronted with future reanalysis of the same data.
 The Living Planet Index (LPI, \url{www.livingplanetindex.org}), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results.
@@
 This would be massive!}
-With updated data released semi-annually alongside new datasets and new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
+% With updated data released semi-annually alongside new datasets and new estimates of decline, 
+A growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
 For examples using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.
@@
 
-We suggest unified training in our proposed workflow, or a similar one \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, would focus on learning to generate questions and then models, and then how to simulate data from them.
+% We suggest unified training in our proposed workflow, or a similar one \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, would focus on learning to generate questions and then models, and then how to simulate data from them.
+\llabel{collab}Although in some cases it may be most expedient for ecologists to collaborate with statisticians who are already well-equipped to adopt our proposed workflow or a similar workflow \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, we recommend that domain scientists who regularly engage in model-driven research efforts receive unified training about how to generate well-formed questions and appropriate models, and then how to simulate data from them.
 Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
@@
 
-\subsection{A tractable alternative to machine learning}
+Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community.
+First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently, Second, open sharing of more comprehensive study design and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community.
+Third, where feasible, alignment on common datasets would provide a standardized basis for model development and comparison, reducing the fragmentation of efforts and facilitating more direct and meaningful evaluations of different modeling approaches.
+Coupled with universal training, these shifts in the sharing and standardization of data and models would create a more collaborative, efficient, and rigorous environment for ecological modeling, ultimately leading to more accurate forecasts and better understanding of complex environmental systems.
+
+% New paragraph by Jim!
+\llabel{studydesignandmore}Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community.
+First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently, Second, open sharing of more comprehensive study design and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community.
+Third, where feasible, alignment on common datasets would provide a standardized basis for model development and comparison, reducing the fragmentation of efforts and facilitating more direct and meaningful evaluations of different modeling approaches.
+Coupled with universal training, these shifts in the sharing and standardization of data and models would create a more collaborative, efficient, and rigorous environment for ecological modeling, ultimately leading to more accurate forecasts and better understanding of complex environmental systems.
+
+\subsection{[Need a better title] A tractable alternative to machine learning}
 

### 2025-07-22 figures/retrodic/simulations_speciesonlyintslope.R
Working on the retrodictive but still not happy ....

@@ -1,0 +1,139 @@
+rm(list = ls());gc()
+if(length(grep("victor", getwd()) > 0)) {
+  wd <- "/home/victor/projects/forecastflows"
+} else if(length(grep("lizzie", getwd()) > 0)) {
+  wd <- "/Users/lizzie/Documents/git/projects/misc/miscmisc/workflowsPhilTrans"
+}  
+
+# Load packages
+require(ape)
+require(geiger)
+require(phytools)
+require(ggplot2)
+require(rstan)
+
+set.seed(1111)
+
+nspecies = 40
+# phytree <- pbtree(n=nspecies, nsim=1, b=1, complete=FALSE, scale=1)
+
+# tree <- pbtree(n = nspecies, scale = 1)
+# tree[["edge.length"]][c(1,10)] <- tree[["edge.length"]][c(1,10)]*2
+# phytree <- force.ultrametric(tree)
+phytree <- rcoal(n = nspecies, scale = 1)
+phytree <- rcoal(n = nspecies, scale = 1)
+plot(phytree)
+
+
+# ------------- #
+# Simulate data #
+# ------------- #
+whichspp <- c(8,15,17,21,28,32,35)
+
+nspecies <- phytree[["Nnode"]]+1 # no. of species
+nobs_perspecies <- round(runif(nspecies, 5, 15)) # no. of different observations per population
+nobs_perspecies[whichspp] <- round(runif(length(whichspp), 90, 140)) # some bias
+spid_perobs <- rep(1:nspecies, times = nobs_perspecies)
+
+years <- c()
+for(np in nobs_perspecies){
+  years_p <- sample(1880:2020, size = np, replace = FALSE)
+  years <- c(years, years_p)
+}
+
+## simulate some intercepts with phylogenetic structure
+lambdaint <- 1
+brootint <- log(100)
+sigmaint <- log(10)
+scaledtree_int <- rescale(phytree, model = "lambda", lambdaint)
+speciesnum <- as.numeric(gsub("t", "", phytree[["tip.label"]]))
+plot.phylo(scaledtree_int,
+           cex = 1, 
+           tip.color = c("#FF7601", "#00809D")[as.numeric(as.numeric(gsub("t", "", phytree[["tip.label"]])) %in% whichspp)+1])
+
+mu_alpha_sp_wphylogeny <- fastBM(scaledtree_int, a = brootint, mu = 0, sig2 = sigmaint ^ 2)
+
[…]

### 2025-07-22 figures/retrodic/simulations_speciesonlyintslope.R
More efforts, still annoyed

@@ -12,7 +12,8 @@ require(phytools)
 require(ggplot2)
 require(rstan)
 
-set.seed(1111)
+# set.seed(1111)
+set.seed(777)
 
 nspecies = 40
 # phytree <- pbtree(n=nspecies, nsim=1, b=1, complete=FALSE, scale=1)
@@ -20,7 +21,7 @@ nspecies = 40
 # tree <- pbtree(n = nspecies, scale = 1)
 # tree[["edge.length"]][c(1,10)] <- tree[["edge.length"]][c(1,10)]*2
 # phytree <- force.ultrametric(tree)
-phytree <- rcoal(n = nspecies, scale = 1)
+# phytree <- rcoal(n = nspecies, scale = 1)
 phytree <- rcoal(n = nspecies, scale = 1)
 plot(phytree)
 
@@ -28,7 +29,9 @@ plot(phytree)
 # ------------- #
 # Simulate data #
 # ------------- #
-whichspp <- c(8,15,17,21,28,32,35)
+# whichspp <- c(8,15,17,21,28,32,35) # seed(1111) with second trees
+
+whichspp <- c(4, 15:17,20, 25)
 
 nspecies <- phytree[["Nnode"]]+1 # no. of species
 nobs_perspecies <- round(runif(nspecies, 5, 15)) # no. of different observations per population
@@ -79,6 +82,7 @@ ggplot(data = data.frame(sp = speciesnum, mu_beta_sp_wphylogeny)) +
 ## simulate counts, here with the WITH phylogenetic structure
 yphylo <- c()
 sigma_obs <- 0.5
+
 for(p in 1:nspecies){
   nobs_p <- nobs_perspecies[p]
   years_p <- years[spid_perobs == p] 
@@ -96,6 +100,16 @@ datasim <- data.frame(
 
 datasim$logyphylo <- log(datasim$yphylo)
 
+dropspp <- c(1:20)
+
+datasimdrop <- data.frame(
+  yphylo[which(!spid_perobs %in% dropspp)],
+  year = years[which(!spid_perobs %in% dropspp)] - 1980,
+  species = as.character(spid_perobs)[which(!spid_perobs %in% dropspp)]
+)
+
+datasimdrop$logyphylo <- log(datasim$yphylo)[which(!spid_perobs %in% dropspp)]
+
 # -------------------------------------------------- #
 # Fit the model against data without phylo structure #
 # -------------------------------------------------- #
[…]

### 2025-07-24 revision/response_r1.tex
Edits by Lizzie finally

@@
 
-We have added thus suggestion (\href{file:forecastflows_r1\#lintarget:collab}{line \ref*{lin:collab}}).
+Good point.
+We have adjusted our language in one place (\href{file:forecastflows_r1\#lintarget:collab1}{line \ref*{lin:collab1}}) and we have added this suggestion also (\href{file:forecastflows_r1\#lintarget:collab}{line \ref*{lin:collab}}).
 

### 2025-07-24 revision/forecastflows_r1.tex
Edits by Lizzie finally

@@
 % \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
-\title{Closing the gap between statistical and scientific workflows for improved inference in ecology} 
+\title{Closing the gap between statistical and scientific workflows for improved inference in ecology} %emw24Jul: Hmm, you say in the letter you changed the title but I am not sure you did. 
 \date{\today}
@@
 New research often suggests that models accounting for these complexities can yield more accurate trends and predictions.
+%emw24Jul: Are you sure you want 'paradigm'?
+I love the word and the sentence but I think we mean more two 'groups' or 'spheres' -- I would switch and then carry the language you use to the next sentence.
 We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons.
-First, the research community is split between two disconnected paradigms: using data to fit simple, trend-like models with few parameters, or developing forecasting models that include complex, mechanistic submodels only indirectly informed by data.
-Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process.
-We propose a principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process.
-We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform ecological modeling.
+First, the research community is split between two disconnected paradigms that prevent uniting a natural connection between trend estimation and forecasting.
+One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, which often include complex, mechanistic sub-models only indirectly informed by data.
+Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions throughout the entire modeling process.
+We propose a principled workflow that can address both of these limitations, by integrating statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process.
+We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could improve our understanding of ecological systems and forecasts for them.
+
+%emw24Jul: I pulled out a few of the best sentences that may help the abstract: 
+% Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date, alongside the skill to forecast future dynamics.
+% Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
+% Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
 
@@
 
-Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on less data.
+Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on fewer data.
 Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic biases.
@@
 Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning.
-These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012, Thuiller2019}.
+These approaches have failed to yield agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012, Thuiller2019}.
 
-\llabel{inferentialgoal}At first glance, these two modeling paradigms may appear to pursue independent inferential goals--- explanation vs.\ prediction---but in reality this distinction is blurred.
-Trend estimation may be nominally focused on explanatory outcomes, but especially because it tends to be carried out using simpler models, it often lacks a clear link with theories that facilitate meaningful causal interpretation.
-Moreover, whether implicitly or explicitly, estimated trends are frequently treated as quantities that project forward into the future, taking on a forecasting role.
-Meanwhile, although more complex forecasting models are certainly built for predictive purposes, the typical use of mechanistic, process-based model components is often justified by the assumption that ``we must understand to predict”---and indeed, such models are often judged by how faithfully they mimic the real-world processes they represent---which contributes to the conflation of explanation with prediction \citep{Shmueli2010}.
+%emw24Jul: Same comment as above about 'paradigms' but here I think you mean 'approaches' so I just changed it. 
+\llabel{inferentialgoal}These two modeling approaches may appear to pursue independent inferential goals---explanation vs.\ prediction---but in reality this distinction is blurred.
+Trend estimation may be nominally focused on explanatory outcomes, yet it rarely links clearly to theories that could provide causal inference, usually because it tends to be carried out using simpler models.
+Moreover,  estimated trends are frequently treated---implicitly and explicitly---as quantities that project forward into the future, taking on a forecasting role.
+%emw24Jul: Nice!
+In contrast, approaches using complex forecasting models are certainly built for predictive purposes, yet these also have an additional agenda.
+Developing and using mechanistic, process-based models is often justified by the assumption that ``we must understand to predict," and such models are often judged by how faithfully they mimic the real-world processes they represent, which conflates explanation with prediction \citep{Shmueli2010}.
 
-We argue that current debates, confusions and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
-Research estimating biodiversity trends has become focused on methodological aspects; the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting.
-At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components.
-These new parts are often disconnected from the original research aim, its data stream, and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).
+%emw24Jul: I edited something out that I do and you should avoid (or try to edit out when I add it): adding too many 'subjunctive/conditional' phrases.
+We ARGUE THAT MANY.
+...
+IN LARGE PART mean ...
+try to pick one (of two here) conditional phrase and drop the rest.
+We argue that many current debates and diverging forecasts are driven by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
+Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting.
+At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components because the current workflow rarely examines the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).
+Thus it encourages new parts that are often disconnected from the original research aim, its data stream, and the previous scientific insights.
 
@@
 	\includegraphics{../figures/figure_modeldataspaces_revised}
-	\caption{Different trajectories through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
-	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses we want to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
-	Panel B illustrates that in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
-	Panel C illustrates the typical process-based model development.
-A lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals), without clear distinction between model building and model calibration and without evaluating the full model structure a priori.
+	\caption{Different trajectories of current (b-c) and proposed (d) workflows through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
+	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
+	Panel B illustrates that, in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
+	Panel C illustrates the typical process-based model development where a lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals) without clear distinction between model building and model calibration and without evaluating the full model structure a priori.
 Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
@@
 
+%emw24Jul: Nice stuff here!
 This reality should drive researchers to use more robust and coherent methods.
@@
 From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}).
-The general idea is to start with a relatively simple model that we could refine later (see example workflow we provide).
+Generally, starting with a relatively simple model that we could refine later makes understanding the model and how it interacts with empirical data easier (see example workflow we provide).
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
 This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).
+%emw24Jul: 'see example workflow we provide' and 'see example workflow' -- can we standardize our mention of this and make it clear where it is?
+Like: `see example workflow we provide as supplement files.'
 
 With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}).
-`Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values (which is straightforward if the parameters are interpretable) and from fake predictor data.
+`Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values, which is straightforward if the parameters are interpretable, and from `fake' (simulated) predictor data.
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, figure \ref{fig:modeldata}d).
@@
 \llabel{repeatsim}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range.
-\llabel{priorPC}In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors, \citealp{Gelman2020}).
+\llabel{priorPC}In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}), while in other frameworks they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).
 
@@
 	\includegraphics{../figures/figure_worflow_wsteps}
-	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data.
+	\caption{The workflow we propose here, which builds from recent advances in workflows \citep[][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024}, focuses on iterative feedbacks between the research question, model and data.
 Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5).
@@
 This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations.
-\llabel{howtoPPC}There is no general way to do this: the goal is to tailor checks that assess key features of the data in regards to our research question \citep{Gelman2020}.
-It's only once all steps have been completed that we can interpret parameter values with respect to our research question.
+\llabel{howtoPPC}This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regards to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
+It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
 The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.
 
-Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
-The adoption of the workflow in macroecological studies---where model building is often informed by patterns in the data---would also make the exploratory analysis more transparent (as an explicit preliminary step for entering the workflow) and would compel researchers to more clearly develop a research question before extensive model fitting.
+Within such a workflow, forecasting emerges as a natural outcome.
+Rather than being a final goal, forecasting only involves jointly modeling new circumstances along with the original data.
+This workflow could also bridge gaps in certain areas between exploratory analyses and developing research questions.
+For example, in macroecological studies model building is often informed by patterns in the data.
+The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting.
+%emw24Jul: possible to ref Fig 1 here?
 
-A key feature of this workflow is the central role of data simulation, which introduces two feedback loops---\llabel{broadcontext}and should be applied not only in a Bayesian framework but across broader modeling approaches.
+A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
 The first feedback arises when we evaluate the model on simulated data.
@@
 Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic.
-We can refine the model to integrate the missing process(es) (if we can identify them) and return to the start of the workflow.
+We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow.
 Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g.
@@
 At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.
+\llabel{broadcontext}These feedback loops are applied currently in only in a Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.
 
@@
 
+%emw24Jul: I removed case study from this draft, I think they're more than case studies and the term no longer helps the paper.
 Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}understand and manage uncertainties and yield new scientific insights.
 This will help refocus the debate on designing new hypotheses, formulating new questions---and guiding efforts to collect new data.
-Here, we illustrate how such a workflow could lead to significant improvements in two case studies: (i) estimating global biodiversity trends and (ii) forecasting future species and ecosystem dynamics using process-based models.
+Here, we illustrate how such a workflow could lead to significant improvements in two related areas: (i) estimating global biodiversity trends and (ii) forecasting species and ecosystem dynamics using process-based models.
 
@@
 
-An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
+%emw24Jul: Cite supp retro check figure here below? 
+We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
 Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process.
@@
 But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model.
+
 We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models.
@@
 Or a need to also find which species are declining most, including those that may have poor or no data?
-From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data.
+From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or `synthetic data generation' currently \citep{Buschke2021,mcrae2025utility}, but could be built into the models fit to the empirical data.
 Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends (see workflow example).
@@
 Ecological forecasting is a broad field with a diverse range of methods.
-Our second case-study focuses on process-based modeling, which is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond.
-Newer models, however, generally incorporate greater complexity (and an ever-growing number of parameters), which can make it difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.
-
+Across these methods, process-based modeling is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond because it is based on mechanistic understanding that should drive accurate out-of-sample predictions.
+%emw24Jul: Maybe here we should say why it's the gold standard?
+See if what I added is okay and below also in a few places, please CHECK!
+...
+This focus on a mechanistic basis, however, generally drives newer models to incorporate greater complexity---and an ever-growing number of parameters.
+This can make it difficult to increase scientific understanding \citep{Franklin2020} and suggest or model potential policies.
 In principle, increasing model complexity can be beneficial by enhancing mechanistic realism, better representing more complex dynamics, and \llabel{quant2}reducing uncertainty by restricting the effective parameter space.
 In practice, however, this is not always the outcome.
-In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}.
+In process-based modeling for climate forecasts, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged over decades of increasing model complexity \citep{Zelinka2020}.
 This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}.
-Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
+Similar concerns arise in ecology, where strong disagreements persist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017} after decades of modeling work.
 These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
 Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
-If a model becomes too complex, it may become a black box,
-and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
+These arguments strss that that the strength of process-based models is their interpretability, and thus their parameters should represent biological processes, and the values of these parameters are of interest.
+If a model becomes too complex, however, it may become a black box, where many parameters are effectively hidden and can take on unreasonable values.
+%emw24Jul: I copied this language from your letter, so you may want to mention that here (very end of letter).
+Further, with so many parameters understanding the sources of uncertainty and how they propagate through the model may becomes nearly impossible (see Box).
 Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
 
-Process-based models used to project species or ecosystem distributions highlight some of these problems.
-Most of the focus is on the model projections---often represented as maps without uncertainties.
+Process-based models used to project species and ecosystem distributions are some of the most important for informing policy, and they highlight many of these problems.
+Most of the focus is on the model projections---often represented as values on maps, presented without uncertainties.
 Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only one of the many different output variables of the model.
-The complexity of these models can make it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability---such work is generally done as a separate effort, exposing potential model problems later on \citep{VanderMeersch2025b}, and not required as a preliminary step.
+The complexity of these models can make it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability.
+Studies of such underlying problems are generally separate; thus potential model problems are usually exposed only later \citep{VanderMeersch2025b}, as opposed to being discovered in preliminary steps before generating forecasts.
 
-Applying the workflow to process-based models would help open the black box.
-Each successive step of model development in the workflow may highlight current problems and a path to solutions.
-Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two, and help expose potential parameter identifiability issues in the model design.
+Applying the workflow to process-based models would help open the black box and thus could identify and fix model problems before forecasting.
+Each successive step of model development in the proposed workflow may highlight problems and a path to solutions.
+Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two, and help expose potential parameter identifiability issues earlier in the model design.
 Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
 The workflow may also highlight model problems by requiring more explicit model calibration (i.e.
-data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model.
-Non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
-The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.
+data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability.
+Once found, non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
+The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating uncertainty through to projections.
 
@@
 
-\centerline{\bf Trend estimation}
+\centerline{\bf Trend estimation} %emw24Jul: I did not have time to stare at the figures to make sure they make sense ... if there are any that need a second set of eyes, let me know and I can. 
 \vspace*{2mm}
@@
 Calibrating submodels separately allows to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.
+%emw24Jul: nice stuff here.
 
@@
 Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge---and where the full model is fitted jointly.
-It would be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimates.
+It would be an opportunity to incoporate mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimates.
 For both trend estimation and forecasting, the workflow would refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model.
@@
 
-When ecologists trained in a limited set of tests require more complex models, they are expected to collaborate with others trained more in model development (though often for highly specific applications, such as wildlife population estimates, where generative models are often predefined).
+When ecologists trained in a limited set of tests require more complex models, they are expected to collaborate with other ecologists \llabel{collab1} trained more in model development (though often for highly specific applications, such as wildlife population estimates, where generative models are often predefined).
 These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models.
-Neither group has fully integrated data simulation into their statistical or scientific workflows.
+None of these groups have fully integrated data simulation into their statistical or scientific workflows.
 Simulation is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data.
@@
 % We suggest unified training in our proposed workflow, or a similar one \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, would focus on learning to generate questions and then models, and then how to simulate data from them.
-\llabel{collab}Although in some cases it may be most expedient for ecologists to collaborate with statisticians who are already well-equipped to adopt our proposed workflow or a similar workflow \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, we recommend that domain scientists who regularly engage in model-driven research efforts receive unified training about how to generate well-formed questions and appropriate models, and then how to simulate data from them.
+We recommend that scientists who regularly engage in model-driven research efforts would benefit from unified training about how to generate well-formed questions and appropriate models, and then how to simulate data from them.
 Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
@@
 Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may see new bridges from their work to empirical data and forecasting.
+\llabel{collab}Such progress could be accelerated through collaboration with statisticians who are already using or otherwise well-equipped to adopt our proposed workflow \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}.
+%emw24Jul: I would say they actually have to be good at doing science to be useful collaborators -- a point Andrew makes often but I think we are tacitly saying that already in second half of this sentence.
 
-Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community.
+%emw24Jul: I am deleting this as it appears IDENTICAL to the below paragraph, but you should check. 
+% Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community.
 First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently, Second, open sharing of more comprehensive study design and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community.
@@
 
-\subsection{[Need a better title] A tractable alternative to machine learning}
+%emw24Jul: Bold proposed changes below.... see what you think. 
+% \subsection{[Need a better title] A tractable alternative to machine learning}
+\emph{Conclusions:}
 
 With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important.
-\llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
+\llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy, especially if the latter lack a robust estimation of their parameters and fall in a complexity trap---a trap with additional costs to interpretability of process-based models.
 Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.
 
-Beyond improving model building and evaluation, our proposed workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
-Process-based models could once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
-Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty.
+%emw24Jul: I feel like we say most of this before, so suggest we cut it. 
+% Beyond improving model building and evaluation, our proposed workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
+% Process-based models could once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
+% Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty.
 Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification---preventing modelers from making biased inferences and unfounded assumptions beyond what the data can support.
@@
 
-\emph{Conclusions:} 
 Today model development in ecology is rarely transparent, which limits how easily the research community can understand models, and thus identify potential issues.

### 2025-07-24 revision/forecastflows_r1.tex
Changes to abstract

@@
 One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, which often include complex, mechanistic sub-models only indirectly informed by data.
-Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions throughout the entire modeling process.
-We propose a principled workflow that can address both of these limitations, by integrating statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process.
-We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could improve our understanding of ecological systems and forecasts for them.
+Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions throughout the modeling process.
+To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting a natural output.
+We show how this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could harmonize currently divided efforts at trend estimation and forecasting to better inform sustainable policies.
 
-%emw24Jul: I pulled out a few of the best sentences that may help the abstract: 
+%emw24Jul: I pulled out a few of the best sentences that may help the abstract: I worked some in, see what you think. 
 % Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date, alongside the skill to forecast future dynamics.

### 2025-07-24 revision/response_r1.tex
Edits done, look for %emw24Jul:

@@
 
+%emw24Jul: Cut the dot dot dot?
 \begin{mybox}
@@
 
-Bayesian models are more and more frequently used in ecology (we cite for example \citealp{Johnson2024}). % I think this one is Bayesian right?
-We clarified that some terms are specific to Bayesian inference but that data simulation still apply for all (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).
+This is a good point, as we agree this could be applied more widely and don't mean to limit the approach to one framework.
+To address this, we have clarified that some terms are specific to Bayesian inference, but that data simulation still apply for all (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).
+% Bayesian models are more and more frequently used in ecology (we cite, for example, \citealp{Johnson2024}).
+% I think this one is Bayesian right?
 
@@
 
-Thank you for this interesting point.
-We have work to clarify the data simulaton step (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}, \href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}), and now highlight one should develop specific posterior predictive checks for the question of interest (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).
-However, we believe an extended discussion of the PPC advantages and inconvenients is outside the word limit we have.
+%emw24Jul: I won't flag all of them, but check for dot dot dot and do a good spellcheck before submitting. 
+Thank you for highlighting this as it does capture some of the complexities of how to apply this workflow.
+To address this, we have worked to clarify the data simulation step (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}, \href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}), and now highlight one should develop specific posterior predictive checks for the question of interest (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).
+As we now state, we believe the best current approach is tailored to the research questions and aims, and thus exactly how conservative posterior predictive checks are may vary in different situations which the researcher should consider.
 
@@
 
-Sorry it was not clear.
+%emw24Jul:  I never use sorry in these letters, it sounds less formal in a bad way to me, but that may be just me. 
+Our apologies that this was not clear.
 Applying the workflow to a process-based model would imply applying it to the model as a whole.
 This would make researchers calibrate the full model jointly, rather than treating submodels separately.
-We clarify this in Box (C).
+We now clarify this in Box (C).
 
@@
 
+%emw24Jul: watch out of too many nominalizations (eg conceptualization) 
 Thank you for this reference.
-We added it and now mention that the conceptualization of explanatory hypothesis has gaining attention in ecology (\href{file:forecastflows_r1\#lintarget:causalinf}{line \ref*{lin:causalinf}}).
+We have added it and now mention the importance of carefully developed hypotheses to drive mathematical model development in ecology (\href{file:forecastflows_r1\#lintarget:causalinf}{line \ref*{lin:causalinf}}).
 
@@
 
-Thank you for spotting these mistakes.
+Thank you for spotting these mistakes. We have corrected them and apologize for the errors.
 
@@
 
+%emw24Jul: If you add the retro check I think this would be a good example, you could also reference the workflow again?
+In fact here and elsewhere perhaps I think we could say in response: Great point, we realized we could also better reference the example workflow we provide as supplemental files, which we now do XX times in the paper.
 ...
@@
 
-...
+... %emw24Jul: Say something here like: Great, as mentioned above we have tried to add more pointers to what we mean and how to apply this workflow through the revised manuscript.
 
@@
 
-Sorry it was not clear. We have clarified this first paragraph.
+Apologies that this was not clear. We have clarified this first paragraph in the revised manuscript.
 
@@
 
-% Hmm, not sure? "For examples using LPI data" and then we talk about several examples right?
-...
+% Hmm, not sure? "For examples using LPI data" and then we talk about several examples right? %emw24Jul: No, 'For example' is the only way we ever say this in English. It's just a phrase. 
+Thanks for catching this. It is fixed now.
 
@@
 Response. %hmm, he's making a good point, maybe we should at least change the section title?
+%emw24Jul: See what you think of my edits! Which I made before reading this (wow).
 
@@
 We also changed the title to reflect the broader inferential goals.
+%emw24Jul: do we mean we changed section title?
+If so, you can ignore my comments in the other file on the paper's title!
 
@@
 
-While we agree that this is an interesting question, we believe it falls outside the scope of this manuscript.
+This is a good point and we could have been clearer on different frameworks that the workflow can be used with.
+We have worked to clarify that in several places in this revision.
+%emw24Jul: Add line numbers?
+I think there are two places we now say this....
+%emw24Jul: try to avoid saying 'not in scope' and 'no space' when possible.
 
@@
 
-We have added an entire paragraph to discuss the methodological aspects, including study/sampling design, that should be more transparently shared along with the use of this workflow (\href{file:forecastflows_r1\#lintarget:studydesignandmore}{line \ref*{lin:studydesignandmore}}).
+This is a great point.
+To address it we have added an entire paragraph that discusses the methodological aspects, including study/sampling design, that should be more transparently shared along with the use of this workflow (\href{file:forecastflows_r1\#lintarget:studydesignandmore}{line \ref*{lin:studydesignandmore}}).
 
@@
 ...
+%emw24Jul: No but this would be a great place be sure to reference the worked example and to say to the reviewer: Good point.
+We provide a worked example, but we realized upon receiving this feedback, that we could better reference the example workflow here and in other places, which we now do.
+...
 
@@
 
-Sorry for this typo.
+Thanks for catching this, our apologies for this typo.
 
@@
 
-We now mention that this step can allow to check our prior assumptions (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}).
+Good point. We now mention that this step can help check prior assumptions (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}).
 
@@
 
-Sorry for this typo.
+Thanks for catching this; it is fixed now.
 
@@
 
-...
+We have edited the text here to clarify.
 
@@
 
-We now mention that the summary statistics should be developped specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).
+We now mention that the summary statistics should be developed specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).
 % I already said this above...
@@
 
-Corrected.
+Yes, and this is now corrected.
 

### 2025-07-24 revision/forecastflows_r1.tex
Edits done, look for %emw24Jul:

@@
 A growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
-For examples using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.
+For example, using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.
 

### 2025-07-24 figures/retrodic/build_tree.R
finally drafting an ugly supp figure

@@ -1,0 +1,34 @@
+rm(list = ls());gc()
+wd <- "~/projects/climategrowthshifts/analysis/pnw"
+setwd(file.path(wd, 'model'))
+util <- new.env()
+source('mcmc_analysis_tools_rstan.R', local=util)
+source('mcmc_visualization_tools.R', local=util)
+setwd(wd)
+
+library(rstan)
+library(dplyr)
+
+# Load treering data
+datasets <- readRDS(file.path(wd, 'input', 'itrdb', 'datasets_summary_usonly.rds'))
+
+# Temporary
+datasets$dataset <- sub("-.*", "", datasets[,'dataset'])
+
+# datasets <- datasets[datasets$last_year >= 1999,] # at least 20 years of observations
+datasets <- datasets[datasets$dataset != 'wa149',] # temporary 
+datasets <- datasets[datasets$dataset != 'ca673',] # temporary
+
+# Temporary, dropping Angiosperms
+todrop <- c('ABLA', 'ABCO', 'ABMA', 'ABPR', 'ABAM', 'ABBI', 'TSME', 'JUSC', 'JUOS', 'JUOC', 'PCEN', 'PCPU',
+            'CADE', 'THPL', 'SEGI', 'CANO', 'CHLA')
+datasets <- datasets[!(datasets$species_code %in% todrop),]
+
+
+source(file.path(wd, 'getphylo.R'))
+
+phy.example <- phy.plants.here
+phy.example[["node.label"]] <- rep("", length(phy.example[["node.label"]])) 
+phy.example[["tip.label"]] <- paste0('sp',  1:length(phy.example[["tip.label"]]))
+
+saveRDS(phy.example, '/home/victor/projects/forecastflows/figures/input/phytree.rds')

### 2025-07-24 figures/retrodic/justanothertest.R

@@ -1,0 +1,137 @@
+wd <- "/home/victor/projects/forecastflows"
+
+library(ape)
+library(phytools)
+library(ggplot2)
+library(rstan)
+rstan_options(auto_write = TRUE)
+
+nspecies <- 80
+nperspecies <- 5
+alpha_root <- 0
+beta_root <- 1.5  
+
+lambda_alpha <- 0.3 
+lambda_beta  <- 0.99
+
+set.seed(12345) #443434
+seeds <- round(runif(100, 1, 9999999),0)
+
+for(s in seeds){
+
+  tree <- pbtree(n = nspecies, scale = 1)
+  # tree[["edge.length"]][1] <- tree[["edge.length"]][1]
+  tree <- force.ultrametric(tree)
+  species <- tree$tip.label
+  plot(tree)
+  
+  corrmat <- vcv(tree, corr = TRUE)
+  
+  corrmat_alpha <- lambda_alpha * corrmat + (1 - lambda_alpha) * diag(nrow(corrmat))
+  corrmat_beta  <- lambda_beta * corrmat + (1 - lambda_beta) * diag(nrow(corrmat))
+  
+  sigma_y <- 0.1
+  sigma_alpha <- 0.3
+  sigma_beta <- 1.5
+  
+  alpha <- MASS::mvrnorm(1, mu = rep(alpha_root, nspecies), Sigma = sigma_alpha^2 * corrmat_alpha)
+  beta  <- MASS::mvrnorm(1, mu = rep(beta_root, nspecies), Sigma = sigma_beta^2 * corrmat_beta)
+  
+  datasim <- data.frame()
+  for(i in 1:nspecies){
+    x <- rnorm(nperspecies, mean = 0, sd = 3)
+    y <- alpha[i] + beta[i] * x + rnorm(nperspecies, mean = 0, sd = sigma_y)
+    datasim <- rbind(
+      datasim,
+      data.frame(speciesname = species[i], species = i, x = x, y = y, alpha = alpha[i], beta = beta[i])
+    )
+  }
+  
+  ggplot(datasim, aes(x = species, y = beta)) +
+    geom_point() +
+    theme_minimal()
+  
+  
+  m <- stan_model(file.path(wd, 'analyses/stan', 'model_onlyspecies.stan'))
+  data <- list(
+    N = nrow(datasim),
+    Nsp = nspecies,
+    y = datasim$y,
+    x = datasim$x,
+    Cphy = corrmat,
[…]

### 2025-07-24 figures/retrodic/laaaastone.R

@@ -1,0 +1,107 @@
+wd <- "/home/victor/projects/forecastflows"
+library(ggplot2)
+library(rstan)
+library(ape)
+library(geiger)
+library(phytools) 
+
+set.seed(19944719)
+tree <- pbtree(n = 10)
+tree <- drop.tip(tree, 't10')
+# tree <- drop.tip(tree, 't3')
+# tree <- drop.tip(tree, 't9')
+tree[["edge.length"]][c(1,10)] <- 20
+tree[["edge.length"]][c(2,11,14)] <- 1
+tree[["edge.length"]][c(15)] <- 0.001
+tree[["edge.length"]][c(16)] <- 0.001
+tree <- force.ultrametric(tree)
+tree <- drop.tip(tree, 't4')
+plot.phylo(tree)
+
+species <- tree$tip.label
+
+nspecies <- length(species)
+alpha_root <- 0
+beta_root <- 1.5  
+lambda_alpha <- 0.99
+lambda_beta  <- 0.99
+
+set.seed(777)
+seeds <- round(runif(100, 1, 9999999),0)
+
+for(s in seeds){
+  set.seed(s)
+  
+  corrmat <- vcv(tree, corr = TRUE)
+  
+  corrmat_alpha <- lambda_alpha * corrmat + (1 - lambda_alpha) * diag(nrow(corrmat))
+  corrmat_beta  <- lambda_beta * corrmat + (1 - lambda_beta) * diag(nrow(corrmat))
+  
+  sigma_y <- 0.1
+  sigma_alpha <- 0.3
+  sigma_beta <- 1.5
+  
+  alpha <- MASS::mvrnorm(1, mu = rep(alpha_root, nspecies), Sigma = sigma_alpha^2 * corrmat_alpha)
+  beta  <- MASS::mvrnorm(1, mu = rep(beta_root, nspecies), Sigma = sigma_beta^2 * corrmat_beta)
+  
+  datasim <- data.frame()
+  spid <- c()
+  for(i in 1:nspecies){
+    # nperspecies <- ifelse(i == 6, 100, 5)
+    nperspecies <- 5
+    x <- rnorm(nperspecies, mean = 0, sd = 3)
+    y <- alpha[i] + beta[i] * x + rnorm(nperspecies, mean = 0, sd = sigma_y)
+    datasim <- rbind(
+      datasim,
[…]

### 2025-07-24 figures/retrodic/last_test.R

@@ -1,0 +1,181 @@
+rm(list = ls());gc()
+if(length(grep("victor", getwd()) > 0)) {
+  wd <- "/home/victor/projects/forecastflows"
+} else if(length(grep("lizzie", getwd()) > 0)) {
+  wd <- "/Users/lizzie/Documents/git/projects/misc/miscmisc/workflowsPhilTrans"
+}  
+library(ggplot2)
+library(rstan)
+set.seed(123456)
+rstan_options(auto_write = TRUE)
+
+phytree <- readRDS(file.path(wd, 'figures/input', 'phytree.rds')) # Lizzie does not have, right?
+phytree[["edge.length"]][39:40] <- phytree[["edge.length"]][39:40]
+phytree <- force.ultrametric(phytree)
+plot(phytree)
+
+if(length(grep("lizzie", getwd()) > 0)){
+  require(ape)
+  require(geiger)
+  require(phytools) 
+  nspecies = 40
+  phytree <- pbtree(n=nspecies, nsim=1, b=1, complete=FALSE,scale=1)
+}
+
+# ------------- #
+# Simulate data #
+# ------------- #
+nspecies <- phytree[["Nnode"]]+1 # no. of species
+nobs_perspecies <- round(runif(nspecies, 5,10)) # no. of different observations per population
+nobs_perspecies[c(17:21)] <- round(runif(5, 90,120)) # some bias
+nobs_perspecies[c(1)] <- round(runif(1, 90,120)) # some bias
+spid_perobs <- rep(1:nspecies, times = nobs_perspecies)
+
+years <- c()
+for(np in nobs_perspecies){
+  years_p <- sample(1900:2020, size = np, replace = FALSE)
+  years <- c(years, years_p)
+}
+
+## species intercepts, with phylogenetic structure
+lambda <- 0.99999
+aroot <- 100
+sigma <- 0.01
+scaledtree_int <- rescale(phytree, model = "lambda", lambda)
+speciesnum <- as.numeric(gsub("sp", "", phytree[["tip.label"]]))
+plot.phylo(scaledtree_int,
+           cex = 1, 
+           tip.color = c("#FF7601", "#00809D")[as.numeric(as.numeric(gsub("sp", "", phytree[["tip.label"]])) %in% 1:6)+1])
+
+mu_alpha_sp_wphylogeny <- fastBM(scaledtree_int, a = aroot, mu = 0, sig2 = sigma ^ 2)
[…]

### 2025-07-24 figures/retrodic/mikefashion/randomseeds.R

@@ -1,0 +1,23 @@
+wd <- '/home/victor/projects/climategrowthshifts/analysis/pnwandmore'
+setwd(file.path(wd, 'model'))
+util <- new.env()
+source('mcmc_analysis_tools_rstan.R', local=util)
+source('mcmc_visualization_tools.R', local=util)
+
+
+wd <- "/home/victor/projects/forecastflows"
+library(ggplot2)
+library(rstan)
+library(phytools)
+library(geiger)
+rstan_options(auto_write = TRUE)
+
+s <- 12345
+
+set.seed(12345678)
+seeds <- round(runif(100, 1, 9999999),0)
+
+for(s in seeds){
+  source(file.path(wd, 'figures/retrodic/mikefashion', 'run_plot.R'))
+}
+

### 2025-07-24 figures/retrodic/mikefashion/run_plot.R

@@ -1,0 +1,205 @@
+nspecies <- phytree[["Nnode"]]+1 # no. of species
+nobs_perspecies <- round(runif(nspecies, 20,30)) # no. of different observations per population
+nobs_perspecies[c(14,20)] <- round(runif(2, 180,221)) # some bias
+# nobs_perspecies[c(1)] <- round(runif(1, 90,120)) # some bias
+spid_perobs <- rep(1:nspecies, times = nobs_perspecies)
+
+years <- c()
+for(np in nobs_perspecies){
+  years_p <- sample(1800:2020, size = np, replace = FALSE)
+  years <- c(years, years_p)
+}
+
+## species intercepts, with phylogenetic structure
+lambda <- 0.7
+aroot <- 5
+sigma <- 1e-1
+scaledtree_int <- rescale(phytree, model = "lambda", lambda)
+speciesnum <- as.numeric(gsub("sp", "", phytree[["tip.label"]]))
+plot.phylo(scaledtree_int,
+           cex = 1, 
+           tip.color = c("#FF7601", "#00809D")[as.numeric(as.numeric(gsub("sp", "", phytree[["tip.label"]])) %in% 1:6)+1])
+mu_alpha_sp_wphylogeny <- fastBM(scaledtree_int, a = aroot, mu = 0, sig2 = sigma ^ 2)
+intercepts <- ggplot(data = data.frame(sp = speciesnum, mu_alpha_sp_wphylogeny)) +
+  geom_boxplot(aes(x = sp, y = mu_alpha_sp_wphylogeny, group = sp, color =  sp %in% 1:6)) +
+  scale_color_manual(values = c("#FF7601", "#00809D")) +
+  theme_classic() +
+  theme(legend.position = 'none')
+
+## slopes with phylogenetic structure
+lambda <- 0.99999
+broot <- -0.1
+sigma <- 5e-3
+scaledtree_slope <- rescale(phytree, model = "lambda", lambda)
+speciesnum <- as.numeric(gsub("sp", "", phytree[["tip.label"]]))
+plot.phylo(scaledtree_slope,
+           cex = 1,
+           tip.color = c("#FF7601", "#00809D")[as.numeric(as.numeric(gsub("sp", "", phytree[["tip.label"]])) %in% 1:6)+1])
+mu_beta_sp_wphylogeny <- fastBM(scaledtree_slope, a = broot, mu = 0, sig2 = sigma ^ 2)
[…]

### 2025-07-24 figures/retrodic/phylo_nophylo_whocares.R

@@ -11,8 +11,8 @@ library(geiger)
 rstan_options(auto_write = TRUE)
 
 phytree <- readRDS(file.path(wd, 'figures/input', 'phytree.rds')) # Lizzie does not have, right?
-# phytree[["edge.length"]][39:40] <- phytree[["edge.length"]][39:40]*100
-# phytree <- force.ultrametric(phytree)
+phytree[["edge.length"]][39:40] <- phytree[["edge.length"]][39:40]*10
+phytree <- force.ultrametric(phytree)
 plot(phytree)
 
 if(length(grep("lizzie", getwd()) > 0)){
@@ -25,14 +25,14 @@ if(length(grep("lizzie", getwd()) > 0)){
 # Simulate data #
 # ------------- #
 
-set.seed(1234567)
+set.seed(12345678)
 seeds <- round(runif(100, 1, 9999999),0)
 
 for(s in seeds){
   set.seed(s)
   
   nspecies <- phytree[["Nnode"]]+1 # no. of species
-  nobs_perspecies <- round(runif(nspecies, 5,10)) # no. of different observations per population
+  nobs_perspecies <- round(runif(nspecies, 20,30)) # no. of different observations per population
   nobs_perspecies[c(14,20)] <- round(runif(2, 180,221)) # some bias
   # nobs_perspecies[c(1)] <- round(runif(1, 90,120)) # some bias
   spid_perobs <- rep(1:nspecies, times = nobs_perspecies)
@@ -60,9 +60,9 @@ for(s in seeds){
     theme(legend.position = 'none')
   
   ## slopes with phylogenetic structure
-  lambda <- 0.95
+  lambda <- 0.99999
   broot <- -0.1
-  sigma <- 5e-3
+  sigma <- 1e-3
   scaledtree_slope <- rescale(phytree, model = "lambda", lambda)
   speciesnum <- as.numeric(gsub("sp", "", phytree[["tip.label"]]))
   plot.phylo(scaledtree_slope,

### 2025-07-24 figures/retrodic/plotme.R

@@ -28,4 +28,4 @@ plotslopes <- ggplot(data = rbind(slopes, slopesphylo)) +
   theme_classic() +
   theme(legend.position = 'none')
 
-ggsave(plotslopes, file = file.path(wd, 'figures','retrodic', 'multisim', paste0(s, '.pdf')))
+ggsave(plotslopes, file = file.path(wd, 'figures','retrodic', 'multisim', paste0(s, '.png')))

### 2025-07-24 figures/retrodic/predpots.R

@@ -1,0 +1,43 @@
+
+
+summ_fit <- data.frame(summary(fit)$summary)
+retrodata <- data.frame(obs = data$y, 
+                        pred = summ_fit[paste0("logy_pred[",1:data$N,"]"), c('X50.')],
+                        predlc = summ_fit[paste0("logy_pred[",1:data$N,"]"), c('X2.5.')],
+                        preduc = summ_fit[paste0("logy_pred[",1:data$N,"]"), c('X97.5.')],
+                        year = data$year,
+                        species = data$spid)
+
+ggplot(data = retrodata) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
+  facet_wrap(~ species, scales = "free") +
+  geom_pointrange(aes(x = log(obs), y = pred, ymin = predlc, ymax = preduc, color = as.character(species)),
+                  size = 0.1, linewidth = 0.2) +
+  theme_classic() +
+  theme(legend.position = 'none')
+
+ggplot(data = retrodata[retrodata$species %in% c(5,12),]) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
+  geom_smooth(aes(x = log(obs), y = pred, color = as.character(species)), method = 'lm', se = FALSE, linewidth = 0.4) +
+  facet_wrap(~ species, scales = "free") +
+  geom_point(aes(x = log(obs), y = pred), size = 3, color = 'white') +
+  geom_pointrange(aes(x = log(obs), y = pred, ymin = predlc, ymax = preduc, color = as.character(species)),
+                  size = 0.1, linewidth = 0.2) +
+  scale_color_manual(values = c('darkblue', 'darkgreen')) +
+  theme_classic() +
+  theme(legend.position = 'none')
+
+
+plot.phylo(phytree, tip.color = c(rep('grey50', 4), 'darkgreen', rep('grey50', 6), 'darkblue', rep('grey50', 8)))
+
+
+ggplot(data = slopes[slopes$species %in% c(5,10),]) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
+  geom_pointrange(aes(ymin = slp_fit.lcl, ymax = slp_fit.ucl, x = true_slopes, 
+                      y = slp_fit, color = species %in% 1:6),
+                  size = 0.2) +
+  theme_classic() +
+  scale_color_manual(values = c("#FF7601", "#00809D")) +
[…]

### 2025-07-24 figures/retrodic/supp_figure.R

@@ -1,0 +1,165 @@
+wd <- "/home/victor/projects/forecastflows"
+library(ggplot2)
+library(patchwork)
+library(rstan)
+library(ape)
+library(geiger)
+library(phytools) 
+
+set.seed(19944719)
+tree <- pbtree(n = 10)
+tree <- drop.tip(tree, 't10')
+# tree <- drop.tip(tree, 't3')
+# tree <- drop.tip(tree, 't9')
+tree[["edge.length"]][c(1,10)] <- 20
+tree[["edge.length"]][c(2,11,14)] <- 1
+tree[["edge.length"]][c(15)] <- 0.001
+tree[["edge.length"]][c(16)] <- 0.001
+tree <- force.ultrametric(tree)
+tree <- drop.tip(tree, 't4')
+plot.phylo(tree)
+
+
+species <- tree$tip.label
+
+nspecies <- length(species)
+alpha_root <- 0
+beta_root <- 1.5  
+lambda_alpha <- 0.99
+lambda_beta  <- 0.99
+
+s <- 3804643
+set.seed(s)
+
+{
+  
+  corrmat <- vcv(tree, corr = TRUE)
+  
+  corrmat_alpha <- lambda_alpha * corrmat + (1 - lambda_alpha) * diag(nrow(corrmat))
+  corrmat_beta  <- lambda_beta * corrmat + (1 - lambda_beta) * diag(nrow(corrmat))
+  
+  sigma_y <- 0.1
+  sigma_alpha <- 0.3
+  sigma_beta <- 1.5
+  
+  alpha <- MASS::mvrnorm(1, mu = rep(alpha_root, nspecies), Sigma = sigma_alpha^2 * corrmat_alpha)
+  beta  <- MASS::mvrnorm(1, mu = rep(beta_root, nspecies), Sigma = sigma_beta^2 * corrmat_beta)
+  
+  datasim <- data.frame()
+  spid <- c()
+  for(i in 1:nspecies){
+    # nperspecies <- ifelse(i == 6, 100, 5)
+    nperspecies <- 5
+    x <- rnorm(nperspecies, mean = 0, sd = 3)
+    y <- alpha[i] + beta[i] * x + rnorm(nperspecies, mean = 0, sd = sigma_y)
+    datasim <- rbind(
+      datasim,
+      data.frame(speciesname = species[i], species = i, x = x, y = y, alpha = alpha[i], beta = beta[i])
+    )
[…]

### 2025-07-24 figures/retrodic/testseed.R

@@ -1,0 +1,257 @@
+rm(list = ls());gc()
+if(length(grep("victor", getwd()) > 0)) {
+  wd <- "/home/victor/projects/forecastflows"
+} else if(length(grep("lizzie", getwd()) > 0)) {
+  wd <- "/Users/lizzie/Documents/git/projects/misc/miscmisc/workflowsPhilTrans"
+}  
+library(ggplot2)
+library(rstan)
+library(phytools)
+library(geiger)
+rstan_options(auto_write = TRUE)
+
+phytree <- readRDS(file.path(wd, 'figures/input', 'phytree.rds')) # Lizzie does not have, right?
+phytree[["edge.length"]][39:40] <- phytree[["edge.length"]][39:40]*10
+phytree <- force.ultrametric(phytree)
+plot(phytree)
+
+if(length(grep("lizzie", getwd()) > 0)){
+  require(ape)
+  nspecies = 40
+  phytree <- pbtree(n=nspecies, nsim=1, b=1, complete=FALSE,scale=1)
+}
+
+# ------------- #
+# Simulate data #
+# ------------- #
+set.seed(s)
+{
+  nspecies <- phytree[["Nnode"]]+1 # no. of species
+  nobs_perspecies <- round(runif(nspecies, 20,30)) # no. of different observations per population
+  nobs_perspecies[c(14,20)] <- round(runif(2, 180,221)) # some bias
+  # nobs_perspecies[c(1)] <- round(runif(1, 90,120)) # some bias
+  spid_perobs <- rep(1:nspecies, times = nobs_perspecies)
+  
+  years <- c()
+  for(np in nobs_perspecies){
+    years_p <- sample(1800:2020, size = np, replace = FALSE)
+    years <- c(years, years_p)
+  }
+  
+  ## species intercepts, with phylogenetic structure
+  lambda <- 0.7
+  aroot <- 5
+  sigma <- 1e-1
+  scaledtree_int <- rescale(phytree, model = "lambda", lambda)
+  speciesnum <- as.numeric(gsub("sp", "", phytree[["tip.label"]]))
+  plot.phylo(scaledtree_int,
+             cex = 1, 
+             tip.color = c("#FF7601", "#00809D")[as.numeric(as.numeric(gsub("sp", "", phytree[["tip.label"]])) %in% 1:6)+1])
[…]

### 2025-07-25 figures/retrodic/simulations_speciesonlyintslope.R
My code that runs but I never got the PPC I wanted, but I will try again!!!

@@ -31,11 +31,12 @@ plot(phytree)
 # ------------- #
 # whichspp <- c(8,15,17,21,28,32,35) # seed(1111) with second trees
 
-whichspp <- c(4, 15:17,20, 25)
+whichphyspp <- c(4, 15:17,20, 25)
+whichspp <- which(as.numeric(gsub("t", "", phytree[["tip.label"]])) %in% whichphyspp)
 
 nspecies <- phytree[["Nnode"]]+1 # no. of species
 nobs_perspecies <- round(runif(nspecies, 5, 15)) # no. of different observations per population
-nobs_perspecies[whichspp] <- round(runif(length(whichspp), 90, 140)) # some bias
+nobs_perspecies[whichspp] <- round(runif(length(whichspp), 90, 140)) # some bias 
 spid_perobs <- rep(1:nspecies, times = nobs_perspecies)
 
 years <- c()
@@ -52,7 +53,7 @@ scaledtree_int <- rescale(phytree, model = "lambda", lambdaint)
 speciesnum <- as.numeric(gsub("t", "", phytree[["tip.label"]]))
 plot.phylo(scaledtree_int,
            cex = 1, 
-           tip.color = c("#FF7601", "#00809D")[as.numeric(as.numeric(gsub("t", "", phytree[["tip.label"]])) %in% whichspp)+1])
+           tip.color = c("#FF7601", "#00809D")[as.numeric(as.numeric(gsub("t", "", phytree[["tip.label"]])) %in% whichphyspp)+1])
 
 mu_alpha_sp_wphylogeny <- fastBM(scaledtree_int, a = brootint, mu = 0, sig2 = sigmaint ^ 2)
 
@@ -99,16 +100,26 @@ datasim <- data.frame(
 )
 
 datasim$logyphylo <- log(datasim$yphylo)
+datasim$whichspp <- NA
+datasim$whichspp[which(datasim$species %in% whichspp)] <- "special"
+datasim$whichspp[which(!datasim$species %in% whichspp)] <- "notspecial"
 
+ggplot() +
+  geom_histogram(aes(x = log(datasim$logyphylo), color=datasim$whichspp), alpha = 0.2)
+
+ggplot(datasim, aes(x=year, y=logyphylo, color=whichspp, group=species)) +
+  geom_point() +
+  facet_wrap(.~species)
+
+if(FALSE){
 dropspp <- c(1:20)
-
 datasimdrop <- data.frame(
[…]

### 2025-07-25 revision/response_r1.tex
last modifications (last, really?) and supp figure

@@
 
-%emw24Jul: Cut the dot dot dot?
-\begin{mybox}
-\emph{...}
-\end{mybox}
-
 We are happy to hear that both reviewers recognized the value of our work and we appreciate the opportunity to revise our manuscript.
@@
 % here I hope we will have a PPC figure
-[Waiting for a nice PPC figure] Response.
+Thank you for this suggestion.
+We have added a figure in the supplementary material to illustrate a retrodictive check with a missing phylogenetic structure.
+We now mention this figure several times throughout the manuscript (\href{file:forecastflows_r1\#lintarget:suppfig1}{lines \ref*{lin:suppfig1}}, \href{file:forecastflows_r1\#lintarget:suppfig2}{\ref*{lin:suppfig2}}, \href{file:forecastflows_r1\#lintarget:suppfig3}{\ref*{lin:suppfig3}}, \href{file:forecastflows_r1\#lintarget:suppfig4}{\ref*{lin:suppfig4}}).
 
@@
 
-%emw24Jul:  I never use sorry in these letters, it sounds less formal in a bad way to me, but that may be just me. 
 Our apologies that this was not clear.
@@
 
-%emw24Jul: watch out of too many nominalizations (eg conceptualization) 
 Thank you for this reference.
@@
 In fact here and elsewhere perhaps I think we could say in response: Great point, we realized we could also better reference the example workflow we provide as supplemental files, which we now do XX times in the paper.
-...
-% Jim: I think the reviewer makes a fair point that this is vague.
-Can we add something here, or even just refer to section 3.2 where (I think) relevant examples are already given in the final paragraph?
+Great point, we realized we could also better reference the example workflow we provide as supplemental files, which we now do (\href{file:forecastflows_r1\#lintarget:exampleworkflow1}{lines \ref*{lin:exampleworkflow1}}, \href{file:forecastflows_r1\#lintarget:exampleworkflow2}{\ref*{lin:exampleworkflow2}}).
+We also illustrate this point in the retrodictive check figure in the supplementary.
 
@@
 
-... %emw24Jul: Say something here like: Great, as mentioned above we have tried to add more pointers to what we mean and how to apply this workflow through the revised manuscript.
+Great, as mentioned above we have tried to add more pointers to what we mean and how to apply this workflow through the revised manuscript.
 
@@
 
-% Hmm, not sure? "For examples using LPI data" and then we talk about several examples right? %emw24Jul: No, 'For example' is the only way we ever say this in English. It's just a phrase. 
 Thanks for catching this. It is fixed now.
@@
 
-Response. %hmm, he's making a good point, maybe we should at least change the section title?
-%emw24Jul: See what you think of my edits! Which I made before reading this (wow).
+We have removed this last section and have expanded our conclusion.
 
@@
 Thank you for this suggestion.
-We added a paragraph in the introduction to discuss the inferential goals of these two approachs (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}).
-We also changed the title to reflect the broader inferential goals.
-%emw24Jul: do we mean we changed section title?
-If so, you can ignore my comments in the other file on the paper's title!
+We added a paragraph in the introduction to discuss the inferential goals of these two approaches (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}).
 
@@
 
-This is a good point and we could have been clearer on different frameworks that the workflow can be used with.
-We have worked to clarify that in several places in this revision.
+This is a good point, and we could have been clearer on different frameworks that the workflow can be used with.
+We have worked to clarify that in several places in this revision (in particular, see Box).
 %emw24Jul: Add line numbers?
@@
 
-We have clarified the data simulaton step.
+We have clarified the data simulation step.
 We believe a first step would be to simulate from one parameter set (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}), and then to repeat the simulations with several parameter sets (\href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}).
@@
 However, in the workflow there is no mention of transparent communication of study design, which if ignored in analysis, could result in discrepancies in results for studies with similar research questions.
-Differences in sampling designs must also be considered prior tointegrating different data sources (if integration is even possible).
+Differences in sampling designs must also be considered prior to integrating different data sources (if integration is even possible).
 The importance of study design/sampling design and its ties to appropriate modeling is not discussed.
-Is itimplicitly assumed that there is alignment between design, question, and model in this workflow?
+Is it implicitly assumed that there is alignment between design, question, and model in this workflow?
 Some mention of the importance of study design would strengthen the work.}
@@
 
-Thank you for this suggestion. We now say that the checks should be developped specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).
+Thank you for this suggestion.
+We now say that the checks should be developed specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}), and now show a simple example of retrodictive check in supplementary material.
 
@@
 
-% Real-data examples... not do-able!
-...
-%emw24Jul: No but this would be a great place be sure to reference the worked example and to say to the reviewer: Good point.
-We provide a worked example, but we realized upon receiving this feedback, that we could better reference the example workflow here and in other places, which we now do.
-...
+Good point.
+We provide a worked example, but we realized upon receiving this feedback, that we could better reference the example workflow here and in other places, which we now do (\href{file:forecastflows_r1\#lintarget:exampleworkflow1}{lines \ref*{lin:exampleworkflow1}}, \href{file:forecastflows_r1\#lintarget:exampleworkflow2}{\ref*{lin:exampleworkflow2}}).
 
@@
 Their parameters should represent biological processes, and their values are of interest.
-The inferential goal of process-based models is both predictive and explanatory (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}), and poor identifiability could reduce this.
+The inferential goal of process-based models is both predictive and explanatory (\href{file:forecastflows_r1\#lintarget:inferentialgoal}{line \ref*{lin:inferentialgoal}}), and poor identifiability could reduce this (\href{file:forecastflows_r1\#lintarget:blackbox}{line \ref*{lin:blackbox}}).
 
-\clearpage
-\bibliography{../forecastflows.bib}
+%\clearpage
+%\bibliography{../forecastflows.bib}
 

### 2025-07-25 revision/forecastflows_r1.tex
last modifications (last, really?) and supp figure

@@
 % \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
-\title{Closing the gap between statistical and scientific workflows for improved inference in ecology} %emw24Jul: Hmm, you say in the letter you changed the title but I am not sure you did. 
+\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} %emw24Jul: Hmm, you say in the letter you changed the title but I am not sure you did. 
 \date{\today}
@@
 We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons.
-First, the research community is split between two disconnected paradigms that prevent uniting a natural connection between trend estimation and forecasting.
+First, the research community is split between two disconnected spheres that prevent uniting a natural connection between trend estimation and forecasting.
 One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, which often include complex, mechanistic sub-models only indirectly informed by data.
@@
 
-%emw24Jul: Same comment as above about 'paradigms' but here I think you mean 'approaches' so I just changed it. 
 \llabel{inferentialgoal}These two modeling approaches may appear to pursue independent inferential goals---explanation vs.\ prediction---but in reality this distinction is blurred.
@@
 Moreover,  estimated trends are frequently treated---implicitly and explicitly---as quantities that project forward into the future, taking on a forecasting role.
-%emw24Jul: Nice!
 In contrast, approaches using complex forecasting models are certainly built for predictive purposes, yet these also have an additional agenda.
-Developing and using mechanistic, process-based models is often justified by the assumption that ``we must understand to predict," and such models are often judged by how faithfully they mimic the real-world processes they represent, which conflates explanation with prediction \citep{Shmueli2010}.
+Developing and using mechanistic, process-based models is often justified by the assumption that ``we must understand to predict", and such models are often judged by how faithfully they mimic the real-world processes they represent, which conflates explanation with prediction \citep{Shmueli2010}.
 
-%emw24Jul: I edited something out that I do and you should avoid (or try to edit out when I add it): adding too many 'subjunctive/conditional' phrases.
-We ARGUE THAT MANY.
-...
-IN LARGE PART mean ...
-try to pick one (of two here) conditional phrase and drop the rest.
 We argue that many current debates and diverging forecasts are driven by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}.
@@
 At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components because the current workflow rarely examines the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).
-Thus it encourages new parts that are often disconnected from the original research aim, its data stream, and the previous scientific insights.
+Thus, it encourages new parts that are often disconnected from the original research aim, its data stream, and the previous scientific insights.
 
@@
 
-%emw24Jul: Nice stuff here!
 This reality should drive researchers to use more robust and coherent methods.
@@
 This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, figure \ref{fig:modeldata}d).
-\llabel{causalinf}This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}.
+\llabel{causalinf} This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}.
 From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}).
-Generally, starting with a relatively simple model that we could refine later makes understanding the model and how it interacts with empirical data easier (see example workflow we provide).
+Generally, starting with a relatively simple model that we could refine later makes understanding the model and how it interacts with empirical data easier \llabel{exampleworkflow1}(see example workflow we provide as a supplementary file).
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
-This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).
-%emw24Jul: 'see example workflow we provide' and 'see example workflow' -- can we standardize our mention of this and make it clear where it is?
-Like: `see example workflow we provide as supplement files.'
+This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterization).
 
@@
 At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not.
-This is also a step for making sure the model is working as expected (see example worklow).
-\llabel{repeatsim}Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range.
-\llabel{priorPC}In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}), while in other frameworks they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).
+This is also a step for making sure the model is working as expected (see example workflow).
+\llabel{repeatsim} Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range.
+\llabel{priorPC} In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}), while in other frameworks they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).
 
@@
 These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}).
-This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations.
-\llabel{howtoPPC}This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regards to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
+This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations \llabel{suppfig1}(see supplementary figure 1).
+\llabel{howtoPPC} This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regard to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
 It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
@@
 For example, in macroecological studies model building is often informed by patterns in the data.
-The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting.
-%emw24Jul: possible to ref Fig 1 here?
+The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting (figure 1).
 
@@
 The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
-Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
-flagging the need to change the model structure---before incorporating empirical observations. 
+Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated), flagging the need to change the model structure---before incorporating empirical observations.
 The second feedback loop comes from the retrodictive check. 
@@
 We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow.
-Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g.
+Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see supplementary figure 1) or observational biases (e.g.
 unbalanced data).
@@
 At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.
-\llabel{broadcontext}These feedback loops are applied currently in only in a Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.
+\llabel{broadcontext}These feedback loops are applied currently in only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.
 
@@
 
-%emw24Jul: I removed case study from this draft, I think they're more than case studies and the term no longer helps the paper.
 Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to accurately \llabel{quant1}understand and manage uncertainties and yield new scientific insights.
@@
 
-%emw24Jul: Cite supp retro check figure here below? 
 We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
@@
 Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected.
-But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model.
+But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model \llabel{suppfig3}(see supplementary figure 1).
 
@@
 From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or `synthetic data generation' currently \citep{Buschke2021,mcrae2025utility}, but could be built into the models fit to the empirical data.
-Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends (see workflow example).
+Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model \llabel{suppfig4}(see supplementary figure 1) and, ultimately, this would help inform our global estimates of mean trends \llabel{exampleworkflow2}(see example workflow we provide as a supplement file).
 For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled.
@@
 Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
-These arguments strss that that the strength of process-based models is their interpretability, and thus their parameters should represent biological processes, and the values of these parameters are of interest.
-If a model becomes too complex, however, it may become a black box, where many parameters are effectively hidden and can take on unreasonable values.
-%emw24Jul: I copied this language from your letter, so you may want to mention that here (very end of letter).
-Further, with so many parameters understanding the sources of uncertainty and how they propagate through the model may becomes nearly impossible (see Box).
+These arguments stress that the strength of process-based models is their interpretability, and thus their parameters should represent biological processes, and the values of these parameters are of interest.
+If a model becomes too complex, however, it may become a black box, \llabel{blackbox}where many parameters are effectively hidden and can take on unreasonable values. 
+Further, with so many parameters understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
 Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
@@
 Most of the focus is on the model projections---often represented as values on maps, presented without uncertainties.
-Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only one of the many different output variables of the model.
+Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only one of the many output variables of the model.
 The complexity of these models can make it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability.
@@
 % In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data.
-In the current workflow for estimating trends over time, a model is typically chosen a priori to fit to data, yielding parameter estimates that are treated as true and often translated directly into published findings and conclusions, without interrogating the model using simulated data and carefully evaluating its performance.
+In the current workflow for estimating trends over time, a model is typically chosen a priori to fit to data, yielding parameter estimates that are treated as true and often translated directly into published findings and conclusions, without interrogating the model using simulated data and carefully evaluating its performance (see example workflow we provide as a supplementary file).
 Any such analysis must be treated with caution, because different model variants can yield very different estimates (hence different conclusions), and published results may not prove robust when confronted with future reanalysis of the same data.
@@
 Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge---and where the full model is fitted jointly.
-It would be an opportunity to incoporate mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimates.
+It would be an opportunity to incorporate mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimates.
 For both trend estimation and forecasting, the workflow would refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model.

### 2025-07-25 revision/supplementary_figure.tex
last modifications (last, really?) and supp figure

\documentclass[11pt]{article}
\usepackage[top=0.20in, bottom=.20in, left=0.5in, right=0.5in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage{tcolorbox}
\usepackage{caption}
\usepackage{todonotes}
\usepackage[font={sf,footnotesize}]{caption}
\usepackage{hyperref}
\newcommand{\llabel}[1]{\hypertarget{lintarget:#1}{}\linelabel{lin:#1}}

\usepackage{lineno}
\renewcommand\linenumberfont{\normalfont\tiny\color{gray}}

\bibliographystyle{besjournals}

\begin{document}

% \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
\title{\large Closing the gap between statistical and scientific workflows for improved forecasts in ecology\\Supplementary material}
\date{}
\author{\small Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
\maketitle
{\footnotesize \noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}

\renewcommand{\thefigure}{S\arabic{figure}}

\begin{figure}[b]
	\centering
    \vspace*{-20cm}
	\includegraphics{../figures/retrodic/retrodictive_check.pdf}
	\caption{A simple illustration of a retrodictive check to highlight a missing structure in a model. Here, we consider ``observations'' simulated from a linear regression model ($y_\text{obs}$), for 8 different species. The species-specific intercepts and slopes are simulated with a phylogenetic structure, i.e. the correlation between two species-specific parameters is proportional to the amount of evolutionary history (as defined by a simulated phylogenetic tree, shown in the left and right parts of the figure) shared by the two species. As a first approach, we use a model with partial pooling on the intercepts and the slopes---assuming no correlation structure---to fit the data. We can simulate data from the fitted model ($y_\text{pred}$), and compare the retrodictions to the observations---i.e. doing a retrodictive check. The check has to be tailored in regard to our research question and our domain expertise: here, we might believe that some species are more closely related than others, and we want to visually check whether we find some apparent distinctions between species groups (``clades''). In the left part of the figure, we find some (small) symptoms of a missing group structure: the upper clade residuals tend to have a positive bias, while the lower clade ones have a negative bias. This retrodictive check indicates that we could refine our model, using the same phylogenetic correlation structure that we used to simulate data (i.e. the same phylogenetic tree). With this new refined model, we can perform a new retrodictive check (right part of the figure): the difference of residual biases between the two distinct clades now appears smaller. \\
	Interestingly, when looking at the interquartile range of these residuals (middle part of the figure), both models (with and without phylogenetic structure) seem less distinct. This illustrates that we may need to spend more time to build a more adapted retrodictive check. The addition of correlation structure in our second model should indeed mostly result in a reduction in species-specific parameter uncertainty rather than an important shift to the $y_\text{pred}$ themselves. One might need to develop a smarter check, for example looking at empirical covariance between species two by two.}
	\label{fig:modeldata}
	\vspace*{2cm}
\end{figure}

\end{document}

### 2025-07-25 revision/supplementary_figure.tex
Some edits to supp figure text and a comment

@@
 	\caption{A simple illustration of a retrodictive check to highlight a missing structure in a model.
-Here, we consider ``observations'' simulated from a linear regression model ($y_\text{obs}$), for 8 different species.
-The species-specific intercepts and slopes are simulated with a phylogenetic structure, i.e.
+Here, we consider ``observations'' simulated from a linear regression model ($y_\text{obs}$) for 8 different species as an example of what could happen using empirical data on time-series observations of abundance for multiple species.
+In this example, we simulated species-specific intercepts and slopes with a phylogenetic structure, i.e.
 the correlation between two species-specific parameters is proportional to the amount of evolutionary history (as defined by a simulated phylogenetic tree, shown in the left and right parts of the figure) shared by the two species.
-As a first approach, we use a model with partial pooling on the intercepts and the slopes---assuming no correlation structure---to fit the data.
-We can simulate data from the fitted model ($y_\text{pred}$), and compare the retrodictions to the observations---i.e.
+
+	As a first approach we may often take in model-fitting of such data, we used a model with partial pooling on the intercepts and the slopes---assuming no correlation structure.
+We then simulated data from the fitted model ($y_\text{pred}$) and compared the retrodictions to the observations---i.e.
 doing a retrodictive check.
-The check has to be tailored in regard to our research question and our domain expertise: here, we might believe that some species are more closely related than others, and we want to visually check whether we find some apparent distinctions between species groups (``clades'').
+The check has to be tailored in regard to our research question and our domain expertise: here, we believe that the shared evolutionary history of species could shape their responses and thus we want to visually check whether we find any apparent differences between species groups (``clades'').
 In the left part of the figure, we find some (small) symptoms of a missing group structure: the upper clade residuals tend to have a positive bias, while the lower clade ones have a negative bias.
-This retrodictive check indicates that we could refine our model, using the same phylogenetic correlation structure that we used to simulate data (i.e.
-the same phylogenetic tree).
+This retrodictive check indicates that we could refine our model, using the phylogenetic correlation structure (in this case, because we have simulated this example, we know this structure, provided by the phylogenetic tree we used, is relevant).
 With this new refined model, we can perform a new retrodictive check (right part of the figure): the difference of residual biases between the two distinct clades now appears smaller.
 \\
-	Interestingly, when looking at the interquartile range of these residuals (middle part of the figure), both models (with and without phylogenetic structure) seem less distinct.
-This illustrates that we may need to spend more time to build a more adapted retrodictive check.
+	Interestingly, when looking at the interquartile range of these residuals (middle part of the figure in cyan and blue, with lines representing IQR and dots the medians), both models (with and without phylogenetic structure) seem less distinct.
+This illustrates that we may want to consider building a more adapted retrodictive check.
 The addition of correlation structure in our second model should indeed mostly result in a reduction in species-specific parameter uncertainty rather than an important shift to the $y_\text{pred}$ themselves.
-One might need to develop a smarter check, for example looking at empirical covariance between species two by two.}
+One might need to develop a smarter check, for example looking at empirical covariance between species two by two.} %emwJul24 -- not sure about last two sentences, could be hard for anyone but us to follow?
 	\label{fig:modeldata}

### 2025-07-25 revision/supplementary_figure.tex
added slopes in the supp figure

@@
     \vspace*{-20cm}
-	\includegraphics{../figures/retrodic/retrodictive_check.pdf}
+	\includegraphics{../figures/retrodic/retrodictive_check_slopes.pdf}
 	\caption{A simple illustration of a retrodictive check to highlight a missing structure in a model.
@@
 the correlation between two species-specific parameters is proportional to the amount of evolutionary history (as defined by a simulated phylogenetic tree, shown in the left and right parts of the figure) shared by the two species.
-
+\\
 	As a first approach we may often take in model-fitting of such data, we used a model with partial pooling on the intercepts and the slopes---assuming no correlation structure.
@@
 doing a retrodictive check.
-The check has to be tailored in regard to our research question and our domain expertise: here, we believe that the shared evolutionary history of species could shape their responses and thus we want to visually check whether we find any apparent differences between species groups (``clades'').
+The check has to be tailored in regard to our research question and our domain expertise: here, we believe that the shared evolutionary history of species could shape their responses, and thus we want to visually check whether we find any apparent differences between species groups (``clades'').
 In the left part of the figure, we find some (small) symptoms of a missing group structure: the upper clade residuals tend to have a positive bias, while the lower clade ones have a negative bias.
@@
 \\
-	Interestingly, when looking at the interquartile range of these residuals (middle part of the figure in cyan and blue, with lines representing IQR and dots the medians), both models (with and without phylogenetic structure) seem less distinct.
-This illustrates that we may want to consider building a more adapted retrodictive check.
-The addition of correlation structure in our second model should indeed mostly result in a reduction in species-specific parameter uncertainty rather than an important shift to the $y_\text{pred}$ themselves.
-One might need to develop a smarter check, for example looking at empirical covariance between species two by two.} %emwJul24 -- not sure about last two sentences, could be hard for anyone but us to follow?
+	However, we do not observe an important shift in the residuals (i.e.in the $y_\text{pred}$ themselves).
+Rather, the addition of correlation structure in our second model mostly result in a reduction in species-specific parameter uncertainty (middle part of the figure in cyan and blue, with lines representing IQR and dots the medians).}
 	\label{fig:modeldata}

### 2025-09-18 revision/r1/response_r1.tex
working on revisions

--- 2025-07-25 revision/response_r1.tex
@@
 To address this, we have worked to clarify the data simulation step (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}, \href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}), and now highlight one should develop specific posterior predictive checks for the question of interest (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}).
-As we now state, we believe the best current approach is tailored to the research questions and aims, and thus exactly how conservative posterior predictive checks are may vary in different situations which the researcher should consider.
+As we now state, we believe the best current approach is tailored to the research questions and aims, and thus exactly how conservative posterior predictive checks are may vary in different situations which the researcher should consider (see also supplementary figure S1).
 
@@
 Thank you for this suggestion.
-We now say that the checks should be developed specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}), and now show a simple example of retrodictive check in supplementary material.
+We now say that the checks should be developed specifically for the research question (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}), and now show a simple example of retrodictive check in supplementary material (figure S1).
 

### 2025-09-18 revision/r2/response_r2.tex
working on revisions

\documentclass[11pt,letter]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{graphicx} % Required for inserting images
\usepackage{xcolor} 
\definecolor{darkgrey}{HTML}{4d4d4d}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{gensymb}
\usepackage{xr-hyper}
\makeatletter
  \long\def\myempty{}
  \def\XR@addURL#1{\XR@@dURL#1\myempty{}{}{}{}{}\\}
  \def\XR@@dURL#1#2#3#4#5#6#7\\{%
    {#1}{#2}%
    \ifx\myempty#6\@empty
      {#3}{#4}{\XR@URL}%
    \else
    \fi
}
\makeatother
\usepackage{hyperref}
\hypersetup{hidelinks,breaklinks=true}
\usepackage{cleveref}
\usepackage[labelfont=bf]{caption}
\bibliographystyle{besjournals}

\RequirePackage[labelfont={bf,sf},%
                font={small, sf}]{caption}

\RequirePackage[labelfont={bf,sf},%
                font={small, sf}]{caption}

\externaldocument{forecastflows_r1}
\newcommand{\lref}[1]{}
% (\href{file:forecastflows_r1\#lintarget:SOMELABEL}{line \ref*{lin:SOMELABEL}})

\DeclareEmphSequence{\itshape\color{darkgrey}}
\usepackage[most]{tcolorbox}
% \newtcolorbox{mybox}[1][]{%
%     blanker, 
%     left skip=0em,
%     left=1em,
%     borderline west={1pt}{2pt}{darkgrey},
%     breakable,
%     #1}
\newtcolorbox{mybox}[1][]{%
    blanker, 
    left skip=0em,
    left=0em,
    borderline west={0pt}{0pt}{white},
    breakable,
    #1}

\begin{document}
\setlength{\parindent}{0cm}
\setlength{\parskip}{7pt}

Editor and reviewer comments (we provide below the full context of the two reviewers' comments) are in \emph{italics}, while our responses are in regular text. \\

{\bf Response to editor's comments:}

We are happy to hear that the three reviewers recognized the value of our work and that the manuscript is ready for publication. We have worked to fix the last small points raised by the reviewers, and we would like to thank them again for their careful reading. Furthermore, we also particularly appreciate the really interesting perspective on the policy implications from M. Authier, and...

{\bf Response to reviewers' comments:}

{\bf Reviewer 1}

\begin{mybox}
\emph{The authors have addressed all of my concerns and I believe the MS is
(nearly) ready for publication. I suggest one more pass for copy editing
to fix small mistakes like inconistencies for example in "figure"
"Figure" and "e.g." "e.g.,".}
\end{mybox}

We thank the reviewer for their positive comments regarding our manuscript. 
We have worked to correct these inconsistencies.

{\bf Reviewer 2}

\begin{mybox}
\emph{All of my major comments were sufficiently addressed in this revision of
the manuscript. I appreciate that the supplement is now referenced in
the main document (I missed it on my last round of review). I have a few
editorial comments for the main text and a few minor comments about the
supplement, but I feel this manuscript is almost ready for publication.}
\end{mybox}

...

\begin{mybox}
\emph{\textbf{Manuscript Minor comments:}\\
Abstract: potential typo in second-to-last sentence. Should "where
forecasting a natural output" be "where forecasting is a natural output"?}
\end{mybox}
% vvdm18sept: is really a typo? I thought it was just a way to say it 
...

\begin{mybox}
\emph{Line 39: should "ignore" be "ignores"}
\end{mybox}

Thanks for catching this, it is fixed now.

\begin{mybox}
\emph{Line 51: "hypotheses" or "a hypothesis" rather than "hypothesis"}
\end{mybox}

Corrected.

\begin{mybox}
\emph{Line 138: there appears to be an extra "in" in "in only in"}
\end{mybox}

Corrected.

\begin{mybox}
\emph{Line 215: comma after parameters}
\end{mybox}

Added.

\begin{mybox}
\emph{A couple typos in Box e.g., "calibrating submodes separately allows to
avoid" missing "us"?}
\end{mybox}

% vvdm18sept: I'm not sure about this one? "Calibrating submodels separately allows to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another" seems fine to me...
...

\begin{mybox}
\emph{\textbf{Supplement Minor comments:}\\
Although the supplement works through an example far simpler than the
types of research questions discussed in the main text, it serves as a
reasonable (and accessible) example of how the workflow could be
implemented. There is a brief discussion towards the end of the
supplement about the "scalability" of the proposed workflow, but it
leaves something to be desired. Have the authors applied this workflow
(or pieces of it) to examples that are more complex that could be cited,
or is that future work?}
\end{mybox}

...

\begin{mybox}
\emph{First green box: typo in RQ.."What is global trend..." should read
"what is the global trend..."}
\end{mybox}

Thank you, corrected.

\begin{mybox}
\emph{Step 2: so assuming populations are exchangeable? Is this reasonable?}
\end{mybox}

We have edited the text here to clarify this assumption:
\begin{quote}
``In this first hierarchical model, we thus assume that populations are exchangeable---an assumption we might refine later."
\end{quote}

\begin{mybox}
\emph{Step 2: priors on sigma\_alpha,sp and sigma\_beta\_sp do not seem
appropriate, these parameters have positive support, so should these be
half-normals (folded normals)?}
\end{mybox}

...

\begin{mybox}
\emph{Stan model not included so could not be assessed (model1\_nc2.stan), I
recommend sharing for transparency.}
\end{mybox}

Sorry we forgot to share the Stan code. We added the code in the workflow example.

\begin{mybox}
\emph{Figure 15, unclear what the posterior predictive distribution is of,
this detail could be added for transparency. Could also add many
posterior predictive datasets to the same plot and then overlay the
observed data to facilitate easier comparisons between y-tilde and y-obs.}
\end{mybox}

% hmm

\end{document}

### 2025-09-18 revision/r1/forecastflows_r1.tex
working on revisions

--- 2025-07-25 revision/forecastflows_r1.tex
@@
 
-% \title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} 
-\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology} %emw24Jul: Hmm, you say in the letter you changed the title but I am not sure you did. 
+\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology}
 \date{\today}
@@
 We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
-Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models.
-Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions.
+Concerns about biodiversity loss and climate change have led to greater demands for ecological models.
+Datasets for developing these models have increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions.
 New research often suggests that models accounting for these complexities can yield more accurate trends and predictions.
-%emw24Jul: Are you sure you want 'paradigm'?
-I love the word and the sentence but I think we mean more two 'groups' or 'spheres' -- I would switch and then carry the language you use to the next sentence.
-We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons.
-First, the research community is split between two disconnected spheres that prevent uniting a natural connection between trend estimation and forecasting.
-One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, which often include complex, mechanistic sub-models only indirectly informed by data.
-Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions throughout the modeling process.
+We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models.
+First, the research community is split between two spheres that prevent uniting a connection between trend estimation and forecasting.
+One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, often including complex, mechanistic sub-models indirectly informed by data.
+Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions.
 To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting a natural output.
@@
 % Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
-
-% jdavies15april: I feel this could be a critical step somehow - sharing simulations could allow cross-validation between models or perhaps cross-calibration ...
-% These problems stem in part from continuing gaps between statistical workflows---where the data processing and model development are often addressed separately from the ecological question and aim---and scientific workflows, where all steps are integrated.
-Yet, as ecologists become increasingly computational the opportunity to close this gap has never been greater.
-We outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights for both trend estimation and forecasting.
-%Combining these changes with more open model and data sharing---and developing new efforts to race the same data---could be transformative for ecological forecasting. 
-%A shift toward universal training in a more robust model building could bridge the gap between process-based and statistical approaches and be transformative for ecological modeling.
 \end{abstract}
@@
 Across these methods, process-based modeling is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond because it is based on mechanistic understanding that should drive accurate out-of-sample predictions.
-%emw24Jul: Maybe here we should say why it's the gold standard?
-See if what I added is okay and below also in a few places, please CHECK!
-...
 This focus on a mechanistic basis, however, generally drives newer models to incorporate greater complexity---and an ever-growing number of parameters.
@@
 
-\centerline{\bf Trend estimation} %emw24Jul: I did not have time to stare at the figures to make sure they make sense ... if there are any that need a second set of eyes, let me know and I can. 
+\centerline{\bf Trend estimation} 
 \vspace*{2mm}
@@
 
-\centerline{\bf A common workflow to bridge trend estimation and forecasting} % jdavies15april: would it be possible to label this panel with components of the LPI to provide a clearer illustration of how the framework is implemented?
-% [FLAG: not sure I understood]
+\centerline{\bf A common workflow to bridge trend estimation and forecasting} 
 \vspace*{-3mm}
@@
 
-%emw24Jul: I am deleting this as it appears IDENTICAL to the below paragraph, but you should check. 
-% Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community.
-First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently, Second, open sharing of more comprehensive study design and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community.
-Third, where feasible, alignment on common datasets would provide a standardized basis for model development and comparison, reducing the fragmentation of efforts and facilitating more direct and meaningful evaluations of different modeling approaches.
-Coupled with universal training, these shifts in the sharing and standardization of data and models would create a more collaborative, efficient, and rigorous environment for ecological modeling, ultimately leading to more accurate forecasts and better understanding of complex environmental systems.
-
 % New paragraph by Jim!

### 2025-09-18 revision/r2/forecastflows_r2.tex
working on revisions

--- 2025-09-18 revision/r1/forecastflows_r1.tex
@@
 Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting.
-At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components because the current workflow rarely examines the model as a functioning whole and thus ignore major problems (e.g., non-identifiability, discussed below).
+At the same time, process-based models developed for forecasting often evolve through the addition of new separate layers or components because the current workflow rarely examines the model as a functioning whole and thus ignores major problems (e.g.
+non-identifiability, discussed below).
 Thus, it encourages new parts that are often disconnected from the original research aim, its data stream, and the previous scientific insights.
@@
 	\centering
-	\includegraphics{../figures/figure_modeldataspaces_revised}
+	\includegraphics{../../figures/figure_modeldataspaces_revised}
 	\caption{Different trajectories of current (b-c) and proposed (d) workflows through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
@@
 
-Quantitative science relies on a model-based framework to confront hypothesis with data \citep{Chamberlin:1965cd}.
+Quantitative science relies on a model-based framework to confront hypotheses with data \citep{Chamberlin:1965cd}.
 In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses.
@@
 Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e.
-post-model/post-data, figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}b).
-For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data).
+post-model/post-data, Figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Figure \ref{fig:modeldata}b).
+For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (Figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data).
 Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
@@
 
-We argue a workflow that moves \llabel{quad4}along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}d) could reduce many of these problems and thus improve ecological science.
-The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}).
+We argue a workflow that moves \llabel{quad4}along the data-model space in a coherent sequence of steps (Figure \ref{fig:modeldata}d) could reduce many of these problems and thus improve ecological science.
+The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, Figure \ref{fig:workflow}).
 This involves making clear assumptions about the most influential drivers, within the specific context of our study.
-This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, figure \ref{fig:modeldata}d).
+This should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error (\llabel{quad5}pre-model/pre-data, Figure \ref{fig:modeldata}d).
 \llabel{causalinf} This step of carefully formulating hypotheses has gained increasing attention in the ecological literature \citep{Grace2020}.
-From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}).
+From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, Figure \ref{fig:workflow}).
 Generally, starting with a relatively simple model that we could refine later makes understanding the model and how it interacts with empirical data easier \llabel{exampleworkflow1}(see example workflow we provide as a supplementary file).
@@
 
-With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}).
+With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, Figure \ref{fig:workflow}).
 `Fake' or `test' data are generated directly from the model by fixing parameters \llabel{firstsim}to some reasonable values, which is straightforward if the parameters are interpretable, and from `fake' (simulated) predictor data.
-We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, figure \ref{fig:modeldata}d).
+We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values (\llabel{quad6}post-model/pre-data, Figure \ref{fig:modeldata}d).
 At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is realistic or not.
@@
     \hspace*{-1.5cm}
-	\includegraphics{../figures/figure_worflow_wsteps}
+	\includegraphics{../../figures/figure_worflow_wsteps}
 	\caption{The workflow we propose here, which builds from recent advances in workflows \citep[][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024}, focuses on iterative feedbacks between the research question, model and data.
@@
 
-Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}).
-This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, figure \ref{fig:modeldata}d).
-These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}).
-This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations \llabel{suppfig1}(see supplementary figure 1).
+Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, Figure \ref{fig:workflow}).
+This way, we obtain parameter estimates constrained by observations (\llabel{quad7}post-model/post-data, Figure \ref{fig:modeldata}d).
+These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, Figure \ref{fig:workflow}).
+This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations \llabel{suppfig1}(see Supplementary Figure 1).
 \llabel{howtoPPC} This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regard to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
@@
 For example, in macroecological studies model building is often informed by patterns in the data.
-The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting (figure 1).
+The adoption of this workflow would make the exploratory stage more transparent---as an explicit preliminary step for entering the workflow---and would compel researchers to more clearly develop a research question before extensive model fitting (Figure 1).
 
@@
 We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow.
-Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see supplementary figure 1) or observational biases (e.g.
+Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see Supplementary Figure 1) or observational biases (e.g.
 unbalanced data).
@@
 At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.
-\llabel{broadcontext}These feedback loops are applied currently in only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.
+\llabel{broadcontext}These feedback loops are applied currently only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.
 
@@
 Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected.
-But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model \llabel{suppfig3}(see supplementary figure 1).
+But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model \llabel{suppfig3}(see Supplementary Figure 1).
 
@@
 From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or `synthetic data generation' currently \citep{Buschke2021,mcrae2025utility}, but could be built into the models fit to the empirical data.
-Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model \llabel{suppfig4}(see supplementary figure 1) and, ultimately, this would help inform our global estimates of mean trends \llabel{exampleworkflow2}(see example workflow we provide as a supplement file).
-For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled.
+Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model \llabel{suppfig4}(see Supplementary Figure 1) and, ultimately, this would help inform our global estimates of mean trends \llabel{exampleworkflow2}(see example workflow we provide as a supplement file).
+For example, certain populations are recovering for very specific reasons (e.g.
+elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled.
 From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.
@@
 If a model becomes too complex, however, it may become a black box, \llabel{blackbox}where many parameters are effectively hidden and can take on unreasonable values. 
-Further, with so many parameters understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
+Further, with so many parameters, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
 Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
@@
 data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability.
-Once found, non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
+Once found, non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g.
+narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
 The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating uncertainty through to projections.
@@
 \begin{minipage}[t]{\linewidth}
-    \includegraphics[width=\linewidth]{../figures/trendestimation_details}
+    \includegraphics[width=\linewidth]{../../figures/trendestimation_details}
 
@@
 In the current workflow for estimating trends over time, a model is typically chosen a priori to fit to data, yielding parameter estimates that are treated as true and often translated directly into published findings and conclusions, without interrogating the model using simulated data and carefully evaluating its performance (see example workflow we provide as a supplementary file).
-Any such analysis must be treated with caution, because different model variants can yield very different estimates (hence different conclusions), and published results may not prove robust when confronted with future reanalysis of the same data.
+Any such analysis must be treated with caution, because different model variants can yield very different estimates (hence different conclusions), and published results may not prove robust when confronted with future reanalysis of the same data (see Figure A above).
 The Living Planet Index (LPI, \url{www.livingplanetindex.org}), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results.
@@
 \begin{minipage}[t]{\linewidth}
-    \includegraphics[width=\linewidth]{../figures/forecasting_details}
+    \includegraphics[width=\linewidth]{../../figures/forecasting_details}
 
@@
 The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred.
-Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see figure B above).
+Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see Figure B above).
 Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory.
@@
 	\vfill
-\includegraphics[width=\linewidth]{../figures/iterativeworkflow_details_revised}
+\includegraphics[width=\linewidth]{../../figures/iterativeworkflow_details_revised}
 \vfill
@@
 Much of ecology still divides the world into training for those who gather data and learn a limited set of pre-built models versus those who develop more complex models.
-In ecological training today, researchers who conduct field and lab studies often learn a limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi-squared test).
+In ecological training today, researchers who conduct field and lab studies often learn a limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g.
+categorical $x$ and $y$ leads to using a chi-squared test).
 
@@
 We recommend that scientists who regularly engage in model-driven research efforts would benefit from unified training about how to generate well-formed questions and appropriate models, and then how to simulate data from them.
-Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
+Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g.
+management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
 Empiricists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models.
@@
 \clearpage
-\bibliography{../forecastflows.bib}
+\bibliography{../../forecastflows.bib}
 

### 2025-09-19 revision/r2/response_r2.tex
starting to answer the very interesting review of referee 3

@@
 
-\externaldocument{forecastflows_r1}
+\externaldocument{forecastflows_r2}
 \newcommand{\lref}[1]{}
@@
 
-...
+We are glad that the reviewer found the manuscript has improved.
+Thank you for this last feedback.
 
@@
 
-% vvdm18sept: I'm not sure about this one?
-"Calibrating submodels separately allows to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another" seems fine to me...
-...
+Changed to: ``Calibrating submodels separately makes possible to avoid the fact that [...]''
 
@@
 implemented. There is a brief discussion towards the end of the
-supplement about the "scalability" of the proposed workflow, but it
+supplement about the ``scalability" of the proposed workflow, but it
 leaves something to be desired. Have the authors applied this workflow
@@
 \begin{mybox}
-\emph{First green box: typo in RQ.."What is global trend..." should read
-"what is the global trend..."}
+\emph{First green box: typo in RQ..``What is global trend..." should read
+``what is the global trend..."}
 \end{mybox}
@@
 
-% hmm
+% Hmm I don't understand?
+
+{\bf Reviewer 3}
+
+\begin{mybox}
+\emph{First, I would like to congratulate the authors for a very clear and well written manuscript.
+The main thesis of the authors is to outline modelling workflows and illustrate in broad brush strikes how to bridge the current gap in ecological forecasting with simulations of synthetic data among the several steps of model building (the pre-data/post-model third corner in their Fig.
+1) to improve confidence in modelling, inference and prediction.
+This is an important contribution, one that echoes many early calls (e.g.
+Thomas 1997): our community in ecology, despite its beefing up with respect to quantitative skills, remains fragmented and disjoint.
+I thus wholeheartedly endorse the main thesis of the authors and agree with it about research.
+There remains one concern that I would like to discuss with the authors, and possibly hear back from them although it is somewhat minor with respect to research but more problematic with respect to policy.
+It should not deflect from publication of the article in Philosophical Transactions of the Royal Society A.}
+\end{mybox}
+
+We are glad that the reviewer found our manuscript well-written, and that he endorses our main messages.
+We are particularly grateful for his detailed comments, with much broader perspectives than just our manuscript.
+We have tried to address most of the comments, but we believe that they deserve longer discussions and could constitute in themselves another paper.
+
+\begin{mybox}
+\emph{The authors have addressed comments from the first round of reviewing.
+Figure 1 is central to their thesis: it is clear and well explained.
+However, I feel the lower left corner, corresponding to the pre-model/post-data area should be some darker grey to highlight that this corner is not overlooked but simply does not make a lot of sense in the current proposal (and in the idealized view of the current scientific process).
+It may sometimes correspond to mindless monitoring if one is uncharitable (Yoccoz et al.
+2001, Legga \& Nagy 2006).
+I believe it should not be greyed out the same colour as the other corners.}
+\end{mybox}
+
+...
+
+\begin{mybox}
+\emph{Moving to the upper right corner (post-model/pre-data), the authors could mention
+Simulation-Based Calibration (Modrák et al. 2025) to point readers to this useful tool for code
+checking, which is also central to workflows. I suspect this omission might be wilful not to
+overload readers. It might still be useful to point SBC on ecologists’ map as new models keep
+being proposed but are insufficiently put to the test beyond one case study in many instances
+(see also Nichols et al. 2019).}  
+\end{mybox}
+
+Thank you for this suggestion. We now mention simulation-based calibration (\href{file:forecastflows_r2\#lintarget:SBC}{line \ref*{lin:SBC}}).
+
+\begin{mybox}
+\emph{With respect to terminology, the authors are mentioning “research spheres” in the abstract,
+which is a catchy term, and one I agree may be more appropriate than that of a research
+programme for what the authors are discussing. What is a research sphere is not articulated
+further by the authors and there must be a “policy sphere” at which the paper is alluding to. In
+fact, where I find the authors’ thesis is less convincing is with respect to policy outreach. In
+policy, vague terminology such a “research sphere” is useful to get discussions started as
+loosely defined terms facilitate interactions. This can also lead to talking past each other but,
+usually, terminology gets to be clarified at some point when misunderstandings are too
+numerous to be ignored (and one get a common understanding document out of a round of
+meetings). My point is that the expectations of the “policy sphere” are usually not completely
+aligned with the “research sphere”. And this realization bears on workflows.}  
+\end{mybox}
+
+...
+
+\begin{mybox}
+\emph{In policy, a trend in abundance is needed as evidence for decision and further actions. Whether
+that trend comes from a phenomenological or mechanistic approach is often of secondary
+interest to policy-makers: what matters is that the trend passes some checks (including
+statistical significance). Much has been written on the topic, especially with respect to
+statistical power. So “outlin[ing] the steps of a universal workflow” (line 48-49) is both novel
+and worthy. Yet I am not convinced that it will solve some of issues that are pregnant in the
+policy sphere. For example, timeliness is important in policy. As the authors are right to point,
+workflows require time to be developed. Sometimes a lot more time is also needed just to agree
+on the workflow, and then there is time and man-power needed for documenting, maintaining
+and updating the workflow. In short, one need a governance structure for the workflow. If the
+goal is to evidence a negative trend because this is the minimum requirement for taking
+corrective actions, then the authors’ proposal, while clearly an improvement for the research
+sphere, will not impress the policy sphere as it is also likely to deliver too late; nor will it
+necessarily lead to better decisions. Awareness, design and publication of workflows will also
+have to happen in relevant policy circles, not just in scientific journals. That we should start
+with scientific journals in the right first steps, but reaching out to the policy sphere will not
+happen on its own either.}  
+\end{mybox}
+
+...
+
+\begin{mybox}
+\emph{I disagree with the framing on lines 7-9: the explanation provided from the two paths is not
+very convincing. The authors seem to have implicitly in mind some global policy, and hence
+justify a need for a global analysis. However, much of conservation is local, national or
+regional. Some surveys are designed and carried out at these scales, but when pooled in a
+global analysis, their analysis can become problematic because of sampling imbalances.
+However, the aim is now to make a global claim but that does not mean the local or regional
+trend was inaccurate in the first place. Yet the global analysis may be taken as evidence of
+biases (more on this term below) in policy fora. I overall think there is an unspoken assumption
+throughout in this opinion piece which is that all policies are somehow global. I see here a
+disconnect between high-profile papers on global trends which get attention precisely because
+they are global (and likely to get many citations and thus are a staple of high impact journals
+too), and a lot of policy which is much more local in scale.}  
+\end{mybox}
+
+We do agree that our framing is not directly adapted to the more local-scale decisions.
+We now mention that the policy needs we refered to are rather ``global'' (\href{file:forecastflows_r2\#lintarget:global}{line \ref*{lin:global}}).
+
+\begin{mybox}
+\emph{To be clearer, I will provide an example (see Authier 2025). Using citizen science data on 724
+species, Callaghan et al. (2021) estimated global population sizes for 9,700 extant bird species
+with modelling. These model-based estimates are not gauged against independent estimates
+for species well-monitored at the global scale. Owing to their large-size and conspicuousness
+several seabirds species can be censused accurately. Although imperfect, these estimates
+represent the best available knowledge for policy (ACAP 2021). Model-based abundances
+were > 20 times higher/smaller than IUCN ones for 45/5 (out of 96 seabird species) and up to
+100 times higher/smaller for 18/1 species (Figure 1). This is particularly concerning for
+endemic species, in which the local and global geographical scales are completely confounded,
+and which can comprise a single small population. A startling example is the Mascarene petrel
+(Pseudobulweria aterrima), a marine megafauna species endemic to Reunion Island in the Indian
+Ocean. Data on Mascarene petrels were used in the model training set and the estimated
+abundance by Callaghan et al. (2021) was several thousands of individuals. The estimated
+population size from local scientists is less than a few hundred individuals (Lopez et al. 2021).
+The global analysis is not helpful in my opinion, and may even hinder local policy which have
+taken a long time to gain attention and momentum.} 
+\end{mybox}
+
+This is a very interesting example.
+As highlighted in your next comment, maybe Callaghan et al.
+would have identified some flaws in their modeling approach if they had spent more time in the pre-data/post-model quadrat?
+If they predicted abundance estimations for 9000 new species outside their data, it seems like their inferential goal was above all prediction---and thus they should have focused on testing this goal in their workflow and should have developed some data simulation steps coherent with this inferential goal.
+It would be a very interesting example applied to conservation to illustrate how step-by-step workflows could help identify these local-to-global discrepancies, and maybe \emph{in fine} reconcile local observations and actions with global trends and policies?
+
+\begin{mybox}
+\emph{Callaghan et al. (2021) are concerned about forking paths and aware of reproducibility issues.
+They described a validation workflow which does not take stock of the pre-data/post-model
+corner but claimed nevertheless accuracy and success in redressing some biases (e.g.
+geographic and taxonomic). More generally, this example illustrates the following science-
+policy workflow: policy-makers read about a new global analysis published in a scientific
+journal. The analysis has a model workflow. Policy-makers look for the modelled-based
+abundance of their pet species and compare those to their knowledge and data. Local, national
+or regional scientists involved in policy are then contacted to explain any discrepancies. If
+there are indeed discrepancies, the scientists have now to explain their origins, looking into the
+model workflow and how it espoused good practices or not. With a growing familiarity of
+what are workflows, this should become more and more seamless in the long run. In the short
+run however, this science-policy workflow (i.e. the linear model of science, see below) is not
+scalable with respect to scientist’s time in my opinion and it will not necessarily help policy if
+scientists (especially non-tenured ones) think that this is time spent not doing field work,
+writing grants and supervising students to further their research.}  
+\end{mybox}
+
+First, we would like to answer about the time it may take for this type of workflow to become more common in research spheres.
+A large portion of scientific results today are investigated and written by non-tenured scientists.
+We have examples around us of graduate students who are likely more inclined than their supervisors to spend time developing carefully checked models and following workflow steps that give them more confidence in their results.
+While their supervisors may not push them to spend more time doing so, they will not prevent them from doing it either.
+Therefore, we believe that this paper---among many others---could actually help spread the use of robust modeling workflows (faster than it seems).
+Although the road between being a graduate student and obtaining a tenure-track position is long and uncertain, those graduate students are soon to become reviewers and submit grants (if they decide to pursue in academia), and interact with decision-makers, at least at a local scale (and even for those who will quit academia).
+
+\begin{mybox}
+\emph{The above scenario (which can be applied to global analyses with or without workflows) is not
+to deflect from the authors’ main point being made that “that an improved workflow that
+required data simulation and retrodictive checks would lead to larger model advances and a
+greater recognition of uncertainty—thus highlighting likely consistency in estimates across
+models—that could better aid policy”. The “could better aid policy” part is simply too vague
+and could betray, in my opinion, that the proposed “universal framework” is implicitly
+embracing the linear model of science (Fernándes 2016) of ‘truth speaks to power’. In other
+words, bad policy decisions are assumed to mainly result from a deficit of knowledge so that
+all that is needed would be better knowledge to naturally lead to good policy decision. I find
+this less and less convincing given that we’ve never had so much data or knowledge than now.
+We know what are the main drivers of biodiversity loss (e.g. Maxwell et al. 2016). We are
+literally awash with models (just think about the number of Species Distribution Models being
+published nowadays). Yet biodiversity, on average, does not fare any better.}  
+\end{mybox}
+
+...
+
+\begin{mybox}
+\emph{While I agree with the authors that the workflow will help research spheres in disciplining
+modelling, I disagree with the policy implications. As highlighted in the box by the authors,
+“[m]odel development is the central step of the process-based workflow, typically requiring
+several years, yet it often remains opaque for anyone who has not worked the model itself.”
+This is not a small issue.}  
+\end{mybox}
+
+We wanted to highlight that process-based model development can sometimes be very long (with or without this workflow), and thus that it would be better to follow coherent steps in model development and checking---given the amount of time process-based model development takes anyway.
+But to be clear, for most applications, applying the workflow do not require years.
+
+\begin{mybox}
+\emph{For example, in the International Whaling Commission, established in
+1946, many scientists are now retiring without new ones stepping in their shoes at the Scientific
+Committee (SC). The IWC SC pioneered some work on simulations (e.g. De la Mare 1986) and
+devised complex population models that were heavily tested to assess whether they would be
+fit-for-purposes. It is ironic that the simulation-savvy and model-based procedures
+encapsulated in the Revised Management Procedure that was developed in the late 1980s and
+early 1990s was never used in practice (for policy reasons, not scientific ones, to make a long
+story short).}  
+\end{mybox}
+
+...
+
+\begin{mybox}
+\emph{The IWC example underlines the long-term commitment that is needed and the bigger
+question is how to square such a commitment in a competitive academic landscape where
+projects, when they are funded, are meant for a couple of years and the research proletariat is
+expected to publish regularly in prestigious journals and be highly mobile with respect to
+institutions. I contend that the incentives for the research and policy spheres are simply not
+aligned enough at the moment. How does one ensure the sort of long-term commitment
+needed in policy where stability of procedures is important (as any management procedures or
+indicators are going to have to jump many hoops to get agreed by all parties) and a steady
+stream of publications in an academic ecosystem where novelty and innovations are a
+premium?}  
+\end{mybox}
+
+...
+
+\begin{mybox}
+\emph{The authors are alluding to the current publishing culture as a barrier but do not elaborate
+much. That is fair to omit in a short paper, and again I agree with their main point with respect to the research spheres.}
+\end{mybox}
+
+...
+
+\begin{mybox}
+\emph{I do think, however, that evidencing a trend serves a different goal in
+policy, and that having a workflow to bridge trend estimation and forecasting is not as urgent
+as the authors imply if the ultimate aim is to enable timely decisions to address biodiversity
+loss. In fact, the insistence on workflow may backfire simply because it can easily lead to either
+distraction or displacement in the sense of Rainer (2012). Distraction may happen because it
+could be easy to ask for more testing of some model features (a new twist to some mechanistic
+parts in the model) and ask that no decision be taken before that an updated workflow is
+completed. This could easily align with the common policy requirement of using the ‘best
+available evidence’, especially if some cutting-edge new results are brought forward at the
+policy table and one is being tasked with updating the current model. Displacement is more
+subtle but can also happen: managing the workflow becomes the policy and simply
+maintaining it absorb most of the available resources. As Rayner (2012) puts it “displacement
+occurs when an organization engages with an issue, but substitutes management of a
+representation of a problem (such as a computer model) for management of the represented
+object or activity.” I think these two issues are real.} 
+\end{mybox}
+
+This is a very interesting question.
+However, we believe that the previous example of Callaghan et al.
+illustrates how bridging the differential goals of understanding and forecasting (or extrapolating to poorly observed species) can actually have important implications for policy.
+Similarly, it does not seem entirely accurate to think that this type of workflow should have been applied earlier (Revised Management Procedure of IWC) while also considering that it might be now 'too late' to highlight its importance (because there is no more time to lose to adress biodiversity loss?).
+We think this issue is linked to other challenges we encouter as scientists today: is there a risk in better quantifying and communicating the uncertainties of our results, if these uncertainties lead decision-makers to avoid making decisions?
+Or, by being the most transparent about our uncertain results (and spending more time applying consistent workflows), do we rather serve not only research but also the policy sphere?
+
+\begin{mybox}
+\emph{I would love to actually read the authors’ thoughts here and ideally be proven wrong that an
+improved workflow would be enough for enabling policy decisions whose beneficial impacts
+would be commensurate with the current magnitude of biodiversity loss7. In particular, I
+would be interested in a better articulation of how they see global studies feed-backing on
+policy which is usually more local, national or regional.}  
+\end{mybox}
+
+\begin{mybox}
+\emph{My remaining points are more concise.
+- Terminology. If the authors are also aiming at the policy sphere, the word bias (e.g. lines
+13, 134) should be wielded with extreme caution as it is easily weaponized. In policy
+circles, the term bias is, in my experience, the kiss of death as it is often interpreted as
+lopsided evidence. The authors are discussing preferential sampling with respect to space,
+time or taxonomy and its consequences in modelling results. Preferential sampling is more
+descriptive than bias, which also has a prescriptive meaning, esp. in policy8. Also, a well-
+designed survey at some spatial scale may become an instance of preferential geographic
+sampling when collated with other surveys and data. This sort of data collation is often the
+case with global trend analyses that the authors seem to have in mind. Also, taxonomic
+biases are bound to happen in global analyses. How do we redress those? That is, what can
+we say about data-poor species? How a richly-parametrized workflow calibrated on data
+collected on a few very well-studied species is going to help the majority of data-poor
+species? Taxonomic gaps have been identified without workflows and how to address
+those gaps point to monitoring and the need of appropriate governance to ensure long-
+term monitoring and inclusion of data-poor species. Research alone may not provide
+enough incentive or impetus for inclusion of data-poor species, especially if what is
+needed for policy about them is descriptive and routine science.}  
+\end{mybox}
+
+\begin{mybox}
+\emph{- ML enthusiasm (lines 298-303): the point is granted for prediction but again evidencing a
+trend for policy is often required for further action. I fail to see how ML, if focused on
+prediction, is going to help. A trend is a useful fiction that is easily understood in policy
+fora where interpretability still has a huge premium. And a trend is very much
+interpretable. I think the authors may be going overboard in their claims if, again, their
+proposal is to have policy-relevance and not only research-relevance (a worthy goal of its
+own).}  
+\end{mybox}
+
+\begin{mybox}
+\emph{- Robustness is great. Policy loves the word but too rarely elaborates on it. Thus, on line 62,
+these methods should be robust against what more precisely?}  
+\end{mybox}
+
+\begin{mybox}
+\emph{- typo on line 138: “These feedback loops are applied currently in only in Bayesian
+frameworks (to our knowledge)...”}
+\end{mybox}
 

### 2025-09-19 revision/r2/forecastflows_r2.tex
starting to answer the very interesting review of referee 3

@@
 
-Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on fewer data.
+Meeting these \llabel{global}global policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on fewer data.
 Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic biases.
@@
 \llabel{repeatsim} Ideally, this data simulation step should be repeated several times with different parameter values, sampled within a reasonable range.
-\llabel{priorPC} In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}), while in other frameworks they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).
+\llabel{priorPC} In a Bayesian framework, simulated parameters can be sampled from the prior distribution (allowing us to check the implications of our priors \citealp{Gelman2020}, \llabel{SBC}and to validate model inferences using simulation-based calibration \citealp{Talts2018_SBC}).
+In other frameworks, they can be sampled from a range of biologically reasonable values (which is very similar to the concept of a prior).
 
@@
 	\centering
-	\vspace*{-0.5cm}
+	\vspace*{-0.4cm}
     \hspace*{-1.5cm}
@@
 The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model.
-Calibrating submodels separately allows to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.
+Calibrating submodels separately makes possible to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.
 %emw24Jul: nice stuff here.

### 2025-09-19 revision/r2/response_r2.tex
slowly continue revisions

@@
 
-...
+Thank you for this suggestion. We now use a darker gray to highlight this corner.
 
@@
 
-Thank you for this suggestion. We now mention simulation-based calibration (\href{file:forecastflows_r2\#lintarget:SBC}{line \ref*{lin:SBC}}).
+We now mention simulation-based calibration (\href{file:forecastflows_r2\#lintarget:SBC}{line \ref*{lin:SBC}}).
 
@@
 \emph{The above scenario (which can be applied to global analyses with or without workflows) is not
-to deflect from the authors’ main point being made that “that an improved workflow that
+to deflect from the authors' main point being made that “that an improved workflow that
 required data simulation and retrodictive checks would lead to larger model advances and a
@@
 and could betray, in my opinion, that the proposed “universal framework” is implicitly
-embracing the linear model of science (Fernándes 2016) of ‘truth speaks to power’. In other
+embracing the linear model of science (Fernándes 2016) of 'truth speaks to power'. In other
 words, bad policy decisions are assumed to mainly result from a deficit of knowledge so that
@@
 
-...
+We agree that the main obstacles---at least at the global scale---for effective climate and biodiversity policies are not a lack of knowledge.
+However, coming back to your previous point (``\emph{Local, national or regional scientists involved in policy are then contacted to explain any discrepancies}''), we believe a different interpretation of the usefulness of the workflow for policy is possible.
+You noted above that scientists and decision-maskers can spend time discussing discrepancies with a high-profile global study---rather than on better decision-making or further research.
+We believe that a more systematic use of this kind of workflow would reduce such discrepancies (between different studies, estimates) and methodological controversies (a point we develop in the manuscript).
+So in this way, an improved workflow could aid policy.
 
@@
 
-...
+We are not familiar with this example.
+Even without considering policy reasons, we do feel however that science sometimes goes in circles---and that great improvements in computational tools since 90s have not necessarily led to more robust models.
 
@@
 
+Sorry it was not clear.
+Our point here was not to share any enthusiasm for machine learning, but rather to highlight the need for an even more careful and coherent workflow to build interpretable models (i.e.
+avoiding  ``complexity trap'' and build models with ``clear mechanistic drivers'').
+We tried to make this point more explicit  (\href{file:forecastflows_r2\#lintarget:clearerML}{line \ref*{lin:clearerML}}).
+
 \begin{mybox}
@@
 
+Thank you for catching this! We fixed this typo.
+
+\begin{mybox}
+\emph{Last, I would like to re-iterate that I view this manuscript as a very useful contribution for the
+research spheres in ecology. It can be published as is for this audience. I do not dispute any of
+the claims made with respect to research, namely that disciplining model building to the
+rationalizing force of workflows is a good thing. I dispute what I think are shortcuts and could
+be perceived as a naïve outlook on the policy process. I do not assume that the authors
+subscribe to the linear model of science but a too quick reading of their opinion piece may
+suggest otherwise, especially the introduction when it refers to policy. For example, if lines 32-
+41 are an accurate description of the research sphere, the policy sphere is then very rational in
+keeping an instrumental outlook to trend detection and favouring simple methods until the
+research sphere gets its workflows fully operational. The path from an improved workflow in
+research to better policy decisions remains to be charted: scientific workflows are likely a
+necessary step, yet one that may not be sufficient.}
+\end{mybox}
+
+...
+
 \end{document}

### 2025-09-19 revision/r2/response_r2.tex
finalize mroe or less revisions

@@
 
-\DeclareEmphSequence{\itshape\color{darkgrey}}
+% \DeclareEmphSequence{\itshape\color{darkgrey}}
 \usepackage[most]{tcolorbox}
@@
     breakable,
+    coltext=darkgrey
     #1}
@@
 (nearly) ready for publication. I suggest one more pass for copy editing
-to fix small mistakes like inconistencies for example in "figure"
-"Figure" and "e.g." "e.g.,".}
+to fix small mistakes like inconistencies for example in ``figure"
+``Figure" and ``e.g." ``e.g.,".}
 \end{mybox}
@@
 \emph{\textbf{Manuscript Minor comments:}\\
-Abstract: potential typo in second-to-last sentence. Should "where
-forecasting a natural output" be "where forecasting is a natural output"?}
+Abstract: potential typo in second-to-last sentence. Should ``where
+forecasting a natural output" be ``where forecasting is a natural output"?}
 \end{mybox}
-% vvdm18sept: is really a typo? I thought it was just a way to say it 
+% vvdm18sept: is really a typo? I thought it was just a way to say it. Lizzie?
 ...
@@
 \begin{mybox}
-\emph{Line 51: "hypotheses" or "a hypothesis" rather than "hypothesis"}
+\emph{Line 51: ``hypotheses" or ``a hypothesis" rather than ``hypothesis"}
 \end{mybox}
@@
 \begin{mybox}
-\emph{Line 138: there appears to be an extra "in" in "in only in"}
+\emph{Line 138: there appears to be an extra ``in" in ``in only in"}
 \end{mybox}
@@
 \begin{mybox}
-\emph{A couple typos in Box e.g., "calibrating submodes separately allows to
+\emph{A couple typos in Box e.g., ``calibrating submodes separately allows to
 avoid" missing "us"?}
@@
 
-...
+... % Lizzie?
 
@@
 
-...
+We now mention that sigmas are bounded to zero.
 
@@
 
-% Hmm I don't understand?
+% Hmm I'm not sure to understand what he wants?
+We now specify that the first posterior predictive distribution we show corresponds to the predicted counts across all species and populations.
 
@@
 We are particularly grateful for his detailed comments, with much broader perspectives than just our manuscript.
-We have tried to address most of the comments, but we believe that they deserve longer discussions and could constitute in themselves another paper.
+We believe that they deserve longer discussions (and could constitute in themselves another paper!), so we will reach out directly to M.
+Authier to share our detailed perspective on this.
 
@@
 sampling when collated with other surveys and data. This sort of data collation is often the
-case with global trend analyses that the authors seem to have in mind. Also, taxonomic
-biases are bound to happen in global analyses. How do we redress those? That is, what can
+case with global trend analyses that the authors seem to have in mind.}  
+\end{mybox}
+
+...
+
+\begin{mybox}
+\emph{Also, taxonomic biases are bound to happen in global analyses. How do we redress those? That is, what can
 we say about data-poor species? How a richly-parametrized workflow calibrated on data
@@
 
-...
+We have made this point clearer (\href{file:forecastflows_r2\#lintarget:robust}{line \ref*{lin:robust}}):
+\begin{quote}
+``This reality should drive researchers to use more coherent methods that remain robust for the intended inferential goals, even with imperfect data and knowledge gaps.''
+\end{quote}
 
@@
 
-...
+We would like to thank the reviewer once again for all these very interesting avenues for discussion---beyond the community of researchers (to whom this paper was primarily addressed).
 

### 2025-09-19 analyses/lpiworkflow.qmd
finalize mroe or less revisions

@@
 
-What is global trend over time in vertebrate species population numbers, both overall and how does it vary across species and populations?
+What is the global trend over time in vertebrate species population numbers, both overall and how does it vary across species and populations?
 
@@
 Populations usually are nested in species in ecological models, so we might consider that to start.
+In this first hierarchical model, we thus assume that populations are exchangeable---an assumption we might refine later.
 We also need to make sure our model yields one global estimate; for this we decide to model species trends over time as part of larger distribution.
@@
 
-$$\log(y_i) \sim \text{normal}(\mu_i, \sigma)$$
+$$\log(y_i) \sim \text{normal}(\mu_i, \ \sigma)$$
 
@@
 
-$$\alpha_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\alpha, \text{sp}} \ , \ \sigma_{\alpha, \text{sp}}) \\
-\mu_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_1} \ , \ \sigma_{\alpha_1}) \\
-\sigma_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_2} \ , \ \sigma_{\alpha_2}) $$
+$$\alpha_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\alpha, \text{sp}}, \ \sigma_{\alpha, \text{sp}})
+\quad \mu_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_1}, \ \sigma_{\alpha_1}) 
+\quad \sigma_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_2}, \ \sigma_{\alpha_2}) $$
 
-$$\beta_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\beta, \text{sp}} \ , \ \sigma_{\beta, \text{sp}}) \\
-\mu_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_1} \ , \ \sigma_{\beta_1}) \\
-\sigma_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_2} \ , \ \sigma_{\beta_2}) $$
+$$\beta_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\beta, \text{sp}}, \ \sigma_{\beta, \text{sp}}) 
+\quad \mu_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_1}, \ \sigma_{\beta_1}) 
+\quad \sigma_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_2}, \ \sigma_{\beta_2}) $$
+
+(where all the $\sigma$ parameters are greater than zero)
 
@@
 Now we code the model from our mathematical notation (above) and fit it to our simulated data (also built from our mathematical model).
+
+```{stan output.var = "compiled_model"}
+
+data {
+  int<lower=1> N;
+  int<lower=1> Nsp;
+  int<lower=1> Npop;
+  vector[N] y;
+  vector[N] year;
+  array[Npop] int<lower=1,upper=Nsp> spid_perpop;
+  array[N] int<lower=1,upper=Npop> popid;
+}
+
+transformed data{
+  vector[N] logy = log(y);
+}
+
+parameters {
+  real mu_alpha1; // overall species intercept
+  real<lower=0> sigma_alpha1;
+
+  real mu_alpha2;
+  real<lower=0> sigma_alpha2;
+
+  array[Nsp] real mu_alpha_sp; // species-level intercepts
+  array[Nsp] real<lower=0> sig_species_alpha;
+
+  array[Npop] real alpha_tilde_pop_sp; // non-centered population intercepts
+
+  real mu_beta1; // overall species trend
+  real<lower=0> sigma_beta1;
+
+  real mu_beta2;
+  real<lower=0> sigma_beta2;
+
+  array[Nsp] real mu_beta_sp; // species-level slopes
+  array[Nsp] real<lower=0> sig_species_beta;
+
+  array[Npop] real beta_tilde_pop_sp; // non-centered population slopes
+
+  real<lower=0> sigma;
+
+}
+
+transformed parameters{
+
+  array[Npop] real alpha_pop_sp;
+  array[Npop] real beta_pop_sp;
+
+  for (p in 1:Npop) {
+
+      alpha_pop_sp[p] = mu_alpha_sp[spid_perpop[p]] 
+      + sig_species_alpha[spid_perpop[p]] *  alpha_tilde_pop_sp[p];
+      beta_pop_sp[p] = mu_beta_sp[spid_perpop[p]] 
+      + sig_species_beta[spid_perpop[p]] *  beta_tilde_pop_sp[p];
+
+  }
+
+}
+
+model {
+
+  vector[N] mu;
+
+  mu_alpha_sp ~ normal(mu_alpha1, sigma_alpha1);
+  sig_species_alpha ~ normal(mu_alpha2, sigma_alpha2);
+
+  mu_beta_sp ~ normal(mu_beta1, sigma_beta1);
+  sig_species_beta ~ normal(mu_beta2, sigma_beta2);
+
+  for (p in 1:Npop) {
+
+    alpha_tilde_pop_sp[p] ~ normal(0,1);
+    beta_tilde_pop_sp[p] ~ normal(0,1);
+
+  }
+
+  for(i in 1:N){
+
+    mu[i] = alpha_pop_sp[popid[i]] + beta_pop_sp[popid[i]] * (year[i] - 1980);
+    logy[i] ~ normal(mu[i], sigma);
+
+  }
+
+  mu_alpha1 ~ normal(0, log(500) / 2.57); // -log(500) < mu_alpha1 < log(500)
+  sigma_alpha1 ~ normal(0, log(100) / 2.57); // 0 < sigma_alpha1 < log(100)
+
+  mu_alpha2 ~ normal(0, log(500) / 2.57); // -log(500) < mu_alpha1 < log(500)
+  sigma_alpha2 ~ normal(0, log(50) / 2.57); // 0 < sigma_alpha1 < log(50)
+
+  mu_beta1 ~ normal(0, 2 / 2.57); // -2 < mu_beta1 < 2
+  sigma_beta1 ~ normal(0, 2 / 2.57); // 0 < sigma_beta1 < 2
+
+  mu_beta2 ~ normal(0, 2 / 2.57); //-2 < mu_beta1 < 2
+  sigma_beta2 ~ normal(0, 2 / 2.57); // 0 < sigma_beta1 < 2
+
+  sigma ~ normal(0, log(100) / 2.57);
+
+}
+
+generated quantities {
+
+  vector[N] logy_pred;
+
+  for(i in 1:N){
+
+    real mu = alpha_pop_sp[popid[i]] 
+    + beta_pop_sp[popid[i]] * (year[i] - 1980);
+    logy_pred[i] = normal_rng(mu, sigma);
+
+  }
+
+}
+
+```
 
@@
 
-Here we start by visually comparing the posterior predictive distribution (red line) to observed data (black dashed line):
+Here we start by visually comparing the posterior predictive distribution of the predicted counts (across all species and populations, red line) to observed data (black dashed line):
 ```{r}

### 2025-09-19 analyses/lpiworkflow.tex
finalize mroe or less revisions

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A sample workflow},
  pdfauthor={Victor; Jonathan; Lizzie},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A sample workflow}
\author{Victor \and Jonathan \and Lizzie}
\date{April 2025}

\begin{document}
\maketitle

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}

To show our workflow in action, we examine a subset of the Living Planet
Index (LPI, download at: https://www.livingplanetindex.org/data\_portal.
This download should include a csv file, here: LPD\_2024\_public.csv,
that we use below). We discuss the LPI in the main text; for more
information, check out \href{www.livingplanetindex.org/}{their website}.

\section{Step 1: Research question}\label{step-1-research-question}

Here we need to focus on exactly what question we are trying to answer,
which will guide us in the next steps. This step is sometimes skipped
over or moved through so quickly that the exact question is not clear,
but is critical for developing a useful model. Many early papers on the
LPI seem focused on something similar to: Are vertebrate species overall
declining on average? And, if so, by how much? We, however, might want
to have a question that more clearly allows any directionality, so we
could start with: What is the global trend for biodiversity in vetebrate
species? This is rather broad and might mean we should develop a model
that will yield one global estimate. In thinking about this, we would
likely realize we don't want just that, we actually want to understand
the variation across different species and perhaps populations and also
have a global estimate. So we might refine the question to:

\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, breakable, colframe=quarto-callout-tip-color-frame, colback=white, opacityback=0, arc=.35mm, bottomrule=.15mm]

What is the global trend over time in vertebrate species population
numbers, both overall and how does it vary across species and
populations?

\end{tcolorbox}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{wd }\OtherTok{\textless{}{-}} \StringTok{"/home/victor/projects/forecastflows/analyses"}

\FunctionTok{set.seed}\NormalTok{(}\DecValTok{20012029}\NormalTok{)}

\FunctionTok{suppressPackageStartupMessages}\NormalTok{(}\FunctionTok{library}\NormalTok{(rstan))}
\FunctionTok{library}\NormalTok{(ggplot2)}

\NormalTok{lpi\_data }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\FunctionTok{file.path}\NormalTok{(wd, }\StringTok{"input"}\NormalTok{, }\StringTok{"LPD\_2024\_public.csv"}\NormalTok{), }\AttributeTok{header=}\NormalTok{T)}
\end{Highlighting}
\end{Shaded}

\section{Step 2: Model building}\label{step-2-model-building}

Next we need to think about how we would build a model to address this
question, with the goal to build the model and evaluate it using
simulated data. There are a lot of demographic models in population
ecology, so we might revisit those and consider a discrete time model
based on population size and stepping through each year of data. We
might also start with a simple linear model, where populations can go
up, down or do nothing.

Using a simple linear gets us started, but now we need to consider how
we will model species and populations. Populations usually are nested in
species in ecological models, so we might consider that to start. In
this first hierarchical model, we thus assume that populations are
exchangeable---an assumption we might refine later. We also need to make
sure our model yields one global estimate; for this we decide to model
species trends over time as part of larger distribution. Our resulting
model would then be:

\[\log(y_i) \sim \text{normal}(\mu_i, \ \sigma)\]

\[\mu_i = \alpha_{\text{pop}[\text{sp}[i]]} + \beta_{\text{pop}[\text{sp}[i]]} * \text{year}_i\]

\[\alpha_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\alpha, \text{sp}}, \ \sigma_{\alpha, \text{sp}})
\quad \mu_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_1}, \ \sigma_{\alpha_1}) 
\quad \sigma_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_2}, \ \sigma_{\alpha_2}) \]

\[\beta_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\beta, \text{sp}}, \ \sigma_{\beta, \text{sp}}) 
\quad \mu_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_1}, \ \sigma_{\beta_1}) 
\quad \sigma_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_2}, \ \sigma_{\beta_2}) \]

(where all the \(\sigma\) parameters are greater than zero)

\section{Step 3: Model evaluation}\label{step-3-model-evaluation}

Now we need to simulate data to test our model. As we start to think
about how to do this, we may become overwhelmed with the scale of the
question we started with. We likely would not feel immediately sure how
to simulate for all different vertebrate species. For example, what are
reasonable starting population sizes (intercepts) for different groups?
As we start to answer this we likely would realize different clades
should have different starting population sizes. We could thus try to
build in increasing complexity immediately, but then we will likely
struggle to see how our model is performing at the scales we care about
-- for each population in each species. It is almost always better to
scale back the question (and potentially dataset) when confronted with
these sorts of problems.

\subsection{Stepping backward\ldots{}}\label{stepping-backward}

Here we would likely feedback (step back) one to two steps. We could
step back to Step 2 and consider adding a phylogeny or other structure
from which we could then simulate data from, or we could step back to
Step 1 and narrow the question (before potentially building it back up.)
As mentioned above, we believe the latter approach is usually better as
working on a small scale first can point out more fundamental problems
versus starting big and then spending much longer to identify
fundamental problems (or potentially missing them).

So here we refine our question to:

\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, breakable, colframe=quarto-callout-tip-color-frame, colback=white, opacityback=0, arc=.35mm, bottomrule=.15mm]

What is trend in mammal species population numbers in southern Africa,
both overall and how does it vary across species and populations?

\end{tcolorbox}

We chose this because mammals are a well studied group compared to other
clades/groups in the dataset, so we will have a better idea of how to
simulate data, and whether our model makes sense.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{lpi\_subset }\OtherTok{\textless{}{-}}\FunctionTok{subset}\NormalTok{(lpi\_data, Region }\SpecialCharTok{==} \StringTok{"Africa"} \SpecialCharTok{\&}\NormalTok{ Class }\SpecialCharTok{==} \StringTok{"Mammalia"} \SpecialCharTok{\&} 
\NormalTok{                      Included.in.LPR2024 }\SpecialCharTok{==} \DecValTok{1}\NormalTok{ )}

\CommentTok{\# subset 5+ counts}
\NormalTok{obs.cnt}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\ControlFlowTok{for}\NormalTok{ (n }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{dim}\NormalTok{(lpi\_subset)[}\DecValTok{1}\NormalTok{])\{}
\NormalTok{  obs.cnt[n]}\OtherTok{\textless{}{-}}\FunctionTok{sum}\NormalTok{(lpi\_subset[n, }\DecValTok{31}\SpecialCharTok{:}\DecValTok{101}\NormalTok{]}\SpecialCharTok{!=}\StringTok{"NULL"}\NormalTok{)}
\NormalTok{\}}
\NormalTok{lpi\_subset}\OtherTok{\textless{}{-}}\NormalTok{lpi\_subset[obs.cnt}\SpecialCharTok{\textgreater{}}\DecValTok{4}\NormalTok{,]}

\CommentTok{\# subset species 10+ populations}
\NormalTok{sp }\OtherTok{\textless{}{-}} \FunctionTok{unique}\NormalTok{(lpi\_subset}\SpecialCharTok{$}\NormalTok{Binomial)}
\NormalTok{sp.cnt}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
\ControlFlowTok{for}\NormalTok{(x }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(sp))\{}
\NormalTok{  Af.tmp}\OtherTok{\textless{}{-}}\FunctionTok{subset}\NormalTok{(lpi\_subset, Binomial }\SpecialCharTok{==}\NormalTok{ sp[x])}
\NormalTok{  sp.cnt[x]}\OtherTok{\textless{}{-}}\FunctionTok{dim}\NormalTok{(Af.tmp)[}\DecValTok{1}\NormalTok{]}
\NormalTok{\}}
\NormalTok{keep.sp}\OtherTok{\textless{}{-}}\NormalTok{sp[sp.cnt}\SpecialCharTok{\textgreater{}}\DecValTok{9}\NormalTok{]}
\NormalTok{lpi\_subset}\OtherTok{\textless{}{-}}\NormalTok{lpi\_subset[lpi\_subset}\SpecialCharTok{$}\NormalTok{Binomial }\SpecialCharTok{\%in\%}\NormalTok{ keep.sp,]}
\end{Highlighting}
\end{Shaded}

\subsection{Simulating data}\label{simulating-data}

Now we start to simulate! We follow the mathematical model we built
above and simulate data from it.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nspecies }\OtherTok{\textless{}{-}} \DecValTok{9} \CommentTok{\# no. of species}
\NormalTok{npops\_persp }\OtherTok{\textless{}{-}} \FunctionTok{round}\NormalTok{(}\FunctionTok{runif}\NormalTok{(nspecies, }\DecValTok{2}\NormalTok{, }\DecValTok{15}\NormalTok{)) }\CommentTok{\# no. of different populations per species}
\NormalTok{npops }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(npops\_persp) }\CommentTok{\# total no. of different populations}
\NormalTok{nobs\_perpop }\OtherTok{\textless{}{-}} \FunctionTok{round}\NormalTok{(}\FunctionTok{runif}\NormalTok{(npops, }\DecValTok{20}\NormalTok{, }\DecValTok{60}\NormalTok{)) }\CommentTok{\# no. of different observations per population}

\NormalTok{spid\_perpop }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{nspecies, }\AttributeTok{times =}\NormalTok{ npops\_persp)}
\NormalTok{spid\_perobs }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(spid\_perpop, }\AttributeTok{times =}\NormalTok{ nobs\_perpop)}
\NormalTok{popid\_perobs }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{npops, }\AttributeTok{times =}\NormalTok{ nobs\_perpop)}

\NormalTok{years }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{()}
\ControlFlowTok{for}\NormalTok{(np }\ControlFlowTok{in}\NormalTok{ nobs\_perpop)\{}
\NormalTok{  years\_p }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(}\DecValTok{1950}\SpecialCharTok{:}\DecValTok{2020}\NormalTok{, }\AttributeTok{size =}\NormalTok{ np, }\AttributeTok{replace =} \ConstantTok{FALSE}\NormalTok{)}
\NormalTok{  years }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(years, years\_p)}
\NormalTok{\}}

\CommentTok{\# nested intercepts}
\NormalTok{mu\_alpha1 }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(}\DecValTok{200}\NormalTok{)}
\NormalTok{sigma\_alpha1 }\OtherTok{\textless{}{-}}  \FunctionTok{log}\NormalTok{(}\DecValTok{50}\NormalTok{)}\SpecialCharTok{/}\FloatTok{2.57}
\NormalTok{mu\_alpha\_sp }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ nspecies, }\AttributeTok{mean =}\NormalTok{ mu\_alpha1, }\AttributeTok{sd =}\NormalTok{ sigma\_alpha1)}
\NormalTok{mu\_alpha2 }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(}\DecValTok{20}\NormalTok{)}
\NormalTok{sigma\_alpha2 }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(}\DecValTok{50}\NormalTok{)}\SpecialCharTok{/}\FloatTok{2.57}
\NormalTok{sig\_species\_alpha }\OtherTok{\textless{}{-}} \FunctionTok{abs}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ nspecies, }\AttributeTok{mean =}\NormalTok{ mu\_alpha2, }
                               \AttributeTok{sd =}\NormalTok{ sigma\_alpha2))}
\NormalTok{alpha\_pop\_sp }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ npops, }\AttributeTok{mean =}\NormalTok{ mu\_alpha\_sp[spid\_perpop], }
                      \AttributeTok{sd =}\NormalTok{ sig\_species\_alpha[spid\_perpop]) }

\CommentTok{\# nested slopes}
\NormalTok{mu\_beta1 }\OtherTok{\textless{}{-}} \FloatTok{0.05}
\NormalTok{sigma\_beta1 }\OtherTok{\textless{}{-}} \FloatTok{0.25}\SpecialCharTok{/}\FloatTok{2.57}
\NormalTok{mu\_beta\_sp }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ nspecies, }\AttributeTok{mean =}\NormalTok{ mu\_beta1, }\AttributeTok{sd =}\NormalTok{ sigma\_beta1)}
\NormalTok{mu\_beta2 }\OtherTok{\textless{}{-}} \DecValTok{0}
\NormalTok{sigma\_beta2 }\OtherTok{\textless{}{-}} \FloatTok{0.05}\SpecialCharTok{/}\FloatTok{2.57}
\NormalTok{sig\_species\_beta }\OtherTok{\textless{}{-}} \FunctionTok{abs}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ nspecies, }\AttributeTok{mean =}\NormalTok{ mu\_beta2, }
                              \AttributeTok{sd =}\NormalTok{ sigma\_beta2))}
\NormalTok{beta\_pop\_sp }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ npops, }\AttributeTok{mean =}\NormalTok{ mu\_beta\_sp[spid\_perpop], }
                     \AttributeTok{sd =}\NormalTok{ sig\_species\_beta[spid\_perpop]) }

\CommentTok{\# save simulated intercepts and slopes}
\NormalTok{pops\_fit }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{()}
\NormalTok{pid }\OtherTok{\textless{}{-}} \DecValTok{1}
\ControlFlowTok{for}\NormalTok{(sp }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{nspecies)\{}
  \ControlFlowTok{for}\NormalTok{(p }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{npops\_persp[sp])\{}
\NormalTok{    pops\_fit }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(pops\_fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{species =}\NormalTok{ sp, }\AttributeTok{pop =}\NormalTok{ pid))}
\NormalTok{    pid }\OtherTok{\textless{}{-}}\NormalTok{ pid }\SpecialCharTok{+} \DecValTok{1}
\NormalTok{  \}}
\NormalTok{\}}
\NormalTok{pops\_fit}\SpecialCharTok{$}\NormalTok{int }\OtherTok{\textless{}{-}}\NormalTok{ alpha\_pop\_sp}
\NormalTok{pops\_fit}\SpecialCharTok{$}\NormalTok{slp }\OtherTok{\textless{}{-}}\NormalTok{ beta\_pop\_sp}

\NormalTok{sigma\_obs }\OtherTok{\textless{}{-}} \FloatTok{0.75} \CommentTok{\# common observational error}

\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{()}
\ControlFlowTok{for}\NormalTok{(p }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{npops)\{}
\NormalTok{  nobs\_p }\OtherTok{\textless{}{-}}\NormalTok{ nobs\_perpop[p]}
\NormalTok{  alpha\_p }\OtherTok{\textless{}{-}}\NormalTok{ alpha\_pop\_sp[p]}
\NormalTok{  beta\_p }\OtherTok{\textless{}{-}}\NormalTok{ beta\_pop\_sp[p]}

\NormalTok{  years\_p }\OtherTok{\textless{}{-}}\NormalTok{ years[popid\_perobs }\SpecialCharTok{==}\NormalTok{ p] }

\NormalTok{  yhat\_p }\OtherTok{\textless{}{-}}\NormalTok{ alpha\_p }\SpecialCharTok{+}\NormalTok{ beta\_p }\SpecialCharTok{*}\NormalTok{ (years\_p }\SpecialCharTok{{-}} \DecValTok{1980}\NormalTok{)}
  \CommentTok{\# yhat\_p \textless{}{-} alpha\_p}
\NormalTok{  y\_p }\OtherTok{\textless{}{-}} \FunctionTok{exp}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\AttributeTok{n =}\NormalTok{ nobs\_p, }\AttributeTok{mean =}\NormalTok{ yhat\_p, }\AttributeTok{sd =}\NormalTok{ sigma\_obs))}
\NormalTok{  y }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(y, y\_p)}
\NormalTok{\}}

\NormalTok{datasim }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
\NormalTok{  y, }
  \AttributeTok{year =}\NormalTok{ years }\SpecialCharTok{{-}} \DecValTok{1980}\NormalTok{,}
  \AttributeTok{species =} \FunctionTok{as.character}\NormalTok{(spid\_perobs),}
  \AttributeTok{population =} \FunctionTok{paste0}\NormalTok{(spid\_perobs,popid\_perobs)}
\NormalTok{)}

\NormalTok{datasim}\SpecialCharTok{$}\NormalTok{logy }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(datasim}\SpecialCharTok{$}\NormalTok{y)}

\FunctionTok{ggplot}\NormalTok{(}\AttributeTok{data =}\NormalTok{ datasim) }\SpecialCharTok{+}
  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ year}\SpecialCharTok{+}\DecValTok{1980}\NormalTok{, }\AttributeTok{y =}\NormalTok{ y, }
                \AttributeTok{group =}\NormalTok{ population,}
                \AttributeTok{color =}\NormalTok{ species)) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{\textquotesingle{}Year\textquotesingle{}}\NormalTok{, }\AttributeTok{y =} \StringTok{\textquotesingle{}Simulated counts\textquotesingle{}}\NormalTok{, }\AttributeTok{colour =} \StringTok{\textquotesingle{}Species\textquotesingle{}}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{lpiworkflow_files/figure-pdf/sim data-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{ggplot}\NormalTok{(}\AttributeTok{data =}\NormalTok{ datasim) }\SpecialCharTok{+}
  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ year}\SpecialCharTok{+}\DecValTok{1980}\NormalTok{, }\AttributeTok{y =}\NormalTok{ logy, }
                \AttributeTok{group =}\NormalTok{ population,}
                \AttributeTok{color =}\NormalTok{ species)) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{\textquotesingle{}Year\textquotesingle{}}\NormalTok{, }\AttributeTok{y =} \StringTok{\textquotesingle{}Simulated counts (log{-}scale)\textquotesingle{}}\NormalTok{, }\AttributeTok{colour =} \StringTok{\textquotesingle{}Species\textquotesingle{}}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{lpiworkflow_files/figure-pdf/sim data-2.pdf}

\subsection{Fit and check model on simulated
data!}\label{fit-and-check-model-on-simulated-data}

Now we code the model from our mathematical notation (above) and fit it
to our simulated data (also built from our mathematical model).

\begin{Shaded}
\begin{Highlighting}[]

\KeywordTok{data}\NormalTok{ \{}
  \DataTypeTok{int}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{1}\NormalTok{\textgreater{} N;}
  \DataTypeTok{int}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{1}\NormalTok{\textgreater{} Nsp;}
  \DataTypeTok{int}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{1}\NormalTok{\textgreater{} Npop;}
  \DataTypeTok{vector}\NormalTok{[N] y;}
  \DataTypeTok{vector}\NormalTok{[N] year;}
  \DataTypeTok{array}\NormalTok{[Npop] }\DataTypeTok{int}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{1}\NormalTok{,}\KeywordTok{upper}\NormalTok{=Nsp\textgreater{} spid\_perpop;}
  \DataTypeTok{array}\NormalTok{[N] }\DataTypeTok{int}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{1}\NormalTok{,}\KeywordTok{upper}\NormalTok{=Npop\textgreater{} popid;}
\NormalTok{\}}

\KeywordTok{transformed data}\NormalTok{\{}
  \DataTypeTok{vector}\NormalTok{[N] logy = log(y);}
\NormalTok{\}}

\KeywordTok{parameters}\NormalTok{ \{}
  \DataTypeTok{real}\NormalTok{ mu\_alpha1; }\CommentTok{// overall species intercept}
  \DataTypeTok{real}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{0}\NormalTok{\textgreater{} sigma\_alpha1; }

  \DataTypeTok{real}\NormalTok{ mu\_alpha2;}
  \DataTypeTok{real}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{0}\NormalTok{\textgreater{} sigma\_alpha2;}

  \DataTypeTok{array}\NormalTok{[Nsp] }\DataTypeTok{real}\NormalTok{ mu\_alpha\_sp; }\CommentTok{// species{-}level intercepts}
  \DataTypeTok{array}\NormalTok{[Nsp] }\DataTypeTok{real}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{0}\NormalTok{\textgreater{} sig\_species\_alpha;}

  \DataTypeTok{array}\NormalTok{[Npop] }\DataTypeTok{real}\NormalTok{ alpha\_tilde\_pop\_sp; }\CommentTok{// non{-}centered population intercepts}

  \DataTypeTok{real}\NormalTok{ mu\_beta1; }\CommentTok{// overall species trend}
  \DataTypeTok{real}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{0}\NormalTok{\textgreater{} sigma\_beta1;}

  \DataTypeTok{real}\NormalTok{ mu\_beta2;}
  \DataTypeTok{real}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{0}\NormalTok{\textgreater{} sigma\_beta2;}

  \DataTypeTok{array}\NormalTok{[Nsp] }\DataTypeTok{real}\NormalTok{ mu\_beta\_sp; }\CommentTok{// species{-}level slopes}
  \DataTypeTok{array}\NormalTok{[Nsp] }\DataTypeTok{real}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{0}\NormalTok{\textgreater{} sig\_species\_beta;}

  \DataTypeTok{array}\NormalTok{[Npop] }\DataTypeTok{real}\NormalTok{ beta\_tilde\_pop\_sp; }\CommentTok{// non{-}centered population slopes}

  \DataTypeTok{real}\NormalTok{\textless{}}\KeywordTok{lower}\NormalTok{=}\DecValTok{0}\NormalTok{\textgreater{} sigma;}

\NormalTok{\}}

\KeywordTok{transformed parameters}\NormalTok{\{}

  \DataTypeTok{array}\NormalTok{[Npop] }\DataTypeTok{real}\NormalTok{ alpha\_pop\_sp;}
  \DataTypeTok{array}\NormalTok{[Npop] }\DataTypeTok{real}\NormalTok{ beta\_pop\_sp;}

  \ControlFlowTok{for}\NormalTok{ (p }\ControlFlowTok{in} \DecValTok{1}\NormalTok{:Npop) \{}

\NormalTok{      alpha\_pop\_sp[p] = mu\_alpha\_sp[spid\_perpop[p]] }
\NormalTok{      + sig\_species\_alpha[spid\_perpop[p]] *  alpha\_tilde\_pop\_sp[p];}
\NormalTok{      beta\_pop\_sp[p] = mu\_beta\_sp[spid\_perpop[p]] }
\NormalTok{      + sig\_species\_beta[spid\_perpop[p]] *  beta\_tilde\_pop\_sp[p];}

\NormalTok{  \}}

\NormalTok{\}}

\KeywordTok{model}\NormalTok{ \{}

  \DataTypeTok{vector}\NormalTok{[N] mu;}

\NormalTok{  mu\_alpha\_sp \textasciitilde{} normal(mu\_alpha1, sigma\_alpha1);}
\NormalTok{  sig\_species\_alpha \textasciitilde{} normal(mu\_alpha2, sigma\_alpha2);}

\NormalTok{  mu\_beta\_sp \textasciitilde{} normal(mu\_beta1, sigma\_beta1);}
\NormalTok{  sig\_species\_beta \textasciitilde{} normal(mu\_beta2, sigma\_beta2);}

  \ControlFlowTok{for}\NormalTok{ (p }\ControlFlowTok{in} \DecValTok{1}\NormalTok{:Npop) \{}

\NormalTok{    alpha\_tilde\_pop\_sp[p] \textasciitilde{} normal(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{);}
\NormalTok{    beta\_tilde\_pop\_sp[p] \textasciitilde{} normal(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{);}

\NormalTok{  \}}

  \ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\NormalTok{:N)\{}

\NormalTok{    mu[i] = alpha\_pop\_sp[popid[i]] + beta\_pop\_sp[popid[i]] * (year[i] {-} }\DecValTok{1980}\NormalTok{);}
\NormalTok{    logy[i] \textasciitilde{} normal(mu[i], sigma);}

\NormalTok{  \}}

\NormalTok{  mu\_alpha1 \textasciitilde{} normal(}\DecValTok{0}\NormalTok{, log(}\DecValTok{500}\NormalTok{) / }\FloatTok{2.57}\NormalTok{); }\CommentTok{// {-}log(500) \textless{} mu\_alpha1 \textless{} log(500)}
\NormalTok{  sigma\_alpha1 \textasciitilde{} normal(}\DecValTok{0}\NormalTok{, log(}\DecValTok{100}\NormalTok{) / }\FloatTok{2.57}\NormalTok{); }\CommentTok{// 0 \textless{} sigma\_alpha1 \textless{} log(100)}

\NormalTok{  mu\_alpha2 \textasciitilde{} normal(}\DecValTok{0}\NormalTok{, log(}\DecValTok{500}\NormalTok{) / }\FloatTok{2.57}\NormalTok{); }\CommentTok{// {-}log(500) \textless{} mu\_alpha1 \textless{} log(500)}
\NormalTok{  sigma\_alpha2 \textasciitilde{} normal(}\DecValTok{0}\NormalTok{, log(}\DecValTok{50}\NormalTok{) / }\FloatTok{2.57}\NormalTok{); }\CommentTok{// 0 \textless{} sigma\_alpha1 \textless{} log(50)}

\NormalTok{  mu\_beta1 \textasciitilde{} normal(}\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{ / }\FloatTok{2.57}\NormalTok{); }\CommentTok{// {-}2 \textless{} mu\_beta1 \textless{} 2}
\NormalTok{  sigma\_beta1 \textasciitilde{} normal(}\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{ / }\FloatTok{2.57}\NormalTok{); }\CommentTok{// 0 \textless{} sigma\_beta1 \textless{} 2}

\NormalTok{  mu\_beta2 \textasciitilde{} normal(}\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{ / }\FloatTok{2.57}\NormalTok{); }\CommentTok{//{-}2 \textless{} mu\_beta1 \textless{} 2}
\NormalTok{  sigma\_beta2 \textasciitilde{} normal(}\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{ / }\FloatTok{2.57}\NormalTok{); }\CommentTok{// 0 \textless{} sigma\_beta1 \textless{} 2}

\NormalTok{  sigma \textasciitilde{} normal(}\DecValTok{0}\NormalTok{, log(}\DecValTok{100}\NormalTok{) / }\FloatTok{2.57}\NormalTok{);}

\NormalTok{\}}

\KeywordTok{generated quantities}\NormalTok{ \{}

  \DataTypeTok{vector}\NormalTok{[N] logy\_pred;}

  \ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\NormalTok{:N)\{}

    \DataTypeTok{real}\NormalTok{ mu = alpha\_pop\_sp[popid[i]] }
\NormalTok{    + beta\_pop\_sp[popid[i]] * (year[i] {-} }\DecValTok{1980}\NormalTok{);}
\NormalTok{    logy\_pred[i] = normal\_rng(mu, sigma);}

\NormalTok{  \}}

\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{runmodel }\OtherTok{\textless{}{-}} \ConstantTok{FALSE}

\NormalTok{data }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}
  \AttributeTok{N =} \FunctionTok{nrow}\NormalTok{(datasim),}
  \AttributeTok{Nsp =}\NormalTok{ nspecies,}
  \AttributeTok{Npop =}\NormalTok{ npops,}
  \AttributeTok{y =}\NormalTok{ y,}
  \AttributeTok{year =}\NormalTok{ years,}
  \AttributeTok{spid\_perpop =}\NormalTok{ spid\_perpop,}
  \AttributeTok{popid =}\NormalTok{ popid\_perobs}
\NormalTok{)}

\ControlFlowTok{if}\NormalTok{(runmodel)\{}
\NormalTok{  fit }\OtherTok{\textless{}{-}} \FunctionTok{stan}\NormalTok{(}\AttributeTok{file =} \StringTok{"\textasciitilde{}/projects/forecastflows/analyses/stan/model1\_nc2.stan"}\NormalTok{,}
            \AttributeTok{data =}\NormalTok{ data,  }\AttributeTok{iter=}\DecValTok{4000}\NormalTok{, }\AttributeTok{warmup=}\DecValTok{3000}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{, }\AttributeTok{seed =} \DecValTok{20012029}\NormalTok{)}
  \FunctionTok{saveRDS}\NormalTok{(fit, }\AttributeTok{file =} \FunctionTok{file.path}\NormalTok{(wd, }\StringTok{"output/fit/simulateddatafit.rds"}\NormalTok{))}
\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{}
\NormalTok{  fit }\OtherTok{\textless{}{-}} \FunctionTok{readRDS}\NormalTok{(}\AttributeTok{file =} \FunctionTok{file.path}\NormalTok{(wd, }\StringTok{"output/fit/simulateddatafit.rds"}\NormalTok{))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Our first goal with simulated data is a simple check of our code (both
the model code and simulated data code), given a high enough sample size
relative to our effect sizes and error, the model should return our
parameters. So our first step is to check this.

First, we start with the parameters, which include an overall estimate
of the trend across species (this parameter would address our question
that we decided on above). The red dashed line represents the true
simulated value:

\includegraphics{lpiworkflow_files/figure-pdf/inference sim data params-1.pdf}

Now we check additional parameters. We find the model does a fair job at
recovering population-level parameters.

\begin{figure}

\begin{minipage}{0.50\linewidth}
\includegraphics{lpiworkflow_files/figure-pdf/inference pop sim data-1.pdf}\end{minipage}%
%
\begin{minipage}{0.50\linewidth}
\includegraphics{lpiworkflow_files/figure-pdf/inference pop sim data-2.pdf}\end{minipage}%

\end{figure}%

What if the model did not do a good job at recovering our parameters?
Then we should likely adjust our sample size up and/or our error lower
to make sure it is can (as a basic code test).

But what if the model can recover parameters only at sample sizes much
higher than our data (or at error sizes much lower than our data)? Then
we would need to take this important information forward in our analysis
and be careful in how we interpret estimates from our model fit to
empirical data (as we have a hint at this stage that getting robust
estimates from our data and model together may be difficult, but we do
not know the actual values of our parameters -- including error and
variance parameters -- from our empirical data so we cannot be sure of
whether we may have these issues yet).

For now, we also take a look at how model fits look against the
simulated data (error bars and ribbon represent the 95\% uncertainty
intervals):

\includegraphics{lpiworkflow_files/figure-pdf/retro sim data-1.pdf}

\includegraphics{lpiworkflow_files/figure-pdf/retro sim data time-1.pdf}

These look pretty good, so we move onto the next step.

\section{Step 4: Fit model to empirical
data}\label{step-4-fit-model-to-empirical-data}

Now that we have designed our question, built our model for it, and
tested our model on simulated data, we can fit the model to our
empirical data. We start by some basic clean-up of the dataframe so we
have years in one column.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{lpi\_subset }\OtherTok{\textless{}{-}}\NormalTok{ lpi\_subset[, }\FunctionTok{c}\NormalTok{(}\StringTok{"ID"}\NormalTok{, }\StringTok{"Genus"}\NormalTok{, }\StringTok{"Species"}\NormalTok{, }\StringTok{"Country"}\NormalTok{, }\StringTok{"Units"}\NormalTok{, }
                             \FunctionTok{paste0}\NormalTok{(}\StringTok{"X"}\NormalTok{, }\DecValTok{1950}\SpecialCharTok{:}\DecValTok{2020}\NormalTok{))]}
\NormalTok{lpi\_subset}\SpecialCharTok{$}\NormalTok{genus\_species }\OtherTok{\textless{}{-}} \FunctionTok{paste}\NormalTok{(lpi\_subset}\SpecialCharTok{$}\NormalTok{Genus, lpi\_subset}\SpecialCharTok{$}\NormalTok{Species)}

\NormalTok{lpi\_subset\_rsh }\OtherTok{\textless{}{-}} 
  \FunctionTok{reshape}\NormalTok{(lpi\_subset, }\AttributeTok{direction =} \StringTok{"long"}\NormalTok{, }\AttributeTok{idvar =} \StringTok{"ID"}\NormalTok{,}
          \AttributeTok{varying =}  \FunctionTok{paste0}\NormalTok{(}\StringTok{"X"}\NormalTok{, }\DecValTok{1950}\SpecialCharTok{:}\DecValTok{2020}\NormalTok{), }\AttributeTok{v.names =} \FunctionTok{c}\NormalTok{(}\StringTok{"value"}\NormalTok{), }
          \AttributeTok{timevar =} \StringTok{"year"}\NormalTok{, }\AttributeTok{times =} \DecValTok{1950}\SpecialCharTok{:}\DecValTok{2020}\NormalTok{)}
\NormalTok{lpi\_subset\_rsh }\OtherTok{\textless{}{-}}\NormalTok{ lpi\_subset\_rsh[lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{value }\SpecialCharTok{!=} \StringTok{"NULL"}\NormalTok{,]}
\NormalTok{lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{value }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{value)}
\end{Highlighting}
\end{Shaded}

The values in the database are not necessarily in the same units, so we
subset to units related to absolute population sizes, which is what we
have mostly been thinking of (some units are in density and we would
need to build our model to incorporate those, but here we start first
with a simpler model and can build up later). We take a quick look at
the data to make sure it looks reasonable based on our domain knowledge.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{units\_to\_keep }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Estimate {-} entire population"}\NormalTok{, }\StringTok{"Estimate {-} entire population (August)"}\NormalTok{, }
                   \StringTok{"Individual"}\NormalTok{, }\StringTok{"individuals"}\NormalTok{, }\StringTok{"Individuals"}\NormalTok{, }\StringTok{"Number of adults"}\NormalTok{, }
                   \StringTok{"number of individuals"}\NormalTok{, }\StringTok{"Number of individuals"}\NormalTok{, }\StringTok{"Number of Individuals"}\NormalTok{, }
                   \StringTok{"Number of rhinos"}\NormalTok{, }\StringTok{"numbers of indviduals"}\NormalTok{, }\StringTok{"population estimate"}\NormalTok{, }
                   \StringTok{"Population estimate"}\NormalTok{, }\StringTok{"Population estimate (individuals)"}\NormalTok{, }
                   \StringTok{"Population estimates"}\NormalTok{, }\StringTok{"Total population size"}\NormalTok{)}

\NormalTok{lpi\_subset\_rsh }\OtherTok{\textless{}{-}}\NormalTok{ lpi\_subset\_rsh[lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{Units }\SpecialCharTok{\%in\%}\NormalTok{ units\_to\_keep,]}

\FunctionTok{ggplot}\NormalTok{(}\AttributeTok{data =}\NormalTok{ lpi\_subset\_rsh) }\SpecialCharTok{+}
  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ year, }\AttributeTok{y =}\NormalTok{ value, }
                \AttributeTok{group =} \FunctionTok{paste0}\NormalTok{(genus\_species, ID),}
                \AttributeTok{color =}\NormalTok{ genus\_species)) }\SpecialCharTok{+}
  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{hcl.colors}\NormalTok{(}\DecValTok{14}\NormalTok{, }\AttributeTok{palette =} \StringTok{"Dark2"}\NormalTok{),}
                     \AttributeTok{breaks =} \FunctionTok{unique}\NormalTok{(lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{genus\_species)[}\FunctionTok{order}\NormalTok{(}\FunctionTok{unique}\NormalTok{(lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{genus\_species))]) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{\textquotesingle{}Year\textquotesingle{}}\NormalTok{, }\AttributeTok{y =} \StringTok{\textquotesingle{}Observed counts\textquotesingle{}}\NormalTok{)}
\FunctionTok{ggplot}\NormalTok{(}\AttributeTok{data =}\NormalTok{ lpi\_subset\_rsh[lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{value }\SpecialCharTok{!=} \DecValTok{0}\NormalTok{,]) }\SpecialCharTok{+}
  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ year, }\AttributeTok{y =} \FunctionTok{log}\NormalTok{(value), }
                \AttributeTok{group =} \FunctionTok{paste0}\NormalTok{(genus\_species, ID),}
                \AttributeTok{color =}\NormalTok{ genus\_species)) }\SpecialCharTok{+}
  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{hcl.colors}\NormalTok{(}\DecValTok{14}\NormalTok{, }\AttributeTok{palette =} \StringTok{"Dark2"}\NormalTok{),}
                     \AttributeTok{breaks =} \FunctionTok{unique}\NormalTok{(lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{genus\_species)[}\FunctionTok{order}\NormalTok{(}\FunctionTok{unique}\NormalTok{(lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{genus\_species))]) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{\textquotesingle{}Year\textquotesingle{}}\NormalTok{, }\AttributeTok{y =} \StringTok{\textquotesingle{}Observed counts (log{-}scale)\textquotesingle{}}\NormalTok{, }\AttributeTok{color =} \StringTok{"Species"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}

\begin{minipage}{\linewidth}
\includegraphics{lpiworkflow_files/figure-pdf/units-1.pdf}\end{minipage}%
\newline
\begin{minipage}{\linewidth}
\includegraphics{lpiworkflow_files/figure-pdf/units-2.pdf}\end{minipage}%

\end{figure}%

Now, we make a few final adjustments and fit the model:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{runmodel }\OtherTok{\textless{}{-}} \ConstantTok{FALSE}

\CommentTok{\# We now also find we have counts of 0, which is tricky for our approach }
\CommentTok{\#(as we cannot evaluate log(0)), we would also flag this to return. }
\CommentTok{\# As we iteratively improve the model, we would need to consider what }
\CommentTok{\# processes generate these zeroes, and include them in our model.}
\NormalTok{lpi\_subset\_rsh }\OtherTok{\textless{}{-}}\NormalTok{ lpi\_subset\_rsh[lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{value }\SpecialCharTok{!=} \DecValTok{0}\NormalTok{,] }

\NormalTok{popids }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{ID =} \FunctionTok{unique}\NormalTok{(lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{ID), }
                     \AttributeTok{popid\_num =} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(}\FunctionTok{unique}\NormalTok{(lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{ID)))}
\NormalTok{spids }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{genus\_species =} \FunctionTok{unique}\NormalTok{(lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{genus\_species), }
                    \AttributeTok{spid\_num =} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(}\FunctionTok{unique}\NormalTok{(lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{genus\_species)))}

\NormalTok{lpi\_subset\_rsh }\OtherTok{\textless{}{-}} \FunctionTok{merge}\NormalTok{(lpi\_subset\_rsh, spids)}
\NormalTok{lpi\_subset\_rsh }\OtherTok{\textless{}{-}} \FunctionTok{merge}\NormalTok{(lpi\_subset\_rsh, popids)}

\NormalTok{spid\_perpop }\OtherTok{\textless{}{-}} \FunctionTok{unique}\NormalTok{(lpi\_subset\_rsh[}\FunctionTok{c}\NormalTok{(}\StringTok{"popid\_num"}\NormalTok{, }\StringTok{"spid\_num"}\NormalTok{)])}
\NormalTok{spid\_perpop }\OtherTok{\textless{}{-}}\NormalTok{ spid\_perpop[}\FunctionTok{order}\NormalTok{(spid\_perpop}\SpecialCharTok{$}\NormalTok{popid\_num),]}

\NormalTok{data }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}
  \AttributeTok{N =} \FunctionTok{nrow}\NormalTok{(lpi\_subset\_rsh),}
  \AttributeTok{Nsp =} \FunctionTok{max}\NormalTok{(spids}\SpecialCharTok{$}\NormalTok{spid\_num),}
  \AttributeTok{Npop =} \FunctionTok{max}\NormalTok{(popids}\SpecialCharTok{$}\NormalTok{popid\_num),}
  \AttributeTok{y =}\NormalTok{ lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{value,}
  \AttributeTok{year =}\NormalTok{ lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{year,}
  \AttributeTok{spid\_perpop =}\NormalTok{ spid\_perpop}\SpecialCharTok{$}\NormalTok{spid\_num,}
  \AttributeTok{popid =}\NormalTok{ lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{popid\_num}
\NormalTok{)}

\ControlFlowTok{if}\NormalTok{(runmodel)\{}
\NormalTok{  fit }\OtherTok{\textless{}{-}} \FunctionTok{stan}\NormalTok{(}\AttributeTok{file =} \StringTok{"\textasciitilde{}/projects/forecastflows/analyses/stan/model1\_nc2.stan"}\NormalTok{,}
            \AttributeTok{data =}\NormalTok{ data,  }\AttributeTok{iter=}\DecValTok{4000}\NormalTok{, }\AttributeTok{warmup=}\DecValTok{3000}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{, }\AttributeTok{seed =} \DecValTok{20012029}\NormalTok{)}
  \FunctionTok{saveRDS}\NormalTok{(fit, }\AttributeTok{file =} \FunctionTok{file.path}\NormalTok{(wd, }\StringTok{"output/fit/truedatafit.rds"}\NormalTok{))}
\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{}
\NormalTok{  fit }\OtherTok{\textless{}{-}} \FunctionTok{readRDS}\NormalTok{(}\AttributeTok{file =} \FunctionTok{file.path}\NormalTok{(wd, }\StringTok{"output/fit/truedatafit.rds"}\NormalTok{))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Here we finally get to see how our model performs on the empirical data
and are a step closer to answering our question, but we also start to
notice some differences in our empirical data from our simulated data.
In preparing our data for the model, we realize we have quite different
amplitudes of variations across populations and species (on a
log-scale), e.g.~for blue wildebeest (each color represents a different
population):

\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-4-1.pdf}

Our model may accommodate this, as we have allowed different populations
to have different intercepts (starting population sizes), but the scale
of this variation is much larger than we simulated so we would again
want to flag this and check how our model performs with these data.

\section{Step 5: Retrodictive and related
checks}\label{step-5-retrodictive-and-related-checks}

Now we use our fitted model parameters to generate predictions, and
compare them to observations. This is similar to the simulating data
step above, but differs in two major ways: (1) we simulate from the
parameters estimated by the model fit to the empirical data (versus
coming up with them based on our expert knowledge) and (2) we need to
check across many predictions from the model to understand the full
uncertainty of the model (and not just one draw from the model).

Here we start by visually comparing the posterior predictive
distribution of the predicted counts (across all species and
populations, red line) to observed data (black dashed line):

\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-5-1.pdf}

We can also look at some posterior replications separately:

\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-6-1.pdf}

We then examine, by species, how well predictions from the model compare
to the observed data; this is similar to many plots where we simply want
to see how well the model seems to predict the data.

\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-7-1.pdf}

To evaluate our model more precisely, we need to check how the model
predicts the data on finer scales. So we start by examining some species
separately, e.g.~here the African bush elephant (each color within a
country represents a different population):

\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-8-1.pdf}

Focusing on populations in South Africa, we see some places where the
linear model appears to fit reasonably well (ID 7673), but also one
population where it does not appear as a very good fit (ID 10010).

\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-9-1.pdf}

The three populations are measured in the same units, but are very
different in their sizes (ranging from a few hundred individuals to more
than 10000), thus they likely represent populations measured across
different spatial scales.

The same holds for other species and other countries, e.g.~the blue
wildebeest in Tanzania:

\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-10-1.pdf}

The four populations are also measured in the same basic units, but
represent likely represent very different populations, spanning ones in
small parks that cannot migrate to the large migrating herds (e.g.,
Serengeti Masai-Mara), though gathering additional metadata to fully
check this would be another important component to add to our workflow
for these data.

\section{Inference}\label{inference}

Once we are confident with our model (which is not truly the case here),
we can look at the parameter values and check what the overall and
species trends appear to be:

\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-11-1.pdf}

\section{Feedbacks}\label{feedbacks}

At this point, we would have a lot of ideas of ways to improve this
model that we would want to follow up on, including (but not limited to)
questioning our assumption of one observational error \(\sigma\) for all
populations, how we model population trends (e.g., there are many ways
to model abundance), and (relatedly) our linearity assumption. One
possibility would be to move to a non-linear regression, and use a more
flexible model that would allow population numbers to stabilize, such as
shown using a logistic model for one wildbeest population in Tanzania:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{subset\_wildebeest }\OtherTok{\textless{}{-}}\NormalTok{ lpi\_subset\_rsh[lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{ID }\SpecialCharTok{==} \DecValTok{5589}\NormalTok{,]}
\NormalTok{subset\_wildebeest}\SpecialCharTok{$}\NormalTok{year }\OtherTok{\textless{}{-}}\NormalTok{ subset\_wildebeest}\SpecialCharTok{$}\NormalTok{year}\DecValTok{{-}1980}
\NormalTok{gompodel }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(}\FunctionTok{log}\NormalTok{(value) }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(year, Asym, b2, b3), }\AttributeTok{data=}\NormalTok{subset\_wildebeest)}
\NormalTok{logisfit }\OtherTok{\textless{}{-}}\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{pred =} \FunctionTok{predict}\NormalTok{(gompodel, }
                                     \AttributeTok{newdata =} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{year =} \FunctionTok{seq}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{25}\NormalTok{,}\DecValTok{30}\NormalTok{,}\DecValTok{1}\NormalTok{))), }
                      \AttributeTok{year =} \FunctionTok{seq}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{25}\NormalTok{,}\DecValTok{30}\NormalTok{,}\DecValTok{1}\NormalTok{))}

\FunctionTok{ggplot}\NormalTok{(}\AttributeTok{data =}\NormalTok{ lpi\_subset\_rsh[lpi\_subset\_rsh}\SpecialCharTok{$}\NormalTok{ID }\SpecialCharTok{==} \DecValTok{5589}\NormalTok{,]) }\SpecialCharTok{+}
  \FunctionTok{facet\_wrap}\NormalTok{( }\SpecialCharTok{\textasciitilde{}}\NormalTok{ ID, }\AttributeTok{scales =} \StringTok{"free\_y"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ year, }\AttributeTok{y =}\NormalTok{ value),}
             \AttributeTok{size =} \FloatTok{0.7}\NormalTok{, }\AttributeTok{color =} \StringTok{"grey"}\NormalTok{) }\SpecialCharTok{+} 
  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ year, }\AttributeTok{y =}\NormalTok{ value),}
             \AttributeTok{linewidth =} \FloatTok{0.3}\NormalTok{, }\AttributeTok{color =} \StringTok{"grey"}\NormalTok{) }\SpecialCharTok{+} 
  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ year }\SpecialCharTok{+} \DecValTok{1980}\NormalTok{, }\AttributeTok{y =} \FunctionTok{exp}\NormalTok{(pred)),}
            \AttributeTok{linewidth =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{data =}\NormalTok{ logisfit, }\AttributeTok{linetype =} \StringTok{\textquotesingle{}dashed\textquotesingle{}}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{() }\SpecialCharTok{+} 
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{y =} \StringTok{\textquotesingle{}Counts\textquotesingle{}}\NormalTok{, }\AttributeTok{x =} \StringTok{\textquotesingle{}Year\textquotesingle{}}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-12-1.pdf}

We also have seen variability in population trends within species (e.g.,
two apparently increasing populations and two apparently decreasing
populations in Tanzania), so we would want to think a lot more about how
we scale up from population to species to capture this variation in a
way that is most useful to our question. For example, may want to think
about adding country, which may capture differences in efforts related
to conservation and thus predict trends (or we may have better metrics
of what drives these trends, but given our global aim, we would ideally
need the predictor for most/all populations) and we may want to think
about how or if we partially pool species and populations within species
(potentially using models other than a normal distribution for pooling).
As we address these potential problems, we would likely begin to add in
more countries, species, and populations.

We expect scaling up to more countries, species and populations may
cause us to think carefully about the data we have in hand and our
approach versus the question we aimed to answer. While many recent
papers treat the LPI data as (effectively) a well stratified and
randomized sample or even a good census, the LPI data is clearly
neither. Thus, models that try to deal with such uneven sampling and
build in more of the structure of data would be more appropriate. For
example, most models for election forecasts now include regularization
via partial pooling, post-stratification and other methods designed to
adjust for the fact that many polls are `non-representative.' Such
non-representative polls may be very similar to LPI data. Our model
includes some regularization via partial pooling, but we may need much
more of an observational model added to our population demography model
to accommodate the uneven sampling. Estimates from such a model may be
more uncertain, but likely also more correct given our combination of
question and data.

This may lead us to realize we really don't have great data for this
question and we would then need to prioritize better data collection
efforts if we want good estimates to answer this question.

\section{The full process}\label{the-full-process}

The small workflow we show here may seem like extra work to those who
have not used it, and it is compared to many ecological workflows in
practice. We argue that the extra effort is worth it given the critical
questions we are addressing, and also that the process becomes much
quicker with practice. The first time may be slow, but the resulting
model will be better for it (even if it is still arguably simple, as the
one we show here is), but the following uses of this iterative workflow
will go faster and very likely yield improved models.

\end{document}

### 2025-09-19 revision/r2/forecastflows_r2.tex
slowly continue revisions
finalize mroe or less revisions

@@
 	\centering
-	\includegraphics{../../figures/figure_modeldataspaces_revised}
+	\includegraphics{../../figures/figure_modeldataspaces_revised2}
 	\caption{Different trajectories of current (b-c) and proposed (d) workflows through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
@@
 
-This reality should drive researchers to use more robust and coherent methods.
+\llabel{robust} This reality should drive researchers to use more coherent methods that remain robust for the intended inferential goals, even with imperfect data and knowledge gaps.
 But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
@@
 With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important.
+\llabel{clearerML} A careful and coherent workflow is essential to preserve the interpretability of non–machine-learning models---their primary advantage.
 \llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy, especially if the latter lack a robust estimation of their parameters and fall in a complexity trap---a trap with additional costs to interpretability of process-based models.

### 2025-09-22 revision/r2/response_r2.tex
Revisions to letter (have not reviewed main text) and started a README

--- 2025-09-19 revision/r2/response_r2.tex
@@
 
-Editor and reviewer comments (we provide below the full context of the two reviewers' comments) are in \emph{italics}, while our responses are in regular text. \\
+Editor and reviewer comments are in \emph{gray italics} (we provide below the full context of the reviewers 1-2 comments below, and extract the parts of reviewer 3' comments that required changes to this manuscript), while our responses are in regular text.
+\\
 
@@
 We have worked to fix the last small points raised by the reviewers, and we would like to thank them again for their careful reading.
-Furthermore, we also particularly appreciate the really interesting perspective on the policy implications from M.
-Authier, and...
+Furthermore, we also particularly appreciate the perspective on the policy implications from M.
+Authier, and will reach out to him to discuss further directly (as both he and we appear to agree that this is perhaps the topic for another paper).
 
@@
 {\bf Reviewer 1}
+% emwSep22: Go through and fix " -> ``
 
@@
 
-Changed to: ``Calibrating submodels separately makes possible to avoid the fact that [...]''
+% emwSep22: I changed this in the text and here; your previous version needs an 'it' (makes it possible) but I wanted to remove 'fact' so adjusted more. 
+Changed to: ``Calibrating submodels separately makes it easier to avoid the likely reality that [...]''
 
@@
 ...
+% emwSep22:  I actually hate the idea that a model needs to be complex, but it seems common among ecologists reviewing our workflow papers ....
+I might just say: We disagree that if the model is not sufficiently complex it leaves `something to be desired.' As we want this workflow to be approachable and leave room for readers to think through how they would advance upon it (to whatever degree of complexity they see as biological relevant), we have left it as is.
 
@@
 
-...
+... % emwSep22: Just add a comment to the Stan code or text regarding how Stan works, and say: we understand this confusion and have updated the text.
 
@@
 
-Sorry we forgot to share the Stan code. We added the code in the workflow example.
+Apologies that we forgot to share the Stan code. We have now added the code in the workflow example.
 
@@
 % Hmm I don't understand?
+% emwSep22:  I think this must be the figure on pg 15 -- we show just one draw of the PPC (red), we could overlap more (like 10-30) in a lighter shade.
+And put the observed data black ON TOP so easier to see.
+I think this would be good to do....
+just one draw of a PPC is okay, but not greatI IMHO.
 
-{\bf Reviewer 3}
+{\bf Reviewer 3} % emwSep22:  I think you should extract only the parts that require a change to the current text.
+I will read the below once you have that done and am happy to read a separate letter to MA.
 

### 2025-09-22 revision/r2/forecastflows_r2.tex
Revisions to letter (have not reviewed main text) and started a README

@@
 
-\llabel{robust} This reality should drive researchers to use more coherent methods that remain robust for the intended inferential goals, even with imperfect data and knowledge gaps.
+This reality should drive researchers to use more robust and coherent methods.
 But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
@@
 The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model.
-Calibrating submodels separately makes possible to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.
+Calibrating submodels separately makes it easier to avoid the likely reality that, if the submodels were fitted together (as a whole), many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.
 %emw24Jul: nice stuff here.

### 2025-09-26 revision/r2/response_r2.tex

--- 2025-09-19 revision/r2/response_r2.tex
@@
 
-... % Lizzie?
+...
+% emwSep22:  I actually hate the idea that a model needs to be complex, but it seems common among ecologists reviewing our workflow papers ....
+I might just say: We disagree that if the model is not sufficiently complex it leaves `something to be desired.' As we want this workflow to be approachable and leave room for readers to think through how they would advance upon it (to whatever degree of complexity they see as biological relevant), we have left it as is.
 
@@
 
-We now mention that sigmas are bounded to zero.
+We understand this confusion and now mention that sigmas are bounded to zero.
 
@@
 
-% Hmm I'm not sure to understand what he wants?
 We now specify that the first posterior predictive distribution we show corresponds to the predicted counts across all species and populations.
+% Hmm I don't understand?
+% emwSep22:  I think this must be the figure on pg 15 -- we show just one draw of the PPC (red), we could overlap more (like 10-30) in a lighter shade.
+And put the observed data black ON TOP so easier to see.
+I think this would be good to do....
+just one draw of a PPC is okay, but not greatI IMHO.
 
-{\bf Reviewer 3}
+{\bf Reviewer 3} % emwSep22:  I think you should extract only the parts that require a change to the current text.
+I will read the below once you have that done and am happy to read a separate letter to MA.
 

### 2025-09-26 revision/r2/response_r2.tex
modify letter

@@
 Authier to share our detailed perspective on this.
+Below we address the comment that require changes to the current text.
 
@@
 ...
+% do we have t ochange something here?
 
-\begin{mybox}
-\emph{In policy, a trend in abundance is needed as evidence for decision and further actions. Whether
-that trend comes from a phenomenological or mechanistic approach is often of secondary
-interest to policy-makers: what matters is that the trend passes some checks (including
-statistical significance). Much has been written on the topic, especially with respect to
-statistical power. So “outlin[ing] the steps of a universal workflow” (line 48-49) is both novel
-and worthy. Yet I am not convinced that it will solve some of issues that are pregnant in the
-policy sphere. For example, timeliness is important in policy. As the authors are right to point,
-workflows require time to be developed. Sometimes a lot more time is also needed just to agree
-on the workflow, and then there is time and man-power needed for documenting, maintaining
-and updating the workflow. In short, one need a governance structure for the workflow. If the
-goal is to evidence a negative trend because this is the minimum requirement for taking
-corrective actions, then the authors’ proposal, while clearly an improvement for the research
-sphere, will not impress the policy sphere as it is also likely to deliver too late; nor will it
-necessarily lead to better decisions. Awareness, design and publication of workflows will also
-have to happen in relevant policy circles, not just in scientific journals. That we should start
-with scientific journals in the right first steps, but reaching out to the policy sphere will not
-happen on its own either.}  
-\end{mybox}
+% \begin{mybox}
+% \emph{In policy, a trend in abundance is needed as evidence for decision and further actions. Whether
+% that trend comes from a phenomenological or mechanistic approach is often of secondary
+% interest to policy-makers: what matters is that the trend passes some checks (including
+% statistical significance). Much has been written on the topic, especially with respect to
+% statistical power. So “outlin[ing] the steps of a universal workflow” (line 48-49) is both novel
+% and worthy. Yet I am not convinced that it will solve some of issues that are pregnant in the
+% policy sphere. For example, timeliness is important in policy. As the authors are right to point,
+% workflows require time to be developed. Sometimes a lot more time is also needed just to agree
+% on the workflow, and then there is time and man-power needed for documenting, maintaining
+% and updating the workflow. In short, one need a governance structure for the workflow. If the
+% goal is to evidence a negative trend because this is the minimum requirement for taking
+% corrective actions, then the authors’ proposal, while clearly an improvement for the research
+% sphere, will not impress the policy sphere as it is also likely to deliver too late; nor will it
+% necessarily lead to better decisions. Awareness, design and publication of workflows will also
+% have to happen in relevant policy circles, not just in scientific journals. That we should start
+% with scientific journals in the right first steps, but reaching out to the policy sphere will not
+% happen on its own either.}  
+% \end{mybox}
 
-...
+% ...
 
@@
 
-\begin{mybox}
-\emph{To be clearer, I will provide an example (see Authier 2025). Using citizen science data on 724
-species, Callaghan et al. (2021) estimated global population sizes for 9,700 extant bird species
-with modelling. These model-based estimates are not gauged against independent estimates
-for species well-monitored at the global scale. Owing to their large-size and conspicuousness
-several seabirds species can be censused accurately. Although imperfect, these estimates
-represent the best available knowledge for policy (ACAP 2021). Model-based abundances
-were > 20 times higher/smaller than IUCN ones for 45/5 (out of 96 seabird species) and up to
-100 times higher/smaller for 18/1 species (Figure 1). This is particularly concerning for
-endemic species, in which the local and global geographical scales are completely confounded,
-and which can comprise a single small population. A startling example is the Mascarene petrel
-(Pseudobulweria aterrima), a marine megafauna species endemic to Reunion Island in the Indian
-Ocean. Data on Mascarene petrels were used in the model training set and the estimated
-abundance by Callaghan et al. (2021) was several thousands of individuals. The estimated
-population size from local scientists is less than a few hundred individuals (Lopez et al. 2021).
-The global analysis is not helpful in my opinion, and may even hinder local policy which have
-taken a long time to gain attention and momentum.} 
-\end{mybox}
+% \begin{mybox}
+% \emph{To be clearer, I will provide an example (see Authier 2025). Using citizen science data on 724
+% species, Callaghan et al. (2021) estimated global population sizes for 9,700 extant bird species
+% with modelling. These model-based estimates are not gauged against independent estimates
+% for species well-monitored at the global scale. Owing to their large-size and conspicuousness
+% several seabirds species can be censused accurately. Although imperfect, these estimates
+% represent the best available knowledge for policy (ACAP 2021). Model-based abundances
+% were > 20 times higher/smaller than IUCN ones for 45/5 (out of 96 seabird species) and up to
+% 100 times higher/smaller for 18/1 species (Figure 1). This is particularly concerning for
+% endemic species, in which the local and global geographical scales are completely confounded,
+% and which can comprise a single small population. A startling example is the Mascarene petrel
+% (Pseudobulweria aterrima), a marine megafauna species endemic to Reunion Island in the Indian
+% Ocean. Data on Mascarene petrels were used in the model training set and the estimated
+% abundance by Callaghan et al. (2021) was several thousands of individuals. The estimated
+% population size from local scientists is less than a few hundred individuals (Lopez et al. 2021).
+% The global analysis is not helpful in my opinion, and may even hinder local policy which have
+% taken a long time to gain attention and momentum.} 
+% \end{mybox}
 
-This is a very interesting example.
+% This is a very interesting example.
 As highlighted in your next comment, maybe Callaghan et al.
@@
 
-\begin{mybox}
-\emph{Callaghan et al. (2021) are concerned about forking paths and aware of reproducibility issues.
-They described a validation workflow which does not take stock of the pre-data/post-model
-corner but claimed nevertheless accuracy and success in redressing some biases (e.g.
-geographic and taxonomic). More generally, this example illustrates the following science-
-policy workflow: policy-makers read about a new global analysis published in a scientific
-journal. The analysis has a model workflow. Policy-makers look for the modelled-based
-abundance of their pet species and compare those to their knowledge and data. Local, national
-or regional scientists involved in policy are then contacted to explain any discrepancies. If
-there are indeed discrepancies, the scientists have now to explain their origins, looking into the
-model workflow and how it espoused good practices or not. With a growing familiarity of
-what are workflows, this should become more and more seamless in the long run. In the short
-run however, this science-policy workflow (i.e. the linear model of science, see below) is not
-scalable with respect to scientist’s time in my opinion and it will not necessarily help policy if
-scientists (especially non-tenured ones) think that this is time spent not doing field work,
-writing grants and supervising students to further their research.}  
-\end{mybox}
+% \begin{mybox}
+% \emph{Callaghan et al. (2021) are concerned about forking paths and aware of reproducibility issues.
+% They described a validation workflow which does not take stock of the pre-data/post-model
+% corner but claimed nevertheless accuracy and success in redressing some biases (e.g.
+% geographic and taxonomic). More generally, this example illustrates the following science-
+% policy workflow: policy-makers read about a new global analysis published in a scientific
+% journal. The analysis has a model workflow. Policy-makers look for the modelled-based
+% abundance of their pet species and compare those to their knowledge and data. Local, national
+% or regional scientists involved in policy are then contacted to explain any discrepancies. If
+% there are indeed discrepancies, the scientists have now to explain their origins, looking into the
+% model workflow and how it espoused good practices or not. With a growing familiarity of
+% what are workflows, this should become more and more seamless in the long run. In the short
+% run however, this science-policy workflow (i.e. the linear model of science, see below) is not
+% scalable with respect to scientist’s time in my opinion and it will not necessarily help policy if
+% scientists (especially non-tenured ones) think that this is time spent not doing field work,
+% writing grants and supervising students to further their research.}  
+% \end{mybox}
 
-First, we would like to answer about the time it may take for this type of workflow to become more common in research spheres.
+% First, we would like to answer about the time it may take for this type of workflow to become more common in research spheres.
 A large portion of scientific results today are investigated and written by non-tenured scientists.
@@
 
-\begin{mybox}
-\emph{The above scenario (which can be applied to global analyses with or without workflows) is not
-to deflect from the authors' main point being made that “that an improved workflow that
-required data simulation and retrodictive checks would lead to larger model advances and a
-greater recognition of uncertainty—thus highlighting likely consistency in estimates across
-models—that could better aid policy”. The “could better aid policy” part is simply too vague
-and could betray, in my opinion, that the proposed “universal framework” is implicitly
-embracing the linear model of science (Fernándes 2016) of 'truth speaks to power'. In other
-words, bad policy decisions are assumed to mainly result from a deficit of knowledge so that
-all that is needed would be better knowledge to naturally lead to good policy decision. I find
-this less and less convincing given that we’ve never had so much data or knowledge than now.
-We know what are the main drivers of biodiversity loss (e.g. Maxwell et al. 2016). We are
-literally awash with models (just think about the number of Species Distribution Models being
-published nowadays). Yet biodiversity, on average, does not fare any better.}  
-\end{mybox}
+% \begin{mybox}
+% \emph{The above scenario (which can be applied to global analyses with or without workflows) is not
+% to deflect from the authors' main point being made that “that an improved workflow that
+% required data simulation and retrodictive checks would lead to larger model advances and a
+% greater recognition of uncertainty—thus highlighting likely consistency in estimates across
+% models—that could better aid policy”. The “could better aid policy” part is simply too vague
+% and could betray, in my opinion, that the proposed “universal framework” is implicitly
+% embracing the linear model of science (Fernándes 2016) of 'truth speaks to power'. In other
+% words, bad policy decisions are assumed to mainly result from a deficit of knowledge so that
+% all that is needed would be better knowledge to naturally lead to good policy decision. I find
+% this less and less convincing given that we’ve never had so much data or knowledge than now.
+% We know what are the main drivers of biodiversity loss (e.g. Maxwell et al. 2016). We are
+% literally awash with models (just think about the number of Species Distribution Models being
+% published nowadays). Yet biodiversity, on average, does not fare any better.}  
+% \end{mybox}
 
-We agree that the main obstacles---at least at the global scale---for effective climate and biodiversity policies are not a lack of knowledge.
+% We agree that the main obstacles---at least at the global scale---for effective climate and biodiversity policies are not a lack of knowledge.
 However, coming back to your previous point (``\emph{Local, national or regional scientists involved in policy are then contacted to explain any discrepancies}''), we believe a different interpretation of the usefulness of the workflow for policy is possible.
@@
 
-\begin{mybox}
-\emph{While I agree with the authors that the workflow will help research spheres in disciplining
-modelling, I disagree with the policy implications. As highlighted in the box by the authors,
-“[m]odel development is the central step of the process-based workflow, typically requiring
-several years, yet it often remains opaque for anyone who has not worked the model itself.”
-This is not a small issue.}  
-\end{mybox}
+% \begin{mybox}
+% \emph{While I agree with the authors that the workflow will help research spheres in disciplining
+% modelling, I disagree with the policy implications. As highlighted in the box by the authors,
+% “[m]odel development is the central step of the process-based workflow, typically requiring
+% several years, yet it often remains opaque for anyone who has not worked the model itself.”
+% This is not a small issue.}  
+% \end{mybox}
 
-We wanted to highlight that process-based model development can sometimes be very long (with or without this workflow), and thus that it would be better to follow coherent steps in model development and checking---given the amount of time process-based model development takes anyway.
+% We wanted to highlight that process-based model development can sometimes be very long (with or without this workflow), and thus that it would be better to follow coherent steps in model development and checking---given the amount of time process-based model development takes anyway.
 But to be clear, for most applications, applying the workflow do not require years.
 
-\begin{mybox}
-\emph{For example, in the International Whaling Commission, established in
-1946, many scientists are now retiring without new ones stepping in their shoes at the Scientific
-Committee (SC). The IWC SC pioneered some work on simulations (e.g. De la Mare 1986) and
-devised complex population models that were heavily tested to assess whether they would be
-fit-for-purposes. It is ironic that the simulation-savvy and model-based procedures
-encapsulated in the Revised Management Procedure that was developed in the late 1980s and
-early 1990s was never used in practice (for policy reasons, not scientific ones, to make a long
-story short).}  
-\end{mybox}
+% \begin{mybox}
+% \emph{For example, in the International Whaling Commission, established in
+% 1946, many scientists are now retiring without new ones stepping in their shoes at the Scientific
+% Committee (SC). The IWC SC pioneered some work on simulations (e.g. De la Mare 1986) and
+% devised complex population models that were heavily tested to assess whether they would be
+% fit-for-purposes. It is ironic that the simulation-savvy and model-based procedures
+% encapsulated in the Revised Management Procedure that was developed in the late 1980s and
+% early 1990s was never used in practice (for policy reasons, not scientific ones, to make a long
+% story short).}  
+% \end{mybox}
 
-We are not familiar with this example.
+% We are not familiar with this example.
 Even without considering policy reasons, we do feel however that science sometimes goes in circles---and that great improvements in computational tools since 90s have not necessarily led to more robust models.
 
-\begin{mybox}
-\emph{The IWC example underlines the long-term commitment that is needed and the bigger
-question is how to square such a commitment in a competitive academic landscape where
-projects, when they are funded, are meant for a couple of years and the research proletariat is
-expected to publish regularly in prestigious journals and be highly mobile with respect to
-institutions. I contend that the incentives for the research and policy spheres are simply not
-aligned enough at the moment. How does one ensure the sort of long-term commitment
-needed in policy where stability of procedures is important (as any management procedures or
-indicators are going to have to jump many hoops to get agreed by all parties) and a steady
-stream of publications in an academic ecosystem where novelty and innovations are a
-premium?}  
-\end{mybox}
+% \begin{mybox}
+% \emph{The IWC example underlines the long-term commitment that is needed and the bigger
+% question is how to square such a commitment in a competitive academic landscape where
+% projects, when they are funded, are meant for a couple of years and the research proletariat is
+% expected to publish regularly in prestigious journals and be highly mobile with respect to
+% institutions. I contend that the incentives for the research and policy spheres are simply not
+% aligned enough at the moment. How does one ensure the sort of long-term commitment
+% needed in policy where stability of procedures is important (as any management procedures or
+% indicators are going to have to jump many hoops to get agreed by all parties) and a steady
+% stream of publications in an academic ecosystem where novelty and innovations are a
+% premium?}  
+% \end{mybox}
 
-...
+% ...
 
-\begin{mybox}
-\emph{The authors are alluding to the current publishing culture as a barrier but do not elaborate
-much. That is fair to omit in a short paper, and again I agree with their main point with respect to the research spheres.}
-\end{mybox}
+% \begin{mybox}
+% \emph{The authors are alluding to the current publishing culture as a barrier but do not elaborate
+% much. That is fair to omit in a short paper, and again I agree with their main point with respect to the research spheres.}
+% \end{mybox}
 
-...
+% ...
 
-\begin{mybox}
-\emph{I do think, however, that evidencing a trend serves a different goal in
-policy, and that having a workflow to bridge trend estimation and forecasting is not as urgent
-as the authors imply if the ultimate aim is to enable timely decisions to address biodiversity
-loss. In fact, the insistence on workflow may backfire simply because it can easily lead to either
-distraction or displacement in the sense of Rainer (2012). Distraction may happen because it
-could be easy to ask for more testing of some model features (a new twist to some mechanistic
-parts in the model) and ask that no decision be taken before that an updated workflow is
-completed. This could easily align with the common policy requirement of using the ‘best
-available evidence’, especially if some cutting-edge new results are brought forward at the
-policy table and one is being tasked with updating the current model. Displacement is more
-subtle but can also happen: managing the workflow becomes the policy and simply
-maintaining it absorb most of the available resources. As Rayner (2012) puts it “displacement
-occurs when an organization engages with an issue, but substitutes management of a
-representation of a problem (such as a computer model) for management of the represented
-object or activity.” I think these two issues are real.} 
-\end{mybox}
+% \begin{mybox}
+% \emph{I do think, however, that evidencing a trend serves a different goal in
+% policy, and that having a workflow to bridge trend estimation and forecasting is not as urgent
+% as the authors imply if the ultimate aim is to enable timely decisions to address biodiversity
+% loss. In fact, the insistence on workflow may backfire simply because it can easily lead to either
+% distraction or displacement in the sense of Rainer (2012). Distraction may happen because it
+% could be easy to ask for more testing of some model features (a new twist to some mechanistic
+% parts in the model) and ask that no decision be taken before that an updated workflow is
+% completed. This could easily align with the common policy requirement of using the ‘best
+% available evidence’, especially if some cutting-edge new results are brought forward at the
+% policy table and one is being tasked with updating the current model. Displacement is more
+% subtle but can also happen: managing the workflow becomes the policy and simply
+% maintaining it absorb most of the available resources. As Rayner (2012) puts it “displacement
+% occurs when an organization engages with an issue, but substitutes management of a
+% representation of a problem (such as a computer model) for management of the represented
+% object or activity.” I think these two issues are real.} 
+% \end{mybox}
 
-This is a very interesting question.
+% This is a very interesting question.
 However, we believe that the previous example of Callaghan et al.
@@
 
-\begin{mybox}
-\emph{I would love to actually read the authors’ thoughts here and ideally be proven wrong that an
-improved workflow would be enough for enabling policy decisions whose beneficial impacts
-would be commensurate with the current magnitude of biodiversity loss7. In particular, I
-would be interested in a better articulation of how they see global studies feed-backing on
-policy which is usually more local, national or regional.}  
-\end{mybox}
+% \begin{mybox}
+% \emph{I would love to actually read the authors’ thoughts here and ideally be proven wrong that an
+% improved workflow would be enough for enabling policy decisions whose beneficial impacts
+% would be commensurate with the current magnitude of biodiversity loss7. In particular, I
+% would be interested in a better articulation of how they see global studies feed-backing on
+% policy which is usually more local, national or regional.}  
+% \end{mybox}
 
-...
+% ...
 
@@
 time or taxonomy and its consequences in modelling results. Preferential sampling is more
-descriptive than bias, which also has a prescriptive meaning, esp. in policy8. Also, a well-
+descriptive than bias, which also has a prescriptive meaning, esp. in policy. Also, a well-
 designed survey at some spatial scale may become an instance of preferential geographic
@@
 ...
+% Hmm... well... I prefer bias. Preferential sampling sounds to me like it's done on purpose
 
@@
 
-...
+The workflow may allow to identify a missing phylogenetic structure that could be used to improve inference for rare species (with information from close taxa).
+Moreover, the workflow can also help identify the limits of our model---e.g.
+through data simulation, we could see that the model does not do a good job at predicting data-poor species.
+This could thus help to identify priorities for data collection.
 

### 2025-09-26 figures/retrodic/supp_figure.R
add missing script for Lizzie---since she loves so much working on this!

@@ -28,7 +28,8 @@ beta_root <- 1.5
 lambda_alpha <- 0.99
 lambda_beta  <- 0.99
 
-s <- 3804643
+# s <- 38046488
+s <- 380977
 set.seed(s)
 
 {
@@ -38,8 +39,8 @@ set.seed(s)
   corrmat_alpha <- lambda_alpha * corrmat + (1 - lambda_alpha) * diag(nrow(corrmat))
   corrmat_beta  <- lambda_beta * corrmat + (1 - lambda_beta) * diag(nrow(corrmat))
   
-  sigma_y <- 0.1
-  sigma_alpha <- 0.3
+  sigma_y <- 2
+  sigma_alpha <- 1.5
   sigma_beta <- 1.5
   
   alpha <- MASS::mvrnorm(1, mu = rep(alpha_root, nspecies), Sigma = sigma_alpha^2 * corrmat_alpha)
@@ -48,8 +49,7 @@ set.seed(s)
   datasim <- data.frame()
   spid <- c()
   for(i in 1:nspecies){
-    # nperspecies <- ifelse(i == 6, 100, 5)
-    nperspecies <- 5
+    nperspecies <- ifelse(i %in% c(1:3), 20, 3)
     x <- rnorm(nperspecies, mean = 0, sd = 3)
     y <- alpha[i] + beta[i] * x + rnorm(nperspecies, mean = 0, sd = sigma_y)
     datasim <- rbind(
@@ -86,7 +86,7 @@ set.seed(s)
   
   par(mfrow=c(1, 2))
   yrep <- rstan::extract(fit, pars = "y_pred")$y_pred 
-  resid_matrix <- sweep(yrep, 2, data$y, FUN = "-")  
+  resid_matrix <- sweep(yrep, 2, data$y, FUN = "-")
   # resid_mean <- colMeans(resid_matrix) 
   residnophylo <- data.frame(mod = "Standard", sp = 1:data$Nsp,  t(sapply(1:data$Nsp, function(sp) quantile(as.numeric(resid_matrix[,data$spid==sp]), probs = c(0.25, 0.5, 0.75)))),
                              Xmean = sapply(1:data$Nsp, function(sp) mean(as.numeric(resid_matrix[,data$spid==sp]))))
@@ -99,66 +99,99 @@ set.seed(s)
   
   
   # Below code for plots from https://www.joelnitta.com/posts/2021-06-02_color-scheme-anc-states/
[…]

### 2025-09-26 analyses/lpiworkflow.qmd
removing one of the PPC figure to make everyone happy!

@@
 
-Here we start by visually comparing the posterior predictive distribution of the predicted counts (across all species and populations, red line) to observed data (black dashed line):
+Here we start by visually comparing the 9 draws from the posterior predictive distribution of counts (across all species and populations, red line) to observed data (black dashed line):
 ```{r}
@@
 #| echo: false
+#| eval: false
 
 names <- paste0('logy_pred[', 1:data$N, ']')
-retrodictions <- unlist(extract(fit, pars = names))
+retrodictions <- extract(fit, pars = names)
+m <- t(sapply(names, function(n) retrodictions[[n]][1:10]))
 
 ggplot() +
-  geom_density(data = data.frame(obs = log(data$y)), aes(x = obs), alpha = 0.5, linetype = 'dashed') +
-  geom_density(data = data.frame(retrodictions), aes(x = retrodictions), alpha = 0.5, col = 'darkred') +
+  geom_density(data = data.frame(retro = m[,1]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .5) +
+  geom_density(data = data.frame(retro = m[,2]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .5) +
+  geom_density(data = data.frame(retro = m[,3]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .5) +
+  geom_density(data = data.frame(retro = m[,4]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .5) +
+  geom_density(data = data.frame(retro = m[,5]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .5) +
+  geom_density(data = data.frame(retro = m[,6]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .5) +
+  geom_density(data = data.frame(retro = m[,7]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .5) +
+  geom_density(data = data.frame(retro = m[,8]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .5) +
+  geom_density(data = data.frame(retro = m[,9]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .5) +
+  geom_density(data = data.frame(retro = m[,10]), aes(x = retro), alpha = 0.2, col = 'darkred') +
+  geom_density(data = data.frame(obs = log(data$y)), aes(x = obs), alpha = 0.2, linetype = 'solid', color = 'white', linewidth = 1.5) +
+  geom_density(data = data.frame(obs = log(data$y)), aes(x = obs), alpha = 1, linetype = 'dashed') +
   theme_bw() +
@@
 ```
-
-We can also look at some posterior replications separately:
 
@@
 
+names <- paste0('logy_pred[', 1:data$N, ']')
 n <- sample(1:4000, 9)
@@
   facet_wrap(~ chain) +
+  geom_density(data = retrodictions_sub_df, aes(x = ret), alpha = 0.5, col = 'darkred') +
   geom_density(data = data.frame(obs = log(data$y)), aes(x = obs), alpha = 0.5, linetype = 'dashed') +
-  geom_density(data = retrodictions_sub_df, aes(x = ret), alpha = 0.5, col = 'darkred') +
   theme_bw() +

### 2025-09-26 analyses/lpiworkflow.tex
removing one of the PPC figure to make everyone happy!

@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, breakable, colframe=quarto-callout-tip-color-frame, colback=white, opacityback=0, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, opacityback=0, breakable, bottomrule=.15mm, toprule=.15mm, leftrule=.75mm, left=2mm, arc=.35mm, rightrule=.15mm, colframe=quarto-callout-tip-color-frame]
 
@@
 
-Here we start by visually comparing the posterior predictive
-distribution of the predicted counts (across all species and
-populations, red line) to observed data (black dashed line):
-
-\includegraphics{lpiworkflow_files/figure-pdf/unnamed-chunk-5-1.pdf}
-
-We can also look at some posterior replications separately:
+Here we start by visually comparing the 9 draws from the posterior
+predictive distribution of counts (across all species and populations,
+red line) to observed data (black dashed line):
 

### 2025-09-26 revision/r2/response_r2.tex
small modification (ppc...)

@@
 
-...
+We disagree that if the model is not sufficiently complex it leaves `something to be desired.' As we want this workflow to be approachable and leave room for readers to think through how they would advance upon it (to whatever degree of complexity they see as biological relevant), we have left it as is.
 % emwSep22:  I actually hate the idea that a model needs to be complex, but it seems common among ecologists reviewing our workflow papers ....
@@
 
-We now specify that the first posterior predictive distribution we show corresponds to the predicted counts across all species and populations.
+We now specify that the first posterior predictive distribution we show corresponds to the predicted counts across all species and populations, and show 9 draws.
 % Hmm I don't understand?

### 2025-09-29 analyses/lpiworkflow.qmd
updating ppc figure

--- 2025-09-19 analyses/lpiworkflow.qmd
@@
 
-Here we start by visually comparing the posterior predictive distribution of the predicted counts (across all species and populations, red line) to observed data (black dashed line):
-```{r}
-#| fig-height: 3
-#| fig-width: 3
-#| echo: false
-
-names <- paste0('logy_pred[', 1:data$N, ']')
-retrodictions <- unlist(extract(fit, pars = names))
-
-ggplot() +
-  geom_density(data = data.frame(obs = log(data$y)), aes(x = obs), alpha = 0.5, linetype = 'dashed') +
-  geom_density(data = data.frame(retrodictions), aes(x = retrodictions), alpha = 0.5, col = 'darkred') +
-  theme_bw() +
-  labs(y = 'Density', x = 'Counts (log-scale)')
-
-```
-
-We can also look at some posterior replications separately:
-
+Here we start by visually comparing 9 draws from the posterior predictive distribution of counts (across all species and populations, red line) to observed data (black dashed line):
 ```{r}
@@
 #| echo: false
+#| eval: true
 
+names <- paste0('logy_pred[', 1:data$N, ']')
 n <- sample(1:4000, 9)
@@
   theme(strip.text.x.top = element_blank())
+
+```
+
+We may also overlay several draws (here, 30) on the same plot:
+```{r}
+#| fig-height: 3
+#| fig-width: 3
+#| echo: false
+#| eval: true
+
+names <- paste0('logy_pred[', 1:data$N, ']')
+retrodictions <- extract(fit, pars = names)
+m <- t(sapply(names, function(n) retrodictions[[n]][1:30]))
+
+ggplot() +
+  geom_density(data = data.frame(retro = m[,1]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,2]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,3]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,4]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,5]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,6]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,7]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,8]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,9]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,10]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,11]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,12]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,13]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,14]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,15]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,16]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,17]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,18]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,19]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,20]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,21]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,22]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,23]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,24]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,25]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,26]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,27]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,28]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,29]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(retro = m[,30]), aes(x = retro), alpha = 0.2, col = 'darkred', linewidth = .1) +
+  geom_density(data = data.frame(obs = log(data$y)), aes(x = obs), alpha = 0.2, linetype = 'solid', color = 'white', linewidth = 1.5) +
+  geom_density(data = data.frame(obs = log(data$y)), aes(x = obs), alpha = 1, linetype = 'solid') +
+  theme_bw() +
+  labs(y = 'Density', x = 'Counts (log-scale)')
 

### 2025-09-29 analyses/lpiworkflow.tex
updating ppc figure

--- 2025-09-19 analyses/lpiworkflow.tex
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, breakable, colframe=quarto-callout-tip-color-frame, colback=white, opacityback=0, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, toprule=.15mm, left=2mm, leftrule=.75mm, colframe=quarto-callout-tip-color-frame, arc=.35mm, rightrule=.15mm, bottomrule=.15mm, colback=white, breakable]
 
@@
 
-Here we start by visually comparing the posterior predictive
-distribution of the predicted counts (across all species and
-populations, red line) to observed data (black dashed line):
+Here we start by visually comparing 9 draws from the posterior
+predictive distribution of counts (across all species and populations,
+red line) to observed data (black dashed line):
 
@@
 
-We can also look at some posterior replications separately:
+We may also overlay several draws (here, 30) on the same plot:
 

### 2025-10-18 revision/r2/forecastflows_r2.tex
Small edits to main text

--- 2025-09-19 revision/r2/forecastflows_r2.tex
@@
 We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models.
-First, the research community is split between two spheres that prevent uniting a connection between trend estimation and forecasting.
+First, the research community is split between two spheres that prevent uniting trend estimation and forecasting.
 One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, often including complex, mechanistic sub-models indirectly informed by data.
@@
 This---which we call a retrodictive check (\citealp{betanworkflow}, also called a \llabel{PPC}posterior predictive check in a Bayesian workflow,  \citealp{Gelman2020})---allows model output to be compared to observations \llabel{suppfig1}(see Supplementary Figure 1).
-\llabel{howtoPPC} This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regard to our research question, but this also means there is no standard way to do this \citep{Gelman2020}.
+\llabel{howtoPPC} This is a powerful approach, in part because checks should be tailored to assess key features of the data and model output together in regard to our research question, but this also means there is no standard approach  \citep{Gelman2020}.
 It is only once all steps have been completed that we can interpret parameter values with respect to our research question.
@@
 We argue that an improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.
-Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process.
+Using the workflow would make it more obvious that what now appear as major discrepancies are shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and would challenge modelers to show major predictive advances, which is not currently part of the process.
 Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected.
@@
 This step builds somewhat on the skills needed for null models \citep{Gotelli:2012oz}, but with a shift in focus towards the specific bespoke model at hand.
-Ecologists have started to use simulated data more to understand potential limitations of their models and data combined \citep{Hilborn1997}, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility if used in an iterative workflow.
+Ecologists have started to use simulated data more to understand potential limitations of their models and data combined \citep{Hilborn1997}, but this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility when used in an iterative workflow.
 
@@
 Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory.
+% emwOct18: I think the issue in the last sentence is partly the passive voice, the negation (easier to avoid = less likely), and the changing of verb tenses, so I switched all this.
+You can switch back if you want, the sentence was not wrong, it was just hard to get through.
 The way models are currently calibrated is likely not a coincidence, but rather a workaround to avoid confronting the full complexity of the model.
-Calibrating submodels separately makes possible to avoid the fact that, if the model were fitted as a whole, many parameters would compensate for one another---revealing structural degeneracies and making the model far more difficult to use.
+By calibrating submodels separately researchers are less likely to confront structural degeneracies in models (often because many parameters can compensate for others); in contrast, fitting submodels together (as a whole) would reveal such issues at a point when they can be addressed.
 %emw24Jul: nice stuff here.
@@
 \llabel{studydesignandmore}Beyond routine application of the workflow and universal training to support it, the full transformative potential of the workflow hinges on additional broader systemic changes within the research community.
-First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently, Second, open sharing of more comprehensive study design and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community.
+First, fostering more open model sharing through standardized repositories and collaborative platforms would accelerate scientific discovery by allowing researchers to build upon, validate, and refine existing models more efficiently.
+Second, open sharing of more comprehensive study and sampling design details associated with datasets would enable more appropriate data integration and model design decisions, and independent validation of these decisions by the community.
 Third, where feasible, alignment on common datasets would provide a standardized basis for model development and comparison, reducing the fragmentation of efforts and facilitating more direct and meaningful evaluations of different modeling approaches.
@@
 \llabel{clearerML} A careful and coherent workflow is essential to preserve the interpretability of non–machine-learning models---their primary advantage.
-\llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy, especially if the latter lack a robust estimation of their parameters and fall in a complexity trap---a trap with additional costs to interpretability of process-based models.
+\llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy, especially if the latter lack robust estimation of their parameters and fall into a complexity trap---a trap with additional costs to interpretability of process-based models.
 Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.
@@
 However, we believe our workflow provides a tractable step to fixing this.
-By focusing on model development more tightly tied to ecological expertise, we argue this workflow should broaden the community that contributes to model development.
+By focusing on model development more tightly tied to ecological expertise, we argue this workflow would broaden the community that contributes to model development.
 It may also help resolve apparent conflicts by identifying where divergences in model predictions emerge, whether from differences in assumptions, model structures, or other aspects of the modeling process.

### 2025-10-18 revision/r2/response_r2.tex
Edits to letter and a couple more to main text

--- 2025-09-26 revision/r2/response_r2.tex
@@
 
-Editor and reviewer comments are in \emph{gray italics} (we provide below the full context of the reviewers 1-2 comments below, and extract the parts of reviewer 3' comments that required changes to this manuscript), while our responses are in regular text.
+Editor and reviewer comments are in \emph{gray italics} (we provide below the full context of the reviewers 1-2 comments below, and extract the parts of reviewer 3's comments that required changes to this manuscript), while our responses are in regular text.
 \\
@@
 
-We are glad that the reviewer found the manuscript has improved.
-Thank you for this last feedback.
+We are glad that the reviewer found the manuscript has improved, and thank them for all their feedback to improve the manuscript.
 
@@
 % vvdm18sept: is really a typo? I thought it was just a way to say it. Lizzie? 
-...
+% emw18Oct: It sounds weird and the reviewer does not like it so I changed it.
+Corrected.
 
@@
 
-...
+We appreciate this point, however, we wanted our workflow to be approachable.
+We specifically wanted to leave room for readers to think through how they would advance upon it (to whatever degree of complexity they see as biological relevant), thus, we have left it as is.
 % emwSep22:  I actually hate the idea that a model needs to be complex, but it seems common among ecologists reviewing our workflow papers ....
@@
 
-We now specify that the first posterior predictive distribution we show corresponds to the predicted counts across all species and populations.
+We now specify that the first posterior predictive distribution we show corresponds to the predicted counts across all species and populations, and show nine (9) draws.
 % Hmm I don't understand?
@@
 Authier to share our detailed perspective on this.
-Below we address the comment that require changes to the current text.
+Below we address the comments that required changes to the current text.
 
@@
 
+\iffalse
 \begin{mybox}
@@
 \end{mybox}
-
+\fi
 ...
 % do we have t ochange something here?
+% emw18Oct: I don't think so, I commented it out.
 
@@
 
-...
+This is a good point, we have clarified our use of this word in the two areas we used it without additional context ( \llabel{bias1}, \llabel{bias1}). % emw18Oct: fix these refs to main text!
 % Hmm... well... I prefer bias. Preferential sampling sounds to me like it's done on purpose
@@
 
-Sorry it was not clear.
+% emw18Oct: friendly reminder to avoid 'sorry' in professional letters IMHO
+Apologies this was not clear.
 Our point here was not to share any enthusiasm for machine learning, but rather to highlight the need for an even more careful and coherent workflow to build interpretable models (i.e.
@@
 
-We would like to thank the reviewer once again for all these very interesting avenues for discussion---beyond the community of researchers (to whom this paper was primarily addressed).
+We would like to thank the reviewer once again for all these very interesting avenues for discussion---and look forward to discussions with him directly.
+% beyond the community of researchers (to whom this paper was primarily addressed).
 

### 2025-10-18 revision/r2/forecastflows_r2.tex
Edits to letter and a couple more to main text

@@
 Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions.
-To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting a natural output.
+To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting is a natural output.
 We show how this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could harmonize currently divided efforts at trend estimation and forecasting to better inform sustainable policies.
@@
 Meeting these \llabel{global}global policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on fewer data.
-Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic biases.
-Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
+Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic sampling biases.\llabel{bias1} Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
 Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning.
@@
 unbalanced data).
-For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step.
+For example, if a researcher realizes their empirical data is geographically biased (for example, more intensive sampling in Europe) \llabel{bias2}, this bias should be built into the model and thus into this data simulation step.
 This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics.

### 2025-11-19 revision/r2/response_r2.tex
fix ref to main text

@@
 \fi
-...
+
 % do we have t ochange something here?
@@
 
-This is a good point, we have clarified our use of this word in the two areas we used it without additional context ( \llabel{bias1}, \llabel{bias1}). % emw18Oct: fix these refs to main text!
+This is a good point, we have clarified our use of this word in the two areas we used it without additional context (lines, \href{file:forecastflows_r2\#lintarget:bias1}{\ref*{lin:bias1}}, \href{file:forecastflows_r2\#lintarget:bias2}{\ref*{lin:bias2}}).
+% emw18Oct: fix these refs to main text!
 % Hmm... well... I prefer bias. Preferential sampling sounds to me like it's done on purpose

### 2025-11-19 revision/r2/forecastflows_r2.tex
fix ref to main text

@@
 unbalanced data).
-For example, if a researcher realizes their empirical data is geographically biased (for example, more intensive sampling in Europe) \llabel{bias2}, this bias should be built into the model and thus into this data simulation step.
+For example, if a researcher realizes their empirical data is geographically biased (for example, more intensive sampling in Europe)\llabel{bias2}, this bias should be built into the model and thus into this data simulation step.
 This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics.

### 2025-11-19 forecastflows_beforemerge.tex
missing files

--- 2025-05-01 forecastflows.tex
@@
 \documentclass[11pt]{article}
-\usepackage[top=0.70in, bottom=.80in, left=1in, right=1in]{geometry}
+\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
 \renewcommand{\baselinestretch}{1.1}
@@
 \usepackage[font=sf]{caption}
-\usepackage{hyperref}
 
@@
 \date{\today}
-\author{Victor Van der Meersch$^{1*}$, James Regetz$^{2}$, T. Jonathan Davies$^{1,3}$ \& EM Wolkovich$^{1}$}
+\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}
 \maketitle
-\noindent $^{1}$ Forest and Conservation Sciences, University of British Columbia, Vancouver, BC V6T 1Z4, Canada\\
-$^{2}$ National Center for Ecological Analysis and Synthesis, 1021 Anacapa St, Santa Barbara, CA 93101, United States\\
-$^{3}$ Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada \\
-$^{*}$ \url{mailto:victor.vandermeersch@ubc.ca}\\
 
-\noindent \emph{For:} Scientific and Statistical Workflow theme issue for \emph{Phil Trans A} as an \emph{Opinion}
+{\bf Deadline:} 1-ish May 2025 (was 1 April 2025, getting in by May 8th would be good!)
+
+\emph{For:} Scientific and Statistical Workflow theme issue for \emph{Phil Trans A} as an \emph{Opinion}
 % They mention a tex template here: https://royalsocietypublishing.org/rsta/for-authors#question4
@@
 Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions.
-New research often suggests that models accounting for these complexities can yield more accurate trends and predictions.
-We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons.
+While new research often suggests that models accounting for these complexities can yield more accurate trends and predictions, we argue that the usual workflows for model fitting in ecology make it impossible to evaluate and compare the models used to generate insights and predictions.
 First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data.
@@
 We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process.
-We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
+We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could be transformative for ecological modeling.
 
@@
 
-% {\noindent \bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
+{\noindent \bf Goal:} Increase awareness of how we can merge statistical and scientific workflows in ecology (especially forecasting) and what we would get out of it.
+\vspace*{0.5cm}
 
@@
 Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on less data.
-Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic biases.
+Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from massive geographic, temporal and taxonomic biases.
 Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
 Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning.
-These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012, Thuiller2019}.
+These approaches have failed to yield clear agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.
 
@@
 Workflows that fully integrate all the steps required to build a model from an ecological question, with evaluation of limitations and potential problems before estimating its parameters and making projections, could reduce many of these problems.
-In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge \citep{McElreath2018, betanworkflow, Gelman2020, Wolkovich2024}.
+In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to understand when, where, and why different models diverge.
 Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
@@
 In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and use this data to inform our model and differentiate between hypotheses.
-This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection \citep{Nosek2018}.
+This method underlies much of the recent pre-registration movement, where hypotheses and methodology are defined prior to data collection.
+%add cites for prereg?
 But this idealized method often does not apply to the reality of ecological research.
@@
 	\caption{Different trajectories through model-data space during model development.
-Ideally, this begins in the pre-model/pre-data quadrant, where the research question is defined.
+Ideally, this begins in the pre-model/pre-data quadrant, where the research question is defined---building on previous theories and on our hypothesis.
 The next step moves to the post-model/pre-data quadrant, where a model is build---independently of the data.
-This is where we should spend a good deal of time evaluating model behavior using simulated data.
+This is where we should spent most of the time, evaluating model behavior using simulated data.
 Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data.
@@
 But the current workflows combined with the challenges ecologists are facing---both in term of data complexity and societal needs---instead may lead to persistent problems.
-Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (figure \ref{fig:modeldata}a).
-For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (figure \ref{fig:modeldata}b).
+% jdavies15april: what challenges?
+forecasting from imperfect data or the urgent need to generate forecasts?
+Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way---without the checks and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Fig.
+\ref{fig:modeldata}a).
+For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models, frequently obscuring the steps underlying model building and parameterization (Fig.
+\ref{fig:modeldata}b).
 Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
+% jdavies15april: is this particularly specific to process based models?
 Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.
+% mention the disconnection between trends and forecasts?
 
-We argue a workflow that moves along the data-model space in a coherent sequence of steps (figure \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
-The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, figure \ref{fig:workflow}).
+% jdavies15april: Could the bottom right be where exploratory data analysis fits in, generating hypotheses that inform the research question?
+
+We argue a workflow that moves along the data-model space in a coherent sequence of steps % with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure?
+Or wait to mention simulation until referencing Fig 2?
+[FLAG: what do you think?
+I agree you should choose on of the two options]
+(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
+The first step of this workflow is to define an explicit research question and formulate hypotheses (Step 1, fig.
+\ref{fig:workflow}).
 This involves making clear assumptions about the most influential drivers, within the specific context of our study.
 This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error.
-From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, figure \ref{fig:workflow}).
-The general idea is to start with a relatively simple model that we could refine later (see example worklow we provide).
+% vvdm16april: to answer Jonathan comment about verbal model
+From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (Step 2, fig.
+\ref{fig:workflow}).
+The general idea is to start with a relatively simple model that we could refine later.
 At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.
@@
 
-With a model in place, the next step focuses on testing and understanding it via data simulation (step 3, figure \ref{fig:workflow}).
+With a model in place, the next step focuses on testing and understanding it via data simulation (Step 3, fig.
+\ref{fig:workflow}).
 `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data.
+% This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model.
+% jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e.
+simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model.
+[FLAG: yes I can understand this point...]
+% For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step.
+[FLAG: so I moved this sentence below, when we say "Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data"
 We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.
 At this stage, the focus is on understanding the model, so we may need to spend time thinking about whether each parameter is reasonable or not.
-This is also a step for making sure the model is working as expected (see example worklow).
+This is also a step for making sure the model is working as expected, so simulated data often would have low error and high sample sizes (to make sure the model returns the parameter values used to generate the simulated data).
+%jr24april not sure what this final sentence means?
 
-Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (step 4, figure \ref{fig:workflow}).
+Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step (Step 4, fig.
+\ref{fig:workflow}).
 This way, we obtain parameter estimates constrained by observations.
-These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (step 5, figure \ref{fig:workflow}).
-This---which we call a retrodictive check (\citealp{betanworkflow}, also sometimes called a posterior predictive check, \citealp{Gelman2020})---allows model output to be compared to observations.
+%emwApr19: I commented out the below as I think linking difficulty in fitting the model to problems with the data is pretty tricky (difficulty in fitting the model can point to many things!) and thus too confusing to include here.
+% Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works.
+This could lead us to either simplify our research question or---ideally---launch new data collection efforts.
+These parameter estimates lead to the second data simulation step, this time using our fitted model parameters to generate predictions (Step 5, fig.
+\ref{fig:workflow}).
+This retrodictive check allows model output to be compared to observations.
+% jregetz16april: do we need to define “retrodictive”?
+% The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.
+% jregetz16april: Not sure what this means?
+jdavies did not like it either!
+(he thought it was a typo), vvdm16april: reformulate:
 The workflow encourages a focus on the full model, where any parameter (such as a trend estimate) must be carefully interpreted alongside others, as all are fundamental components that shape both inference and forecasting.
-
 Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.
@@
 
+\begin{figure}[h]
+	\centering
+    \hspace*{-1.5cm}
+	\includegraphics{figures/figure_worflow_wsteps}
+	\caption{Our workflow builds on others \citep{Betancourt2020, Wolkovich2024, McElreath2018, Schoot2021} to highlight the integrated nature of stuff in ecology...} 
+	\label{fig:workflow}
+\end{figure}
+
+% More details, what's special! Feedbacks! %jrApr25 I think this paragraph reads really well!
 A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
-The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
+The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
 Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated),
-flagging the need to change the model structure---before incorporating empirical observations. 
+flagging the need to change the model structure---before incorporating real observations. 
 The second feedback loop comes from the retrodictive check. 
 Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic.
-We can refine the model to integrate the missing process(es) (if we can identify them) and return to the start of the workflow.
+We can refine the model to integrate the missing process(es) (if we can identify them) and restart the workflow.
 Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g.
@@
 
-\begin{figure}[h]
-	\centering
-	\vspace*{-0.1cm}
-    \hspace*{-1.5cm}
-	\includegraphics{figures/figure_worflow_wsteps}
-	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data.
-Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5).
-Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
-The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.}
-	\vspace*{-0.5cm}
-	\label{fig:workflow}
-\end{figure}
-
 \section{The workflow in practice}
 
-Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties and yield new scientific insights.
+%emw6Apr: I like the shift to 'parameter estimation' instead of trend estimation, but I wonder if we could make that shift clearer above.
+Maybe somewhere we say that the workflow would  encourage a focus on the full model and its parameters: For trend estimates, the trend becomes just one parameter that must be carefully interpreted alongside others .
+Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow could highlight the best opportunities to reduce uncertainties through new scientific insights, toward the most critical steps.
 This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.
@@
 
+%jregetz16april: One thing I’m stuck on (and could’ve written this note earlier too).
+Is “trends” really essential?
+Or is this just shorthand for “simple, one/few parameter, non-mechanistic forecasting models”?
+If yes, can we generalize that in the MS overall, or at least say so up front and then say e.g.
+“we’ll refer to these as trend models for brevity”?
+If no, and something about trend estimates is specifically key to the paper, can we say more about *that*?
+[FLAG]
+%emwApr19: I think the only key part is that biodiversity trends and trends over time are big in forecasting ...
+as long as we don't lose that I think it would be great to make the suggested change -- Jim, can you see if you can edit it on the next round?
+
 Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends over time, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}.
-While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be fewer and they would be more rapidly resolved through use of an improved workflow.
+While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced and more rapidly resolved through use of an improved workflow.
 % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.
@@
 
+% expand on 'highlight major missing components in these models'? ADD example?
+% Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models and in turn debate them.. 
+% Mention elephants rebounding...
+
+% Trends - improvements!
 Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates.
-Returning to the example of a global estimate of trends in vertebrate populations of species over time (see Box) and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
+Returning to the example of a global estimate of trends in vertebrate populations of species over time and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate?
+% jdavies15april: This is what the LPI wants to be (like a index fund in the stock market tracking trends over time) , but this is often overlooked.
 Or a need to also find which species are declining most, including those that may have poor or no data?
 From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data.
-Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends (see workflow example).
+Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends.
 For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled.
@@
 Our second case-study focuses on process-based modeling, which is often considered the gold standard for forecasting in ecology \citep{Urban2016, Pilowsky2022} and beyond.
-Newer models, however, generally incorporate greater complexity (and an ever-growing number of parameters), which can make it difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.
+Newer models generally incorporate greater complexity and an ever-growing number of parameters, making it more difficult to increase scientific understanding \citep{Franklin2020}, and suggest or model potential policies.
 
@@
 These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
-Some researchers now advocate for simplifying models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. 
-If a model becomes too complex, it may become a black box,
-and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible (see Box).
+Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}.
+\todo[size=\tiny]{jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
+Reply: we agree but not really relevant to PBM?
+Maybe it's not clear enough we are focusing on PBM in this section, which we tried to change above, see if it works}
+If a model becomes too complex, it may become a black box, % mention black box here to line up with the beginning of next paragraph
+and understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
 Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.
+\todo[size=\tiny]{jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty ....
+Reply: Same as just above.}
 
@@
 Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only one of the many different output variables of the model.
-The complexity of these models can make it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability---such work is generally done as a separate effort, exposing potential model problems later on \citep{VanderMeersch2025b}, and not required as a preliminary step.
+%The model building and calibration steps are usually limited to the Methods section, and estimated parameter values are rarely discussed \citep[e.g.][]{Conradi2024, VanderMeersch2025a}.
+The complexity of these models often makes it difficult to assess how well the potential parameter space was searched, and whether there was any potential non-identifiability---such work is generally done as a separate effort, exposing potential model problems later on \citep{VanderMeersch2025b}, and not required as a preliminary step.
+\todo[inline, size=\tiny]{For above, addressing: jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties?
+This could help illustrate potential degeneracies etc.
+Reply: Tried something, but not sure it works so feedback much appreciated!}
 
+%emw19Apr -- Can we just get away with using non-identifiable and save the reader also needing to digest degenerate?
+(https://betanalpha.github.io/assets/case_studies/identifiability.html)?
+I tried that on this draft, see how it works!
 Applying the workflow to process-based models would help open the black box.
 Each successive step of model development in the workflow may highlight current problems and a path to solutions.
-Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two, and help expose potential parameter identifiability issues in the model design.
+Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential parameter identifiability issues in the model design.
+%jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below)
 Uncovering identifiability issues would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity.
 The workflow may also highlight model problems by requiring more explicit model calibration (i.e.
-data fitting), which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model.
+data fitting), % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
+which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model.
 Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity).
@@
 % Trends - outline current problem
-In the current workflow for estimating trends over time, a new model with a new estimate often leads to a paper (see figure A above) because ecologists spend less time interrogating their models with simulated data, or their model performance fit to empirical data.
-The Living Planet Index (LPI, \url{www.livingplanetindex.org}), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results.
+
+In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data.
+The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe, is emblematic of these conflicting results.
 % \todo[]{could highlight how influential the LPI has become (recent reports suggest a 73\% ave.
@@
 This would be massive!}
-With updated data released semi-annually alongside new datasets and new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
-For examples using LPI data, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed many previous results \citep{Johnson2024}.
+With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach.
+For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}.
+% \todo[]{davies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations.
+We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends.
+For this we need to predict outside of the set}
 
@@
 Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions.
-Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque for anyone who has not worked the model itself.
+Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque from an external perspective.
 The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred.
-Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes (see figure B above).
-Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce certain observations or theory.
-The way models are currently calibrated is likely not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
+Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes.
+Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e., fixed to a value found in the literature) or tuned to reproduce some observations or theory.
+The way models are currently calibrated is not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.
+% jdavies15april: is this more specific to process based  models?
+[FLAG: hmm yes again?
+Same point as above, maybe it's not clear we're talking mostly about PBM?]
 
@@
 Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge.
-It would be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimates.
+It would be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimations.
 For both trend estimation and forecasting, the workflow would refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model.
@@
 
-\clearpage
 \section{Barriers and opportunities}
@@
 We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
+% One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed.
+This may be especially true for those who see science rewarded mostly through the shear number of publications.
+But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help.
+% jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
 The first is pressure to publish quickly, which can be at odds with the reality that good model fitting is inherently iterative and takes time.
 The second is reluctance to embrace adoption of open science practices that ensure modeling efforts are fully transparent and reproducible.
-We believe adopting a workflow such as the one we propose may change this reluctance, and aligns with ongoing efforts to place increasing value on research that is carefully developed, openly collaborative (including both data and code), and transparent about areas of uncertainty.
+It is imperative that we continue to place increasing value on research that is carefully developed, openly collaborative (including both data and code), and transparent about areas of uncertainty.
+Especially in ecology where samples sizes are low and effects are likely non-linear and complex, we believe adopting a workflow such as the one we propose here will pay off in the longer term, both for individual researchers focused on making enduring contributions, and for the community as a whole focused on improving future data collection and model development (Fig.
+\ref{fig:workflow}).
 
@@
 Much of ecology still divides the world into training for those who gather data and learn a limited set of pre-built models versus those who develop more complex models.
-In ecological training today, researchers who conduct field and lab studies often learn a limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi-squared test).
+In ecological training today, researchers who conduct field and lab studies often learn a highly limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).
 
@@
 These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models.
+% process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms
 Neither group has fully integrated data simulation into their statistical or scientific workflows.
@@
 
-We suggest unified training in our proposed workflow, or a similar one \citep{betanworkflow,Gelman2020, grinsztajn2021,vandeschoot2021}, would focus on learning to generate questions and then models, and then how to simulate data from them.
+Instead of siloing ecological training into experimental design and tests that may match certain designs (though rarely do for ecological data, CITES) versus model design and development, we suggest unified training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them.
+This would mean training to link the ecological processes they study with the mathematical models that may describe them.
 Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates.
-Empiricists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models.
-Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may see new bridges from their work to empirical data and forecasting.
+Empiricists % change suggested by JD, was empirical ecologists before...
+ would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models.
+Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting.
+% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
+Those focused on learning complex statistical models may find many others in the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.
 
@@
 
-With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important.
-Machine learning will likely surpass process-based models for forecasting accuracy if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
-Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.
+In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made.
+Machine learning will likely surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability.
+Similarly, engaging in statistical trend analysis, especially when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
+% jdavies15april: nothing intrinsically bad about ML, but sometimes the goal is just to understand the system, not make predictions or forecasts.
 
-Beyond improving model building and evaluation, our proposed workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
-Process-based models could once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
+Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them.
+Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
 Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty.
-Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification---preventing modelers from making biased inferences and unfounded assumptions beyond what the data can support.
-It would thus define a clear and limited context in which the model should apply, and limit discussion of adding increasing complexity.
+Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.
+It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.
 
-\emph{Conclusions:} 
+% jdavies15april: This could be the Concluding section highlighting the general advances provided by adopting your framework (and perhaps a better place for text currently at end of section 3.2).
 Today model development in ecology is rarely transparent, which limits how easily the research community can understand models, and thus identify potential issues.
-Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate has become mired in methodological considerations.
+Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations.
 However, we believe our workflow provides a tractable step to fixing this.
@@
 
-% This shift would present a significant challenge---as it would likely reveal many issues related to non-identifiability in models and data limitations before achieving robust inference.
+This shift would present a significant challenge---as it would likely reveal many issues related to non-identifiability in models and data limitations before achieving robust inference.
 But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.
